### 2024-03-11 main.tex
Update on Overleaf.

\documentclass{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with `a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{array}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\title{Finch: A Datastructure-Driven Array Programming Language}
\author{Willow Ahrens, Teo Collins, Radha Patel, Changwan Hong, Saman Amarasinghe}

\begin{document}
\maketitle

\begin{abstract}
From FORTRAN to Numpy, arrays have revolutionized how we express computation.
Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology. 
However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. 
\end{abstract}

\section{Introduction}

- Fortran supported lots of control structures, no data structures, just dense arrays.

- We tried to emulate complex data using only dense arrays but with complicated control structures.

- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side

- We’re bringing these together, both need to work together.

\subsection{Contributions}

\begin{enumerate}
\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
\item <Performance Contributions>
\end{enumerate}

\begin{table}[h!]
\centering
\begin{tabular}{l|ccccc}
\textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
\hline
Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
Multiple LHS            & \checkmark &            & \checkmark & \checkmark & \checkmark \\
Affine Indices          & \checkmark &            &            & \checkmark & \checkmark \\
Recurrence              & \checkmark &            &            &            &            \\
If-Conditions and Masks & \checkmark & \checkmark &            & \checkmark & \checkmark \\
Scatter Gather          & \checkmark &            &            & \checkmark & \checkmark \\
Early Break             &            & \checkmark &            &   \checkmark         & \checkmark \\
\end{tabular}
\caption{Feature support across various tools.}
\label{tab:features}
\end{table}

\begin{table}[h!]
\centering
\begin{tabular}{l|ccccc}
\textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
\hline
Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
Padded                   & \checkmark &            &            &            & \checkmark \\
One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
Sparse                   &            & \checkmark &            &            & \checkmark \\
Run-length               &            &            &            &            & \checkmark \\
Symmetric                &            &            &            &            & \checkmark \\
Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
Ragged                   &            &            & \checkmark &            & \checkmark \\
\end{tabular}
\caption{Support for various data structures across tools.}
\label{tab:data_structures}
\end{table}

\section{Additional Topics}

\begin{enumerate}
\item Concordant iteration
\item Dimensionalization
\item Bounds analysis/queries
\item Performance warnings

\end{enumerate}

\section{Evaluation}

\subsection{Data-Driven Performance Engineering}
\subsubsection{Sparse-Sparse Matrix Multiply}
Examples that demonstrate performance engineering in a datastructure-driven model

\subsubsection{SpMV}

\subsection{Programming over flexible data}

\subsubsection{RLE Blur}

-Compare against the zip/unzip version

- Which Datasets? Check taco-rle or looplets.
(perhaps consider if a neural network is the right call here)

\subsubsection{Histogram}

\subsubsection{Graph Applications}
Bellman Ford, BFS, Pagerank, Triangle Counting.
	Are there special graph structures or kernels we do better on?

\subsubsection{High-level kernel fusion}
Find an example where fusing the python interface gives a big speedup over non-fused kernels.

\bibliographystyle{alpha}
\bibliography{bibliography}

\end{document}

### 2024-03-11 main.tex
add plots

@@
 
+%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
+
+\begin{figure}
+    \begin{minipage}[t]{0.5\textwidth}
+        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
+    \end{minipage}%
+    \begin{minipage}[t]{0.5\textwidth}
+        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
+    \end{minipage}
+    \caption{Performance of SpMV across various tools.}
+\end{figure}
+
 \subsection{Programming over flexible data}
@@
 \subsubsection{Graph Applications}
-Bellman Ford, BFS, Pagerank, Triangle Counting.
-	Are there special graph structures or kernels we do better on?
+\begin{figure}
+	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
+	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
+    \caption{Performance of graph apps across various tools.}
+\end{figure}
 

### 2024-03-14 main.tex
Update on Overleaf.
cool

@@
 
+\begin{figure}
+    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
+    \caption{Performance of SpMV by Finch format.}
+\end{figure}
+
 \subsection{Programming over flexible data}
 
-\subsubsection{RLE Blur}
+\subsubsection{Image Morphology}
 
--Compare against the zip/unzip version
+\begin{figure}
+	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
+    \caption{Performance of Finch on erosion task (4 iterations).}
+\end{figure}
 
-- Which Datasets? Check taco-rle or looplets.
-(perhaps consider if a neural network is the right call here)
+\begin{figure}
+	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
+    \caption{Performance of Finch on masked histogram task.}
+\end{figure}
 
-\subsubsection{Histogram}
+\begin{figure}
+	\includegraphics[width=\linewidth]{fill_speedup_over_opencv.png}
+    \caption{Performance of Finch on flood fill task.}
+\end{figure}
 
-\subsubsection{Graph Applications}
+\subsubsection{Graph Analytics}
 \begin{figure}

### 2024-03-14 main.tex
some more stuff

@@
 
-- Fortran supported lots of control structures, no data structures, just dense arrays.
-
-- We tried to emulate complex data using only dense arrays but with complicated control structures.
-
-- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
-
-- We’re bringing these together, both need to work together.
-
+%- Fortran supported lots of control structures, no data structures, just dense arrays.
+%
+%- We tried to emulate complex data using only dense arrays but with complicated control structures.
+%
+%- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
+%
+%- We’re bringing these together, both need to work together.
 \subsection{Contributions}
@@
 \begin{enumerate}
-\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
+\item A rich structured array programming language with for-loops
+and complex control flow constructs at the same level of productivity
+of dense arrays. To our knowledge, the Finch programming language is the first
+to support if-conditions, early breaks, and multiple left hand sides over
+structured data, as well as complex accesses such as affine indexing or scatter/gather.
+\item More complex array structures than ever before. A complete level-by-level
+structure-description language for expressing the structure of data
+hierarchically. The first such set of formats to efficiently capture banded,
+triangular, run-length-encoded, or sparse datasets, and any combination thereof.
+\item The Finch compiler specializes programs to data structures in a
+predictable, deterministic approach, making it easier to search the complex
+space of programs and datastructures to find an appropriate fit for a given
+application. The tensor interface makes it easy to extend Finch to new level
+formats. A unique tensor lifecycle model enables polymorphism by analyzing the
+appropriate stages to insert simple, overloadable, interface functions such as
+initialization or finalization. 
+\item A high-level array programming language and fusion interface for
+operations such as map, broadcast, or reduce that can be compiled to efficient
+code using the previous loop-level abstractions.
+\item We evaluate the productivity of our language in several case studies,
+showing that Finch can be used to accelerate a wide range of applications, 
+from classic operations such as spmv and spgemm, to more complex applications such as image processing and graph analytics.
+We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
+%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
 The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets.
 The formats capture many use cases, from random updates to sequential construction.
-\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions.
+%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions.
 The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
-\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
-\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
-\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
-\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
-\item <Performance Contributions>
+%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
+%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
+%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
+%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
+%\item <Performance Contributions>
 \end{enumerate}
@@
 
-\section{Additional Topics}
+\section{Background}
+\subsection{Looplets}
+\subsection{FiberTrees}
+
+\subsection{Concordant Iteration}
+
+\subsection{Protocols}
+
+\section{The Finch Language}
+
+\subsection{Syntax and Semantics}
+
+\section{The Finch Compiler}
+
+\subsection{Dimensionalization}
+
+\subsection{Concordization}
+
+\subsection{Bounds Analysis}
+
+\subsection{Performance Warnings}
+
+\subsection{Wrapperization}
+
+\subsection{Simplification and Algebraic Transformations}
+
+\section{The Tensor Interface}
+
+\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+
+\subsection{Level Abstraction}
 
 \begin{enumerate}
-\item Concordant iteration
-\item Dimensionalization
-\item Bounds analysis/queries
-\item Performance warnings
+\item fibers
+\item assembly
+\item reassembly
+\end{enumerate}
 
+\subsection{Core Level Langage Primitives}
+\begin{enumerate}
+\item SparseList
+\item SparseDict
+\item ...
 \end{enumerate}
+
+\subsection{Wrapper Tensors}
+
+\subsection{Scalars}
+
+\subsubsection{Sparse Scalars}
+\subsubsection{Early Break Scalars}
+
+\section{The Finch High-Level API (Needs a Name)}
+
+\subsection{Finch Logic}
+
+\subsection{Finch Interpreter}
+
+\subsection{Lowering}
+\subsubsection{Heuristic Optimization}
 

### 2024-03-14 main.tex
Update on Overleaf.

@@
 \title{Finch: A Datastructure-Driven Array Programming Language}
-\author{Willow Ahrens, Teo Collins, Radha Patel, Changwan Hong, Saman Amarasinghe}
+\author{Willow Ahrens, Teo Collin, Radha Patel, Kyle Deeds, Changwan Hong, Saman Amarasinghe}
 
@@
 initialization or finalization.
-\item A high-level array programming language and fusion interface for
-operations such as map, broadcast, or reduce that can be compiled to efficient
-code using the previous loop-level abstractions.
+
 \item We evaluate the productivity of our language in several case studies,
@@
 We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
+Additionally, as a case study, a high-level array programming language and fusion interface for
+operations such as map, broadcast, or reduce that can be compiled to efficient
+code using the previous loop-level abstractions.
 %\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
@@
 
-\section{The Finch Compiler}
-
-\subsection{Dimensionalization}
-
-\subsection{Concordization}
-
-\subsection{Bounds Analysis}
-
-\subsection{Performance Warnings}
-
-\subsection{Wrapperization}
-
-\subsection{Simplification and Algebraic Transformations}
-
 \section{The Tensor Interface}
@@
 
-\section{The Finch High-Level API (Needs a Name)}
+\section{The Finch Compiler}
 
-\subsection{Finch Logic}
+\subsection{Dimensionalization}
 
-\subsection{Finch Interpreter}
+\subsection{Concordization}
 
-\subsection{Lowering}
-\subsubsection{Heuristic Optimization}
+\subsection{Bounds Analysis}
+
+\subsection{Performance Warnings}
+
+\subsection{Wrapperization}
+
+\subsection{Simplification and Algebraic Transformations}
 
@@
 
-\subsubsection{High-level kernel fusion}
+\subsection{Implementing Numpy Array API in Finch}
+\subsubsection{The Finch High-Level API (Needs a Name)}
+
+\subsubsection{Finch Logic}
+
+\subsubsection{Finch Interpreter}
+
+\subsubsection{Lowering}
+\subsubsection{Heuristic Optimization}
+
 Find an example where fusing the python interface gives a big speedup over non-fused kernels.
+
+matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
+in-place.
+dot((v^t .* u), w)) vs. 
+(v^t .* dot(u, w))
 

### 2024-03-14 main.tex
Update on Overleaf.

@@
 \begin{document}
+\newcommand\teo[1]{\textcolor{red}{teo:#1}}
+
 \maketitle
@@
 
-\section{The Tensor Interface}
-
-\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
-
-\subsection{Level Abstraction}
-
-\begin{enumerate}
-\item fibers
-\item assembly
-\item reassembly
-\end{enumerate}
-
-\subsection{Core Level Langage Primitives}
-\begin{enumerate}
-\item SparseList
-\item SparseDict
-\item ...
-\end{enumerate}
+\section{Bridging Looplets and Finch: The Tensor Interface}
+\teo{I am doing stuff in this section.}
+\input{sections/tensors.tex}
 
@@
 \subsubsection{Sparse-Sparse Matrix Multiply}
+
 Examples that demonstrate performance engineering in a datastructure-driven model
@@
 \subsubsection{SpMV}
+Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data.
+More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats.
+Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
+% I'm not sure if the right place for this is here - maybe should come earlier?
+
+Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns.
+Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.
+
+The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks.
+The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
+In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
+
+We compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
+We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
+to test the OSKI interface [cite], Ahrens et al.
+to test a variable block row format partitioning strategy [cite], and Kjolstad et al.
+to test the TACO library [cite].
+In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
+The dense vector is randomly generated.
+We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures 1 and 2.
+
+\subsubsection{SparseList Format}
+\subsubsection{SparseVBL Format}
+The SparseVBL format is similar to the SparseList format, but contiguous fibers are stored together as blocks.
+Thus, it is most worthwhile to use when the matrix has a blocked structure—i.e.
+where there tend to be stretches of consecutive nonzero values.
+We found that the SparseVBL format performed the best for matrices with a clear block structure (list specific matrices) or an imperfect blocked structure (list specific matrices).
+\subsubsection{Pattern Format}
+\subsubsection{Symmetric SpMV}
+Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix.
+We can apply this strategy with any level format.
+Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
+However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
+Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO .
+Notably, there is a 1.9x speedup for the HB/saylr4 matrix.
+\subsubsection{4D Blocked SpMV}
 

### 2024-03-14 sections/tensors.tex
Update on Overleaf.

Looplets provide descriptions of iterators. 
%
The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.

%
As we will see
\begin{enumerate}
    \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
    \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
    \item how do we ask for the right iterator and how do we change/manage their state?
\end{enumerate}
Need to Make these problems clear by this point, hope that read anticipate them.

Solutions:
\begin{enumerate}
    \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
    \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
\end{enumerate}

\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}

\subsection{Level Abstraction}

\begin{enumerate}
\item fibers
\item assembly
\item reassembly
\end{enumerate}

\subsection{Core Level Langage Primitives}
\begin{enumerate}
\item SparseList
\item SparseDict
\item ...
\end{enumerate}

### 2024-03-15 main.tex
cool

--- 2024-03-14 main.tex
@@
-\documentclass{article}
+%%
+%% This is file `sample-acmsmall.tex',
+%% generated with the docstrip utility.
+%%
+%% The original source files were:
+%%
+%% samples.dtx  (with options: `all,journal,bibtex,acmsmall')
+%% 
+%% IMPORTANT NOTICE:
+%% 
+%% For the copyright see the source file.
+%% 
+%% Any modified versions of this file must be renamed
+%% with new filenames distinct from sample-acmsmall.tex.
+%% 
+%% For distribution of the original source see the terms
+%% for copying and modification in the file samples.dtx.
+%% 
+%% This generated file may be distributed as long as the
+%% original source files, as listed above, are part of the
+%% same distribution. (The sources need not necessarily be
+%% in the same archive or directory.)
+%%
+%%
+%% Commands for TeXCount
+%TC:macro \cite [option:text,text]
+%TC:macro \citep [option:text,text]
+%TC:macro \citet [option:text,text]
+%TC:envir table 0 1
+%TC:envir table* 0 1
+%TC:envir tabular [ignore] word
+%TC:envir displaymath 0 word
+%TC:envir math 0 word
+%TC:envir comment 0 0
+%%
+%%
+%% The first command in your LaTeX source must be the \documentclass
+%% command.
+%%
+%% For submission and review of your manuscript please change the
+%% command to \documentclass[manuscript, screen, review]{acmart}.
+%%
+%% When submitting camera ready or to TAPS, please change the command
+%% to \documentclass[sigconf]{acmart} or whichever template is required
+%% for your publication.
+%%
+%%
+\documentclass[acmsmall]{acmart}
+
+%%
+%% \BibTeX command to typeset BibTeX logo in the docs
+\AtBeginDocument{%
+  \providecommand\BibTeX{{%
+    Bib\TeX}}}
+
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{acmlicensed}
+\copyrightyear{2024}
+\acmYear{2024}
+\acmDOI{XXXXXXX.XXXXXXX}
+
+%%
+%% These commands are for a JOURNAL article.
+\acmJournal{JACM}
+\acmVolume{37}
+\acmNumber{4}
+\acmArticle{111}
+\acmMonth{8}
+
+%%
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% For managing citations, it is recommended to use bibliography
+%% files in BibTeX format.
+%%
+%% You can then either use BibTeX with the ACM-Reference-Format style,
+%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
+%% support for advanced citation of software artefact from the
+%% biblatex-software package, also separately available on CTAN.
+%%
+%% Look at the sample-*-biblatex.tex files for templates showcasing
+%% the biblatex styles.
+%%
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+%% If you are preparing content for an event
+%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
+%% citations and references.
+%% Uncommenting
+%% the next command will enable that style.
+%%\citestyle{acmauthoryear}
 
@@
 
+%%
+%% end of the preamble, start of the body of the document source.
+\begin{document}
+
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
 \title{Finch: A Datastructure-Driven Array Programming Language}
-\author{Willow Ahrens, Teo Collins, Radha Patel, Changwan Hong, Saman Amarasinghe}
 
-\begin{document}
-\maketitle
+%%
+%% The "author" command and its associated commands are used to define
+%% the authors and their affiliations.
+%% Of note is the shared affiliation of the first two authors, and the
+%% "authornote" and "authornotemark" commands
+%% used to denote shared contribution to the research.
+\author{Willow Ahrens}
+\affiliation{%
+  \institution{MIT CSAIL}
+  \city{Cambridge}
+  \state{Massachusetts}
+  \country{USA}}
+\email{willow@csail.mit.edu}
 
+\author{Teo Collins}
+\affiliation{%
+  \institution{MIT CSAIL}
+  \city{Cambridge}
+  \state{Massachusetts}
+  \country{USA}}
+\email{teoc@mit.edu}
+
+\author{Radha Patel}
+\affiliation{%
+  \institution{MIT CSAIL}
+  \city{Cambridge}
+  \state{Massachusetts}
+  \country{USA}}
+\email{rrpatel@mit.edu}
+
+\author{Kyle Deeds}
+\affiliation{%
+  \institution{University of Washington}
+  \city{Seattle}
+  \state{Washington}
+  \country{USA}}
+\email{kdeeds@cs.washington.edu}
+
+\author{Changwan Hong}
+\affiliation{%
+  \institution{MIT CSAIL}
+  \city{Cambridge}
+  \state{Massachusetts}
+  \country{USA}}
+\email{changwan@mit.edu}
+
+\author{Saman Amarasinghe}
+\affiliation{%
+  \institution{MIT CSAIL}
+  \city{Cambridge}
+  \state{Massachusetts}
+  \country{USA}}
+\email{saman@csail.mit.edu}
+
+%%
+%% By default, the full list of authors will be used in the page
+%% headers. Often, this list is too long, and will overlap
+%% other information printed in the page headers. This command allows
+%% the author to define a more concise list
+%% of authors' names for this purpose.
+\renewcommand{\shortauthors}{Ahrens et al.}
+
+%%
+%% The abstract is a short summary of the work to be presented in the
+%% article.
 \begin{abstract}
@@
 \end{abstract}
+
+%%
+%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
+%% Please copy and paste the code instead of the example below.
+%%
+\begin{CCSXML}
+<ccs2012>
+ <concept>
+  <concept_id>00000000.0000000.0000000</concept_id>
+  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_significance>500</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>00000000.00000000.00000000</concept_id>
+  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_significance>300</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>00000000.00000000.00000000</concept_id>
+  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+ <concept>
+  <concept_id>00000000.00000000.00000000</concept_id>
+  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_significance>100</concept_significance>
+ </concept>
+</ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
+
+%%
+%% Keywords. The author(s) should pick words that accurately describe
+%% the work being presented. Separate the keywords with commas.
+\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
+  Your, Paper}
+
+\received{20 February 2007}
+\received[revised]{12 March 2009}
+\received[accepted]{5 June 2009}
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
+\maketitle
 
@@
 
-\bibliographystyle{alpha}
-\bibliography{bibliography}
+%%
+%% The acknowledgments section is defined using the "acks" environment
+%% (and NOT an unnumbered section). This ensures the proper
+%% identification of the section in the article metadata, and the
+%% consistent spelling of the heading.
+\begin{acks}
+    To Mateusz, Hameer, and Jaeyeon for their excellent programming contributions to the Finch codebase.
+\end{acks}
+
+%%
+%% The next two lines define the bibliography style to be used, and
+%% the bibliography file.
+\bibliographystyle{ACM-Reference-Format}
+\bibliography{bibliography.bib}
+
+%%
+%% If your work has an appendix, this is the place to put it.
+\appendix
 
 \end{document}
+\endinput

### 2024-03-15 main.tex
appears fixed

@@
 
-% Set page size and margins
-% Replace `letterpaper' with `a4paper' for UK/EU standard size
-\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
-
 % Useful packages
@@
 \usepackage{graphicx}
-\usepackage{amssymb}
+%\usepackage{amssymb}
 \usepackage{array}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+%\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+
+\newcommand\teo[1]{\textcolor{red}{teo:#1}}
 
@@
 
+\section{Bridging Looplets and Finch: The Tensor Interface}
+\teo{I am doing stuff in this section.}
+\input{sections/tensors.tex}
+
+\subsection{Wrapper Tensors}
+
+\subsection{Scalars}
+
+\subsubsection{Sparse Scalars}
+\subsubsection{Early Break Scalars}
+
 \section{The Finch Compiler}
@@
 
-\section{The Tensor Interface}
-
-\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
-
-\subsection{Level Abstraction}
-
-\begin{enumerate}
-\item fibers
-\item assembly
-\item reassembly
-\end{enumerate}
-
-\subsection{Core Level Langage Primitives}
-\begin{enumerate}
-\item SparseList
-\item SparseDict
-\item ...
-\end{enumerate}
-
-\subsection{Wrapper Tensors}
-
-\subsection{Scalars}
-
-\subsubsection{Sparse Scalars}
-\subsubsection{Early Break Scalars}
-
-\section{The Finch High-Level API (Needs a Name)}
-
-\subsection{Finch Logic}
-
-\subsection{Finch Interpreter}
-
-\subsection{Lowering}
-\subsubsection{Heuristic Optimization}
-
 \section{Evaluation}
@@
 
-\subsubsection{High-level kernel fusion}
+\subsection{Implementing Numpy Array API in Finch}
+\subsubsection{The Finch High-Level API (Needs a Name)}
+
+\subsubsection{Finch Logic}
+
+\subsubsection{Finch Interpreter}
+
+\subsubsection{Lowering}
+\subsubsection{Heuristic Optimization}
+
 Find an example where fusing the python interface gives a big speedup over non-fused kernels.
+
+%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
+%in-place.
+%dot((v^t .* u), w)) vs. 
+%(v^t .* dot(u, w))
 

### 2024-03-18 main.tex
Update on Overleaf.

@@
 \usepackage{array}
+\usepackage{minted}
+
+% Configuration for minted package
+\setminted{
+    style=colorful, % Set code style to colorful
+    fontsize=\scriptsize,
+    linenos, % Enable line numbers
+    breaklines, % Enable line breaks
+    breakanywhere, % Allow line breaks anywhere
+    autogobble,
+}
 %\usepackage[colorlinks=true, allcolors=blue]{hyperref}
@@
 
-\author{Teo Collins}
+\author{Teodoro Fields Collin}
 \affiliation{%
@@
 %
-%- We’re bringing these together, both need to work together.
+%- We’re bringing these together, because if both of these can work together, we can more efficiently search the space of programs
+
+\textbf{Finch makes high-performance computing easier with compilers that adapt programs to the structure of data}.
+Currently, experts must hand write variations of even the simplest kernels, like matrix multiply, for each data structure and architecture to get performance.
+This makes performance engineering an expensive, repetitive task.
+Several popular programs compute the same thing on different data.
+
+\textbf{Unstructured data is the easiest way to program, but the world is full of structured data}.
+Multidimensional dense (unstructured) arrays are the most important abstraction in computer science.
+Since their introduction in the 1950s with Fortran and Lisp, the simplicity of dense arrays has enabled extensive innovation, from compiler optimizations to convenience layers.
+However, structure is everywhere.
+Sparse arrays (which store only nonzero elements) can describe networks and simulations.
+Run-length encoding describes images, geometry and databases (such as a list of transactions with the date field all the same).
+Symmetry, bands, padding, and blocks are all natural properties which arise in the course of scientific computing, simulating the world around us.
+
+\textbf{Structured data support is fragmented and incomplete}.
+Programmers today must choose between program flexibility and datastructure flexibility.
+Hand-written solutions are collected in diverse libraries like MKL, OpenCV, or SciPy.
+However, libraries will only ever support a subset of programs on a subset of datastructure combinations.
+Special-purpose compilers like TACO or CoRa support a select subset of data structures (only sparse, or only ragged arrays) in highly constrained classes of programs such as tensor contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}.
+
+\textbf{Finch handles the two main challenges for structured array data}:
+\begin{itemize}
+\item \textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory.
+For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value.
+These zero regions or repeated regions are optimization opportunities.
+Here, we show how Finch combines programs with datastructure descriptions to generate efficient, datastructure-specific code.
+
+\item \textbf{Performance is highly structure dependent: Structure aware kernels are dramatically faster than unstructured ones, but the landscape of implementation decisions is also dramatically unpredictable}.
+For example, sparse kernels don’t need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
+Finch allows you to search it more efficiently.
+\end{itemize}
+
+\subsection{Finch: Compiling Structured Data}
+The finch array programming language adapts complex programs to array datastructures to produce efficient code.
+Array programming is the core abstraction behind many of the modern miracles of computing (e.g.
+neural networks, scientific simulation, database processing).
+Finch supports \textit{both} flexible programming constructs and diverse data structures.
+Finch supports loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+
+Crucially, Finch is able to adapt simple, unstructured programs to the structure of the input data.
+It does this using Looplets, a new language that decomposes datastructure iterators hierarchically.
+Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program.
+Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
+can be implemented using simple compiler passes like term rewriting and constant propagation.
+
 \subsection{Contributions}
@@
 \begin{enumerate}
+\item More complex array structures than ever before. A complete level-by-level
+structure-description language for expressing the structure of data
+hierarchically. The first such set of formats to efficiently capture banded,
+triangular, run-length-encoded, or sparse datasets, and any combination thereof.
 \item A rich structured array programming language with for-loops
@@
 structured data, as well as complex accesses such as affine indexing or scatter/gather.
-\item More complex array structures than ever before. A complete level-by-level
-structure-description language for expressing the structure of data
-hierarchically. The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
-\item The Finch compiler specializes programs to data structures in a
-predictable, deterministic approach, making it easier to search the complex
-space of programs and datastructures to find an appropriate fit for a given
-application. The tensor interface makes it easy to extend Finch to new level
-formats. A unique tensor lifecycle model enables polymorphism by analyzing the
-appropriate stages to insert simple, overloadable, interface functions such as
+\item A compiler that specializes programs to data structures 
+predictably, facilitating an expressive language that makes it easier to search the complex
+space of algorithms and datastructures.
+It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
 initialization or finalization.
-
 \item We evaluate the productivity of our language in several case studies,
 showing that Finch can be used to accelerate a wide range of applications, 
-from classic operations such as spmv and spgemm, to more complex applications such as image processing and graph analytics.
-We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
-Additionally, as a case study, a high-level array programming language and fusion interface for
-operations such as map, broadcast, or reduce that can be compiled to efficient
-code using the previous loop-level abstractions.
+from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
+%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
+Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
 %\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
@@
 
+\begin{figure}
+	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
+    \caption{Performance of Finch on erosion task (4 iterations).}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
+    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension.
+Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense.
+Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
+    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
+\end{figure}
+
 Examples that demonstrate performance engineering in a datastructure-driven model
@@
 
-\subsubsection{SparseList Format}
-\subsubsection{SparseVBL Format}
-The SparseVBL format is similar to the SparseList format, but contiguous fibers are stored together as blocks.
-Thus, it is most worthwhile to use when the matrix has a blocked structure—i.e.
-where there tend to be stretches of consecutive nonzero values.
-We found that the SparseVBL format performed the best for matrices with a clear block structure (list specific matrices) or an imperfect blocked structure (list specific matrices).
-\subsubsection{Pattern Format}
+% TODO: make this be 3 columns
+\begin{figure}
+    \begin{minipage}[t]{0.315\textwidth}
+        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \begin{minted}{julia}
+            @finch begin
+                y .= 0
+                for j = _, i = _
+                    y[i] += A[i, j] * x[j]
+                end
+                return y
+            end
+        \end{minted}
+    \end{minipage}%
+    \begin{minipage}[t]{0.315\textwidth}
+        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \begin{minted}{julia}
+            @finch begin
+                y .= 0
+                for j = _, i = _
+                    y[j] += A[i, j] * x[i]
+                end
+                return y
+            end
+        \end{minted}
+    \end{minipage}
+    \begin{minipage}[t]{0.36\textwidth}
+        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \begin{minted}{julia}
+            @finch begin
+                y .= 0
+                for j = _
+                    let x_j = x[j]
+                        y_j .= 0
+                        for i = _
+                            let A_ij = A[i, j]
+                                y[i] += x_j * A_ij
+                                y_j[] += A_ij * x[i]
+                            end
+                        end
+                        y[j] += y_j[] + diag[j] * x_j
+                    end
+                end
+                return y
+            end
+        \end{minted}
+    \end{minipage}
+    \caption{Performance of SpMV across various tools.}
+\end{figure}
+
+\subsubsection{Tensor Formats}
+We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix.
+To represent scalar values, we benchmarked with both the Element and Pattern levels.
+The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices.
+The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.
+
+We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
+Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
+Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.
+
+There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean.
+The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.
+
 \subsubsection{Symmetric SpMV}
@@
 \begin{figure}
+	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
+    \caption{Performance of Finch on erosion task (2 iterations).}
+\end{figure}
+
+\begin{figure}
 	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
@@
 
-\begin{figure}
-	\includegraphics[width=\linewidth]{fill_speedup_over_opencv.png}
-    \caption{Performance of Finch on flood fill task.}
-\end{figure}
-
 \subsubsection{Graph Analytics}
@@
 \end{figure}
+
+Push-Pull BFS in Finch:
+\begin{minted}{julia}
+V = Tensor(Dense(Element(false)))
+P = Tensor(Dense(Element(0)))
+F = Tensor(SparseByteMap(Pattern()))
+_F = Tensor(SparseByteMap(Pattern()))
+A = Tensor(Dense(SparseList(Pattern())))
+AT = Tensor(Dense(SparseList(Pattern())))
+
+function finch_bfs_push_kernel(_F, F, A, V, P)
+    @finch begin
+        _F .= false
+        for j=_, k=_
+            if F[j] && A[k, j] && !(V[k])
+                _F[k] |= true
+                P[k] <<choose(0)>>= j #Only set the parent for this vertex
+            end
+        end
+        return _F
+    end
+end
+
+function finch_bfs_pull_kernel(_F, F, AT, V, P)
+    p = ShortCircuitScalar{0}()
+    @finch begin
+        _F .= false
+        for k=_
+            if !V[k]
+                p .= 0
+                for j=_
+                    if F[follow(j)] && AT[j, k]
+                        p[] <<choose(0)>>= j #Only set the parent for this vertex
+                    end
+                end
+                if p[] != 0
+                    _F[k] |= true
+                    P[k] = p[]
+                end
+            end
+        end
+        return _F
+    end
+end)
+\end{minted}
 

### 2024-03-18 sections/tensors.tex
Update on Overleaf.

@@
 The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
+%
+At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
+%
+First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is much easier to manage when considering a single dimensional iterator. 
+%
+We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
+%
+Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
+%
+So we require a bridge that can provide different iteration orders depending on the context.
+%
+Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.
 
+To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
 %
-As we will see
-\begin{enumerate}
-    \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
-    \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
-    \item how do we ask for the right iterator and how do we change/manage their state?
-\end{enumerate}
-Need to Make these problems clear by this point, hope that read anticipate them.
+Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
+%
+We recapitulate the essential details here.
+%
+Our second abstraction adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation and limiting the amount of specific work.
+%
+Previous efforts to compiler a greater variety of sparse tensor programs left this bridge untouched ~\cite{henry2021compilation, won2023unified}.
+%
+Our third abstraction limits our set of sparse tensor programs to accommodate a natural restriction on when multi-dimensional data structures can be modified.
+%
+% As we will see
+% \begin{enumerate}
+%     \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
+%     \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
+%     \item how do we ask for the right iterator and how do we change/manage their state?
+% \end{enumerate}
+% Need to Make these problems clear by this point, hope that read anticipate them.
 
-Solutions:
-\begin{enumerate}
-    \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
-    \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
-\end{enumerate}
-
-\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+% Solutions:
+% \begin{enumerate}
+%     \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
+%     \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
+% \end{enumerate}
 
@@
 
-\subsection{Core Level Langage Primitives}
+\subsection{Instatiation Interface}
+
+\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+
+\subsection{Core Level Language Primitives}
 \begin{enumerate}

### 2024-03-18 main.tex
fix

@@
 
+Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
+\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
+Arrays are also universal across programming languages, from their introduction
+in Fortran in 1957 to present-day languages like Python
+\cite{backus_fortran_1957}, keeping more-or-less the same semantics. Modern
+array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
+and Halide have pushed the limits of productive data processing with arrays,
+fueling breakthroughs in machine learning, scientific computing, image
+processing, and more.  \cite{harris_array_2020, virtanen_scipy_2020,
+moler_history_2020, abadi_tensorflow_2016,
+paszke_pytorch_2019,ragan-kelley_halide_2013}.  These frameworks have been the
+subject of extensive industry investment to enable performant implementations,
+and often operate at the peak capacity of the hardware they run on
+\cite{lo_roofline_2015}.
+
+The success and ubiquity of arrays is likely due to their simplicity. Since
+their introduction, multidimensional arrays have represented dense, rectilinear,
+integer grids of points. By \textbf{dense}, we mean that every entry the array
+represents is stored faithfully at a location determined by a simple row or
+column-major mapping from multidimensional space to linear memory. This
+simplicity enables extensive interoperability, convenience layers, and
+optimizations by breaking the abstraction barrier between array representation
+and array storage.  Compilers understand dense array computations across many
+programming constructs, such as for and while loops, breaks, parallelism,
+caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
+loop-carry-dependencies, and more. Several optimizations have been developed for
+dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
+interchange.  However, while dense arrays are the easiest way to program, the
+world is not all dense.
+
+Our world is full of structured arrays.  Sparse arrays (which store only nonzero
+elements) can describe networks and simulations. Run-length encoding describes
+images and masks, geometry, and databases (such as a list of transactions with
+the date field all the same). Symmetry, bands, padding, and blocks are all
+natural matrix properties which arise in the course of scientific computing or
+simulations of the natural world. Operators, such as convolution, can be
+expressed as structured arrays. For example, a convolution with a filter can be
+expressed as a matrix multiplication with the toeplitz matrix of all the
+circular shifts of the filter.
+
+%Sparse arrays
+% - Networks (graphblas, graph frameworks)
+% - Simulations (BLAS, symblas, banded blas, etc)
+% - Databases (database engines)
+%Run-length encoding
+%Symmetry
+%Bands
+%Padding
+%Blocks
+
+%Bring in structure
+% - old arrays are Dense, rectilinear, integer grids
+% - The world is not like that
+% - Sparsity, runs of repeated values, symmetry
+% - Lots of citations, perhaps a few figures from my powerpoint
+
+%Existing solutions do not scale to the complexity of the problem
+% Must choose between program flexibility and datastructure flexibility
+% Dense arrays have tons of programming constructs
+% Sparse array frameworks have very limited functionality, discuss a few key related works
+
+% Finch Supports both! The combination of the two leads us to a new data-structure driven programming model.
+% In this model, we can search the space of programs more efficiently, and we can more efficiently search the space of datastructures.
+% Programs are specified separately from datastructures.
+% Certain concepts are expressed in programs, others in data
+% At a high level, whenever we need to match corresponding keys in a program, those keys become array indices, and the datastructure becomes a sparse array
+
 %- Fortran supported lots of control structures, no data structures, just dense arrays.
@@
 Several popular programs compute the same thing on different data.
-
-\textbf{Unstructured data is the easiest way to program, but the world is full of structured data}.
-Multidimensional dense (unstructured) arrays are the most important abstraction in computer science.
-Since their introduction in the 1950s with Fortran and Lisp, the simplicity of dense arrays has enabled extensive innovation, from compiler optimizations to convenience layers.
-However, structure is everywhere.
-Sparse arrays (which store only nonzero elements) can describe networks and simulations.
-Run-length encoding describes images, geometry and databases (such as a list of transactions with the date field all the same).
-Symmetry, bands, padding, and blocks are all natural properties which arise in the course of scientific computing, simulating the world around us.
+\cite{ahrens_looplets_2023}.
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{bibliography.bib}
+\bibliography{FinchOOPSLA.bib}
 
@@
 
+\section Frequently Requested Finch kernels
+
 \end{document}

### 2024-03-19 main.tex
add suitesparse bfs and intro

@@
 \newcommand\teo[1]{\textcolor{red}{teo:#1}}
+\newcommand\help[1]{\textcolor{green}{Help Needed:#1}}
 
@@
 
+%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
 Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
@@
 
-Our world is full of structured arrays.  Sparse arrays (which store only nonzero
-elements) can describe networks and simulations. Run-length encoding describes
-images and masks, geometry, and databases (such as a list of transactions with
-the date field all the same). Symmetry, bands, padding, and blocks are all
-natural matrix properties which arise in the course of scientific computing or
-simulations of the natural world. Operators, such as convolution, can be
-expressed as structured arrays. For example, a convolution with a filter can be
-expressed as a matrix multiplication with the toeplitz matrix of all the
-circular shifts of the filter.
-
+%Bring in structure
+% - old arrays are Dense, rectilinear, integer grids
+% - The world is not like that
+% - Sparsity, runs of repeated values, symmetry
+% - Lots of citations, perhaps a few figures from my powerpoint
 %Sparse arrays
@@
 %Blocks
+\help{This section needs citations, please help expand it and make it more
+convincing. A figure could be nice.} Our world is full of structured arrays.
+Sparse arrays (which store only nonzero elements) describe networks and
+simulations. Run-length encoding describes images and masks, geometry, and
+databases (such as a list of transactions with the date field all the same)
+Symmetry, bands, padding, and blocks are all natural matrix properties which
+arise in the course of scientific computing or simulations of the natural world.
+Operators, such as convolution, can be expressed as structured arrays. For
+example, a convolution with a filter can be expressed as a matrix multiplication
+with the toeplitz matrix of all the circular shifts of the filter.
 
-%Bring in structure
-% - old arrays are Dense, rectilinear, integer grids
-% - The world is not like that
-% - Sparsity, runs of repeated values, symmetry
-% - Lots of citations, perhaps a few figures from my powerpoint
+\help{This section needs citations}
+\textbf{Currently, support for structured data is fragmented and incomplete}.
+Experts must hand write variations of even the simplest kernels, like matrix
+multiply, for each data structure and architecture to get performance.
+Implementations must choose a small set of features to support well, resulting
+in a compromise between \textbf{program flexibility} and \textbf{datastructure
+flexibility}. Hand-written solutions are collected in diverse libraries like
+MKL, OpenCV, or SciPy. However, libraries will only ever support a subset of
+programs on a subset of datastructure combinations. Even the most advanced
+libraries, such as the GraphBLAS, which support a wide variety of sparse
+operations over various semirings always lack support for other features, such
+as tensors, fused outputs, or runs of repeated values. While dense array
+compilers support an enormous variety of program constructs like early break and
+multiple left hand sides, they only support dense arrays.  Special-purpose
+compilers like TACO or CoRa which support a select subset of structured data
+structures (only sparse, or only ragged arrays) must compromise by greatly
+constraining the classes of programs which they support, such as tensor
+contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. This trade-off is
+visualized in figure \ref{fig:datavsprogram}.
 
-%Existing solutions do not scale to the complexity of the problem
-% Must choose between program flexibility and datastructure flexibility
-% Dense arrays have tons of programming constructs
-% Sparse array frameworks have very limited functionality, discuss a few key related works
+\begin{table}[h!]
+  \help{This table needs citations, as well as a little bit of double checking}
+  \centering
+  \begin{tabular}{l|ccccc}
+  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
+  \hline
+  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  High-Level API           &            & \checkmark &            &            & \checkmark \\
+  Automatic API Fusion     &            &            &            &            & \checkmark \\
+%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
+  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
+  Recurrence               & \checkmark &            &            &            &            \\
+  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
+  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
+  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
+  \end{tabular}
+  \caption{Feature support across various tools.}
+  \label{tab:features}
+  \end{table}
 
-% Finch Supports both! The combination of the two leads us to a new data-structure driven programming model.
-% In this model, we can search the space of programs more efficiently, and we can more efficiently search the space of datastructures.
-% Programs are specified separately from datastructures.
-% Certain concepts are expressed in programs, others in data
-% At a high level, whenever we need to match corresponding keys in a program, those keys become array indices, and the datastructure becomes a sparse array
+  \begin{table}[h!]
+  \centering
+  \begin{tabular}{l|ccccc}
+  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
+  \hline
+  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  Padded                   & \checkmark &            &            &            & \checkmark \\
+  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
+  Sparse                   &            & \checkmark &            &            & \checkmark \\
+  Run-length               &            &            &            &            & \checkmark \\
+  Symmetric                &            &            &            &            & \checkmark \\
+  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
+  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
+  Ragged                   &            &            & \checkmark &            & \checkmark \\
+  \end{tabular}
+  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
+  \label{tab:data_structures}
+\end{table}
 
-%- Fortran supported lots of control structures, no data structures, just dense arrays.
-%
-%- We tried to emulate complex data using only dense arrays but with complicated control structures.
-%
-%- Recently, people tried to build frameworks to support structured data, but gave up a lot of program side
-%
-%- We’re bringing these together, because if both of these can work together, we can more efficiently search the space of programs
+We see two main challenges to writing efficient code over structured data.
 
-\textbf{Finch makes high-performance computing easier with compilers that adapt programs to the structure of data}.
-Currently, experts must hand write variations of even the simplest kernels, like matrix multiply, for each data structure and architecture to get performance.
-This makes performance engineering an expensive, repetitive task.
-Several popular programs compute the same thing on different data.
-\cite{ahrens_looplets_2023}.
+\textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory.
+For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value.
+These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.
 
-\textbf{Structured data support is fragmented and incomplete}.
-Programmers today must choose between program flexibility and datastructure flexibility.
-Hand-written solutions are collected in diverse libraries like MKL, OpenCV, or SciPy.
-However, libraries will only ever support a subset of programs on a subset of datastructure combinations.
-Special-purpose compilers like TACO or CoRa support a select subset of data structures (only sparse, or only ragged arrays) in highly constrained classes of programs such as tensor contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}.
+\textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable.
+For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
 
-\textbf{Finch handles the two main challenges for structured array data}:
-\begin{itemize}
-\item \textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory.
-For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value.
-These zero regions or repeated regions are optimization opportunities.
-Here, we show how Finch combines programs with datastructure descriptions to generate efficient, datastructure-specific code.
+In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.
+In this model, performance engineers can more efficiently search the
+complex landscape of programs and datastructures to find the best implementation.
+In this programming model, we can express certain concepts in programs, and others in data.
+For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
+While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
 
-\item \textbf{Performance is highly structure dependent: Structure aware kernels are dramatically faster than unstructured ones, but the landscape of implementation decisions is also dramatically unpredictable}.
-For example, sparse kernels don’t need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
-Finch allows you to search it more efficiently.
-\end{itemize}
-
-\subsection{Finch: Compiling Structured Data}
-The finch array programming language adapts complex programs to array datastructures to produce efficient code.
-Array programming is the core abstraction behind many of the modern miracles of computing (e.g.
-neural networks, scientific simulation, database processing).
-Finch supports \textit{both} flexible programming constructs and diverse data structures.
-Finch supports loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
-
-Crucially, Finch is able to adapt simple, unstructured programs to the structure of the input data.
-It does this using Looplets, a new language that decomposes datastructure iterators hierarchically.
+We develop a new array programming language, Finch, which supports \textit{both}
+flexible programming constructs and diverse data structures. Finch supports a
+familiar programming language of loops, statements, if conditions, breaks, etc,
+over a wide variety of array structures, such as sparsity, run-length-encoding,
+symmetry, triangles, padding, or blocks. The Finch compiler uses the structure
+of the data to generate efficient implementations of these programs.
+%It does this using Looplets, a new language that decomposes datastructure iterators hierarchically.
 Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program.
@@
 \end{enumerate}
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{l|ccccc}
-\textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
-\hline
-Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-Multiple LHS            & \checkmark &            & \checkmark & \checkmark & \checkmark \\
-Affine Indices          & \checkmark &            &            & \checkmark & \checkmark \\
-Recurrence              & \checkmark &            &            &            &            \\
-If-Conditions and Masks & \checkmark & \checkmark &            & \checkmark & \checkmark \\
-Scatter Gather          & \checkmark &            &            & \checkmark & \checkmark \\
-Early Break             &            & \checkmark &            &   \checkmark         & \checkmark \\
-\end{tabular}
-\caption{Feature support across various tools.}
-\label{tab:features}
-\end{table}
-
-\begin{table}[h!]
-\centering
-\begin{tabular}{l|ccccc}
-\textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
-\hline
-Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-Padded                   & \checkmark &            &            &            & \checkmark \\
-One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
-Sparse                   &            & \checkmark &            &            & \checkmark \\
-Run-length               &            &            &            &            & \checkmark \\
-Symmetric                &            &            &            &            & \checkmark \\
-Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
-Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
-Ragged                   &            &            & \checkmark &            & \checkmark \\
-\end{tabular}
-\caption{Support for various data structures across tools.}
-\label{tab:data_structures}
-\end{table}
 

### 2024-03-19 main.tex
Data goes first
Update on Overleaf.

@@
 \cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
+%
 Arrays are also universal across programming languages, from their introduction
 in Fortran in 1957 to present-day languages like Python
-\cite{backus_fortran_1957}, keeping more-or-less the same semantics. Modern
+\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
+%
+Modern
 array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
@@
 fueling breakthroughs in machine learning, scientific computing, image
-processing, and more.  \cite{harris_array_2020, virtanen_scipy_2020,
+processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
 moler_history_2020, abadi_tensorflow_2016,
-paszke_pytorch_2019,ragan-kelley_halide_2013}.  These frameworks have been the
+paszke_pytorch_2019,ragan-kelley_halide_2013}.
+%
+These frameworks have been the
 subject of extensive industry investment to enable performant implementations,
@@
 
-The success and ubiquity of arrays is likely due to their simplicity. Since
+The success and ubiquity of arrays is likely due to their simplicity. 
+%
+Since
 their introduction, multidimensional arrays have represented dense, rectilinear,
-integer grids of points. By \textbf{dense}, we mean that every entry the array
+integer grids of points. 
+%
+By \textbf{dense}, we mean that every entry the array
 represents is stored faithfully at a location determined by a simple row or
-column-major mapping from multidimensional space to linear memory. This
+column-major mapping from multidimensional space to linear memory.
+%
+This
 simplicity enables extensive interoperability, convenience layers, and
 optimizations by breaking the abstraction barrier between array representation
-and array storage.  Compilers understand dense array computations across many
+and array storage.  
+%
+Compilers understand dense array computations across many
 programming constructs, such as for and while loops, breaks, parallelism,
@@
 
-\section{The Finch Language}
-
-\subsection{Syntax and Semantics}
-
 \section{Bridging Looplets and Finch: The Tensor Interface}
@@
 \subsubsection{Early Break Scalars}
+
+\section{The Finch Language}
+
+\subsection{Syntax and Semantics}
 

### 2024-03-19 main.tex
Update on Overleaf.

@@
 %
-By \textbf{dense}, we mean that every entry the array
-represents is stored faithfully at a location determined by a simple row or
-column-major mapping from multidimensional space to linear memory.
+By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
 %
@@
 dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
-interchange.  However, while dense arrays are the easiest way to program, the
-world is not all dense.
+interchange.
+%
+However, while dense arrays are the easiest way to program, the world is not all dense.
 
@@
 \help{This section needs citations, please help expand it and make it more
-convincing. A figure could be nice.} Our world is full of structured arrays.
+convincing. A figure could be nice.} 
+\begin{figure}
+	\includegraphics[width=\linewidth]{example_structures.png}
+    \caption{some example structures that might be nice to put into a figure}
+\end{figure}
+Our world is full of structured arrays.
+%
+
 Sparse arrays (which store only nonzero elements) describe networks and
-simulations. Run-length encoding describes images and masks, geometry, and
+simulations.
+%
+Run-length encoding describes images and masks, geometry, and
 databases (such as a list of transactions with the date field all the same)
@@
 arise in the course of scientific computing or simulations of the natural world.
-Operators, such as convolution, can be expressed as structured arrays. For
+%
+
+Operators, such as convolution, can be expressed as structured arrays.
+%
+For
 example, a convolution with a filter can be expressed as a matrix multiplication
@@
 \textbf{Currently, support for structured data is fragmented and incomplete}.
+%
 Experts must hand write variations of even the simplest kernels, like matrix
 multiply, for each data structure and architecture to get performance.
+%
 Implementations must choose a small set of features to support well, resulting
 in a compromise between \textbf{program flexibility} and \textbf{datastructure
-flexibility}. Hand-written solutions are collected in diverse libraries like
-MKL, OpenCV, or SciPy. However, libraries will only ever support a subset of
+flexibility}.
+%
+Hand-written solutions are collected in diverse libraries like
+MKL, OpenCV, or SciPy. 
+%
+However, libraries will only ever support a subset of
 programs on a subset of datastructure combinations. Even the most advanced
@@
 operations over various semirings always lack support for other features, such
-as tensors, fused outputs, or runs of repeated values. While dense array
+as tensors, fused outputs, or runs of repeated values. 
+%
+While dense array
 compilers support an enormous variety of program constructs like early break and
-multiple left hand sides, they only support dense arrays.  Special-purpose
+multiple left hand sides, they only support dense arrays.  
+%
+Special-purpose
 compilers like TACO or CoRa which support a select subset of structured data
@@
 constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. This trade-off is
+contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. 
+%
+This trade-off is
 visualized in figure \ref{fig:datavsprogram}.
@@
 \subsection{Looplets}
+\begin{figure}
+	\includegraphics[width=\linewidth]{example_looplet_figure.png}
+    \caption{Looplets visualized, along with their semantics}
+\end{figure}
+
 \subsection{FiberTrees}

### 2024-03-19 main.tex
cool

@@
 of the data to generate efficient implementations of these programs.
-%It does this using Looplets, a new language that decomposes datastructure iterators hierarchically.
-Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program.
-Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
-can be implemented using simple compiler passes like term rewriting and constant propagation.
 
@@
 \subsection{Looplets}
+Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically.
+Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously.
+In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space.
+Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
+can be implemented using simple compiler passes like term rewriting and constant propagation.
+
+We begin a formal description of looplets here:
+
 \begin{figure}
@@
 
+\help{The main points of this section are:}
+
+\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}
+
+\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}
+
+\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}
+
+\help{4.
+Take a look at the figure and point out that Finch is competitive with TACO on inner and gustavson, and that for outer product we see what we would expect from TACO: i.e.
+Finch Outer with dense performs similarly to TACO with dense, and Finch Outer with sparse performs much better as the matrix size increases.}
+
+\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.
+
+\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
 \begin{figure}
-	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (4 iterations).}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
+	\includegraphics[width=\linewidth]{spgemm_small_speedup_log_scale.png}
     \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension.
@@
 
+\help{In this section, explain what erosion is, link a few cool images that show erosion, and then explain what the finch kernel looks like.
+in particular, be sure to point out that we're doing unrolled convolution, and how that maps to the generated code (merging shifted sparse iterators).}
+
+\help{Explain the role of formats in this kernel, how sparseRLE(PAttern) is really the right thing to use here}
+
+\help{Explain that finch can support a bitwise version, and that finch can mask the bitwise kernel too to get performance with a small change to the code}
+
+\help{Point out that masks with SparseRLE can sometimes perform better because they lift the masking out of the inner loop.
+Talk about hist and also the fact that Finch supports scatter, here and in spgemm.}
+
 \begin{figure}
@@
 \subsubsection{Graph Analytics}
+\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
+choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
+We may need to explain a little bit about what push pull is.
+For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
+
 \begin{figure}
@@
     end
-end)
+end
 \end{minted}

### 2024-03-19 main.tex
fix
Update on Overleaf.

--- 2024-03-19 main.tex
@@
 
-\newcommand\teo[1]{\textcolor{red}{teo:#1}}
-\newcommand\help[1]{\textcolor{green}{Help Needed:#1}}
+\newcommand\teo[1]{\textcolor{red}{teo: #1}}
+\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
+\newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
 
@@
 \maketitle
-
-\section{Introduction}
-
-%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
-Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
-\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
-%
-Arrays are also universal across programming languages, from their introduction
-in Fortran in 1957 to present-day languages like Python
-\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
-%
-Modern
-array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
-and Halide have pushed the limits of productive data processing with arrays,
-fueling breakthroughs in machine learning, scientific computing, image
-processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
-moler_history_2020, abadi_tensorflow_2016,
-paszke_pytorch_2019,ragan-kelley_halide_2013}.
-%
-These frameworks have been the
-subject of extensive industry investment to enable performant implementations,
-and often operate at the peak capacity of the hardware they run on
-\cite{lo_roofline_2015}.
-
-The success and ubiquity of arrays is likely due to their simplicity. 
-%
-Since
-their introduction, multidimensional arrays have represented dense, rectilinear,
-integer grids of points. 
-%
-By \textbf{dense}, we mean that every entry the array
-represents is stored faithfully at a location determined by a simple row or
-column-major mapping from multidimensional space to linear memory.
-%
-This
-simplicity enables extensive interoperability, convenience layers, and
-optimizations by breaking the abstraction barrier between array representation
-and array storage.  
-%
-Compilers understand dense array computations across many
-programming constructs, such as for and while loops, breaks, parallelism,
-caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
-loop-carry-dependencies, and more. Several optimizations have been developed for
-dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
-interchange.  However, while dense arrays are the easiest way to program, the
-world is not all dense.
-
-%Bring in structure
-% - old arrays are Dense, rectilinear, integer grids
-% - The world is not like that
-% - Sparsity, runs of repeated values, symmetry
-% - Lots of citations, perhaps a few figures from my powerpoint
-%Sparse arrays
-% - Networks (graphblas, graph frameworks)
-% - Simulations (BLAS, symblas, banded blas, etc)
-% - Databases (database engines)
-%Run-length encoding
-%Symmetry
-%Bands
-%Padding
-%Blocks
-\help{This section needs citations, please help expand it and make it more
-convincing. A figure could be nice.} Our world is full of structured arrays.
-Sparse arrays (which store only nonzero elements) describe networks and
-simulations. Run-length encoding describes images and masks, geometry, and
-databases (such as a list of transactions with the date field all the same)
-Symmetry, bands, padding, and blocks are all natural matrix properties which
-arise in the course of scientific computing or simulations of the natural world.
-Operators, such as convolution, can be expressed as structured arrays. For
-example, a convolution with a filter can be expressed as a matrix multiplication
-with the toeplitz matrix of all the circular shifts of the filter.
-
-\help{This section needs citations}
-\textbf{Currently, support for structured data is fragmented and incomplete}.
-Experts must hand write variations of even the simplest kernels, like matrix
-multiply, for each data structure and architecture to get performance.
-Implementations must choose a small set of features to support well, resulting
-in a compromise between \textbf{program flexibility} and \textbf{datastructure
-flexibility}. Hand-written solutions are collected in diverse libraries like
-MKL, OpenCV, or SciPy. However, libraries will only ever support a subset of
-programs on a subset of datastructure combinations. Even the most advanced
-libraries, such as the GraphBLAS, which support a wide variety of sparse
-operations over various semirings always lack support for other features, such
-as tensors, fused outputs, or runs of repeated values. While dense array
-compilers support an enormous variety of program constructs like early break and
-multiple left hand sides, they only support dense arrays.  Special-purpose
-compilers like TACO or CoRa which support a select subset of structured data
-structures (only sparse, or only ragged arrays) must compromise by greatly
-constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. This trade-off is
-visualized in figure \ref{fig:datavsprogram}.
-
-\begin{table}[h!]
-  \help{This table needs citations, as well as a little bit of double checking}
-  \centering
-  \begin{tabular}{l|ccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
-  \hline
-  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  High-Level API           &            & \checkmark &            &            & \checkmark \\
-  Automatic API Fusion     &            &            &            &            & \checkmark \\
-%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
-  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
-  Recurrence               & \checkmark &            &            &            &            \\
-  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
-  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
-  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
-  \end{tabular}
-  \caption{Feature support across various tools.}
-  \label{tab:features}
-  \end{table}
-
-  \begin{table}[h!]
-  \centering
-  \begin{tabular}{l|ccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
-  \hline
-  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  Padded                   & \checkmark &            &            &            & \checkmark \\
-  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
-  Sparse                   &            & \checkmark &            &            & \checkmark \\
-  Run-length               &            &            &            &            & \checkmark \\
-  Symmetric                &            &            &            &            & \checkmark \\
-  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
-  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
-  Ragged                   &            &            & \checkmark &            & \checkmark \\
-  \end{tabular}
-  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
-  \label{tab:data_structures}
-\end{table}
-
-We see two main challenges to writing efficient code over structured data.
-
-\textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory.
-For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value.
-These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.
-
-\textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable.
-For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
-
-In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.
-In this model, performance engineers can more efficiently search the
-complex landscape of programs and datastructures to find the best implementation.
-In this programming model, we can express certain concepts in programs, and others in data.
-For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
-While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
-
-We develop a new array programming language, Finch, which supports \textit{both}
-flexible programming constructs and diverse data structures. Finch supports a
-familiar programming language of loops, statements, if conditions, breaks, etc,
-over a wide variety of array structures, such as sparsity, run-length-encoding,
-symmetry, triangles, padding, or blocks. The Finch compiler uses the structure
-of the data to generate efficient implementations of these programs.
-%It does this using Looplets, a new language that decomposes datastructure iterators hierarchically.
-Looplets represent the control-flow structures needed to iterate over them simultaneously with other structures in the program.
-Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
-can be implemented using simple compiler passes like term rewriting and constant propagation.
-
-\subsection{Contributions}
-
-\begin{enumerate}
-\item More complex array structures than ever before. A complete level-by-level
-structure-description language for expressing the structure of data
-hierarchically. The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
-\item A rich structured array programming language with for-loops
-and complex control flow constructs at the same level of productivity
-of dense arrays. To our knowledge, the Finch programming language is the first
-to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather.
-\item A compiler that specializes programs to data structures 
-predictably, facilitating an expressive language that makes it easier to search the complex
-space of algorithms and datastructures.
-It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
-initialization or finalization.
-\item We evaluate the productivity of our language in several case studies,
-showing that Finch can be used to accelerate a wide range of applications, 
-from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
-%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
-Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
-%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
-The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets.
-The formats capture many use cases, from random updates to sequential construction.
-%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions.
-The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
-%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
-%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
-%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
-%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
-%\item <Performance Contributions>
-\end{enumerate}
-
-\section{Background}
-\subsection{Looplets}
-\subsection{FiberTrees}
-
-\subsection{Concordant Iteration}
-
-\subsection{Protocols}
-
-\section{Bridging Looplets and Finch: The Tensor Interface}
-\teo{I am doing stuff in this section.}
-\input{sections/tensors.tex}
-
-\subsection{Wrapper Tensors}
-
-\subsection{Scalars}
-
-\subsubsection{Sparse Scalars}
-\subsubsection{Early Break Scalars}
-
-\section{The Finch Language}
-
-\subsection{Syntax and Semantics}
-
-\section{The Finch Compiler}
-
-\subsection{Dimensionalization}
-
-\subsection{Concordization}
-
-\subsection{Bounds Analysis}
-
-\subsection{Performance Warnings}
-
-\subsection{Wrapperization}
-
-\subsection{Simplification and Algebraic Transformations}
-
-\section{Evaluation}
-
-\subsection{Data-Driven Performance Engineering}
-\subsubsection{Sparse-Sparse Matrix Multiply}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (4 iterations).}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
-    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension.
-Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense.
-Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
-    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
-\end{figure}
-
-Examples that demonstrate performance engineering in a datastructure-driven model
-
-\subsubsection{SpMV}
-Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data.
-More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats.
-Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
-% I'm not sure if the right place for this is here - maybe should come earlier?
-
-Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns.
-Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.
-
-The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks.
-The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
-In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
-
-We compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface [cite], Ahrens et al.
-to test a variable block row format partitioning strategy [cite], and Kjolstad et al.
-to test the TACO library [cite].
-In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures 1 and 2.
-
-% TODO: make this be 3 columns
-\begin{figure}
-    \begin{minipage}[t]{0.315\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            @finch begin
-                y .= 0
-                for j = _, i = _
-                    y[i] += A[i, j] * x[j]
-                end
-                return y
-            end
-        \end{minted}
-    \end{minipage}%
-    \begin{minipage}[t]{0.315\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            @finch begin
-                y .= 0
-                for j = _, i = _
-                    y[j] += A[i, j] * x[i]
-                end
-                return y
-            end
-        \end{minted}
-    \end{minipage}
-    \begin{minipage}[t]{0.36\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            @finch begin
-                y .= 0
-                for j = _
-                    let x_j = x[j]
-                        y_j .= 0
-                        for i = _
-                            let A_ij = A[i, j]
-                                y[i] += x_j * A_ij
-                                y_j[] += A_ij * x[i]
-                            end
-                        end
-                        y[j] += y_j[] + diag[j] * x_j
-                    end
-                end
-                return y
-            end
-        \end{minted}
-    \end{minipage}
-    \caption{Performance of SpMV across various tools.}
-\end{figure}
-
-\subsubsection{Tensor Formats}
-We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix.
-To represent scalar values, we benchmarked with both the Element and Pattern levels.
-The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices.
-The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.
-
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean.
-The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.
-
-\subsubsection{Symmetric SpMV}
-Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix.
-We can apply this strategy with any level format.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO .
-Notably, there is a 1.9x speedup for the HB/saylr4 matrix.
-\subsubsection{4D Blocked SpMV}
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \begin{minipage}[t]{0.5\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
-    \end{minipage}%
-    \begin{minipage}[t]{0.5\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
-    \end{minipage}
-    \caption{Performance of SpMV across various tools.}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
-    \caption{Performance of SpMV by Finch format.}
-\end{figure}
-
-\subsection{Programming over flexible data}
-
-\subsubsection{Image Morphology}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (2 iterations).}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (4 iterations).}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
-    \caption{Performance of Finch on masked histogram task.}
-\end{figure}
-
-\subsubsection{Graph Analytics}
-\begin{figure}
-	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
-	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \caption{Performance of graph apps across various tools.}
-\end{figure}
-
-Push-Pull BFS in Finch:
-\begin{minted}{julia}
-V = Tensor(Dense(Element(false)))
-P = Tensor(Dense(Element(0)))
-F = Tensor(SparseByteMap(Pattern()))
-_F = Tensor(SparseByteMap(Pattern()))
-A = Tensor(Dense(SparseList(Pattern())))
-AT = Tensor(Dense(SparseList(Pattern())))
-
-function finch_bfs_push_kernel(_F, F, A, V, P)
-    @finch begin
-        _F .= false
-        for j=_, k=_
-            if F[j] && A[k, j] && !(V[k])
-                _F[k] |= true
-                P[k] <<choose(0)>>= j #Only set the parent for this vertex
-            end
-        end
-        return _F
-    end
-end
-
-function finch_bfs_pull_kernel(_F, F, AT, V, P)
-    p = ShortCircuitScalar{0}()
-    @finch begin
-        _F .= false
-        for k=_
-            if !V[k]
-                p .= 0
-                for j=_
-                    if F[follow(j)] && AT[j, k]
-                        p[] <<choose(0)>>= j #Only set the parent for this vertex
-                    end
-                end
-                if p[] != 0
-                    _F[k] |= true
-                    P[k] = p[]
-                end
-            end
-        end
-        return _F
-    end
-end)
-\end{minted}
-
-\subsection{Implementing Numpy Array API in Finch}
-\subsubsection{The Finch High-Level API (Needs a Name)}
-
-\subsubsection{Finch Logic}
-
-\subsubsection{Finch Interpreter}
-
-\subsubsection{Lowering}
-\subsubsection{Heuristic Optimization}
-
-Find an example where fusing the python interface gives a big speedup over non-fused kernels.
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
+\input{sections/intro}
+\input{sections/background}
+\input{sections/lang}
+\input{sections/tensors}
+\input{sections/compiler}
+\input{sections/eval}
 %%
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{FinchOOPSLA.bib}
+\bibliography{FinchOOPSLA.bib, FinchOOPSLAOverleaf.bib}
 

### 2024-03-19 sections/background.tex
Update on Overleaf.

\section{Background}
\subsection{Looplets}
Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space. Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc. can be implemented using simple compiler passes like term rewriting and constant propagation.

We begin a formal description of looplets here:

\begin{figure}
	\includegraphics[width=\linewidth]{example_looplet_figure.png}
    \caption{Looplets visualized, along with their semantics}
\end{figure}

\subsection{FiberTrees}

\subsection{Concordant Iteration}

\subsection{Protocols}

### 2024-03-19 sections/compiler.tex
Update on Overleaf.

\section{The Finch Compiler}

\subsection{Dimensionalization}

\subsection{Concordization}

\subsection{Bounds Analysis}

\subsection{Performance Warnings}

\subsection{Wrapperization}

\subsection{Simplification and Algebraic Transformations}

### 2024-03-19 sections/eval.tex
Update on Overleaf.

\section{Evaluation}

\subsection{Data-Driven Performance Engineering}
\subsubsection{Sparse-Sparse Matrix Multiply}

\help{The main points of this section are:}

\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}

\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}

\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}

\help{4. Take a look at the figure and point out that Finch is competitive with TACO on inner and gustavson, and that for outer product we see what we would expect from TACO: i.e. Finch Outer with dense performs similarly to TACO with dense, and Finch Outer with sparse performs much better as the matrix size increases.}

\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.

\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_small_speedup_log_scale.png}
    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension. Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense. Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
\end{figure}

Examples that demonstrate performance engineering in a datastructure-driven model

\subsubsection{SpMV}
Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data. More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats. Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix. % I'm not sure if the right place for this is here - maybe should come earlier?

Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns. Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.

The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks. The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats. In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.

We compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.  We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al. to test the OSKI interface [cite], Ahrens et al. to test a variable block row format partitioning strategy [cite], and Kjolstad et al. to test the TACO library [cite]. In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases. The dense vector is randomly generated. We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures 1 and 2.

% TODO: make this be 3 columns
\begin{figure}
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[i] += A[i, j] * x[j]
                end
                return y
            end
        \end{minted}
    \end{minipage}%
    \begin{minipage}[t]{0.315\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _, i = _
                    y[j] += A[i, j] * x[i]
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \begin{minipage}[t]{0.36\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \begin{minted}{julia}
            @finch begin
                y .= 0
                for j = _
                    let x_j = x[j]
                        y_j .= 0
                        for i = _
                            let A_ij = A[i, j]
                                y[i] += x_j * A_ij
                                y_j[] += A_ij * x[i]
                            end
                        end
                        y[j] += y_j[] + diag[j] * x_j
                    end
                end
                return y
            end
        \end{minted}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
\end{figure}

\subsubsection{Tensor Formats}
We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix. To represent scalar values, we benchmarked with both the Element and Pattern levels. The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices. The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.

We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor. Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation. Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.

There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean. The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.

\subsubsection{Symmetric SpMV}
Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix. We can apply this strategy with any level format. Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry. However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage. Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO . Notably, there is a 1.9x speedup for the HB/saylr4 matrix. 
\subsubsection{4D Blocked SpMV}

%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png

\begin{figure}
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
    \end{minipage}%
    \begin{minipage}[t]{0.5\textwidth}
        \vspace{0pt} % Add this to ensure top alignment within minipage
        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
    \end{minipage}
    \caption{Performance of SpMV across various tools.}
\end{figure}

\begin{figure}
    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
    \caption{Performance of SpMV by Finch format.}
\end{figure}

\subsection{Programming over flexible data}

\subsubsection{Image Morphology}

\help{In this section, explain what erosion is, link a few cool images that show erosion, and then explain what the finch kernel looks like. in particular, be sure to point out that we're doing unrolled convolution, and how that maps to the generated code (merging shifted sparse iterators).}

\help{Explain the role of formats in this kernel, how sparseRLE(PAttern) is really the right thing to use here}

\help{Explain that finch can support a bitwise version, and that finch can mask the bitwise kernel too to get performance with a small change to the code}

\help{Point out that masks with SparseRLE can sometimes perform better because they lift the masking out of the inner loop. Talk about hist and also the fact that Finch supports scatter, here and in spgemm.}

\begin{figure}
	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (2 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
    \caption{Performance of Finch on erosion task (4 iterations).}
\end{figure}

\begin{figure}
	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
    \caption{Performance of Finch on masked histogram task.}
\end{figure}

\subsubsection{Graph Analytics}
\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e. choose), and that Finch can do early break, and also the different loop orders and multiple outputs. We may need to explain a little bit about what push pull is. For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}

\begin{figure}
	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
    \caption{Performance of graph apps across various tools.}
\end{figure}

Push-Pull BFS in Finch:
\begin{minted}{julia}
V = Tensor(Dense(Element(false)))
P = Tensor(Dense(Element(0)))
F = Tensor(SparseByteMap(Pattern()))
_F = Tensor(SparseByteMap(Pattern()))
A = Tensor(Dense(SparseList(Pattern())))
AT = Tensor(Dense(SparseList(Pattern())))

function finch_bfs_push_kernel(_F, F, A, V, P)
    @finch begin
        _F .= false
        for j=_, k=_
            if F[j] && A[k, j] && !(V[k])
                _F[k] |= true
                P[k] <<choose(0)>>= j #Only set the parent for this vertex
            end
        end
        return _F
    end
end

function finch_bfs_pull_kernel(_F, F, AT, V, P)
    p = ShortCircuitScalar{0}()
    @finch begin
        _F .= false
        for k=_
            if !V[k]
                p .= 0
                for j=_
                    if F[follow(j)] && AT[j, k]
                        p[] <<choose(0)>>= j #Only set the parent for this vertex
                    end
                end
                if p[] != 0
                    _F[k] |= true
                    P[k] = p[]
                end
            end
        end
        return _F
    end
end
\end{minted}

\subsection{Implementing Numpy Array API in Finch}
Experiments:
- elementwise op of 3 things (fused, unfused, julia broadcast, duckdb)
- Matrix multiply (with the insert indices)
- Triangle Counting
- SDDMM

\subsubsection{The Finch High-Level API (Needs a Name)}

\subsubsection{Finch Logic}

\subsubsection{Finch Interpreter}

\subsubsection{Lowering}
\subsubsection{Heuristic Optimization}

Find an example where fusing the python interface gives a big speedup over non-fused kernels.

%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
%in-place.
%dot((v^t .* u), w)) vs. 
%(v^t .* dot(u, w))

### 2024-03-19 sections/intro.tex
Update on Overleaf.

\section{Introduction}

%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
%
Arrays are also universal across programming languages, from their introduction
in Fortran in 1957 to present-day languages like Python
\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
%
Modern
array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
and Halide have pushed the limits of productive data processing with arrays,
fueling breakthroughs in machine learning, scientific computing, image
processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
moler_history_2020, abadi_tensorflow_2016,
paszke_pytorch_2019,ragan-kelley_halide_2013}.
%
These frameworks have been the
subject of extensive industry investment to enable performant implementations,
and often operate at the peak capacity of the hardware they run on
\cite{lo_roofline_2015}.

The success and ubiquity of arrays is likely due to their simplicity. 
%
Since
their introduction, multidimensional arrays have represented dense, rectilinear,
integer grids of points. 
%
By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
%
This
simplicity enables extensive interoperability, convenience layers, and
optimizations by breaking the abstraction barrier between array representation
and array storage.  
%
Compilers understand dense array computations across many
programming constructs, such as for and while loops, breaks, parallelism,
caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
loop-carry-dependencies, and more. Several optimizations have been developed for
dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
interchange.
%
However, while dense arrays are the easiest way to program, the world is not all dense.

%Bring in structure
% - old arrays are Dense, rectilinear, integer grids
% - The world is not like that
% - Sparsity, runs of repeated values, symmetry
% - Lots of citations, perhaps a few figures from my powerpoint
%Sparse arrays
% - Networks (graphblas, graph frameworks)
% - Simulations (BLAS, symblas, banded blas, etc)
% - Databases (database engines)
%Run-length encoding
%Symmetry
%Bands
%Padding
%Blocks
\help{This section needs citations, please help expand it and make it more
convincing. A figure could be nice.} 
\begin{figure}
	\includegraphics[width=\linewidth]{example_structures.png}
    \caption{some example structures that might be nice to put into a figure}
\end{figure}
Our world is full of structured arrays.
%

Sparse arrays (which store only nonzero elements) describe networks and
simulations.
%
Run-length encoding describes images and masks, geometry, and
databases (such as a list of transactions with the date field all the same)
Symmetry, bands, padding, and blocks are all natural matrix properties which
arise in the course of scientific computing or simulations of the natural world.
%

Operators, such as convolution, can be expressed as structured arrays.
%
For
example, a convolution with a filter can be expressed as a matrix multiplication
with the toeplitz matrix of all the circular shifts of the filter.

\help{This section needs citations}
\textbf{Currently, support for structured data is fragmented and incomplete}.
%
Experts must hand write variations of even the simplest kernels, like matrix
multiply, for each data structure and architecture to get performance.
%
Implementations must choose a small set of features to support well, resulting
in a compromise between \textbf{program flexibility} and \textbf{datastructure
flexibility}.
%
Hand-written solutions are collected in diverse libraries like
MKL, OpenCV, or SciPy. 
%
However, libraries will only ever support a subset of
programs on a subset of datastructure combinations. Even the most advanced
libraries, such as the GraphBLAS, which support a wide variety of sparse
operations over various semirings always lack support for other features, such
as tensors, fused outputs, or runs of repeated values. 
%
While dense array
compilers support an enormous variety of program constructs like early break and
multiple left hand sides, they only support dense arrays.  
%
Special-purpose
compilers like TACO or CoRa which support a select subset of structured data
structures (only sparse, or only ragged arrays) must compromise by greatly
constraining the classes of programs which they support, such as tensor
contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. 
%
This trade-off is
visualized in figure \ref{fig:datavsprogram}.

\begin{table}[h!]
  \help{This table needs citations, as well as a little bit of double checking}
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  High-Level API           &            & \checkmark &            &            & \checkmark \\
  Automatic API Fusion     &            &            &            &            & \checkmark \\
%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
  Recurrence               & \checkmark &            &            &            &            \\
  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
  \end{tabular}
  \caption{Feature support across various tools.}
  \label{tab:features}
  \end{table}

  \begin{table}[h!]
  \centering
  \begin{tabular}{l|ccccc}
  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
  \hline
  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
  Padded                   & \checkmark &            &            &            & \checkmark \\
  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
  Sparse                   &            & \checkmark &            &            & \checkmark \\
  Run-length               &            &            &            &            & \checkmark \\
  Symmetric                &            &            &            &            & \checkmark \\
  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
  Ragged                   &            &            & \checkmark &            & \checkmark \\
  \end{tabular}
  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
  \label{tab:data_structures}
\end{table}

We see two main challenges to writing efficient code over structured data.

\textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory. For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value. These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.

\textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable. For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.

In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code. In this model, performance engineers can more efficiently search the
complex landscape of programs and datastructures to find the best implementation. In this programming model, we can express certain concepts in programs, and others in data. For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.

We develop a new array programming language, Finch, which supports \textit{both}
flexible programming constructs and diverse data structures. Finch supports a
familiar programming language of loops, statements, if conditions, breaks, etc,
over a wide variety of array structures, such as sparsity, run-length-encoding,
symmetry, triangles, padding, or blocks. The Finch compiler uses the structure
of the data to generate efficient implementations of these programs.

\subsection{Contributions}

\begin{enumerate}
\item More complex array structures than ever before. A complete level-by-level
structure-description language for expressing the structure of data
hierarchically. The first such set of formats to efficiently capture banded,
triangular, run-length-encoded, or sparse datasets, and any combination thereof.
\item A rich structured array programming language with for-loops
and complex control flow constructs at the same level of productivity
of dense arrays. To our knowledge, the Finch programming language is the first
to support if-conditions, early breaks, and multiple left hand sides over
structured data, as well as complex accesses such as affine indexing or scatter/gather.
\item A compiler that specializes programs to data structures 
predictably, facilitating an expressive language that makes it easier to search the complex
space of algorithms and datastructures. It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
initialization or finalization.
\item We evaluate the productivity of our language in several case studies,
showing that Finch can be used to accelerate a wide range of applications, 
from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels. Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions. The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets. The formats capture many use cases, from random updates to sequential construction.
%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions. The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
%\item <Performance Contributions>
\end{enumerate}

### 2024-03-19 sections/lang.tex
Update on Overleaf.

\section{The Finch Language}

\subsection{Syntax and Semantics}

### 2024-03-19 sections/tensors.tex
Update on Overleaf.

@@
+\section{Bridging Looplets and Finch: The Tensor Interface}
+\teo{I am doing stuff in this section.}
 Looplets provide descriptions of iterators. 
@@
 %
-Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
+Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format, sparseTIR}.
 %

### 2024-03-20 main.tex
Update on Overleaf.

@@
 \usepackage{minted}
+\usepackage{ifthen}
 
@@
 
+\ifthenelse{1=1}{
+\newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
+\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
 \newcommand\teo[1]{\textcolor{red}{teo: #1}}
-\newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
 \newcommand\willow[1]{\textcolor{purple}{Willow: #1}}
+\newcommand{\saman}[1]{\textcolor{orange}{Saman: #1}}
+\newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
+}{
+\newcommand{\TODO}[1]{}
+\newcommand{\help}[1]{}
+\newcommand{\saman}[1]{}
+\newcommand{\teo}{}
+\newcommand{\willow}{}
+\newcommand{\leftoff}{}
+}
 
@@
 Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
+\saman{Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. 
+Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Arrays are the highest-
+performing datastructure with a long history of investment and innovation, from hardware support to compiler
+technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
+However, a large portion of the data, either obtained from nature or generated by humans, have structure.  \\
+Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!).
+So far this has failed in both scope of the ambition and execution of comparable features.
+Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
+ Thus, focusing only on sparsity limits the scope needed to handle a vast array of data.
+Second, the arrays and loops abstraction introduced in FORTRAN supports a lot more complex control flow not supported by einsum based modern sparse array compilers.
+For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
+In this paper, we attempt to truly bring computing on structured data to the same Sparse tensor algebra can be put on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
+}
 \end{abstract}
@@
 \input{sections/background}
+\input{sections/tensors}
 \input{sections/lang}
-\input{sections/tensors}
 \input{sections/compiler}

### 2024-03-20 main.tex
Update on Overleaf.

@@
 For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-In this paper, we attempt to truly bring computing on structured data to the same Sparse tensor algebra can be put on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
+
+ In this paper, we describe a compiler, Finch,
+which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
+programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
+Our approach enables new loop optimizations across multiple domains, unifying techniques such as
+sparse tensors, databases, and lossless compression
+We hope to truly bring computing on structured data to the same Sparse tensor algebra can be put on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
 }

### 2024-03-20 main.tex
Update on Overleaf.

--- 2024-03-20 main.tex
@@
 For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-In this paper, we attempt to truly bring computing on structured data to the same Sparse tensor algebra can be put on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
-}
+\\
+ In this paper, we describe a compiler, Finch,
+which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
+programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
+Our approach enables new loop optimizations across multiple domains, unifying techniques such as
+sparse tensors, databases, and lossless compression
+We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  }
 \end{abstract}

### 2024-03-20 sections/intro.tex
Update on Overleaf.

@@
 
-In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.
-In this model, performance engineers can more efficiently search the
+In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
+TACO did something similar?} In this model, performance engineers can more efficiently search the
 complex landscape of programs and datastructures to find the best implementation.
@@
 structure-description language for expressing the structure of data
-hierarchically. The first such set of formats to efficiently capture banded,
+hierarchically.\saman{level formats were done in TACO.} The first such set of formats to efficiently capture banded,
 triangular, run-length-encoded, or sparse datasets, and any combination thereof.

### 2024-03-20 sections/intro.tex
Update on Overleaf.

@@
 to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather.
+structured data, as well as complex accesses such as affine indexing or scatter/gather. \saman{for beyond dense...don't want to sound like we are the fist to do for dense. }
 \item A compiler that specializes programs to data structures 
-predictably, facilitating an expressive language that makes it easier to search the complex
+predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex
 space of algorithms and datastructures.

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 Notably, there is a 1.9x speedup for the HB/saylr4 matrix.
+
 \subsubsection{4D Blocked SpMV}
+Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
+Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
+We represent the blocked matrix with a SparseList as its second level (i.e.
+of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
+Then, we perform SpMV on each b x b block individually.
+Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
+This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
+
+We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix with a blocked matrix of size 10x10 and a dense and randomly generated vector.
+We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
+
+Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
+In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features.
+Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
+For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
+
+\begin{minted}{julia}
+            @finch begin
+                for J = _
+                    for I = _
+                        for j = 1:10
+                            for i = 1:10
+                                let _i = (I - 1) * 10 + i
+                                    y[_i] += block_A[i, j, I, J] * block_x[j, J]
+                                end
+                            end
+                        end
+                    end
+                end
+            end
+\end{minted}
 
@@
 \subsection{Implementing Numpy Array API in Finch}
-Experiments:
-- elementwise op of 3 things (fused, unfused, julia broadcast, duckdb)
-- Matrix multiply (with the insert indices)
-- Triangle Counting
-- SDDMM
 

### 2024-03-20 main.tex
Update on Overleaf.

@@
 
+\newcommand{\HIDE}{}
+
 %%
@@
 \begin{abstract}
-From FORTRAN to Numpy, arrays have revolutionized how we express computation.
+\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.
 Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.
@@
 Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
-\saman{Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. 
-Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Arrays are the highest-
-performing datastructure with a long history of investment and innovation, from hardware support to compiler
+}
+
+Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
+Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
+Multi-dimensional rrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
 technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
-However, a large portion of the data, either obtained from nature or generated by humans, have structure.  \\
+However, a large portion of the data, either obtained from nature or generated by humans, have structure.
+
 Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!).
@@
 For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-\\
+
  In this paper, we describe a compiler, Finch,
@@
 sparse tensors, databases, and lossless compression
-We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  }
+We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
 \end{abstract}

### 2024-03-20 main.tex
Update on Overleaf.

@@
 \newcommand{\saman}[1]{}
-\newcommand{\teo}{}
-\newcommand{\willow}{}
+\newcommand{\teo}[1]{}
+\newcommand{\willow}[1]{}
 \newcommand{\leftoff}{}
@@
 
-\newcommand{\HIDE}{}
+\newcommand{\HIDE}[1]{}
 
@@
 Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
-Multi-dimensional rrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
-technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
+Furthermore, multi-dimensional arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
+technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
 However, a large portion of the data, either obtained from nature or generated by humans, have structure.

### 2024-03-20 main.tex
Update on Overleaf.

@@
 \newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
+\newcommand\changwan[1]{\textcolor{green}{Help Needed: #1}}
 }{
@@
 Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
-Furthermore, multi-dimensional arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
+Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler
 technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
@@
  Thus, focusing only on sparsity limits the scope needed to handle a vast array of data.
-Second, the arrays and loops abstraction introduced in FORTRAN supports a lot more complex control flow not supported by einsum based modern sparse array compilers.
+Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.
 For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

### 2024-03-20 main.tex
Update on Overleaf.

@@
 \newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
-\newcommand\changwan[1]{\textcolor{green}{Help Needed: #1}}
+\newcommand{\changwan[1]{\textcolor{blue}{CW: #1}}
 }{
@@
 which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
-programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
+programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays.
 Our approach enables new loop optimizations across multiple domains, unifying techniques such as
-sparse tensors, databases, and lossless compression
+sparse tensors, databases, and lossless compression.
 We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  

### 2024-03-20 main.tex
Update on Overleaf.
not quite sure what I'm doing

--- 2024-03-20 main.tex
@@
 \newcommand{\saman}[1]{}
-\newcommand{\teo}[1]{}
-\newcommand{\willow}[1]{}
+\newcommand{\teo}{}
+\newcommand{\willow}{}
 \newcommand{\leftoff}{}
@@
 
-\newcommand{\HIDE}[1]{}
+\newcommand{\finchextent}{\ensuremath{\text{\textbf{extent}}}}
+\newcommand{\finchliteral}{\ensuremath{\text{\textbf{literal}}}}
+\newcommand{\finchvalue}{\ensuremath{\text{\textbf{value}}}}
+\newcommand{\finchtensor}{\ensuremath{\text{\textbf{tensor}}}}
+\newcommand{\finchindex}{\ensuremath{\text{\textbf{index}}}}
+\newcommand{\finchvar}{\ensuremath{\text{\textbf{variable}}}}
+\newcommand{\finchcall}{\ensuremath{\text{\textbf{call}}}}
+\newcommand{\finchaccess}{\ensuremath{\text{\textbf{access}}}}
+\newcommand{\finchread}{\ensuremath{\text{\textbf{read}}}}
+\newcommand{\finchassign}{\ensuremath{\text{\textbf{assign}}}}
+\newcommand{\finchupdate}{\ensuremath{\text{\textbf{update}}}}
+\newcommand{\finchloop}{\ensuremath{\text{\textbf{loop}}}}
+\newcommand{\finchdeclare}{\ensuremath{\text{\textbf{declare}}}}
+\newcommand{\finchdefine}{\ensuremath{\text{\textbf{define}}}}
+\newcommand{\finchsieve}{\ensuremath{\text{\textbf{sieve}}}}
+\newcommand{\finchblock}{\ensuremath{\text{\textbf{block}}}}
 
@@
 \begin{abstract}
-\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.
+From FORTRAN to Numpy, arrays have revolutionized how we express computation.
 Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.
@@
 Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
-}
-
-Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
-Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
-Furthermore, multi-dimensional arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler
-technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
-However, a large portion of the data, either obtained from nature or generated by humans, have structure.
-
+\saman{Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays. 
+Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy. Arrays are the highest-
+performing datastructure with a long history of investment and innovation, from hardware support to compiler
+technology.  Arrays in these, and most prominent systems, can only handle dense rectilinear integer grids.
+However, a large portion of the data, either obtained from nature or generated by humans, have structure.  \\
 Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!).
@@
 For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-
+\\
  In this paper, we describe a compiler, Finch,
@@
 sparse tensors, databases, and lossless compression
-We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
+We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  }
 \end{abstract}

### 2024-03-20 sections/lang.tex
Update on Overleaf.
not quite sure what I'm doing

@@
 \subsection{Syntax and Semantics}
+
+\begin{align*}
+    \finchliteral := val \in \mathbb{V}
+\end{align*}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchliteral(val \in \mathbb{V}) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+name
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+name
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchtensor(name \in \mathbb{S}) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+name
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchindex(name \in \mathbb{S}) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+name
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchvar(name \in \mathbb{S}) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+name
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchextent(a \in E, b \in E) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+a : b
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchcall(f \in E, args\ldots \in E) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+f(args...)
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchaccess(tns \in E, idxs\ldots \in E) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+tns[idxs...]
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchassign(lhs \in A, op \in E, rhs \in E) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+tns[idxs...] <<op>>= rhs
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchloop(idx \in I, range \in E, body \in S) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+for idx = range; body end
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchdefine(var \in V, val \in E, body \in S) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+let var = val; body end
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchsieve(cond \in E, body \in S) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+if cond
+    body
+end
+\end{minted}
+\end{minipage}
+
+\noindent\begin{minipage}{.5\linewidth}
+\raggedleft $\finchblock(bodies\ldots \in S) :=$~
+\end{minipage}%
+\begin{minipage}{.5\linewidth}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+begin; bodies... end
+\end{minted}
+\end{minipage}
+
+\subsection{Wrapper Tensors}
+
+\subsection{Scalars}
+
+\subsubsection{Sparse Scalars}
+\subsubsection{Early Break Scalars}

### 2024-03-20 sections/lang.tex
good start

@@
 
-\begin{align*}
-    \finchliteral := val \in \mathbb{V}
-\end{align*}
-
+\begin{minipage}{0.6\linewidth}
     \noindent\begin{minipage}{.5\linewidth}
@@
     \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
+    \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
+    val
     \end{minted}
@@
     \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
+    \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
+    ex :: type
     \end{minted}
@@
     \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+    \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
     name
@@
     \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+    \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
     a : b
@@
     \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+    \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
     f(args...)
     \end{minted}
+    \end{minipage}
 \end{minipage}
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
 tns[idxs...]
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
 tns[idxs...] <<op>>= rhs
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
 for idx = range; body end
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
 let var = val; body end
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-if cond
-    body
-end
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
+if cond; body end
 \end{minted}
@@
 \begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
+\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
 begin; bodies... end

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 \subsection{Data-Driven Performance Engineering}
-\subsubsection{Sparse-Sparse Matrix Multiply}
+\subsubsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
+\changwan{
+Explain what is SpGEMM. why SpGEMM hard? (not sure)
+three approaches. Asym time complexity?
+}
 
@@
 Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
-% I'm not sure if the right place for this is here - maybe should come earlier?
 
@@
 The dense vector is randomly generated.
-We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures 1 and 2.
+We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.
 
@@
 We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like $exdata\_1$, $TSOPF\_RS\_b678\_c1$, and $heart3$ performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
+Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
 Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.
@@
     \caption{Performance of SpMV across various tools.}
+    \label{spmv_sorted}
 \end{figure}
@@
     \caption{Performance of SpMV by Finch format.}
+    \label{spmv_grouped}
 \end{figure}

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
+Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, 1.55, and 1.30 relative to TACO and of 1.47, 2.13, and 1.42 relative to the SparseList Finch SpMV implementation.
 Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.

### 2024-03-20 sections/eval.tex
Update on Overleaf.

--- 2024-03-20 sections/eval.tex
@@
 The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.
+We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}
 
 We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, X, and Y relative to TACO and of 1.47, X, and Y relative to the SparseList Finch SpMV implementation.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of A and B relative to TACO and a speedup of A and B relative to the SparseList Finch SpMV implementation.
+Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, 1.55, and 1.30 relative to TACO and of 1.47, 2.13, and 1.42 relative to the fastest SparseList Finch SpMV implementation.
+Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of 1.98 and 1.64 relative to TACO and a speedup of 1.47 and 1.24 relative to the fastest SparseList Finch SpMV implementation.
 

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.
-We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}
+We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
 

### 2024-03-20 sections/lang.tex
add some more variables and stuff
hmm

@@
 
-\begin{minipage}{0.6\linewidth}
-    \noindent\begin{minipage}{.5\linewidth}
+% Define variables for left and right minipage widths
+\newlength{\leftwidth}
+\setlength{\leftwidth}{.4\linewidth}
+\newlength{\rightwidth}
+\setlength{\rightwidth}{.6\linewidth}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $\mathbb{V} :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+the set of all values.~
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $\mathbb{S} :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+the set of all symbols.
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $\mathbb{T} :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+the set of all types.
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $I :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+$\finchindex$
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $A :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+$\finchaccess$
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $V :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+$\finchvar$
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $E :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+$\finchliteral | \finchvalue | \finchindex | \finchvar | \finchextent | \finchcall | \finchaccess$
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
+\raggedleft $S :=$~
+\end{minipage}%
+\begin{minipage}{\rightwidth}
+$\finchassign | \finchloop | \finchdefine | \finchsieve | \finchblock$
+\end{minipage}
+
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchliteral(val \in \mathbb{V}) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchtensor(name \in \mathbb{S}) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchindex(name \in \mathbb{S}) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchvar(name \in \mathbb{S}) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchextent(a \in E, b \in E) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-    \noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchcall(f \in E, args\ldots \in E) :=$~
 \end{minipage}%
-    \begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 \end{minipage}
-\end{minipage}
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchaccess(tns \in E, idxs\ldots \in E) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchassign(lhs \in A, op \in E, rhs \in E) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
-tns[idxs...] <<op>>= rhs
+lhs <<op>>= rhs
 \end{minted}
@@
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchloop(idx \in I, range \in E, body \in S) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchdefine(var \in V, val \in E, body \in S) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchsieve(cond \in E, body \in S) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}
@@
 
-\noindent\begin{minipage}{.5\linewidth}
+\noindent\begin{minipage}{\leftwidth}
 \raggedleft $\finchblock(bodies\ldots \in S) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
+\begin{minipage}{\rightwidth}
 \begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{julia}

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
+Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
 
@@
 Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, 1.55, and 1.30 relative to TACO and of 1.47, 2.13, and 1.42 relative to the fastest SparseList Finch SpMV implementation.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the $toeplitz\_large\_band$ and the $toeplitz\_medium\_band$ matrices having a speedup of 1.98 and 1.64 relative to TACO and a speedup of 1.47 and 1.24 relative to the fastest SparseList Finch SpMV implementation.
+Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO and a speedup of 1.47 and 1.24 relative to the fastest SparseList Finch SpMV implementation.
 
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean.
-The matrices in the SNAP collection which represent graph datasets are boolean and had an average speedup of X over TACO and X over the same Finch program, but with an Element format.
+There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
+In particular, Finch Pattern SpMV had for had an average speedup of X over TACO and X over the same Finch program, but with an Element format.
 

### 2024-03-20 sections/eval.tex
Update on Overleaf.

@@
 
-The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks.
+The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
 The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
@@
 There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-In particular, Finch Pattern SpMV had for had an average speedup of X over TACO and X over the same Finch program, but with an Element format.
+For example, Finch Pattern SpMV had a speedup of 2.51 over TACO for SNAP/email-Eu-core and a speedup of X over the same Finch program, but with an Element format.
 
@@
 However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.3x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO .
-Notably, there is a 1.9x speedup for the HB/saylr4 matrix.
+Symmetric SpMV with the SparseList level format in Finch results in an average of 1.30x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO.
+Notably, there is a 1.90x speedup for the HB/saylr4 matrix over TACO.
 
@@
 Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features.
+In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
 Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.

### 2024-03-20 sections/eval.tex

@@
 \subsubsection{SpMV}
+\willow{We might want to start this section off by saying spmv is a bandwidth bound kernel, and historically therefore subject to many format modifications to improve throughput and reduce the associated control flow}
 Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data.
 More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats.
-Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
+\willow{say this faster} Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
 
 Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns.
-Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.
+\willow{this might show up in the intro, but if you wanna say this, maybe give an exmaple here too (e.g.
+symmetry)} Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.
 
-The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
-The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
+\willow{these paragraphs could be merged and condensed.
+Perhaps some of this belongs elsewhere} The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
+\willow{The following two sentences are the most important and relevant so far} The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
 In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
 
-We compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
+\willow{fix the citations, put Figure \ref{fig:spmv} earlier, We display speedup relative to ...
+} Figures \ref{spmv_sorted} and \ref{spmv_grouped} compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
+We display speedup relative to row major or column major TACO performance, whichever was faster.
 We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
@@
 The dense vector is randomly generated.
-We depict the performance of SpMV across the aforementioned tools and compare to the fastest Finch format for that particular dataset in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.
+We tested a wide variety of finch formats, described in the following subsection; the performance displayed for Finch on each dataset is the fastest among the formats we tested.
 
@@
 \subsubsection{Tensor Formats}
-We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix.
+\willow{could this be a table?} We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix.
 To represent scalar values, we benchmarked with both the Element and Pattern levels.
@@
 
-We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix with a blocked matrix of size 10x10 and a dense and randomly generated vector.
+We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
 We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.

### 2024-03-20 sections/lang.tex
syntax

--- 2024-03-20 sections/lang.tex
@@
 
+\noindent % This ensures the minipages fill the page width
+\begin{minipage}{0.4\linewidth}
 \begin{align*}
-    \finchliteral := val \in \mathbb{V}
+    \finchliteral(val \in \mathbb{V}) &:= \text{\mintinline{julia}{val}} \\
+    \finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) &:= \text{\mintinline{julia}{ex :: type}} \\
+    \finchtensor(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchindex(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchvar(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchextent(a \in E, b \in E) &:= \text{\mintinline{julia}{a : b}} \\
+    \finchcall(f \in E, args\ldots \in E) &:= \text{\mintinline{julia}{f(args...)}} \\
+    \finchaccess(tns \in T, idxs\ldots \in E) &:= \text{\mintinline{julia}{tns[idxs...]}} \\
+    \finchfreeze(tns \in T) &:= \text{\mintinline{julia}{@freeze(tns)}} \\
+    \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
 \end{align*}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchliteral(val \in \mathbb{V}) :=$~
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) :=$~
+\begin{minipage}{0.5\linewidth}
+\begin{align*}
+    \mathbb{V} &:= \text{the set of all values.} \\
+    \mathbb{S} &:= \text{the set of all symbols.} \\
+    \mathbb{T} &:= \text{the set of all types.} \\
+    I &:= \finchindex \\
+    A &:= \finchaccess \\
+    V &:= \finchvar \\
+    T &:= \finchtensor \\
+    E &:= \finchliteral \mid \finchvalue \mid \finchindex \\
+        &\quad \mid \finchvar \mid \finchextent \mid \finchcall \mid \finchaccess \\
+    S &:= \finchassign \mid \finchloop \mid \finchdefine \\
+        &\quad \mid \finchsieve \mid \finchblock \\
+\end{align*}
 \end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchtensor(name \in \mathbb{S}) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchindex(name \in \mathbb{S}) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchvar(name \in \mathbb{S}) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-name
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchextent(a \in E, b \in E) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-a : b
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchcall(f \in E, args\ldots \in E) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-f(args...)
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchaccess(tns \in E, idxs\ldots \in E) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-tns[idxs...]
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchassign(lhs \in A, op \in E, rhs \in E) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-tns[idxs...] <<op>>= rhs
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchloop(idx \in I, range \in E, body \in S) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-for idx = range; body end
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchdefine(var \in V, val \in E, body \in S) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-let var = val; body end
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchsieve(cond \in E, body \in S) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-if cond
-    body
-end
-\end{minted}
-\end{minipage}
-
-\noindent\begin{minipage}{.5\linewidth}
-\raggedleft $\finchblock(bodies\ldots \in S) :=$~
-\end{minipage}%
-\begin{minipage}{.5\linewidth}
-\begin{minted}[autogobble, linenos=false, fontsize=\normalsize]{latex}
-begin; bodies... end
-\end{minted}
-\end{minipage}
+\begin{align*}
+    \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
+    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs \texttt{<<op>>=} rhs}} \\
+    \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
+    \finchdefine(var \in V, val \in E, body \in S) &:= \text{\mintinline{julia}{let var = val; body end}} \\
+    \finchsieve(cond \in E, body \in S) &:= \text{\mintinline{julia}{if cond; body end}} \\
+    \finchblock(bodies\ldots \in S) &:= \text{\mintinline{julia}{begin; bodies... end}}
+\end{align*}
 

### 2024-03-20 sections/lang.tex
some more stuff

@@
     \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
+    \finchmode(tns \in T) &:= \text{\mintinline{julia}{@mode(tns)}} \\
 \end{align*}
@@
 
+\[
+    \frac{}{\langle\finchliteral(val), (e, t)\rangle \rightarrow val} \quad
+    \frac{}{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle} \quad
+\]
+\[
+    \frac{}{\langle\finchvar(name), (e, t)\rangle \rightarrow e(\finchvar(name))} \quad
+    \frac{}{\langle\finchindex(name), (e, t)\rangle \rightarrow e(\finchindex(name))}
+\]
+\[
+    \frac{\langle args_i, (e, t) \rangle \Rightarrow vals_i \quad \langle f, (e, t) \rangle \Rightarrow g}
+    {\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
+\]
+\[
+\frac{\langle head, s_1 \rangle \rightarrow s_2}
+{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle} \quad
+\frac{}{\langle\finchblock(), s\rangle \rightarrow s}
+\]
+\[
+\frac{\langle cond, s \rangle \Rightarrow true}
+{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}\quad
+\frac{\langle cond, s \rangle \Rightarrow false}
+{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
+\]
+\[
+\frac{\langle val, (e, t) \rangle \rightarrow val'}
+{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
+\]
+\[
+\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns' \quad  \langle init, s \rangle \rightarrow init' \quad  \langle dims_i, s \rangle \rightarrow dims_i'}
+{\langle\finchdeclare(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle declare(tns', init', dims'...), t \rangle)}
+\]
+\[
+\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
+{\langle\finchthaw(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+\]
+\[
+\frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
+{\langle\finchfreeze(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+\]
+
 \subsection{Wrapper Tensors}

### 2024-03-21 sections/lang.tex
more semantics

@@
 \frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
-{\langle\finchthaw(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+{\langle\finchthaw(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
 \]
@@
 \frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
-{\langle\finchfreeze(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+\]
+
+$E$ is an evaluation context that applies anywhere in the syntax tree.
+
+\[
+\frac{e(tns) = tns' \langle unfurl(tns'), t\rangle \rightarrow \langle tns'', t'\rangle}
+{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), (e, t)\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), (e, t')\rangle}
+\]
+
+\[
+\frac{}
+{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(phase(range, body), tail...), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
+\]
+
+\[
+\frac{}
+{\langle \finchloop(i, ext, E[\finchaccess(sequence(), j..., i)]), s\rangle \rightarrow \langle \finchblock(), s\rangle}
+\]
+
+\[
+\frac{}
+{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(next, body), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
 \]

### 2024-03-21 sections/tensors.tex
Update on Overleaf.

@@
 %
-First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is much easier to manage when considering a single dimensional iterator. 
+First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
 %

### 2024-03-21 sections/tensors.tex
Update on Overleaf.

@@
 %
-Our second abstraction adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation and limiting the amount of specific work.
+Our next two abstractions adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
 %
-Previous efforts to compiler a greater variety of sparse tensor programs left this bridge untouched ~\cite{henry2021compilation, won2023unified}.
+Previous efforts to compiler a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified}.
 %
-Our third abstraction limits our set of sparse tensor programs to accommodate a natural restriction on when multi-dimensional data structures can be modified.
 %

### 2024-03-21 sections/tensors.tex
Update on Overleaf.

@@
 %
-Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format, sparseTIR}.
+Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient ,chou2022compilation, chou2018format}.
 %
@@
 %
-Previous efforts to compiler a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified}.
+Previous efforts to compiler a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
 %

### 2024-03-21 sections/tensors.tex
Update on Overleaf.

@@
 %
-Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient ,chou2022compilation, chou2018format}.
+Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
 %
@@
 %
+We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
 %
-% As we will see
-% \begin{enumerate}
-%     \item Compilation substitutes in different iterators depending on context, mainly the traversal pattern of the iterator (oh god).
-%     \item Less obviously, iterators have other state that needs to be tracked, such as when we write stuff.
-%     \item how do we ask for the right iterator and how do we change/manage their state?
-% \end{enumerate}
-% Need to Make these problems clear by this point, hope that read anticipate them.
-
-% Solutions:
-% \begin{enumerate}
-%     \item Einsum like: limit when state can change, leading to a natural lifecycle for a tensor.
-%     \item simple interface to go from an index expression into an iterator that decomposes in a way tensor compilers are used to.
-% \end{enumerate}
+These interfaces add to the level abstraction.
 
@@
 \begin{enumerate}
-\item fibers
+\item fibers and levels
 \item assembly

### 2024-03-21 sections/intro.tex
Update on Overleaf.

@@
 Run-length encoding describes images and masks, geometry, and
-databases (such as a list of transactions with the date field all the same)
-Symmetry, bands, padding, and blocks are all natural matrix properties which
-arise in the course of scientific computing or simulations of the natural world.
+databases (such as a list of transactions with the date field all the same).
 %
-
+Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing as well as in intermediate structures in many linear solvers.
+%
 Operators, such as convolution, can be expressed as structured arrays.
 %
-For
-example, a convolution with a filter can be expressed as a matrix multiplication
-with the toeplitz matrix of all the circular shifts of the filter.
-
+For example, a convolution with a filter can be expressed as a matrix multiplication
+with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
+%
 \help{This section needs citations}

### 2024-03-21 sections/intro.tex
Update on Overleaf.

@@
 \begin{enumerate}
-\item More complex array structures than ever before. A complete level-by-level
+\item 
+\teo{How about:
+A Looplet based level-format abstraction.
+Although many systems (TACO, Taichi, SPF, Ebb) features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
+}
+
+More complex array structures than ever before. A complete level-by-level
 structure-description language for expressing the structure of data

### 2024-03-21 sections/intro.tex
Update on Overleaf.

@@
 Although many systems (TACO, Taichi, SPF, Ebb) features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
+The first such set of formats to efficiently capture banded,
+triangular, run-length-encoded, or sparse datasets, and any combination thereof.
 }
@@
 predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex
-space of algorithms and datastructures.
-It is easy to implement new datastructures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
+space of algorithms and data structures.
+It is easy to implement new data structures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
 initialization or finalization.
@@
 from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
+\teo{Do we really want to say productivity here?
+What word do other similar papers use?}
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.

### 2024-03-21 sections/intro.tex
Update on Overleaf.

@@
 A Looplet based level-format abstraction.
-Although many systems (TACO, Taichi, SPF, Ebb) features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
+Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
 The first such set of formats to efficiently capture banded,

### 2024-03-21 sections/lang.tex
everything except jumpers

@@
 \end{align*}
+
+the state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
+environment and another state $t$ corresponding to the embedding language.
 
@@
 \frac{}
-{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(phase(range, body), tail...), j..., i)]), s\rangle
-    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
+{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
+\]
+
+\[
+\frac{e(i) = i' \quad \langle seek(i'), t \rangle \rightarrow t'}
+{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle
+    }{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
+\]
+
+\[
+\frac{\langle cond, t \rangle \Rightarrow true}
+{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle
+    }{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
+\]
+
+\[
+\frac{\langle cond, t \rangle \Rightarrow false}
+{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle
+    }{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
 \]
@@
 \frac{}
-{\langle \finchloop(i, ext, E[\finchaccess(sequence(), j..., i)]), s\rangle \rightarrow \langle \finchblock(), s\rangle}
+{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchloop(i, extent(max(a, c), min(b, d)), E[\finchaccess(body, j..., i)]), s\rangle}}
 \]
@@
 \[
-\frac{}
-{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(next, body), j..., i)]), s\rangle
-    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
+\frac{\langle \finchloop(i, ext, E[\finchaccess(head, j..., i)]), s\rangle \rightarrow s'}
+{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(head, tail), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(tail, j..., i)]), s'\rangle}}
+\]
+
+\[
+\frac{\langle seek(a), t \rangle \rightarrow t' \quad }
+{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle
+    }{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
+\]
+
+\[
+\frac{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t') \quad \langle next, t' \rangle \rightarrow t''}
+{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle
+    }{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(thunk(next, stepper(body, next)), j..., i)]), (e', t'')\rangle}}
 \]

### 2024-03-21 sections/eval.tex
Update on Overleaf.

@@
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
+
+\begin{table}[htbp]
+    \centering
+    \caption{SpMV Tensor Formats}
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \hline
+        SparseList & sparse matrices & Vuduc, Ahrens, TACO, synthetic \\
+        \hline
+        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, synthetic  \\
+        \hline 
+        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, synthetic \\ 
+        \hline 
+        SparsePoint & matrices with one value per row & synthetic \\
+        \hline
+        Pattern* & matrices with boolean values & SNAP collection \\
+        \hline 
+        \footnote{test}
+    \end{tabular}
+\end{table}
 

### 2024-03-21 sections/eval.tex
Update on Overleaf.

--- 2024-03-20 sections/eval.tex
@@
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
+
+\begin{table}[htbp]
+    \centering
+    \caption{SpMV Tensor Formats}
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \hline
+        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
+        \hline
+        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
+        \hline 
+        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
+        \hline 
+        SparsePoint & matrices with one value per row & synthetic \\
+        \hline
+        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
+        \hline
+        Pattern* & matrices with boolean values & SNAP  \\
+        \hline 
+    \end{tabular}
+    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
+\end{table}
+
+\begin{table}[htbp]
+    \centering
+    \caption{SpMV Tensor Formats}
+    \begin{tabular}{|l|l|l|l|l|}
+        \hline
+        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \hline
+        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
+        \hline
+        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
+        \hline 
+        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
+        \hline 
+        SparsePoint & matrices with one value per row & synthetic \\
+        \hline
+        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
+        \hline
+        Pattern* & matrices with boolean values & SNAP  \\
+        \hline 
+    \end{tabular}
+    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
+\end{table}
 

### 2024-03-21 sections/eval.tex
Update on Overleaf.

@@
     \caption{SpMV Tensor Formats}
-    \begin{tabular}{|l|l|l|l|l|}
+    \begin{tabular}{|l|l|l|l|}
         \hline
-        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
         \hline
-        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
-        \hline
-        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
-        \hline 
-        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
-        \hline 
-        SparsePoint & matrices with one value per row & synthetic \\
-        \hline
-        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
-        \hline
-        Pattern* & matrices with boolean values & SNAP  \\
+        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
+        \cline{3-4} 
+        & & Pattern & sparse, boolean-valued matrices \\
+        \cline{2-4} 
+        & \multirow{2}{*}{SparseVBL} & Element & real-valued matrices with a blocked structure \\
+        \cline{3-4} 
+        & & Pattern & boolean-valued matrices with a blocked structure \\
+        \cline{2-4}
+        & \multirow{2}{*}{SparseBand} & Element & real-valued matrices with a diagonal band \\
+        \cline{3-4} 
+        & & Pattern & boolean-valued matrices with a diagonal band \\
         \hline 
     \end{tabular}
-    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
+
 \end{table}

### 2024-03-21 sections/eval.tex
Update on Overleaf.

--- 2024-03-21 sections/eval.tex
@@
     \caption{SpMV Tensor Formats}
-    \begin{tabular}{|l|l|l|l|l|}
+    \begin{tabular}{|l|l|l|l|}
         \hline
-        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
         \hline
-        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
-        \hline
-        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
-        \hline 
-        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
-        \hline 
-        SparsePoint & matrices with one value per row & synthetic \\
-        \hline
-        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
-        \hline
-        Pattern* & matrices with boolean values & SNAP  \\
+        \multirow{7}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
+        \cline{3-4} 
+        & & Pattern & sparse, boolean-valued matrices \\
+        \cline{2-4} 
+        & \multirow{2}{*}{SparseVBL} & Element & real-valued matrices with blocked structure \\
+        \cline{3-4} 
+        & & Pattern & boolean-valued matrices with blocked structure \\
+        \cline{2-4}
+        & \multirow{2}{*}{SparseBand} & Element & real-valued matrices with diagonal band \\
+        \cline{3-4} 
+        & & Pattern & boolean-valued matrices with diagonal band \\
+        \cline{2-4}
+        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
+        \cline{3-4} 
+        & & Pattern & boolean-valued matrices with one value per row \\
         \hline 
     \end{tabular}
-    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
+
 \end{table}

### 2024-03-21 sections/eval.tex
Update on Overleaf.

--- 2024-03-21 sections/eval.tex
@@
 \subsubsection{Tensor Formats}
-\willow{could this be a table?} We benchmarked the SpMV kernel using the Dense level format for the outer level and the SparseList, SparseVBL, SparseBand, and SparsePoint Finch levels for the inner level of the sparse matrix.
-To represent scalar values, we benchmarked with both the Element and Pattern levels.
-The SparseList, SparseVBL, and SparseBand formats were functional for all matrices, but the SparsePoint format was applicable only for the synthetic permutation matrices.
-The Element format was applicable to represent scalars in all matrices, while the Pattern level was applicable only for boolean matrices.
-We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
+\willow{could this be a table?} We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
@@
     \centering
+    \label{tensor_formats}
     \caption{SpMV Tensor Formats}
-    \begin{tabular}{|l|l|l|}
+    \begin{tabular}{|l|l|l|l|}
         \hline
-        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
+        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
         \hline
-        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
-        \hline
-        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
-        \hline 
-        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
-        \hline 
-        SparsePoint & matrices with one value per row & synthetic \\
-        \hline
-        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
-        \hline
-        Pattern* & matrices with boolean values & SNAP  \\
+        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
+        \cline{3-4} 
+        & & Pattern & sparse, boolean-valued matrices \\
+        \cline{2-4} 
+        & SparseVBL & Element & real-valued matrices with blocked structure \\
+        \cline{2-4}
+        & SparseBand & Element & real-valued matrices with diagonal band \\
+        \cline{2-4}
+        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
+        \cline{3-4} 
+        & & Pattern & matrix with runs of true or false \\
         \hline 
     \end{tabular}
-    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
-\end{table}
 
-\begin{table}[htbp]
-    \centering
-    \caption{SpMV Tensor Formats}
-    \begin{tabular}{|l|l|l|l|l|}
-        \hline
-        \textbf{Tensor Format} & \textbf{Style of Matrix} & \textbf{Evaluation Datasets} \\
-        \hline
-        SparseList & sparse matrices & Vuduc, Ahrens, TACO, SNAP, synthetic \\
-        \hline
-        SparseVBL & matrices with a clear blocked structure & Vuduc, Ahrens, TACO, SNAP, synthetic  \\
-        \hline 
-        SparseBand & matrices with a diagonal band & Vuduc, Ahrens, TACO, SNAP, synthetic \\ 
-        \hline 
-        SparsePoint & matrices with one value per row & synthetic \\
-        \hline
-        Element* & matrices with real values & Vuduc, Ahrens, TACO, synthetic \\
-        \hline
-        Pattern* & matrices with boolean values & SNAP  \\
-        \hline 
-    \end{tabular}
-    \footnotesize *Used in conjunction with aforementioned formats as the innermost level to represent scalar values.
 \end{table}

### 2024-03-21 sections/eval.tex
Update on Overleaf.

@@
 \subsubsection{Tensor Formats}
-\willow{could this be a table?} We tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
+For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
@@
     \centering
-    \label{tensor_formats}
+    \label{spmv_tensor_formats}
     \caption{SpMV Tensor Formats}

### 2024-03-21 sections/background.tex
fix

@@
 \subsection{Looplets}
-Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically.
-Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously.
-In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space.
-Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
-can be implemented using simple compiler passes like term rewriting and constant propagation.
+Finch represents iteration patterns using Looplets, a language that decomposes
+datastructure iterators hierarchically. Looplets represent the control-flow
+structures needed to iterate over any given datastructure, or multiple
+datastructures simultaneously. In particular, looplets are good at lifting code
+to the highest loop level that it's needed and subdividing iteration
+hierarchically in coordinate space. Because looplets are compiled with
+progressive lowering, structure-specific mathematical optimizations such as
+integrals, multiply by zero, etc. can be implemented using simple compiler
+passes like term rewriting and constant propagation during the intermediate
+lowering stages. \cite{ahrens_looplets_2023}
 
-We begin a formal description of looplets here:
+In this work, we partially simplify the presentation of Looplets to focus on the
+semantics, rather than precise implementation. Since Finch is designed to be
+embedded into another language, we allow for several of the looplets to utilize
+functions that modify the state of that language. It is assumed that if a
+looplet introduces a variable to be used in a child looplet, the child looplet
+will not modify that variaable.
 
-\begin{figure}
-	\includegraphics[width=\linewidth]{example_looplet_figure.png}
-    \caption{Looplets visualized, along with their semantics}
+\begin{figure}[ht]
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchlookup(seek, body)$} The Lookup Looplet represents a
+        randomly accessible region of an iterator. The body of the lookup is
+        understood to have one less dimension than the lookup itself, as we have
+        already ``looked up'' that index in the tensor by the time we reach the
+        body. \texttt{seek(i)} is a function that updates state to the given
+        index.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-lookup.png}
+    \end{minipage}
+
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchrun(body)$} The Run Looplet represents a constant
+        region of an iterator. The body of the run is understood to have one
+        less dimension than the lookup itself, as all of the bodies are
+        identical.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-run.png}
+    \end{minipage}
+
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchphase(c:d, body)$} The Phase Looplet represents a
+        restriction of the range on which a loop should execute, and allows us
+        to succinctly express the ranges on which children of compound looplets
+        are defined.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-phase.png}
+    \end{minipage}
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchswitch(cond, head, tail)$} The Switch Looplet allows
+        us to specialize the body of a looplet based on a condition, evaluated
+        in the embedding context. If the condition is true, we use `head`,
+        otherwise we use `tail`. Switch has a high lowering priority so we can
+        see what's inside of it and lower that appropriately. This also lifts
+        the condition as high as possible into the loop nest. The condition is
+        assumed to evaluate to a boolean.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-switch.png}
+    \end{minipage}
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchthunk(preamble, body, epilogue)$} The Thunk Looplet
+        allows us to cache certain computations in the state under which the
+        body will execute. This is useful for computing and caching the results
+        of expensive computations.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-thunk.png}
+    \end{minipage}
+
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchsequence(head, tail)$} The Sequence looplet represents the
+        concatenation of two looplets. Both arguments must be phase looplets, and
+        are assumed to be nonoverlapping, covering, and in order.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-sequence.png}
+    \end{minipage}
+
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchspike(body, tail)$} The Spike Looplet represents a run
+        followed by a single value. In this paper, Spike will be considered a
+        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
+        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
+        handled with special care, since they are an opportunity to align the
+        final run to the end of the root loop extent, without using any special
+        bounds inference.
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-spike.png}
+    \end{minipage}
+
+    \begin{minipage}[c]{0.6\linewidth}
+        \paragraph{$\finchstepper([seek], next, body)$} The stepper looplet
+        represents a variable number of looplets, concatenated. Since our
+        looplets may be skipped over due to conditions or various rewrites, the
+        $seek$ function allows us to fast-forward the state to the start of the
+        root loop extent when it comes time to lower the stepper. The $next$
+        function advances the state to the next iteration of the stepper. 
+    \end{minipage}%
+    \begin{minipage}[c]{0.4\linewidth}
+        \centering
+        \includegraphics[scale=0.25]{Looplets-stepper.png}
+    \end{minipage}
+    \caption{The looplet language, as understood in a correct execution of a Finch program.}
 \end{figure}

### 2024-03-21 export_looplet_figures.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+file=Looplets # File name excluding extension
+
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file-lookup.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 1 --output "$file-run.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 2 --output "$file-spike.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 3 --output "$file-sequence.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 4 --output "$file-stepper.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 5 --output "$file-phase.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 6 --output "$file-switch.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 7 --output "$file-thunk.png" "$file.drawio"

### 2024-03-21 sections/eval.tex
Update on Overleaf.

@@
 \subsubsection{SpMV}
-\willow{We might want to start this section off by saying spmv is a bandwidth bound kernel, and historically therefore subject to many format modifications to improve throughput and reduce the associated control flow}
-Finch provides many flexible level formats to efficiently capture a variety of patterns in datasets, encompassing both the structure and type of data.
-More specifically, these formats can represent banded, triangular, run-length-encoded, blocked, hashed, and boolean data, as well as several other formats.
-\willow{say this faster} Finch also provides the control flow necessary to manipulate the order in which data in these flexible level formats is read and written, enabling us to take advantage of multiple structural patterns concurrently—for example, we can exploit both sparsity and symmetry by using a sparse level format and restricting data reads to one triangle of a matrix.
+The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
+SpMV is a bandwidth bound kernel, and thus historically has been subject to many modifications to improve throughput and reduce the associated control flow [cite].
+The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
 
-Structure in data arises both naturally, due to the chemical and physical properties of matter, and artificially via mathematical operations that induce particular patterns.
-\willow{this might show up in the intro, but if you wanna say this, maybe give an exmaple here too (e.g.
-symmetry)} Closely tailoring a storage format to a particular data pattern enables us to reduce the amount of stored values, make data accesses more efficient, and take advantage of spatial locality, resulting in more performant code.
+In this case study, we highlight a few different Finch formats, as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
+Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g.
+sparsity \textit{and} symmetry).
 
-\willow{these paragraphs could be merged and condensed.
-Perhaps some of this belongs elsewhere} The sparse matrix-vector multiplication kernel is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
-\willow{The following two sentences are the most important and relevant so far} The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
-In this case study, we highlight a few different Finch formats and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
-
-\willow{fix the citations, put Figure \ref{fig:spmv} earlier, We display speedup relative to ...
-} Figures \ref{spmv_sorted} and \ref{spmv_grouped} compare Finch’s performance to that of TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We display speedup relative to row major or column major TACO performance, whichever was faster.
+We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.
 We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
@@
 to test the TACO library [cite].
-In addition, we also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
+Additionally, we included the SNAP graph collection to test with boolean matrices.
+We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
 The dense vector is randomly generated.
-We tested a wide variety of finch formats, described in the following subsection; the performance displayed for Finch on each dataset is the fastest among the formats we tested.
+We tested a wide variety of Finch formats; the performance displayed for Finch on each dataset is the fastest among the formats we tested.
 
@@
 \subsubsection{Tensor Formats}
-For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations (labeled with a "R" and "C", respectively, where applicable in Figure \ref{spmv_grouped}) with Finch.
+For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations with Finch.
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
@@
     \centering
+    \caption{SpMV Tensor Formats}
     \label{spmv_tensor_formats}
-    \caption{SpMV Tensor Formats}
     \begin{tabular}{|l|l|l|l|}
@@
 
-\subsubsection{4D Blocked SpMV}
-Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
+% Perhaps add the following section again in a future iteration of paper
+% \subsubsection{4D Blocked SpMV}
+% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
 Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
@@
 
-We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
+% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
 We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
 
-Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
+% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
 In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
@@
     \label{spmv_grouped}
+    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
+"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
 \end{figure}

### 2024-03-21 sections/lang.tex
Update on Overleaf.

@@
 
+\begin{figure}
+    \centering
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = read}
+    \hypo{e(tns) = tns'}
+    \infer3{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+    \end{prooftree}
+    \hfill
+    \vspace{2em}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = update}
+    \hypo{e(tns) = tns'}
+    \infer3{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+    \end{prooftree}
+      \vspace{2em}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{tns\not\in e}
+    \hypo{\langle init, s \rangle \Rightarrow v}
+    \hypo{\forall d_{i}\in dims, \langle init, d_{i} \rangle \Rightarrow v_{i}}
+    \infer4{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns, v, v_{0}, \dots)), t \rangle)}
+    \end{prooftree}
+
+    \caption{Rules for evaluating life cycle functions.}
+    \label{fig:enter-label}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \begin{prooftree}
+    \hypo{\langle val, (e, t) \rangle
+    \rightarrow val'}
+    \infer1{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
+    \end{prooftree}
+    \vspace{2em}
+    \caption{Basic Evaluation Rules}
+    \label{fig:enter-label}
+\end{figure}
+
 \[
@@
 \]
-\[
-\frac{\langle val, (e, t) \rangle \rightarrow val'}
-{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns' \quad  \langle init, s \rangle \rightarrow init' \quad  \langle dims_i, s \rangle \rightarrow dims_i'}
-{\langle\finchdeclare(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle declare(tns', init', dims'...), t \rangle)}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
-{\langle\finchthaw(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
-{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
-\]
 
 $E$ is an evaluation context that applies anywhere in the syntax tree.
-
-\[
-\frac{e(tns) = tns' \langle unfurl(tns'), t\rangle \rightarrow \langle tns'', t'\rangle}
-{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), (e, t)\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), (e, t')\rangle}
-\]
 

### 2024-03-21 sections/tensors.tex
Update on Overleaf.
more on tensors

@@
 \section{Bridging Looplets and Finch: The Tensor Interface}
-\teo{I am doing stuff in this section.}
-Looplets provide descriptions of iterators. 
+
 %
@@
 %
-Previous efforts to compiler a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
+Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
 %
@@
 %
-These interfaces add to the level abstraction.
+These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
 
 \subsection{Level Abstraction}
+Fiber-tree style tensor abstractions have been the subject of extensive study
+\cite{sze2017efficient, chou2022compilation, chou2018format}.  The underlying
+idea is to represent a multi-dimensional tensor as a nested vector
+datastructure, where each level of the nesting corresponds to a dimension of the
+tensor. Thus, a matrix would be represented as a vector of vectors. This kind of
+abstraction lends itself to representing sparse tensors if we vary the type of
+vector used at each level in a tree. Thus, a sparse matrix might be represented
+as a dense vector of sparse vectors. The vector of subtensors in this
+abstraction is referred to as a \textbf{fiber}. Prior fiber-tree representations
+focus on sparsity (where only the nonzero elements are represented) and treat
+sparse vectors as sets of represented points. Since our fiber-tree
+represesentation must handle other kinds of structure, such as diagonal,
+repeated, or constant values, we instead view each fiber as a mapping from
+indices into a space of subfibers.
+
+Instead of storing the data for each subfiber separately, most sparse tensor
+formats such as CSR, DCSR, and COO usually store the data for all fibers in a
+level contiguously. In this way, we can think of a level as a bulk allocator for
+fibers. Continuing the analogy, we can think of each fiber as being
+disambiguated by a \textbf{position}, or an index into the bulk pool of
+subfibers. The mapping $f$ from indices to subfibers is thus a mapping from an
+index and a position in a level to a subposition in a sublevel.
+Figure~\ref{fig:help} shows a simple example of a level as a fiber allocator.
+
+Our simplified view of a level is enabled by our use of looplets to represent
+the structure within each fiber. In fact, our level interface requires only
+5 highly general operations, described below:
+
+\paragraph[$declare(tns, init, dims...)$]
 

### 2024-03-21 sections/lang.tex
more rules ugh

--- 2024-03-21 sections/lang.tex
@@
 
-\[
-    \frac{}{\langle\finchliteral(val), (e, t)\rangle \rightarrow val} \quad
-    \frac{}{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle} \quad
-\]
-\[
-    \frac{}{\langle\finchvar(name), (e, t)\rangle \rightarrow e(\finchvar(name))} \quad
-    \frac{}{\langle\finchindex(name), (e, t)\rangle \rightarrow e(\finchindex(name))}
-\]
-\[
-    \frac{\langle args_i, (e, t) \rangle \Rightarrow vals_i \quad \langle f, (e, t) \rangle \Rightarrow g}
-    {\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
-\]
-\[
-\frac{\langle head, s_1 \rangle \rightarrow s_2}
-{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle} \quad
-\frac{}{\langle\finchblock(), s\rangle \rightarrow s}
-\]
-\[
-\frac{\langle cond, s \rangle \Rightarrow true}
-{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}\quad
-\frac{\langle cond, s \rangle \Rightarrow false}
-{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
-\]
-\[
-\frac{\langle val, (e, t) \rangle \rightarrow val'}
-{\langle\finchdefine(var, val, body), (e, t)\rangle \rightarrow \langle body, (e[var \mapsto val'], t) \rangle}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns' \quad  \langle init, s \rangle \rightarrow init' \quad  \langle dims_i, s \rangle \rightarrow dims_i'}
-{\langle\finchdeclare(tns, init, dims...), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle declare(tns', init', dims'...), t \rangle)}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = read \quad e(tns) = tns'}
-{\langle\finchthaw(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
-\]
-\[
-\frac{s = (e, t) \quad e(\finchmode(tns)) = update \quad e(tns) = tns'}
-{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
-\]
+the state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
+environment and another state $t$ corresponding to the embedding language.
+
+\begin{figure}
+    \centering
+    \small
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = read}
+    \hypo{e(tns) = tns'}
+    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = update}
+    \hypo{e(tns) = tns'}
+    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+    \end{prooftree}
+      \vspace{12pt}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(tns) = tns'}
+    \hypo{\langle init, s \rangle \Rightarrow init'}
+    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
+    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
+    \end{prooftree}
+
+    \caption{Rules for evaluating life cycle functions.}
+    \label{fig:enter-label}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \small
+
+    \begin{prooftree}
+    \hypo{\langle val, (e, t) \rangle
+    \rightarrow val'}
+    \infer1[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
+    \end{prooftree}
+    \hfill
+    \begin{prooftree}
+        \hypo{\langle args_i, (e, t) \rangle \Rightarrow vals_i}
+        \hypo{\langle f, (e, t) \rangle \Rightarrow g}
+        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}
+        \hypo{}
+        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
+    \end{prooftree}
+    \hfill
+    \begin{prooftree}
+        \hypo{}
+        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}
+        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
+    \end{prooftree}
+    \hfill
+    \begin{prooftree}
+        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}
+        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
+        \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
+    \end{prooftree}
+    \hfill
+    \begin{prooftree}
+        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}
+        \hypo{\langle cond, s \rangle \Rightarrow true}
+        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
+    \end{prooftree}
+    \hfill
+    \begin{prooftree}
+        \hypo{\langle cond, s \rangle \Rightarrow false}
+        \infer1[$SieveFalse$]{\langle\finchsieve(cond, body), s\rangle \rightarrow s}
+    \end{prooftree}
+
+    \caption{Basic Evaluation Rules}
+    \label{fig:enter-label}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \small
 
 $E$ is an evaluation context that applies anywhere in the syntax tree.
+\vspace{12pt}
 
-\[
-\frac{e(tns) = tns' \langle unfurl(tns'), t\rangle \rightarrow \langle tns'', t'\rangle}
-{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), (e, t)\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), (e, t')\rangle}
-\]
+\begin{prooftree}
+    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
+\end{prooftree}
+\hfill
+\begin{prooftree}
+    \hypo{e(i) = i'}
+    \hypo{\langle seek(i'), t \rangle \rightarrow t'}
+    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
+\end{prooftree}
+\vspace{12pt}
 
-\[
-\frac{}
-{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(phase(range, body), tail...), j..., i)]), s\rangle
-    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
-\]
+\begin{prooftree}
+    \hypo{\langle cond, t \rangle \Rightarrow true}
+    \infer1[]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
+\end{prooftree}
+\hfill
+\begin{prooftree}
+    \hypo{\langle cond, t \rangle \Rightarrow false}
+    \infer1[$Switch$]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
+\end{prooftree}
+\vspace{12pt}
 
-\[
-\frac{}
-{\langle \finchloop(i, ext, E[\finchaccess(sequence(), j..., i)]), s\rangle \rightarrow \langle \finchblock(), s\rangle}
-\]
+\begin{prooftree}
+    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
+\end{prooftree}
+\vspace{12pt}
 
-\[
-\frac{}
-{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(next, body), j..., i)]), s\rangle
-    }{\rightarrow \langle \finchblock(\finchloop(i, ext \cup range, E[\finchaccess(body, j..., i)]), \finchloop(i, ext, E[\finchaccess(sequence(tail...), j..., i)])), s\rangle}}
-\]
+\begin{prooftree}
+    \hypo{\langle \finchloop(i, ext, E[\finchaccess(head, j..., i)]), s\rangle \rightarrow s'}
+    \infer1[$Sequence$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(sequence(head, tail), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(tail, j..., i)]), s'\rangle}}
+\end{prooftree}
+\vspace{12pt}
+
+\begin{prooftree}
+    \hypo{\langle seek(a), t \rangle \rightarrow t'}
+    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
+\end{prooftree}
+\vspace{12pt}
+
+\begin{prooftree}
+    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
+    \hypo{\langle next, t' \rangle \rightarrow t''}
+    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
+\end{prooftree}
+
+    \caption{Looplet Rules}
+    \label{fig:enter-label}
+\end{figure}
 

### 2024-03-22 sections/intro.tex
Update on Overleaf.

@@
 convincing. A figure could be nice.} 
+\teo{I Added citations/heft but we maybe need to improve the figure?}
 \begin{figure}
@@
 %
-
-Sparse arrays (which store only nonzero elements) describe networks and
-simulations.
+Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
 %
 Run-length encoding describes images and masks, geometry, and
-databases (such as a list of transactions with the date field all the same).
+databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
 %
-Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing as well as in intermediate structures in many linear solvers.
+Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g.
+higher order FEMs) as well as in intermediate structures in many linear solvers (e.g.
+GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
+%
+In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
 %
@@
 with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
+
 %
@@
 Experts must hand write variations of even the simplest kernels, like matrix
-multiply, for each data structure and architecture to get performance.
+multiply, for each data structure/data set and architecture to get performance.
 %
 Implementations must choose a small set of features to support well, resulting
-in a compromise between \textbf{program flexibility} and \textbf{datastructure
+in a compromise between \textbf{program flexibility} and \textbf{data structure
 flexibility}.
@@
 Hand-written solutions are collected in diverse libraries like
-MKL, OpenCV, or SciPy. 
+MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
 %
 However, libraries will only ever support a subset of
-programs on a subset of datastructure combinations. Even the most advanced
+programs on a subset of data structure combinations.
+%
+Even the most advanced
 libraries, such as the GraphBLAS, which support a wide variety of sparse
-operations over various semirings always lack support for other features, such
-as tensors, fused outputs, or runs of repeated values. 
+operations over various semi-rings always lack support for other features, such
+as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
 %
@@
 compilers support an enormous variety of program constructs like early break and
-multiple left hand sides, they only support dense arrays.  
+multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
 %
 Special-purpose
-compilers like TACO or CoRa which support a select subset of structured data
+compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
 structures (only sparse, or only ragged arrays) must compromise by greatly
 constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, fegade_cora_2022}. 
+contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
 %

### 2024-03-22 sections/intro.tex
Update on Overleaf.

@@
 %
-This trade-off is
-visualized in figure \ref{fig:datavsprogram}.
+This trade-off is visualized in figure \ref{fig:datavsprogram}.
+%
+We hypothesize that
 
@@
   \begin{tabular}{l|ccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
+  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
   \hline
-  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  High-Level API           &            & \checkmark &            &            & \checkmark \\
-  Automatic API Fusion     &            &            &            &            & \checkmark \\
-%  Parallelism              & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  Multiple LHS             & \checkmark &            & \checkmark & \checkmark & \checkmark \\
-  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark \\
-  Recurrence               & \checkmark &            &            &            &            \\
-  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark & \checkmark \\
-  Scatter Gather           & \checkmark &            &            & \checkmark & \checkmark \\
-  Early Break              &            & \checkmark &            &   \checkmark         & \checkmark \\
+  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
+  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
+%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
+  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
+  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
+  Recurrence               & \checkmark &            &            &            &            &           \\
+  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
+  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
+  Early Break              &            & \checkmark &            & \checkmark &         & \checkmark \\
   \end{tabular}
@@
   \begin{tabular}{l|ccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{Finch} \\
+  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
   \hline
-  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  Padded                   & \checkmark &            &            &            & \checkmark \\
-  One Sparse               &            & \checkmark &            & \checkmark & \checkmark \\
-  Sparse                   &            & \checkmark &            &            & \checkmark \\
-  Run-length               &            &            &            &            & \checkmark \\
-  Symmetric                &            &            &            &            & \checkmark \\
-  Regular Sparse Blocks    &            & \checkmark &            &            & \checkmark \\
-  Irregular Sparse Blocks  &            &            &            &            & \checkmark \\
-  Ragged                   &            &            & \checkmark &            & \checkmark \\
+  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  Padded                   & \checkmark &            &            &            &            & \checkmark \\
+  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
+  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
+  Run-length               &            &            &            &            &            & \checkmark \\
+  Symmetric                &            &            &            &            & \checkmkar & \checkmark \\
+  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
+  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
+  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
   \end{tabular}

### 2024-03-22 main.tex
syntax
some more stuff
more semantics
Update on Overleaf.
fix
more rules ugh

--- 2024-03-20 main.tex
@@
 %\usepackage{amssymb}
+\usepackage{mathtools}
 \usepackage{array}
@@
 \usepackage{ifthen}
+\usepackage{ebproof}
+\ebproofset{label separation=0pt}
+\usepackage{multirow}
 
@@
 }
+\setmintedinline{fontsize=\small}
 %\usepackage[colorlinks=true, allcolors=blue]{hyperref}
@@
 \newcommand\help[1]{\textcolor{green}{Help Needed: #1}}
+\newcommand\kyle[1]{\textcolor{olive}{Kyle: #1}}
 \newcommand\teo[1]{\textcolor{red}{teo: #1}}
@@
 \newcommand{\leftoff}{\textcolor{red}{\textbf{LEFT OFF}}}
-\newcommand{\changwan[1]{\textcolor{blue}{CW: #1}}
+\newcommand\changwan[1]{\textcolor{blue}{CW: #1}}
 }{
@@
 \newcommand{\saman}[1]{}
+\newcommand{\kyle}[1]{}
 \newcommand{\teo}[1]{}
@@
 }
+
+\newcommand{\finchextent}{\ensuremath{\text{\textbf{extent}}}}
+\newcommand{\finchliteral}{\ensuremath{\text{\textbf{literal}}}}
+\newcommand{\finchvalue}{\ensuremath{\text{\textbf{value}}}}
+\newcommand{\finchtensor}{\ensuremath{\text{\textbf{tensor}}}}
+\newcommand{\finchmode}{\ensuremath{\text{\textbf{mode}}}}
+\newcommand{\finchindex}{\ensuremath{\text{\textbf{index}}}}
+\newcommand{\finchvar}{\ensuremath{\text{\textbf{variable}}}}
+\newcommand{\finchcall}{\ensuremath{\text{\textbf{call}}}}
+\newcommand{\finchaccess}{\ensuremath{\text{\textbf{access}}}}
+\newcommand{\finchread}{\ensuremath{\text{\textbf{read}}}}
+\newcommand{\finchassign}{\ensuremath{\text{\textbf{assign}}}}
+\newcommand{\finchupdate}{\ensuremath{\text{\textbf{update}}}}
+\newcommand{\finchloop}{\ensuremath{\text{\textbf{loop}}}}
+\newcommand{\finchdefine}{\ensuremath{\text{\textbf{define}}}}
+\newcommand{\finchsieve}{\ensuremath{\text{\textbf{sieve}}}}
+\newcommand{\finchblock}{\ensuremath{\text{\textbf{block}}}}
+\newcommand{\finchdeclare}{\ensuremath{\text{\textbf{declare}}}}
+\newcommand{\finchfreeze}{\ensuremath{\text{\textbf{freeze}}}}
+\newcommand{\finchthaw}{\ensuremath{\text{\textbf{thaw}}}}
+
+\newcommand{\finchthunk}{\ensuremath{\text{\textbf{thunk}}}}
+\newcommand{\finchphase}{\ensuremath{\text{\textbf{phase}}}}
+\newcommand{\finchswitch}{\ensuremath{\text{\textbf{switch}}}}
+\newcommand{\finchlookup}{\ensuremath{\text{\textbf{lookup}}}}
+\newcommand{\finchrun}{\ensuremath{\text{\textbf{run}}}}
+\newcommand{\finchspike}{\ensuremath{\text{\textbf{spike}}}}
+\newcommand{\finchsequence}{\ensuremath{\text{\textbf{sequence}}}}
+\newcommand{\finchstepper}{\ensuremath{\text{\textbf{stepper}}}}
+
+\newcommand{\finchplan}{\ensuremath{\text{\textbf{plan}}}}
+\newcommand{\finchquery}{\ensuremath{\text{\textbf{query}}}}
+\newcommand{\finchtable}{\ensuremath{\text{\textbf{table}}}}
+\newcommand{\finchmapjoin}{\ensuremath{\text{\textbf{mapjoin}}}}
+\newcommand{\finchaggregate}{\ensuremath{\text{\textbf{aggregate}}}}
+\newcommand{\finchrelabel}{\ensuremath{\text{\textbf{relabel}}}}
+\newcommand{\finchreorder}{\ensuremath{\text{\textbf{reorder}}}}
 
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Finch: A Datastructure-Driven Array Programming Language}
+\title{Finch: Sparse and Structured Array Programming with Control Flow}
 
@@
 Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
- Thus, focusing only on sparsity limits the scope needed to handle a vast array of data.
+ Thus, focusing only on sparsity limits the scope needed to handle a vast array of data \kyle{"vast array" is a good pun here, but a bit confusing to read}.
 Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 \subsection{Implementing Numpy Array API in Finch}
+In the past decade, the adoption of the python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
+They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
+Further, existing work doesn't support the fusion of arbitrary operators which can have a drastic impact on performance as we show in Fig.
+(\kyle{PUT FIG HERE}).
+We believe that a flexible, online compiler like Finch which can manage the complexity of arbitrary operations between inputs with a wide variety of formats is the secret ingredient needed to make the array API performant for structured data.
+
+However, there is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both outputs and the algorithm used to produce those outputs.
+The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
+Unfortu
+
+Due to the wide variety of formats and protocols supported by Finch, it allows for a large array of potential impleme
 
@@
 
+\begin{align*}
+    \finchreorder(name, idxs...) := 
+\end{align*}
+
 \subsubsection{Finch Interpreter}
@@
 \subsubsection{Lowering}
+
 \subsubsection{Heuristic Optimization}

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 \subsection{Implementing Numpy Array API in Finch}
-In the past decade, the adoption of the python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
+In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
 They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
@@
 
-However, there is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both outputs and the algorithm used to produce those outputs.
+There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both outputs and the algorithm used to produce those outputs.
 The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
-Unfortu
+This approach requires a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
+For example,
 

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 
-There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both outputs and the algorithm used to produce those outputs.
+There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both the output and the algorithm.
 The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
-This approach requires a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
-For example,
-
-Due to the wide variety of formats and protocols supported by Finch, it allows for a large array of potential impleme
+This approach results in a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
+To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user.
+This is implemented through 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which lowers Finch Logic programs to one or more Finch programs while making heuristic decisions about output format, loop ordering, and protocols.
 

### 2024-03-22 sections/intro.tex
Update on Overleaf.
make it build

@@
 %
-We hypothesize that
 
@@
   \centering
-  \begin{tabular}{l|ccccc}
+  \begin{tabular}{l|cccccc}
   \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
@@
   Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
-  Early Break              &            & \checkmark &            & \checkmark &         & \checkmark \\
+  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
   \end{tabular}
@@
 For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
+
+\teo{Data structure driven array program performance engineering?}
 

### 2024-03-22 sections/tensors.tex
more rules ugh
cool
make it build

@@
 
+Our view of a level as a fiber allocator implies an allocation function
+$assemble(tns, pos_start:pos_stop)$, which allocates fibers at positions
+$pos_start:pos_stop$ in the level. We don't specify a de-allocation function,
+instead relying on initialization to reset the fiber if it needs to be reused.
+
+\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+
 Our simplified view of a level is enabled by our use of looplets to represent
@@
 
-\paragraph[$declare(tns, init, dims...)$]
-
-\begin{enumerate}
-\item fibers and levels
-\item assembly
-\item reassembly
-\end{enumerate}
-
-\subsection{Instatiation Interface}
-
-\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.
 

### 2024-03-22 sections/intro.tex
build

@@
   Run-length               &            &            &            &            &            & \checkmark \\
-  Symmetric                &            &            &            &            & \checkmkar & \checkmark \\
+  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
   Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 The dense vector is randomly generated.
-We tested a wide variety of Finch formats; the performance displayed for Finch on each dataset is the fastest among the formats we tested.
+We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset in Figure \ref{spmv_grouped} is the fastest among the formats and programs we tested.
+Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
 
@@
     \end{minipage}
-    \caption{Performance of SpMV across various tools.}
+    \caption{Finch SpMV Programs}
+    \label{spmv_programs}
 \end{figure}
-
-\subsubsection{Tensor Formats}
-For the tensor formats in Table \ref{spmv_tensor_formats}, we tested with both the row-major and the column-major implementations with Finch.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
 
@@
     \end{tabular}
-
 \end{table}
 
+\subsubsection{Tensor Formats}
 We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
@@
 For example, Finch Pattern SpMV had a speedup of 2.51 over TACO for SNAP/email-Eu-core and a speedup of X over the same Finch program, but with an Element format.
+
+\begin{table}[htbp]
+    \centering
+    \caption{SpMV Tensor Formats}
+    \label{spmv_tensor_formats}
+    \begin{tabular}{|l|l|l|}
+        \hline
+        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions}\\
+        \hline
+        & HB/saylr4 & 3,564 x 3,564 (22316) \\
+        \hline 
+    \end{tabular}
+\end{table}
 
@@
 
-\subsection{Implementing Numpy Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
+\subsection{Implementing Numpy's Array API in Finch}
+In the past decade, the adoption of the Numpy Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
 They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 
-In this case study, we highlight a few different Finch formats, as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
+In this case study, we highlight a few different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
 Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g.
@@
 We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format in comparison to both TACO and the Finch SparseList formats (which we consider the closest analog to the sparse matrix format used in TACO and other libraries) with speedups of 2.16, 1.55, and 1.30 relative to TACO and of 1.47, 2.13, and 1.42 relative to the fastest SparseList Finch SpMV implementation.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO and a speedup of 1.47 and 1.24 relative to the fastest SparseList Finch SpMV implementation.
+We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
+
+Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
+Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
 
 There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, Finch Pattern SpMV had a speedup of 2.51 over TACO for SNAP/email-Eu-core and a speedup of X over the same Finch program, but with an Element format.
+For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
 
@@
     \centering
-    \caption{SpMV Tensor Formats}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|}
+    \caption{SpMV Sample Matrices}
+    \label{spmv_sample_matrices}
+    \begin{tabular}{|m{3cm}|c|c|c|}
         \hline
-        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions}\\
+        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions} & \textbf{Best Finch Format}\\
         \hline
-        & HB/saylr4 & 3,564 x 3,564 (22316) \\
+        \includegraphics[width=3cm]{spmv_matrices/saylr4.png} & HB/saylr4 & 3,564 x 3,564 (22,316) & Symmetric SparseList \\
+        \hline 
+        \includegraphics[width=3cm]{spmv_matrices/heart3.png} & Norris/heart3 & 2,339 x 2,339 (680,341) & SparseVBL \\
+        \hline 
+        \includegraphics[width=3cm]{spmv_matrices/as-735.png} & SNAP/as-735 & 7,716 x 7,716 (26,467) & Symmetric SparseList-Pattern \\
         \hline
@@
 Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-However, the regular Finch SpMV program has better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.30x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.15x over TACO.
-Notably, there is a 1.90x speedup for the HB/saylr4 matrix over TACO.
+However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
+Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
+Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
 
@@
 "R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
+The baseline Finch format is unsymmetric Dense(SparseList(Element)).
 \end{figure}

### 2024-03-22 sections/tensors.tex
figure changes

@@
 %
-Our next two abstractions adds to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
-%
-Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
+Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
 %
@@
 These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
+%
+Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
 
@@
 index and a position in a level to a subposition in a sublevel.
-Figure~\ref{fig:help} shows a simple example of a level as a fiber allocator.
+Figure~\ref{fig:levelsvsfibers} shows a simple example of a level as a pool of fibers.
 
-Our view of a level as a fiber allocator implies an allocation function
-$assemble(tns, pos_start:pos_stop)$, which allocates fibers at positions
-$pos_start:pos_stop$ in the level. We don't specify a de-allocation function,
-instead relying on initialization to reset the fiber if it needs to be reused.
+When we need to refer to a particular fiber at position $p$ in the level $l$, we
+may write $fiber(l, p)$. Note that the formation of fibers from levels is lazy,
+and the data underlying each fiber is managed entirely by the level, so the
+level may choose to overlap the storage between different fibers. Thus, the only
+unique data associated with $fiber(l, p)$ is the position $p$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.45\linewidth]{LevelsVsFibers-matrix.png}\hfill%
+    \includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
+    \caption{Levels and fiber tree representations of a sparse matrix and a sparse tensor.
+On left, a matrix is represented in a fibertree corresponding to CSC format, with a dense outer level and a sparse inner level.
+On right, a tensor is represented in a fibertree with two sparse outer levels, and a dense inner level.
+Note that the element levels in this case form the leaves of the tree.}
+    \label{fig:levelsvsfibers}
+\end{figure}
 
@@
 
+Our view of a level as a fiber allocator implies an allocation function
+$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
+$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation function,
+instead relying on initialization to reset the fiber if it needs to be reused.
+
 \paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
+\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
+\paragraph{$freeze(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.
 

### 2024-03-22 export_looplet_figures.sh

@@ -9,3 +9,8 @@ file=Looplets # File name excluding extension
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 5 --output "$file-phase.png" "$file.drawio"
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 6 --output "$file-switch.png" "$file.drawio"
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 7 --output "$file-thunk.png" "$file.drawio"
+
+file=LevelsVsFibers # File name excluding extension
+
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file-tensor.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 1 --output "$file-matrix.png" "$file.drawio"

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
         \hline 
+        \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} & toeplitz\_large\_band  & 10,000 x 10,000 (1,999,900) & SparseBand \\
+        \hline
         \includegraphics[width=3cm]{spmv_matrices/as-735.png} & SNAP/as-735 & 7,716 x 7,716 (26,467) & Symmetric SparseList-Pattern \\
@@
 For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-\begin{minted}{julia}
-            @finch begin
-                for J = _
-                    for I = _
-                        for j = 1:10
-                            for i = 1:10
-                                let _i = (I - 1) * 10 + i
-                                    y[_i] += block_A[i, j, I, J] * block_x[j, J]
-                                end
-                            end
-                        end
-                    end
-                end
-            end
-\end{minted}
 

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 \subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Numpy Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
+In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
 They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
@@
 
-\subsubsection{The Finch High-Level API (Needs a Name)}
+To get a sense of the Python Array API, we provide a few examples of their
 

### 2024-03-22 sections/tensors.tex
some notes
some stuff

@@
 
+%What needs to be said about the tensors? It's basically a few main points:
+%1. (DONE) What is the level abstraction 
+%2. (Done) We have identified 8 key level structures to represent most combinations of runs or pinpoints.
+%2. a. beautiful figure with datas as rows of larger structures
+%2. b. How do we write to each of these, in order? Randomly? random access? several important datastructures that show up along the way.
+%3. What is "Unfurling"? and how does it help us represent each structure? refer to looplet decompositions of each case in the earlier figure
+%4. What are lifecycles? How do lifecycles help us keep sane? Why are they necessary for correctness?
+
 \subsection{Level Abstraction}
@@
 indices into a space of subfibers.
+
+%From the docs:
+%Finch represents tensors hierarchically in a tree, where each node in the tree is a vector of subtensors and the leaves are the elements.
+Thus, a matrix is analogous to a vector of vectors, and a 3-tensor is analogous to a vector of vectors of vectors.
+The vectors at each level of the tensor all have the same structure, which can be selected by the user.
+%In a Finch tensor tree, the child of each node is selected by an array index.
+All of the children at the same level will use the same format and share the same storage.
+Finch is column major, so in an expression A[i_1, ..., i_N], the rightmost dimension i_N corresponds to the root level of the tree, and the leftmost dimension i_1 corresponds to the leaf level.
+%We refer to a node in the tree as a subfiber.
+All of the nodes at the same level are stored in the same datastructure, and disambiguated by an integer position.
+in the above example, there are three levels: the rootmost level contains only one subfiber, the root.
+The middle level has 3 subfibers, one for each column.
+The leafmost level has 12 subfibers, one for each element of the array.
+For example, the first level is A_fbr.lvl, and we can represent it's third position as SubFiber(A_fbr.lvl.lvl, 3).
+The second level is A_fbr.lvl.lvl, and we can access it's 9th position as SubFiber(A_fbr.lvl.lvl.lvl, 9).
+For instructional purposes, you can use parentheses to call a subfiber on an index to select among children of a subfiber.
+%When we print the tree in text, positions are numbered from top to bottom. However, if we visualize our tree with the root at the top, positions range from left to right:
+%Dense Format Index Tree
+%Because our array is sparse, (mostly zero, or another fill value), it would be more efficient to store only the nonzero values.
+In Finch, each level is represented with a different format.
+A sparse level only stores non-fill values.
+This time, we'll use a tensor constructor with sl (for "SparseList of nonzeros") instead of d (for "Dense"):
+%CSC Format Index Tree
+%Our Dense(SparseList(Element(0.0))) format is also known as "CSC" and is equivalent to SparseMatrixCSC.
+The Tensor function will perform a zero-cost copy between Finch fibers and sparse matrices, when available.
+CSC is an excellent general-purpose representation when we expect most of the columns to have a few nonzeros.
+However, when most of the columns are entirely fill (a situation known as hypersparsity), it is better to compress the root level as well:
+%DCSC Format Index Tree
+%Here we see that the entirely zero column has also been compressed. The SparseList(SparseList(Element(0.0))) format is also known as "DCSC".
+%The COO format is compact and straightforward, but doesn't support random access.
+For random access, one should use the SparseHash format.
+A full listing of supported formats is described after a rough description of shared common internals of level, relating to types and storage.
+%All levels have a postype, typically denoted as Tp in the constructors, used for internal pointer types but accessible by the function:
+%postype(lvl)
 
@@
 
+\subsection{The 8 Key Level Structures}
+    The main benefits of specializing to structure come from the following properties of the data:
+    \begin{enumerate}
+        \item[Sparsity] Sparse data is data that is mostly zero, or some other
+        fill value. When we specialize on this data, we can use annihilation ($x
+        * 0 = 0$), identity ($x * 1 = 1$), and other constant propagation
+        * properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
+        redundant work.
+
+        \item[Blocks] Blocked data is a subset of sparse data where the nonzeros
+        are clustered and occur adjacent to one another. This provides us with
+        two opportunities: We can avoid storing the locations of the nonzeros
+        individually, and we can use a much more efficient randomly accessible
+        iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.
+
+        \item[Runs] Runs of repeated values may occur in dense or sparse code,
+        cutting down on storage and allowing us to use integration rules such as 
+        \mintinline{julia}{for i = 1:n; s += x; end} $\rightarrow$
+        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops.
+
+        \item[Singletons] When we have only one non-fill region in sparse data,
+        we can use this fact to avoid a loop and instead just compile code for
+        only the single region \cite{ahrens_looplets_2023, ghorbani2023compiling}.
+    \end{enumerate}
+
+    In Finch, we have identified 8 key level structures that can represent all
+    of the relevant combinations of these properties, summarized in Table
+    \ref{tab:TypesOfStructure}. We examine each structure in turn, describing
+    some of the key properties and potential use cases of each. In this sense,
+    the structures we consider are exhaustive. We can represent a wide variety
+    of hierarchical tensor structures by combining these level structures in a
+    tree, as shown in Figure~\ref{fig:levels}.
+    \begin{table}
+        \centering
+        \begin{tabular}{|c|c|c|c|l|}
+            \hline
+            \textbf{Sparse} & \textbf{Blocked} & \textbf{Runs} & \textbf{Singletons} & \textbf{Corresponding Format} \\
+            \hline
+             &  &  &  & Dense \\
+            \hline
+             &  & \checkmark &  & DenseRLE \\
+            \hline
+            \checkmark &  &  &  & Sparse \\
+            \hline
+            \checkmark &  &  & \checkmark & SparsePinpoint \\
+            \hline
+            \checkmark &  & \checkmark &  & SparseRLE \\
+            \hline
+            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
+            \hline
+            \checkmark & \checkmark &  &  & SparseVBL \\
+            \hline
+            \checkmark & \checkmark &  & \checkmark & SparseBand \\
+            \hline
+        \end{tabular}
+        \caption{All combinations of relevant structural properties and their
+        corresponding formats.  Note that blocks and runs need not be considered
+        together because we must store a run length for each run, and so there
+        isn't a significant storage benefit to combining them. Blocks and
+        singletons only make sense in the contex of sparsity, so we don't
+        consider them together either. We omit such combinations
+        from the exhaustive table.}
+        \label{tab:TypesOfStructure}
+    \end{table}
+
+    \begin{figure}
+        \centering
+        %\includegraphics[width=0.45\linewidth]{structuraldiversity.png}\hfill%
+        \caption{This figure is a placeholder, but will eventually contain some very beguiling trees of levels.}
+        \label{fig:structuraldiversity}
+    \end{figure}
+
 \subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 
-To get a sense of the Python Array API, we provide a few examples of their
+To get a sense of the Python Array API, we provide a few examples of required functions which we will then translate to Finch Logic statements.
 
@@
 \begin{align*}
-    \finchreorder(name, idxs...) := 
+    \finchplan(queries...) &  \finchquery(name, expr) & \finchreorder(expr, idxs...)\\
+    \finchrelabel(expr, idxs...)& \finchmapjoin(op, exprs...) & \finchaggregate(op, expr)\\
+    & \finchtable(tns, idxs...) & \\
 \end{align*}

### 2024-03-22 sections/tensors.tex
some fixes

@@
         fill value. When we specialize on this data, we can use annihilation ($x
-        * 0 = 0$), identity ($x * 1 = 1$), and other constant propagation
-        * properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
+        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
+        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
         redundant work.
@@
         two opportunities: We can avoid storing the locations of the nonzeros
-        individually, and we can use a much more efficient randomly accessible
+        individually, and we can use more efficient randomly accessible
         iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.
@@
         cutting down on storage and allowing us to use integration rules such as 
-        \mintinline{julia}{for i = 1:n; s += x; end} $\rightarrow$
-        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops.
+        \mintinline{julia}{for i = 1:n; s += x end} $\rightarrow$
+        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops \cite{donenfeld_unified_2022,ahrens_looplets_2023}.
 
         \item[Singletons] When we have only one non-fill region in sparse data,
-        we can use this fact to avoid a loop and instead just compile code for
-        only the single region \cite{ahrens_looplets_2023, ghorbani2023compiling}.
+        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
     \end{enumerate}

### 2024-03-22 sections/eval.tex
Update on Overleaf.

--- 2024-03-22 sections/eval.tex
@@
 
-To get a sense of the Python Array API, we provide a few examples of their
-
 \subsubsection{Finch Logic}
+The Finch Logic language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. In this analogy,
 
 \begin{align*}
-    \finchreorder(name, idxs...) := 
+    \finchplan(queries...) &  \quad\quad\quad \finchquery(name, expr) & \finchreorder(expr, idxs...)\\
+    \finchrelabel(expr, idxs...)& \quad\quad\quad \finchmapjoin(op, exprs...)  & \finchaggregate(op, expr)\\
+    \quad\quad\quad & \quad\quad\quad \finchtable(tns, idxs...)\quad\quad\quad  & \quad\quad\quad \\
 \end{align*}

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 \subsubsection{Finch Logic}
-The Finch Logic language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. In this analogy,
+The logical language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. This means that ev
 

### 2024-03-22 sections/eval.tex
Update on Overleaf.

@@
 \subsubsection{Finch Logic}
-The logical language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. This means that ev
+The qulogical language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors.
+This means that every operator in the language (with the exception of $\finchtable$ which is analagous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor.
+The uniformity of types within this language allows for straightfor
 

### 2024-03-23 sections/eval.tex
Update on Overleaf.

--- 2024-03-22 sections/eval.tex
@@
 We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface [cite], Ahrens et al.
-to test a variable block row format partitioning strategy [cite], and Kjolstad et al.
-to test the TACO library [cite].
+to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
+to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
+to test the TACO library \cite{kjolstad_tensor_2017}.
 Additionally, we included the SNAP graph collection to test with boolean matrices.
@@
 \subsubsection{Finch Logic}
-The logical language takes inspiration from relational algebra while incorporating the ordered nature of dimensions in sparse tensors. This means that ev
+The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
+This means that every operator (with the exception of $\finchtable$ and $\finchalias$ which are analogous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor.
+Conceptually, an indexed tensor can be thought of as a relation with an ordering on the attributes, and we include the $\finchreorder$ operator to manipulate the order of these attributes.
+This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.
+
+To express materialization, reuse of common sub-expressions, and multiple outputs, we define queries and plans.
+A query is the assignment of the output of an expression to a name.
+We use this assignment to denote that we materialize an expression in memory.
+Later queries can then access the result of earlier queries through the alias operator which allows multiple queries to benefit from a shared computation.
+A plan is a sequence of queries and it produces a set of tensors as output.
 
 \begin{align*}
-    \finchplan(queries...) &  \quad\quad\quad \finchquery(name, expr) & \finchreorder(expr, idxs...)\\
-    \finchrelabel(expr, idxs...)& \quad\quad\quad \finchmapjoin(op, exprs...)  & \finchaggregate(op, expr)\\
-    \quad\quad\quad & \quad\quad\quad \finchtable(tns, idxs...)\quad\quad\quad  & \quad\quad\quad \\
+    \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
+    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \finchaggregate(op, expr)\\
+    \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+     expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias\quad\quad
+\end{align*}
+
+Given this language, we now describe how to define a few example functions from the API as Finch Logic expressions, 
+\begin{align*}
+\text{sum(M, dims=[2])} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchrelabel(M, i_1,\ldots,i_d), i_2)), out)\\
+\text{matmul(A, B)} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)), out)\\
 \end{align*}
@@
 \subsubsection{Finch Interpreter} 
+\textbf{Lowering}
 
-\subsubsection{Lowering}
+\textbf{Heuristic Optimization}
 
-\subsubsection{Heuristic Optimization}
+\subsubsection{Finch Interpreter}
 

### 2024-03-24 sections/intro.tex
Update on Overleaf.

@@
 \end{table}
-
-We see two main challenges to writing efficient code over structured data.
+\teo{ New paragraph:
+Many of these special purpose compilers build on some conception of an iteration space: TACO has merge lattices, Polyhedral compilers rely on the polyhedral model, Halide and Corra uses rectangular intervals, and so on.
+%
+And we believe the limitations of these systems can be traced back to this conception.
+%
+For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
+%
+And many iteration space concepts simply don't allow for skipping work that isn't simply the complement of an interval.
+%
+In our work, we will show how we can use Looplets as our main tool to manage the iteration spaces of our computations and data structures.
+%
+However, to support as many data structures and iteration spaces as is implied above, we face two challenges if we want to also produce efficient code over structured data.
+}
+%We see two main challenges to writing efficient code over structured data.
 
@@
 \teo{Data structure driven array program performance engineering?}
+
+%% Modification: We don't propose to fix these directly, but we do propose a programming model where they can be dealt with 
+%% By supporting the features above,
+
+\teo{Proposed new paragraph: We don't propose to solve these problems automatically, but we do propose a new programming model where these issues are solvable.
+%
+By combining many data structures and iteration spaces into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
+%
+We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
+%
+(Note quite sure how much comes after this)
+(We should say the predicability is generating only the non-zeros!
+We leverage the intiial stuff of Finch to say we will in order generate the code that only grabs the non-zeros, modified in certain ways based on the fine grained ordering of computation)
+(Use the below comment as an example - where the old paragraph talked about merging two lists.)
+}
 

### 2024-03-24 sections/intro.tex
Update on Overleaf.

@@
 %
-And we believe the limitations of these systems can be traced back to this conception.
+And we believe the limitations of these systems can be traced back to this conception.} \willow{I agree with what this sentence is saying, but I'm concerned that the sentiment is unclear, because here it may seem like we say "iteration space bad" and later say "iteration space good".
+Can we perhaps adjust here to say may of these special purpose compilers use overly restrictive models of the iteration space of the program, but that Finch is good because it uses general control flow to express a wider variety of iteration spaces.
+At a high level, we need to define "iteration space" or "control flow" or "Looplets" if we want to use them in the intro, and depending on how much we want to use these words we may want to avoid them here.}
 %
-For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
+\teo{For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
 %

### 2024-03-24 export_looplet_figures.sh
all done!

@@ -14,3 +14,6 @@ file=LevelsVsFibers # File name excluding extension
 
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file-tensor.png" "$file.drawio"
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 1 --output "$file-matrix.png" "$file.drawio"
+
+file=Structures # File name excluding extension
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file.png" "$file.drawio"

### 2024-03-25 sections/intro.tex
Update on Overleaf.

@@
 %
+\teo{Since the gap between the array storage and the array representation is minimal, dense arrays offer extensive compiler optimizations and extensive convenience.
+%
+Compilers understand dense array computation over
+}
+%
 This
@@
 %Blocks
-\help{This section needs citations, please help expand it and make it more
-convincing. A figure could be nice.} 
-\teo{I Added citations/heft but we maybe need to improve the figure?}
+
 \begin{figure}
@@
 \teo{ New paragraph:
-Many of these special purpose compilers build on some conception of an iteration space: TACO has merge lattices, Polyhedral compilers rely on the polyhedral model, Halide and Corra uses rectangular intervals, and so on.
+Note: This first sentence is still problematic - we need to say something other than tightly integrated so that we can be different from that. But I am not sure what.
+
+Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with the class of supported data structures.
 %
-And we believe the limitations of these systems can be traced back to this conception.} \willow{I agree with what this sentence is saying, but I'm concerned that the sentiment is unclear, because here it may seem like we say "iteration space bad" and later say "iteration space good".
+For example, TACO's merge lattices allows it to represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
+%
+Taichi represents general control flow with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
+%
+Furthermore, the limits of a DSL can be traced to the tight coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
+%
+In this work, we will choose find a new representation of a program's data structures and control flows that covers a wide range of programs.
+%
+However, before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
+
+}
+
+%build on some conception of an iteration space: TACO has merge lattices, Polyhedral compilers rely on the polyhedral model, Halide and Corra uses rectangular intervals, and so on.
+%
+%And we believe the limitations of these systems can be traced back to their specific conceptions.} \willow{I agree with what this sentence is saying, but I'm concerned that the sentiment is unclear, because here it may seem like we say "iteration space bad" and later say "iteration space good".
 Can we perhaps adjust here to say may of these special purpose compilers use overly restrictive models of the iteration space of the program, but that Finch is good because it uses general control flow to express a wider variety of iteration spaces.

### 2024-03-25 sections/intro.tex
cool

@@
 \teo{ New paragraph:
-Note: This first sentence is still problematic - we need to say something other than tightly integrated so that we can be different from that. But I am not sure what.
-
+Note: This first sentence is still problematic; we need to say something other than tightly integrated so that we can be different from that. But I am not sure what.
+%
 Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with the class of supported data structures.
@@
 However, before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
-
 }

### 2024-03-25 sections/tensors.tex
cool

@@
     of hierarchical tensor structures by combining these level structures in a
-    tree, as shown in Figure~\ref{fig:levels}.
+    tree, as shown in Figure~\ref{fig:structuraldiversity}.
     \begin{table}
@@
         \centering
-        %\includegraphics[width=0.45\linewidth]{structuraldiversity.png}\hfill%
-        \caption{This figure is a placeholder, but will eventually contain some very beguiling trees of levels.}
+        \includegraphics[width=\linewidth]{Structures.png}\hfill%
+        \caption{Several examples of matrix structures represented using the
+        level structures identified in Table~\ref{tab:TypesOfStructure}.
+        Comparing this figure to \cite[Figure 3]{ahrens_looplets_2023}, we see
+        that a level-by-level structural decomposition is diagrammed together
+        with the Looplets.}
         \label{fig:structuraldiversity}
@@
 Our simplified view of a level is enabled by our use of looplets to represent
-the structure within each fiber. In fact, our level interface requires only
-5 highly general operations, described below:
+the structure within each fiber. In fact, our level interface requires only 5
+highly general operations, described below.
+
+The first three of these functions, declare, freeze, and thaw, have to do with
+managing when tensors can be assumed mutable or immutable. As we use Looplets to
+represent iteration over a tensor, we must restrict the mutability of tensors
+while we iterate over them. For example, if a tensor declares it has a constant
+region from $i = 2:5$, but some other part of the computation modifies the
+tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
+write correct Looplet code if we can assume that the tensor is immutable while
+we are reading from it. Thus, we introduce the notion that a tensor can be in
+read-only mode or update-only mode.  In read-only mode, the tensor may only
+appear in the right-hand side of assignments. In update-only mode, the tensor
+may only appear in the left-hand side of an assignment, either being overwritten
+or incremented by some operator.  We can switch between these modes using freeze
+and thaw functions. The declare function is used to allocate a tensor,
+initialize it to some specified size and value, and leave it in update-only
+mode.
+
+The unfurl function is used to manage iteration over a subfiber. At the point
+when it comes time to iterate over a tensor, be in on the left or right hand
+side of an assignment, we call the unfurl function to precompute whatever state
+and datastructures are necessary to return a looplet nest that would iterate
+over that level of a tensor. The unfurl function is called directly before
+iterating over the corresponding loop, so it has access to any state variables introduced
+by freezing or thawing the tensor.
 
@@
 $assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
-$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation function,
-instead relying on initialization to reset the fiber if it needs to be reused.
+$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
+function, instead relying on initialization to reset the fiber if it needs to be
+reused. While all of the previous functions are used to manage the lifecycle and
+iteration over a general tensor, the assemble function is quite specific to the
+level abstraction, and the notion of positions within sublevels. Note: it was an
+intentional choice to hold the parent level responsible for managing the
+data of the sublevels, which positions they allocate, etc. This allows the parent
+level to reuse allocation logic from internal index datastructures. For example,
+a sparse level might use a list of indices to store which nonzeros are present,
+and when it comes time to resize that list, it could also call assemble to resize the
+sublevel, reducing the number of branches in the code. The assemble function
+lends itself particularly to a "vector doubling" allocation approach, which we
+have found to be effective and flexible when managing the allocation
+of sparse left hand sides. This benefit is made clear in our evaluation section,
+where we see that prior systems like TACO do not support all possible loop
+orderings and format combinations for sparse matrix multiply because they do
+not have a flexible enough allocation strategy, instead using a two-phase approach
+which requires computing a complicated closed-form kernel to iterate over the
+data twice to determine the number of required output nonzeros.
 
-\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
-\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$. 
-\paragraph{$freeze(lvl, init, dims...)$} Declares the level to hold subtensors of size $dims$ and an initial value of $init$.
+\begin{figure}
+    \raggedright
+\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
+of size $dims$ and an initial value of $init$. Requires the level to be in
+read-only mode.
+\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
+level for reading. Requires the level to be in update-only mode.
+\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
+to be in read-only mode.
+\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
+$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
+returns a looplet nest over the values in the read-only fiber.  When $mode =
+\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
+fiber. Often, skipping over mutable locations allows the level to know which
+locations must be stored. Additionally, a dirty bit may be used to communicate
+whether the mutable subfiber has been written to, which allows the parent fiber
+to know whether the subfiber must be stored explicitly.
+\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
+level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
+only ever called on unassembled positions, but some levels (such as dense levels
+or bytemaps) may support reassembly.
+\caption{The five functions that define a level.}
+\end{figure}
 
-\subsection{Core Level Language Primitives}
+\subsection{Level Formats}
+Having described the 8 basic level structures and the functions that define a
+level format, we can now detail the key level structures that we support, some
+of the challenges in their implementation, and how we overcome them.
+
 \begin{enumerate}

### 2024-03-25 sections/tensors.tex
more tensors

@@
 Having described the 8 basic level structures and the functions that define a
-level format, we can now detail the key level structures that we support, some
-of the challenges in their implementation, and how we overcome them.
+level format, we can describe the concrete level formats supported by Finch,
+some of the challenges in their implementation, and how we overcome them. Note
+that these formats are not meant to be exhaustive; we will later give
+recommendations for additional level formats to be implemented and explain how
+users can also add their own custom formats. Note that all of the levels below store their shape and a sublevel.
 
 \begin{enumerate}
-\item SparseList
-\item SparseDict
-\item ...
+\item[Dense]
+    The dense format is the simplest format, and simply maps $fiber(l, p)[i]
+    \rightarrow fiber(l.lvl, p * l.shape + i)$. This format is used to store dense data,
+    and is often a convenient format for the root level of a tensor. Because
+    of its simplicity, freezing and thawing the level are no-ops.
+\item[DenseRLE]
+    The DenseRLE format is used to represent runs of repeated values. Notice
+    that a challenge arises when trying to write to a level with runs: it is
+    difficult to merge duplicate runs. Such a scenario might arise when merging
+    runs of subfibers of length 3, representing colors in an image.  Ideally, we
+    would be able to detect duplicate subfibers and merge them on the fly, but
+    we cannot determine which subfibers are equal because we cannot read them
+    while the sublevel is in update-only mode. Instead, we merge the duplicates
+    during the freeze phase. We $freeze$ the main sublevel, $declare$ a separate
+    sublevel as a buffer to store the deduplicated subfibers, and we can then
+    compare each of the subfibers in the main level, copying the deduplicated
+    subfibers into the buffer:
+    \begin{enumerate}
+        \item[$right$] A vector of indices corresponding to the end of each run in the level
+        \item[$ptr$] A vector of delimiters such that $right[ptr[p]:ptr[p+1] - 1]$ is the set of delimiters in the subfiber at position $p$.
+        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
     \end{enumerate}
+\item[SparseList]
+    The sparse list format is the simplest sparse format, and is used to
+    construct the popular CSR, CSC, DCSR, DCSC, and CSF formats. The sparse list
+    format consists of the following fields:
+    \begin{enumerate}
+        \item[$idx$] A vector of nonzero indices in the level
+        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
+    \end{enumerate}
+    All of Finch's sparse formats use a dirty bit during writing to determine
+    whether the sublevel has been modified from it's default fill value and
+    thus, whether it needs to be stored.
+\item[SparsePinpoint]
+    Because the SparsePinpoint format will only ever store one nonzero in each subfiber,
+    we do not need the $ptr$ field.
+    \begin{enumerate}
+        \item[$idx$] A vector of nonzero indices in the level, one for each subfiber
+    \end{enumerate}
+\item[SparseRLE]
+    Similar to the DenseRLE level, but because the runs are stored on top of a background of fill values, we must also store the start of each run.
+    \begin{enumerate}
+        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
+        \item[$right$] A vector of indices corresponding to the end of each run in the level
+        \item[$ptr$] A vector of delimiters such that $left[ptr[p]:ptr[p+1] - 1]$ and $right[ptr[p]:ptr[p+1] - 1]$ are the set of delimiters in the subfiber at position $p$.
+        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
+    \end{enumerate}
+\item[SparseInterval]
+    Similar to the SparseRLE level, but $ptr$ is redundant because we only store
+    one run. We also don't bother with deduplication here as we cannot store any
+    intermediate results with duplicates.
+    \begin{enumerate}
+        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
+        \item[$right$] A vector of indices corresponding to the end of each run in the level
+    \end{enumerate}
+\item[SparseVBL]
+    The SparseVBL format is used to represent blocked data. The format consists
+    of the following fields:
+    \begin{enumerate}
+        \item[$idx$] A vector of indices at the end of each nonzero block in the level
+        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of block terminals in the subfiber at position $p$.
+        \item[$ofs$] A vector of delimiters such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ in subfiber $p$.
+    \end{enumerate}
+\item[SparseBand]
+    Like SparseVBL, but only one block is stored in each subfiber.
+        \item[$idx$] The index at the end of the block in the level
+        \item[$ofs$] A vector of delimiters such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block in subfiber $p$.
+\end{enumerate}
+
+\help{we need also to talk about pattern and element}
+
+\subsection{Flexible Level Formats}
+The levels we described in the previous section are all oriented towards bulk,
+sequential creation of formats. However, when users want to be able to write out
+of order (which is a common requirement arising from loop order or from the
+problem itself, it occurs in our spgemm algorithms and our histogram example in
+the evaluation section), we must use more complicated datastructures like hash
+tables and trees to support the random access. Because these datastructures are
+more complex and have a higher implementation burden and performance overhead,
+we only support random access construction of sparse or dense structures.  We
+can use these two more general structures as intermediates to convert to our
+more specialized structures later.
+
+%Future
+%work might involve supporting random access construction for additional
+%structures, especially those with asymptotic benefits such as SparseRLE, with
+%the understanding that we can use more general structures as intermediates to
+%convert to more specialized structures later.
+
+%This leave

### 2024-03-25 sections/eval.tex
Update on Overleaf.

@@
     \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \finchaggregate(op, expr)\\
-    \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \\
+    \finchaggregate(op, expr)\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
      expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias\quad\quad
@@
 
-Given this language, we now describe how to define a few example functions from the API as Finch Logic expressions, 
+Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language,
 \begin{align*}
@@
 \end{align*}
+Notably, these plans do not specify
 

### 2024-03-25 sections/eval.tex
Update on Overleaf.

@@
     \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \\
-    \finchaggregate(op, expr)\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias\quad\quad
+    \finchreformat(expr, format) \quad\quad\quad \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \\
+    \finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+     expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad\quad\quad\quad
 \end{align*}

### 2024-03-25 sections/tensors.tex
done with tensors

@@
 
-%Future
-%work might involve supporting random access construction for additional
-%structures, especially those with asymptotic benefits such as SparseRLE, with
-%the understanding that we can use more general structures as intermediates to
-%convert to more specialized structures later.
+\begin{enumerate}
+\item[SparseHash]
+    The sparse hash format uses a hash table to store the locations of nonzeros,
+    and sorts the unique indices for iteration during the freeze phase. This
+    allows for efficient random access, but not incremental construction, as the
+    freeze phase runs in time proportional to the number of nonzeros in the
+    entire level.
+    \begin{enumerate}
+        \item[$idx$] A vector of nonzero indices in the level
+        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
+        \item[$tbl$] A hash table used for construction of the level.
+    \end{enumerate}
+\item[SparseBytemap]
+    The SparseBytemap format uses a bytemap to store which locations have been
+    written to. Unlike the SparseHash format, the bytemap assembles the entire
+    space of possible subfibers. This accelerates random access in the format,
+    but requires a high memory overhead. Because we don't want to reallocate all
+    of the memory in each iteration, the declaration of this format instead
+    re-assembles only the dirty locations in the tensor. This format is
+    analogous to the default workspace format used by TACO.
+    \begin{enumerate}
+        \item[$idx$] A vector of nonzero indices in the level
+        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
+        \item[$tbl$] A dense array of booleans used to mark dirty locations in the tensor.
+    \end{enumerate}
+\end{enumerate}
 
-%This leave
+As future work, we also recommend the implementation of a tree-based format, as
+this would enable incremental construction. Currently, the cost of freezing and
+thawing is usually proportional to the number of stored fibers, so an operation
+that sets a single value in a tensor would be prohibitively expensive (i.e.
+$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
+intermediate randomly accessible structure, because (unlike blocks or
+singletons), runs have asymptotic benefits. However, these cases were not
+necessary for our case studies.

### 2024-03-25 sections/eval.tex
Update on Overleaf.

--- 2024-03-25 sections/eval.tex
@@
     \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-    \finchrelabel(expr, idxs...) \quad\quad\quad \finchmapjoin(op, exprs...) \quad\quad\quad \\
-    \finchaggregate(op, expr)\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     expr:= \finchreorder | \finchrelabel |\finchmapjoin | \finchaggregate | \finchtable | \finchalias\quad\quad
+     \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
+    \finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+     expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
 \end{align*}

### 2024-03-25 sections/eval.tex
Update on Overleaf.

@@
 \end{align*}
-Notably, these plans do not specify
+Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
+In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
+However, for larger or more complex programs, it would be important to apply a cost-based optimization
 

### 2024-03-25 sections/lang.tex
adding captions, tying things together.

@@
 
+The syntax of Finch is displayed in Figure \ref{fig:syntax}. We
+
+\begin{figure}
 \noindent % This ensures the minipages fill the page width
@@
 \end{align*}
+\caption{The syntax of the finch language. Note that instead of a classic EBNF
+form, we have constrained the classes of subterms on the left hand side to allow
+for descriptive names of the arguments. Compare this grammar to the Concrete
+Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
+addition of multiple left hand sides through code blocks, and the explicit
+initialization and finalization of tensors.}\label{fig:syntax}
+\end{figure}
 

### 2024-03-25 sections/eval.tex
Update on Overleaf.

@@
 In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
-However, for larger or more complex programs, it would be important to apply a cost-based optimization
+However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
 
 \subsubsection{Finch Interpreter}
+
 \textbf{Lowering}
@@
 
+\begin{figure}
+	\includegraphics[width=\linewidth]{triangle_count_speedup_over_duckdb.png}
+    \caption{Performance of Finch on triangle counting.}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{elementwise_speedup_over_duckdb.png}
+    \caption{Performance of Finch on elementwise (A+B)*C kernel.}
+\end{figure}
+
+\begin{figure}
+	\includegraphics[width=\linewidth]{sddmm_speedup_over_duckdb.png}
+    \caption{Performance of Finch on SDDMM.}
+\end{figure}
+
 %matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,

### 2024-03-25 sections/eval.tex
Update on Overleaf.

@@
 
-\subsubsection{Finch Interpreter}
+\subsubsection{Standardizing \& Heuristic Optimization}
+Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting.
+This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e.
+tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
+The former allows the interpreter to identify the loop order for each kernel.
+The second determines the output format for each intermediate.
+The last one guarantees that the innermost expression can be computed as a single kernel.
 
-\textbf{Lowering}
-
-\textbf{Heuristic Optimization}
+During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first.
+Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias.
+At this point, the first semantic requirement is satisfied.
+Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required.
+The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.
 
 \subsubsection{Finch Interpreter} 
+Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process.
+First, the output format is identified by unpacking the outer $\finchreformat$ statement.
+Next, the inner expression of the is unpacked to identify the aggregate operator and to convert the pointwise expression into a Finch expression.
+At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor.
+Lastly, the concordant loop order is identified and instantiated.
+At this point, the lowered query can be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
 
-Find an example where fusing the python interface gives a big speedup over non-fused kernels.
+\subsubsection{The Benefits of Kernel Fusion}
 
@@
 	\includegraphics[width=\linewidth]{elementwise_speedup_over_duckdb.png}
-    \caption{Performance of Finch on elementwise (A+B)*C kernel.}
+    \caption{Performance of Finch on elementwise (A+B)*C kernel as the sparsity of C varies.}
 \end{figure}
@@
 %(v^t .* dot(u, w))
+
+\willow{Note: I may want to explain format inference somewhere in here, but I'll have to get to it a little later, perhaps after the paper deadline}

### 2024-03-25 sections/intro.tex
Update on Overleaf.

@@
 \teo{ New paragraph:
-Note: This first sentence is still problematic; we need to say something other than tightly integrated so that we can be different from that. But I am not sure what.
 %
-Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with the class of supported data structures.
+Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with a narrow class of supported data structures.
 %
@@
 %
-Furthermore, the limits of a DSL can be traced to the tight coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
+Furthermore, the limits and advantages of a DSL can be traced to the specific coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
 %

### 2024-03-26 sections/lang.tex
cool

@@
 
-\subsection{Syntax and Semantics}
+\subsection{Syntax}
 
-The syntax of Finch is displayed in Figure \ref{fig:syntax}. We
+The syntax of Finch is displayed in Figure \ref{fig:syntax}. The Finch syntax
+mirrors most imperative languages with for-loops and control flow. Notable
+inclusions to the language include \mintinline{julia}{for},
+\mintinline{julia}{let}, blocks of code with \mintinline{julia}{if}, and 
+the lifecycle functions that let us declare, freeze, and thaw tensors.
+
+Finch uses looplets to lower \mintinline{julia}{for}-loops, though a sparse
+program with complicated looplets should produce output tensors with the same
+semantic value as a dense program with ordinary for-loops.  We include the
+semantics of how looplets are lowered as it affects the structure of the results
+and the asymptotic complexity of the program. Some applications view the
+structure of tensors as semantically meaningful.
+
+The \mintinline{julia}{let} statement (marked by the \finchdefine AST node)
+allows us to define tensor variables and reuse their value in multiple
+subexpressions without passing the variable through a tensor. Although it is too
+complex in this work to propagate constants through tensors, we can propagate
+constants (including 0) through a define statement. This allows the compiler to
+build up complicated functions of known operators and infer annihilation
+properties through them.It also allows use to eliminate common subexpressions,
+or tensor accesses, by reusing a tensor value in multiple places (such as the
+symmetric spmv in Figure \ref{fib:spmv_programs}). As another example, in the
+following program we don't need to define sparsity and annihilator properties of
+$f(a + b) * (a - b)$, as we can derive this from the more basic operators, and
+the code only reads from $A$ and $B$ once.
+\begin{minted}{julia}
+    let a = A[i, j]
+        let b = B[i, j]
+            C[i, j] = (a + b) * (a - b)
+        end
+    end
+\end{minted}
+
+We allow the use of blocks of code to describe multiple operations. This allows
+us to support both multiple outputs and temporary tensors, seen most prominently
+in our implementation of gustavson's algorithm for sparse-sparse matrix multiply
+and breadth-first-search. Our breadth-first seach also uses an
+\mintinline{julia}{if} statement to avoid operating on vertices outside the
+frontier. We have found that \mintinline{julia}{if} is particularly useful in
+our model for masking to avoid work.
+
+Finally, we add some explicit functions to advance the lifecycle of a tensor.
+Note that the user would not typically write \mintinline{julia}{@freeze} or
+\mintinline{julia}{@thaw} explicitly as those can be inferred through a separate
+pass.
 
@@
 
-the state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
-environment and another state $t$ corresponding to the embedding language.
+\subsection{Semantics}
 
-\begin{figure}
-    \centering
-    \small
+We break our description of the semantics of Finch into three parts, those
+relating to lowering the language itself in Figure~\ref{fig:semantics_core},
+those pertaining to the lifecycle of tensors in
+Figure~\ref{fig:semantics_lifecycle}, and the semantics of the looplets in
+Figure~\ref{fig:semantics_looplets}. Note that in addition to the rules seen
+here, we also apply extensive symbolic simplification of the program whenever we
+can. In the implementaiton of the language we have heuristics for when to run a
+simplification pass, but for the purposes of understanding the semantics of the
+language we should understand simplification rules as occuring whenever possible
+before continuing to execute the program as normal.
 
-    \begin{prooftree}  
-    \hypo{s = (e, t)}
-    \hypo{e(\finchmode(tns)) = read}
-    \hypo{e(tns) = tns'}
-    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
-    \end{prooftree}
-    \vspace{12pt}
-
-    \begin{prooftree}  
-    \hypo{s = (e, t)}
-    \hypo{e(\finchmode(tns)) = update}
-    \hypo{e(tns) = tns'}
-    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
-    \end{prooftree}
-      \vspace{12pt}
-
-    \begin{prooftree}  
-    \hypo{s = (e, t)}
-    \hypo{e(tns) = tns'}
-    \hypo{\langle init, s \rangle \Rightarrow init'}
-    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
-    \end{prooftree}
-
-    \caption{Rules for evaluating life cycle functions.}
-    \label{fig:enter-label}
-\end{figure}
+Note that both Looplets and Finch are designed to be embedded into a larger
+language, and a separate state variable is introduced for this language.
 
@@
 
-    \caption{Basic Evaluation Rules}
-    \label{fig:enter-label}
+    \caption{Basic evaluation semantics, roughly defining most of these language
+    constructs to function similarly to their classical definitions. Note that
+    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
+    environment and another state $t$ corresponding to the state in the
+    host language. Several Looplets introduce variables into the embedding
+    language, which may be read when evaluating the \finchvalue node.}
+    \label{fig:semantics_core}
 \end{figure}
@@
 
-$E$ is an evaluation context that applies anywhere in the syntax tree.
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(tns) = tns'}
+    \hypo{\langle init, s \rangle \Rightarrow init'}
+    \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
+    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = update}
+    \hypo{e(tns) = tns'}
+    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+    \end{prooftree}
+    \vspace{12pt}
+
+    \begin{prooftree}  
+    \hypo{s = (e, t)}
+    \hypo{e(\finchmode(tns)) = read}
+    \hypo{e(tns) = tns'}
+    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+    \end{prooftree}
+
+    \caption{Lifecycle evaluation semantics. The state $s$ of the finch compiler
+    is a tuple $(e, t)$ of a variable value environment and another state $t$
+    corresponding to the state in the host language. All of the lifecycle
+    functions are designed to be implemented and executed in the host language,
+    but these semantics enforce that each of these functions may update state in
+    the host language and flip the mode of the tensor between read and update.}
+    \label{fig:semantics_lifecycle}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \small
+
 \vspace{12pt}
@@
 \end{prooftree}
+\hfill
+\begin{prooftree}
+    \hypo{\langle node, algebra \rangle \rightarrow node'}
+    \infer1[$Simplify$]{\langle E[node], s\rangle \rightarrow \langle E[node'], s\rangle}
+\end{prooftree}
 \vspace{12pt}
@@
 
-    \caption{Looplet Rules}
-    \label{fig:enter-label}
+    \caption{Looplet evaluation semantics. 
+    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
+    environment and another state $t$ corresponding to the state in the host
+    language.  Note that $E$ is an evaluation context that applies anywhere in
+    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
+    hoist conditions and subranges out of loops. However, this also means we must specify
+    the priority in which we apply looplet rules, which is as follows:
+    $Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup$
+    Many looplets, most notably the thunk looplet, introduce variables into the
+    host language environment.  While variables introduced by a looplet may be
+    modified by the looplet itself (steppers often increment some state
+    variables), we forbid child looplets from modifying any state variables that
+    they didn't introduce. This allows us to treat the \finchvalue node as a
+    constant.
+    %
+    Note: the $Simplify$ rule references $algebra$, which is our variable
+    defining a set of straightforward simplification rules. These rules include
+    simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
+    as constant propagation. We omit the full set of rules for brevity, but
+    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
+    \label{fig:semantics_looplets}
 \end{figure}
-
-\subsection{Wrapper Tensors}
-
-\subsection{Scalars}
-
-\subsubsection{Sparse Scalars}
-\subsubsection{Early Break Scalars}

### 2024-03-26 sections/intro.tex
Update on Overleaf.

@@
 %
-Many of these special purpose compilers represent a program's control flow via a specialized data structure that is tightly integrated with a narrow class of supported data structures.
+Many of these special purpose compilers represent a program's control flow via a specialized representation that is tightly integrated with a narrow class of supported data structures.
 %
-For example, TACO's merge lattices allows it to represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
+For example, TACO's merge lattices represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
 %

### 2024-03-26 sections/intro.tex
Update on Overleaf.

@@
 %
-This trade-off is visualized in figure \ref{fig:datavsprogram}.
+This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
 %
@@
 \end{table}
+
 \teo{ New paragraph:
@@
 %
-For example, TACO's merge lattices represent iterations over subsets of rectangular grids where non-zero values are produced whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
+For example, TACO's merge lattices represent iterations over subsets of rectangular integer grids where computations produce non-zero values whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
 %
-Taichi represents general control flow with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
+Taichi enriches single static assignment form with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
 %
@@
 %
-In this work, we will choose find a new representation of a program's data structures and control flows that covers a wide range of programs.
+In this work, we will choose  a new representation of a program's data structures and control flows that covers a wide range of programs.
 %
-However, before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
-}
+However, simply combining a large variety of structured data with a large variety of programs does not necessarily yield a performance advantages.
+%
+To find and produce efficient programs for structured data, we must overcome two challenges that occur when we intersect complex control flow with structured data:}
+
+%before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
 
@@
 %
-\teo{For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
-%
-And many iteration space concepts simply don't allow for skipping work that isn't simply the complement of an interval.
-%
-In our work, we will show how we can use Looplets as our main tool to manage the iteration spaces of our computations and data structures.
-%
-However, to support as many data structures and iteration spaces as is implied above, we face two challenges if we want to also produce efficient code over structured data.
-}
+% \teo{For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
+% %
+% And many iteration space concepts simply don't allow for skipping work that isn't simply the complement of an interval.
+% %
+% In our work, we will show how we can use Looplets as our main tool to manage the iteration spaces of our computations and data structures.
+% %
+% However, to support as many data structures and iteration spaces as is implied above, we face two challenges if we want to also produce efficient code over structured data.
+% }
 %We see two main challenges to writing efficient code over structured data.
 
-\textbf{Optimizations are specific to the indirection and patterns in datastructures}: These structures break the simple mapping between array elements and where they are stored in memory.
+\textbf{Optimizations are specific to the indirection and patterns in data structures}: 
+%
+These structures break the simple mapping between array elements and where they are stored in memory.
+%
 For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value. 
+%
 These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.
@@
 \textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable.
+%
 For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
 
-\teo{Data structure driven array program performance engineering?}
+\teo{We could add a third here: making intersections worth it.}
+%\teo{Data structure driven array program performance engineering?}
 
@@
 
-\teo{Proposed new paragraph: We don't propose to solve these problems automatically, but we do propose a new programming model where these issues are solvable.
+\teo{Proposed new paragraph:}
+
+\teo{We don't propose to solve these problems automatically, but we do propose a new programming language where these issues are solvable, by virtue of the precise manner in which we combine structured data and control flow.}
 %
-By combining many data structures and iteration spaces into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
+\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into an abstraction: Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
 %
-We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
+\teo{For example, if a programmer wanted to merge the even indices of two sorted arrays, a programmer could represent by point-wise multiplying two sparse vectors to produce another sparse vector and then set the even indices two zero or the programmer could only point-wise multiply the two sparse vectors under an if condition that selects even indices.
+(This crucial part needs improvement)}
 %
-(Note quite sure how much comes after this)
-(We should say the predicability is generating only the non-zeros!
+\teo{Via index expressions, multiple levels of loops and tensors, and more complex filtering expressions, we can imagine many variation of this problem with fine grained distinctions about when various operations are performed.}
+
+\teo{
+Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
+%
+Finch supports \textit{both} flexible programming constructs and diverse data structures. 
+%
+Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+%
+The Finch compiler uses the structure
+of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
+
+%% Many variations of this problem.
+%By combining many data structures and control flows into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
+%
+%We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
+%
+%(Note quite sure how much comes after this)
+%We should say the predicability is generating only the non-zeros!
 We leverage the intiial stuff of Finch to say we will in order generate the code that only grabs the non-zeros, modified in certain ways based on the fine grained ordering of computation)
-(Use the below comment as an example - where the old paragraph talked about merging two lists.)
-}
+%(Use the below comment as an example - where the old paragraph talked about merging two lists.)
 
@@
 A Looplet based level-format abstraction.
+%
 Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
+%
 The first such set of formats to efficiently capture banded,
@@
 and complex control flow constructs at the same level of productivity
-of dense arrays. To our knowledge, the Finch programming language is the first 
+of dense arrays. 
+%
+To our knowledge, the Finch programming language is the first 
 to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather. \saman{for beyond dense...don't want to sound like we are the fist to do for dense. } 
+structured data, as well as complex accesses such as affine indexing or scatter/gather. 
+%
+\saman{for beyond dense...don't want to sound like we are the fist to do for dense. } 
 \item A compiler that specializes programs to data structures 
-predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex
-space of algorithms and data structures.
-It is easy to implement new data structures with a simple interface; the compiler analyzes the tensor lifecycle and inserts insert simple, overloadable, interface functions such as
-initialization or finalization.
+predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
+%
+%
+\teo{
+The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
+%
+By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
+%
+We observe that this strategy naturally extends the dense solution.
+}
+\teo{I think we can back it up in two ways:
+1. We need to modify the semantics to focus when on when things are converted into Looplets
+2.
+We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations.
+If we do triangle counting or something like that, we could maybe talk about that a bit.
+}
+\teo{
+\item Extensibility of the language and cmopiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
+}
+
 \item We evaluate the productivity of our language in several case studies,

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 
+\changwan{
+To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
+For scale-free graphs where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.
+Push traversal starts from the source node, pushing neighboring nodes into the queue.
+Pull traversal starts from the target node, exploring predecessors (parents).
+
+}
+
 \begin{figure}

### 2024-03-26 sections/eval.tex
Update on Overleaf.

--- 2024-03-25 sections/eval.tex
@@
 
+\changwan{explain roadnet and scale free graphs here}
+To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
+We For scale-free graphs where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.
+Push traversal starts from the source node, pushing neighboring nodes into the queue.
+Pull traversal starts from the target node, exploring predecessors (parents).
+This approach uses early-break in pull traversal, which is used to terminate the traversal prematurely once the target node is encountered, to avoid unnecessary edge traversal.
+
+Figure~\cite{graph_result} (a) \changwan{add (a) in fig} shows BFS speedup over Graphs.jl, a BFS implementation with push traversal of Julia.
+`finch\_push\_only` only uses push traversal, whereas `finch\_push\_pull` applies direction-optimization used in GraphBLAS.
+As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
+
 \begin{figure}
@@
     \caption{Performance of graph apps across various tools.}
+     \label{graph_result}
 \end{figure}

### 2024-03-26 sections/compiler.tex
okay, I'm calling it
something like that
add some stuff about wrapper arrays

@@
 
+\subsection{Finch Normal Form}
+
+Our semantics in Figure
+\ref{fig:semantics_core,fig:semantics_lifecycle,fig:semantics_looplets} is only
+well-defined on some programs. We define a particular class of programs on which
+our semantics are well-defined, and refer to it as \textbf{Finch Normal Form}.
+The properties of such a program are as follows:
+\begin{enumerate}
+    \item[Access with Indices] Though Finch allows general expressions in an
+    access (i.e. `A[i + j]` or `A[I[i]]`), the normal form restricts to allow only indices in 
+    accesses (i.e. `A[i]`), rather than more general expressions.
+    \item[Evaluable Dimensions] Loop dimensions and declaration dimensions must
+    be evaluable at the time we compile them, so we restrict the normal form to
+    programs whose loop dimensions and declaration dimensions are extents with
+    limits defined in the scope of the corresponding loop or declaration
+    statement.
+    \item[Concordant] Finch is column-major by default, and the normal form
+    requires the order of indices in an access to match the order in which loops
+    are nested around it.  For example,
+    \mintinline{julia}{for j = _; for i = _; s[] += A[i, j] end end}
+    is concordant but
+    \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end} is not.
+    \item[Lifecycle Constraints] Tensors in read mode may appear on the right
+    hand side only. Tensors in update mode may appear on the left hand side
+    only. To make it easier to statically analyze lifecycle constraints, we
+    restrict tensors to only change modes in the same scope in which they were
+    defined.
+\end{enumerate}
+
+The subsequent sections will explain how programs that violate each of these
+constraints can be rewritten to programs that satisfy them, and thus how we can
+support such a wide variety of programs. For example, we can write nonconcordant
+programs like  \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end}
+by adding a loop to randomly access \mintinline{julia}{A} or adjusting the
+storage order of \mintinline{julia}{A} by adding a lazy transposition wrapper.
+
+\subsection{Wrapperization}
+
+    Many fancy operations on indices can be resolved by introducing equivalent
+    \textbf{wrapper arrays} which modify the behavior of the tensors they wrap,
+    or by introducing \textbf{mask arrays} which replace index expressions like
+    \mintinline{julia}{i == j} with their equivalent masks (in this case, a
+    diagonal mask tensor).
+
+    All wrapper arrays are eventually unwrapped by the compiler as we lower
+    them, some earlier than others. For example, the $swizzle$ array wraps a
+    tensor and permutes the indices of an access when it is unwrapped during the
+    wrapperization pass. On the other hand, the $offset$ array wraps a tensor
+    and shifts all of the ranges declared in Figure \ref{fig:semantics_looplets} by one.
+    The implementation burden for a wrapper tensor is to implement a suitable
+    program modification during the wrapperization procedure and then unwrap the wrapper, or to 
+    implement the looplet functions of \ref{fig:semantics_looplets} with some minor modifications
+    to shift dimensions, for example. Wrapper arrays are summarized in Table \ref{tab:wrappers}
+
+    Mask arrays have a more straightforward implementation using static looplets
+    that are constructed during the unfurl step. They are summarized in table \ref{tab:masks}. Mask tensors
+    allow us to lift computations with masks to the level of the loop, without modifying the loop directly.
+
+    For example, \begin{minted}{julia}
+    for i=_, j=_
+        if i <= j
+            s[] += A[i, j]
+        end
+    end
+    \end{minted}
+
+    would likely compile to something like
+
+    \begin{minted}{julia}
+    for i = 1:n
+        for j = 1:i
+            s[] += A[i, j]
+        end
+    end
+    \end{minted}
+
+    \begin{table}[h]
+        \centering
+        \begin{tabular}{|>{\raggedright\arraybackslash}p{0.2\linewidth}|>{\raggedright\arraybackslash}p{0.8\linewidth}|}
+        \hline
+        \textbf{Transformation Example} & \textbf{Description} \\
+        \hline
+        A[i + a] $\rightarrow$ offset(A, 1)[i] & Creates an OffsetArray such that \texttt{offset(tns, delta...)[i...] == tns[i .+ delta...]}. \\
+        \hline
+        A[i + x] $\rightarrow$ toeplitz(A, 1)[i, x] & Creates a ToeplitzArray, adding a dimension that shifts another dimension of the original tensor.
+The added dimensions are produced during a call to Unfurl, when a lookup looplet is emitted for the first dimension.
+\\
+        \hline
+        A[(a:b)(i)] $\rightarrow$ window(A, a:b)[i] & Creates a WindowedArray, representing a view into another tensor. This wrapper returns a different size of tensor. \\
+        \hline
+        A[\~i] $\rightarrow$ permissive(A)[i] & Creates a PermissiveArray, allowing for out-of-bounds access or padding. This tensor returns no dimensions as its size. \\
+        \hline
+        A[p(i)] $\rightarrow$ protocolize(A, p)[i] & Accesses dimension n with protocol protos[n], allowing for advanced iteration protocols. \\
+        \hline
+        A[p(i)] $\rightarrow$ swizzle(A, perm)[i] & A lazily transposed array, swizzle(A, perm)[idx...] is transformed to A[idx[perm]...] during wrapperization. \\
+        \hline
+        \end{tabular}
+        \caption{Wrapper arrays and some example indexing sugar they enable.}
+    \label{table:wrappers}
+    \end{table}
+
+    \begin{table}[h]
+        \centering
+        \begin{tabular}{|>{\raggedright\arraybackslash}p{0.5\linewidth}|>{\raggedright\arraybackslash}p{0.5\linewidth}|}
+        \hline
+        \textbf{Transformation Example} & \textbf{Description} \\
+        \hline
+        $i < j \rightarrow$ UpTriMask()[i, j - 1] & Upper triangular mask, true if $i < j$. \\
+        \hline
+        $i \geq j \rightarrow$ LoTriMask()[i, j] & Lower triangular mask, true if $i \geq j$. \\
+        \hline
+        $l \leq i < j \rightarrow$ Bandmask()[i, l, h - 1] & Banded mask, true for elements within a specified band. \\
+        \hline
+        $i == j \rightarrow$ DiagMask()[i, j] & Diagonal mask, true if $i == j$. \\
+        \hline
+        $i \neq j \rightarrow$ !(DiagMask()[i, j]) & Inverse diagonal mask, true if $i \neq j$. \\
+        \hline
+        chunkmask(b) & Chunk mask, for chunked tensor access. True if $b \times (j - 1) < i \leq b \times j$. \\
+        \hline
+        \end{tabular}
+        \caption{Mask tensors and some example cases they are inserted.}
+        \label{table:masks}
+    \end{table}
+
 \subsection{Dimensionalization}
+    Looplets typically require the dimension of the loop extent to match the
+    dimensions of the tensor.
 
 \subsection{Concordization}
+
+\subsection{Life Cycles and Scopes}
+
+\subsection{Simplification and Algebraic Transformations}
 
@@
 \subsection{Performance Warnings}
-
-\subsection{Wrapperization}
-
-\subsection{Simplification and Algebraic Transformations}

### 2024-03-26 sections/eval.tex

@@
 
-\changwan{explain roadnet and scale free graphs here}
 To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-We For scale-free graphs where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+We used graphs in GraphBLAS~\cite{}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
 Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.

### 2024-03-26 sections/compiler.tex
add dimensionalization

@@
 
+    However, it is cumbersome to write the dimensions
+    in loop programs, and most tensor compilers have a means of specifying the
+    dimensions automatically. We refer to the process of discovering suitable
+    dimensions as \textbf{dimensionalization}. \willow{Teo, please explain
+    alternative approaches}>
+
+    In finch, we use a straightforward algorithm to automatically determine
+    suitable dimensions for loops and declaration statements (output tensors).
+    Finch determines the dimension of a loop index i from all of the tensors
+    using i in an access, as well as the bounds in the loop itself, and operates
+    similarly for declarations.
+
+    Our algorithm uses the following principles, assuming dimension types form a
+    lattice:
+    \begin{enumerate}
+        \item We assign dimensions to indices.
+        \item Using an index in an access “hints” that the index should have the corresponding dimension.
+        \item Loop dimensions are equal to the “meet” of all hints in the loop body
+        and any existing dimensions in the loop bounds. The meet usually asserts
+        that dimensions match, but may also e.g. propagate info about parallelization
+        \item The \mintinline{julia}{_} symbol represents a dimensionless quantity at the bottom of the dimension lattice.
+        \item We assign dimensions to declarations.
+        \item Left hand side (updating) tensor access “hint” the size of that tensor
+        \item The dimensions of a declaration are the “meet” of all hints from
+        the declaration to the first read.
+        \item The new dimensions of the declared tensor are used when the tensor is on the right hand side (reading)
+        access.
+    \end{enumerate}
+
+    For example, consider the following code
+
+    \begin{minted}{julia}
+    A = fsprand(3, 4, 0.5)
+    B = fsprand(4, 5, 0.5)
+    C = Tensor(Dense(SparseList(Element(0.0))))
+    @finch begin
+    C .= 0
+    for i = 1:3
+        for j = _
+            for k = _
+                C[i, j] += A[i, k] * B[k, j]
+            end
+        end
+    end
+    \end{minted}
+
+    In the above code, the second dimension of A must match the first dimension of B.
+Also, the first dimension of A must match the i loop dimension, 1:3.
+Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5).
+If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.
+
+    Dimensionalization occurs after wrapper arrays are de-sugared. You can
+    therefore exempt a tensor from dimensionalization by wrapping the
+    corresponding index in ~ to produce a "PermissiveArray". For example,
+
+    \begin{minted}{julia}
+    @finch begin
+    y .= 0
+    for i = 1:3
+        y[~i] += x[i]
+    end
+    \end{minted}
+
+    does not set the dimension of y, and y does not participate in dimensionalization.
+
 \subsection{Concordization}
+    Finch uses
 

### 2024-03-26 sections/compiler.tex
concordization

@@
 \subsection{Concordization}
-    Finch uses
+    After wrapperization, all arrays are normalized to column-major ordering and
+    the unrecognized expressions are left in the access expressions. At that
+    point, we run a pass over the code to make the program concordant.
+
+    We make expressions concordant by inserting loops with one iteration. For example, the following expression can be made concordant by adding a loop for the out of order access to \texttt{i}.
+    \begin{minted}{julia}
+        @finch begin
+        for i = _
+            for j = _
+                for k = _
+                    s[] += A[i, k, j]
+                end
+            end
+        end
+    \end{minted}
+
+    \begin{minted}{julia}
+        @finch begin
+        for i = _
+            for j = _
+                for k = _
+                    for l = _
+                        if i == l #diagonal mask
+                            s[] += A[l, k, j]
+                        end
+                    end
+                end
+            end
+        end
+    \end{minted}
+
+    We can concordize unrecognized index expressions using a similar approach:
+
+    \begin{minted}{julia}
+        for i = _
+            A[I[i]] += 1
+        end
+    \end{minted}
+
+    \begin{minted}{julia}
+        for i = _
+            for j = _
+                if j == I[i]
+                    A[j] += 1
+                end
+            end
+        end
+    \end{minted}
 

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 `finch\_push\_only` only uses push traversal, whereas `finch\_push\_pull` applies direction-optimization used in GraphBLAS.
-As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
+Finch supports push/pull traversal, early break, so it supports direction-optimization.
+Finch also supports As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
+\changwan{how much average speedup?}
 

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 Finch supports push/pull traversal, early break, so it supports direction-optimization.
-Finch also supports As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
+Finch also supports the different loop orders and multiple outputs, which also As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
 \changwan{how much average speedup?}

### 2024-03-26 sections/eval.tex
Update on Overleaf.

--- 2024-03-26 sections/eval.tex
@@
 \subsubsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\changwan{
-Explain what is SpGEMM. why SpGEMM hard? (not sure)
-three approaches. Asym time complexity?
-}
+%\changwan{
+%Explain what is SpGEMM. why SpGEMM hard? (not sure)
+%three approaches. Asym time complexity?
+%}
 
@@
 `finch\_push\_only` only uses push traversal, whereas `finch\_push\_pull` applies direction-optimization used in GraphBLAS.
+Finch supports push/pull traversal, early break, so it supports direction-optimization.
+Finch additionaly supports the different loop orders and multiple outputs, which also improves the performance.
 As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
+\changwan{how much average speedup?} GraphBLAS sometimes better than Finch since GraphBLAS incorporates all hardwired optimizations.
+We believe we will achieve a better performance if we incorporate those optimizations.
 

### 2024-03-26 sections/eval.tex
fixes
Update on Overleaf.

@@
 %}
+
+Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics.
+SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
+Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
+Gustavson product optimizes memory access by storing one matrix in row-major order and the other in column-major order.
+Outer product computes the product of all pairs of elements from the input matrices.
+Our results indicate that inner and Gustavson products perform competitively, while outer product exhibits different characteristics, especially with sparse matrices.
 
@@
 
-\willow{Note: I may want to explain format inference somewhere in here, but I'll have to get to it a little later, perhaps after the paper deadline}
+\willow{Note: I may want to explain format inference somewhere in here, but I'll
+have to get to it a little later, perhaps after the paper deadline}
+\willow{Note: I think it would be cool to include something about how we support
+numerically stable norms and argmin in this model}

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
+The inner product, outer product, and Gustavson's method are three different approaches to performing SpGEMM. 
 Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
-Gustavson product optimizes memory access by storing one matrix in row-major order and the other in column-major order.
 Outer product computes the product of all pairs of elements from the input matrices.
-Our results indicate that inner and Gustavson products perform competitively, while outer product exhibits different characteristics, especially with sparse matrices.
 

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
-The inner product, outer product, and Gustavson's method are three different approaches to performing SpGEMM. 
+The inner product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
 Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
 Outer product computes the product of all pairs of elements from the input matrices.
+
+More detailed explanation is available in Section 2.2 of ~\cite{} \changwan{gamma}.
 

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 Outer product computes the product of all pairs of elements from the input matrices.
+More detailed explanation is available in Section 2.2 of ~\cite{} \changwan{gamma}.
+\changwan{Inner, asym bad, }
 
-More detailed explanation is available in Section 2.2 of ~\cite{} \changwan{gamma}.
+Figure~\ref{spgemm_small} 
+TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
 
@@
 \begin{figure}
+    \label{spgemm_small}
 	\includegraphics[width=\linewidth]{spgemm_small_speedup_log_scale.png}
@@
 \begin{figure}
+\label{spgemm_large}
 	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
@@
 Lastly, the concordant loop order is identified and instantiated.
-At this point, the lowered query can be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
+The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
 

### 2024-03-26 main.tex
some stuff
Update on Overleaf.
fix
addressing comments
anonymize

@@
 %%
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall,screen,review,anonymous]{acmart}
 
@@
 \usepackage[english]{babel}
-
+\usepackage{subcaption}
 % Useful packages
@@
 \usepackage{multirow}
+%
 
@@
 \newcommand{\finchtable}{\ensuremath{\text{\textbf{table}}}}
+\newcommand{\finchalias}{\ensuremath{\text{\textbf{alias}}}}
 \newcommand{\finchmapjoin}{\ensuremath{\text{\textbf{mapjoin}}}}
@@
 \newcommand{\finchreorder}{\ensuremath{\text{\textbf{reorder}}}}
+\newcommand{\finchreformat}{\ensuremath{\text{\textbf{reformat}}}}
 
@@
 Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
- Thus, focusing only on sparsity limits the scope needed to handle a vast array of data \kyle{"vast array" is a good pun here, but a bit confusing to read}.
+Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data.
 Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.
@@
 Our approach enables new loop optimizations across multiple domains, unifying techniques such as
-sparse tensors, databases, and lossless compression.
+sparse tensors, databases, and lossless compression\kyle{nit: "unifying techniques in areas such as"}.
 We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{FinchOOPSLA.bib, FinchOOPSLAOverleaf.bib}
+\bibliography{FinchOOPSLAWillow.bib, FinchOOPSLAOverleaf.bib}
 

### 2024-03-26 sections/intro.tex
Update on Overleaf.
fixes
fix

@@
 %
-Compilers understand dense array computation over
+Compilers understand dense array computation over cross many
+programming constructs, such as for and while loops, breaks, parallelism,
+caching, prefetching, multiple outputs, scatters, gathers, vectorization,
+loop-carry-dependencies, and more.
 }
@@
 programming constructs, such as for and while loops, breaks, parallelism,
-caching, prefetching, multiple outputs, scatters, gathers, vectoriation,
+caching, prefetching, multiple outputs, scatters, gathers, vectorization,
 loop-carry-dependencies, and more. Several optimizations have been developed for
@@
 %
-\help{This section needs citations}
 \textbf{Currently, support for structured data is fragmented and incomplete}.
@@
 \begin{table}[h!]
-  \help{This table needs citations, as well as a little bit of double checking}
   \centering
+  \scriptsize
   \begin{tabular}{l|cccccc}
@@
 \teo{Proposed new paragraph:}
-
+\teo{Add: }
 \teo{We don't propose to solve these problems automatically, but we do propose a new programming language where these issues are solvable, by virtue of the precise manner in which we combine structured data and control flow.}
 %
-\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into an abstraction: Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
+\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into a single abstraction} Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
 %
@@
 of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
+
+% \begin{figure*}[t!]
+%     \centering
+%     \begin{subfigure}[t][0.2\textwidth]
+%         \begin{minted}{julia}
+%     for i = _
+%         if i % 2 == 0
+%             c[i] = a[i] * b[i]
+%         end
+%     end
+%         \end{minted}
+%         \caption{}
+%         \label{fig:inter1}
+%     \end{subfigure}
+
+%     \begin{subfigure}[t][0.2\textwidth]
+%         \begin{minted}{julia}
+%     for i = _
+%         if i % 2 == 0
+%             ap[i] = a[i]
+%         end
+%     end
+%     for i = _
+%         c[i] = ap[i] * b[i]
+%     end
+%         \end{minted}
+%         \caption{}
+%         \label{fig:inter2}
+%     \end{subfigure}
+%     \begin{subfigure}[t][0.2\textwidth]
+%         \begin{minted}{julia}
+%     for i = _
+%         cp[i] = ap[i] * b[i]
+%     end
+%     for i = _
+%         if i % 2 == 0
+%             c[i] = cp[i]
+%         end
+%     end
+%         \end{minted}
+%         \caption{}
+%         \label{fig:inter3}
+%     \end{subfigure}
+%     \caption{Caption}
+%     \label{fig:intersections}
+% \end{figure*}
 
@@
 
-\item We evaluate the productivity of our language in several case studies,
-showing that Finch can be used to accelerate a wide range of applications, 
-from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
-\teo{Do we really want to say productivity here?
-What word do other similar papers use?}
+\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.

### 2024-03-26 sections/eval.tex
Update on Overleaf.

@@
 This means that every operator (with the exception of $\finchtable$ and $\finchalias$ which are analogous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the attributes, and we include the $\finchreorder$ operator to manipulate the order of these attributes.
+Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate the order.
 This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.
@@
 
-Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language,
+Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language.
+Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can manipulate an index as a scalar to implement $argmin$.
+For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
 \begin{align*}
-\text{sum(M, dims=[2])} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchrelabel(M, i_1,\ldots,i_d), i_2)), out)\\
-\text{matmul(A, B)} &\rightarrow \finchplan(\finchquery(out,\finchaggregate(\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)), out)\\
+&\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+&\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+&\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
 \end{align*}

### 2024-03-26 sections/intro.tex
Update on Overleaf.

@@
 paszke_pytorch_2019,ragan-kelley_halide_2013}.
-%
+% EASY CUT
 These frameworks have been the
@@
 
-The success and ubiquity of arrays is likely due to their simplicity. 
+The success and ubiquity of arrays is largely due to their simplicity. 
 %
-Since
-their introduction, multidimensional arrays have represented dense, rectilinear,
+Since their introduction, multidimensional arrays have represented dense, rectilinear,
 integer grids of points. 
@@
 %
-\teo{Since the gap between the array storage and the array representation is minimal, dense arrays offer extensive compiler optimizations and extensive convenience.
+Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
 %
-Compilers understand dense array computation over cross many
-programming constructs, such as for and while loops, breaks, parallelism,
-caching, prefetching, multiple outputs, scatters, gathers, vectorization,
-loop-carry-dependencies, and more.
-}
-%
-This
-simplicity enables extensive interoperability, convenience layers, and
-optimizations by breaking the abstraction barrier between array representation
-and array storage.  
+%This
+%simplicity enables extensive interoperability, convenience layers, and
+%optimizations by breaking the abstraction barrier between array representation
+%and array storage.  
 %
@@
 %
-However, while dense arrays are the easiest way to program, the world is not all dense.
+However, while dense arrays are the easiest way to program for performance, the world is not all dense.
 
@@
 
-\teo{ New paragraph:
 %
-Many of these special purpose compilers represent a program's control flow via a specialized representation that is tightly integrated with a narrow class of supported data structures.
+Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
+%many of these special purpose compilers represent a program's control flow via a specialized representation that is tightly integrated with a narrow class of supported data structures.
 %
-For example, TACO's merge lattices represent iterations over subsets of rectangular integer grids where computations produce non-zero values whereas the polyhedral model allows various compilers to represent dense computations on irregular regions~\cite{kjolstad_tensor_2017, grosser2012polly}.
+For example, TACO merge lattices represent boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
 %
-Taichi enriches single static assignment form with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
+The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
 %
-Furthermore, the limits and advantages of a DSL can be traced to the specific coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
+Taichi enriches single static assignment form with specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
+% TODO: Make me shorter
+The reason these systems tightly couple their control flow to narrow classes of data structures is because producing efficient programs over structured data requires overcoming two challenges that occur when we intersect complex control flow with structured data:
+%Taichi enriches single static assignment form with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
 %
-In this work, we will choose  a new representation of a program's data structures and control flows that covers a wide range of programs.
+%Furthermore, the limits and advantages of a DSL can be traced to the specific coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
 %
-However, simply combining a large variety of structured data with a large variety of programs does not necessarily yield a performance advantages.
+%In this work, we will choose  a new representation of a program's data structures and control flows that covers a wide range of programs.
 %
-To find and produce efficient programs for structured data, we must overcome two challenges that occur when we intersect complex control flow with structured data:}
+%However, simply combining a large variety of structured data with a large variety of programs does not necessarily yield a performance advantages.
+%
+%After all previous implementations choose to tightly couple to a specific data structure in part to ensure the data structures are iterated over efficiently.
+%
+%To find and produce efficient programs for structured data, we must overcome two challenges that occur when we intersect complex control flow with structured data:
 
@@
 
-\textbf{Performance is highly structure dependent}: Structure aware kernels are dramatically faster than unstructured ones, and the landscape of implementation decisions is dramatically unpredictable.
+\textbf{Performance on structured data is highly algorithm dependent}: The landscape of implementation decisions is dramatically unpredictable. 
 %
-For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
+For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens2022autoscheduling, zhang2021gamma}.
+% TODO: Add Gamma
+%For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
+This means that performance engineering for such kernels requires the exploration of a large design space, changing the algorithm as well as the data structures.
 
-\teo{We could add a third here: making intersections worth it.}
-%\teo{Data structure driven array program performance engineering?}
+In this work, we propose a new programming language, Finch, which supports \textit{both} flexible control flow and diverse data structures.
+%
+Finch facilitates a programming model which resolves the challenges of computing over structured arrays by combining control flow and data structures into a common representation where they can be co-optimized.
+%
+In particular, Finch automatically specializes the control flow to the data so that performance engineers can focus on experimenting with many algorithms.
+%
+Finch supports a familiar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+%
+This support would be useless without the appropriate level of structural specialization; Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
+%
 
-%% Modification: We don't propose to fix these directly, but we do propose a programming model where they can be dealt with 
-%% By supporting the features above,
+As an example, a programmer might explore the different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices).
+The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
+%The Finch compiler uses the structure of the data to generate efficient implementations of these programs.
 
-\teo{Proposed new paragraph:}
-\teo{Add: }
-\teo{We don't propose to solve these problems automatically, but we do propose a new programming language where these issues are solvable, by virtue of the precise manner in which we combine structured data and control flow.}
+%\teo{Proposed new paragraph:}
+%\teo{Add: }
+%\teo{We don't propose to solve these problems automatically, but we do propose a new programming language where these issues are solvable, by virtue of the precise manner in which we combine structured data and control flow.}
 %
-\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into a single abstraction} Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
+%\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into a single abstraction} Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
 %
-\teo{For example, if a programmer wanted to merge the even indices of two sorted arrays, a programmer could represent by point-wise multiplying two sparse vectors to produce another sparse vector and then set the even indices two zero or the programmer could only point-wise multiply the two sparse vectors under an if condition that selects even indices.
+%\teo{For example, if a programmer wanted to merge the even indices of two sorted arrays, a programmer could represent by point-wise multiplying two sparse vectors to produce another sparse vector and then set the even indices two zero or the programmer could only point-wise multiply the two sparse vectors under an if condition that selects even indices.
 (This crucial part needs improvement)}
 %
-\teo{Via index expressions, multiple levels of loops and tensors, and more complex filtering expressions, we can imagine many variation of this problem with fine grained distinctions about when various operations are performed.}
+%\teo{Via index expressions, multiple levels of loops and tensors, and more complex filtering expressions, we can imagine many variation of this problem with fine grained distinctions about when various operations are performed.}
 
-\teo{
-Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
-%
-Finch supports \textit{both} flexible programming constructs and diverse data structures. 
-%
-Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
-%
-The Finch compiler uses the structure
-of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
+\begin{minipage}{0.333\linewidth}
+\begin{minted}{julia}
+     for i = _
+         if i % 2 == 0
+             c[i] = a[i] * b[i]
+         end
+     end
+\end{minted}
+\end{minipage}%
+\begin{minipage}{0.333\linewidth}
+\begin{minted}{julia}
+     for i = _
+         if i % 2 == 0
+             ap[i] = a[i]
+         end
+     end
+     for i = _
+         c[i] = ap[i] * b[i]
+     end
+\end{minted}
+\end{minipage}%
+\begin{minipage}{0.333\linewidth}
+\begin{minted}{julia}
+     for i = _
+         cp[i] = ap[i] * b[i]
+     end
+     for i = _
+         if i % 2 == 0
+             c[i] = cp[i]
+         end
+     end
+\end{minted}
+\end{minipage}%
 
-% \begin{figure*}[t!]
-%     \centering
-%     \begin{subfigure}[t][0.2\textwidth]
-%         \begin{minted}{julia}
-%     for i = _
-%         if i % 2 == 0
-%             c[i] = a[i] * b[i]
-%         end
-%     end
-%         \end{minted}
-%         \caption{}
-%         \label{fig:inter1}
-%     \end{subfigure}
+\begin{wrapfigure}{r}{0.2\linewidth}
+\begin{minted}{julia}
+     for i = _
+         cp[i] = ap[i] * b[i]
+     end
+     for i = _
+         if i % 2 == 0
+             c[i] = cp[i]
+         end
+     end
+\end{minted}
+\begin{minted}{julia}
+     for i = _
+         if i % 2 == 0
+             c[i] = a[i] * b[i]
+         end
+     end
+\end{minted}
+\begin{minted}{julia}
+     for i = _
+         if i % 2 == 0
+             ap[i] = a[i]
+         end
+     end
+     for i = _
+         c[i] = ap[i] * b[i]
+     end
+\end{minted}
+\end{wrapfigure}
 
-%     \begin{subfigure}[t][0.2\textwidth]
-%         \begin{minted}{julia}
-%     for i = _
-%         if i % 2 == 0
-%             ap[i] = a[i]
-%         end
-%     end
-%     for i = _
-%         c[i] = ap[i] * b[i]
-%     end
-%         \end{minted}
-%         \caption{}
-%         \label{fig:inter2}
-%     \end{subfigure}
-%     \begin{subfigure}[t][0.2\textwidth]
-%         \begin{minted}{julia}
-%     for i = _
-%         cp[i] = ap[i] * b[i]
-%     end
-%     for i = _
-%         if i % 2 == 0
-%             c[i] = cp[i]
-%         end
-%     end
-%         \end{minted}
-%         \caption{}
-%         \label{fig:inter3}
-%     \end{subfigure}
-%     \caption{Caption}
-%     \label{fig:intersections}
-% \end{figure*}
+% \teo{
+% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
+% %
+% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
+% %
+% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+% %
+% The Finch compiler uses the structure
+% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
 
@@
 
-In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
+%In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
 TACO did something similar?} In this model, performance engineers can more efficiently search the
-complex landscape of programs and datastructures to find the best implementation.
+%complex landscape of programs and datastructures to find the best implementation.
 In this programming model, we can express certain concepts in programs, and others in data.
 For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
-While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
-
-We develop a new array programming language, Finch, which supports \textit{both}
-flexible programming constructs and diverse data structures. Finch supports a
-familiar programming language of loops, statements, if conditions, breaks, etc,
-over a wide variety of array structures, such as sparsity, run-length-encoding,
-symmetry, triangles, padding, or blocks. The Finch compiler uses the structure
-of the data to generate efficient implementations of these programs.
+%While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
 
@@
 \teo{
-\item Extensibility of the language and cmopiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
+\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
 }

### 2024-03-26 sections/cut.tex
Clean up intro after discussions

--- 2024-03-26 sections/intro.tex
@@
-\section{Introduction}
+% \teo{
+% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
+% %
+% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
+% %
+% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+% %
+% The Finch compiler uses the structure
+% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
 
-%Array programming is the core abstraction behind many of the modern miracles of computing (e.g. neural networks, scientific simulation, database processing). 
-Arrays are the most fundamental abstraction in computer science. Arrays and lists are often the first-taught datastructure
-\cite[Chapter 2.2]{abelson_structure_1996}, \cite[Chapter 2.2]{knuth_art_1997}.
+%% Many variations of this problem.
+%By combining many data structures and control flows into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
 %
-Arrays are also universal across programming languages, from their introduction
-in Fortran in 1957 to present-day languages like Python
-\cite{backus_fortran_1957}, keeping more-or-less the same semantics.
+%We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
 %
-Modern
-array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
-and Halide have pushed the limits of productive data processing with arrays,
-fueling breakthroughs in machine learning, scientific computing, image
-processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
-moler_history_2020, abadi_tensorflow_2016,
-paszke_pytorch_2019,ragan-kelley_halide_2013}.
-% EASY CUT
-These frameworks have been the
-subject of extensive industry investment to enable performant implementations,
-and often operate at the peak capacity of the hardware they run on
-\cite{lo_roofline_2015}.
+%(Note quite sure how much comes after this)
+%We should say the predicability is generating only the non-zeros!
+We leverage the intiial stuff of Finch to say we will in order generate the code that only grabs the non-zeros, modified in certain ways based on the fine grained ordering of computation)
+%(Use the below comment as an example - where the old paragraph talked about merging two lists.)
 
-The success and ubiquity of arrays is largely due to their simplicity. 
-%
-Since their introduction, multidimensional arrays have represented dense, rectilinear,
-integer grids of points. 
-%
-By \textbf{dense}, we mean that indices are mapped to value via a simple formula relating multidimensional space to linear memory.
-%
-Consequently, dense arrays offer extensive compiler optimizations and many convenient interfaces.
-%
-%This
-%simplicity enables extensive interoperability, convenience layers, and
-%optimizations by breaking the abstraction barrier between array representation
-%and array storage.  
-%
-Compilers understand dense array computations across many
-programming constructs, such as for and while loops, breaks, parallelism,
-caching, prefetching, multiple outputs, scatters, gathers, vectorization,
-loop-carry-dependencies, and more. Several optimizations have been developed for
-dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
-interchange.
-%
-However, while dense arrays are the easiest way to program for performance, the world is not all dense.
+%In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
+TACO did something similar?} In this model, performance engineers can more efficiently search the
+%complex landscape of programs and datastructures to find the best implementation.
+In this programming model, we can express certain concepts in programs, and others in data.
+For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
+%While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
 
-%Bring in structure
-% - old arrays are Dense, rectilinear, integer grids
-% - The world is not like that
-% - Sparsity, runs of repeated values, symmetry
-% - Lots of citations, perhaps a few figures from my powerpoint
-%Sparse arrays
-% - Networks (graphblas, graph frameworks)
-% - Simulations (BLAS, symblas, banded blas, etc)
-% - Databases (database engines)
-%Run-length encoding
-%Symmetry
-%Bands
-%Padding
-%Blocks
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{example_structures.png}
-    \caption{some example structures that might be nice to put into a figure}
-\end{figure}
-Our world is full of structured arrays.
-%
-Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
-%
-Run-length encoding describes images and masks, geometry, and
-databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
-%
-Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g.
-higher order FEMs) as well as in intermediate structures in many linear solvers (e.g.
-GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
-%
-In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
-%
-Operators, such as convolution, can be expressed as structured arrays.
-%
-For example, a convolution with a filter can be expressed as a matrix multiplication
-with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
-
-%
-\textbf{Currently, support for structured data is fragmented and incomplete}.
-%
-Experts must hand write variations of even the simplest kernels, like matrix
-multiply, for each data structure/data set and architecture to get performance.
-%
-Implementations must choose a small set of features to support well, resulting
-in a compromise between \textbf{program flexibility} and \textbf{data structure
-flexibility}.
-%
-Hand-written solutions are collected in diverse libraries like
-MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
-%
-However, libraries will only ever support a subset of
-programs on a subset of data structure combinations.
-%
-Even the most advanced
-libraries, such as the GraphBLAS, which support a wide variety of sparse
-operations over various semi-rings always lack support for other features, such
-as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
-%
-While dense array
-compilers support an enormous variety of program constructs like early break and
-multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
-%
-Special-purpose
-compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
-structures (only sparse, or only ragged arrays) must compromise by greatly
-constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
-%
-This trade-off is visualized in Tables \ref{tab:features} and \ref{tab:data_structures}.
-%
-
-\begin{table}[h!]
-  \centering
-  \scriptsize
-  \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
-  \hline
-  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
-  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
-%  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
-  Multiple LHS             & \checkmark &            & \checkmark & \checkmark &            & \checkmark \\
-  Affine Indices           & \checkmark &            &            & \checkmark & \checkmark & \checkmark \\
-  Recurrence               & \checkmark &            &            &            &            &           \\
-  If-Conditions and Masks  & \checkmark & \checkmark &            & \checkmark &            & \checkmark \\
-  Scatter Gather           & \checkmark &            &            & \checkmark &            &\checkmark \\
-  Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
-  \end{tabular}
-  \caption{Feature support across various tools.}
-  \label{tab:features}
-  \end{table}
-
-  \begin{table}[h!]
-  \centering
-  \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
-  \hline
-  Dense                    & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  Padded                   & \checkmark &            &            &            &            & \checkmark \\
-  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
-  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
-  Run-length               &            &            &            &            &            & \checkmark \\
-  Symmetric                &            &            &            &            & \checkmark & \checkmark \\
-  Regular Sparse Blocks    &            & \checkmark &            &            &            & \checkmark \\
-  Irregular Sparse Blocks  &            &            &            &            &            &\checkmark \\
-  Ragged                   &            &            & \checkmark &            &            & \checkmark \\
-  \end{tabular}
-  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
-  \label{tab:data_structures}
-\end{table}
-
-%
-Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
-%many of these special purpose compilers represent a program's control flow via a specialized representation that is tightly integrated with a narrow class of supported data structures.
-%
-For example, TACO merge lattices represent boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
-%
-The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
-%
-Taichi enriches single static assignment form with specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
-% TODO: Make me shorter
-The reason these systems tightly couple their control flow to narrow classes of data structures is because producing efficient programs over structured data requires overcoming two challenges that occur when we intersect complex control flow with structured data:
 %Taichi enriches single static assignment form with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
@@
 %We see two main challenges to writing efficient code over structured data.
-
-\textbf{Optimizations are specific to the indirection and patterns in data structures}: 
-%
-These structures break the simple mapping between array elements and where they are stored in memory.
-%
-For example, sparse arrays store lists of which coordinates are nonzero, whereas run-length-encoded arrays map several pixels to the same color value. 
-%
-These zero regions or repeated regions are optimization opportunities, and we must adapt the program to avoid repetitive work on these regions by referencing the stored structure.
-
-\textbf{Performance on structured data is highly algorithm dependent}: The landscape of implementation decisions is dramatically unpredictable. 
-%
-For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens2022autoscheduling, zhang2021gamma}.
-% TODO: Add Gamma
-%For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
-This means that performance engineering for such kernels requires the exploration of a large design space, changing the algorithm as well as the data structures.
-
-In this work, we propose a new programming language, Finch, which supports \textit{both} flexible control flow and diverse data structures.
-%
-Finch facilitates a programming model which resolves the challenges of computing over structured arrays by combining control flow and data structures into a common representation where they can be co-optimized.
-%
-In particular, Finch automatically specializes the control flow to the data so that performance engineers can focus on experimenting with many algorithms.
-%
-Finch supports a familiar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
-%
-This support would be useless without the appropriate level of structural specialization; Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
-%
-
-As an example, a programmer might explore the different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices).
-The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
-%The Finch compiler uses the structure of the data to generate efficient implementations of these programs.
-
-%\teo{Proposed new paragraph:}
-%\teo{Add: }
-%\teo{We don't propose to solve these problems automatically, but we do propose a new programming language where these issues are solvable, by virtue of the precise manner in which we combine structured data and control flow.}
-%
-%\teo{Our language, Finch, combines structured data and control flow in a predictable manner by lowering both of them into a single abstraction} Looplets~\cite{ahrens_looplets_2023}, which combines structured iterators into a single control flow that only produces the relevant portions of the output.}
-%
-%\teo{For example, if a programmer wanted to merge the even indices of two sorted arrays, a programmer could represent by point-wise multiplying two sparse vectors to produce another sparse vector and then set the even indices two zero or the programmer could only point-wise multiply the two sparse vectors under an if condition that selects even indices.
-(This crucial part needs improvement)}
-%
-%\teo{Via index expressions, multiple levels of loops and tensors, and more complex filtering expressions, we can imagine many variation of this problem with fine grained distinctions about when various operations are performed.}
-
-\begin{minipage}{0.333\linewidth}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             c[i] = a[i] * b[i]
-         end
-     end
-\end{minted}
-\end{minipage}%
-\begin{minipage}{0.333\linewidth}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             ap[i] = a[i]
-         end
-     end
-     for i = _
-         c[i] = ap[i] * b[i]
-     end
-\end{minted}
-\end{minipage}%
-\begin{minipage}{0.333\linewidth}
-\begin{minted}{julia}
-     for i = _
-         cp[i] = ap[i] * b[i]
-     end
-     for i = _
-         if i % 2 == 0
-             c[i] = cp[i]
-         end
-     end
-\end{minted}
-\end{minipage}%
-
-\begin{wrapfigure}{r}{0.2\linewidth}
-\begin{minted}{julia}
-     for i = _
-         cp[i] = ap[i] * b[i]
-     end
-     for i = _
-         if i % 2 == 0
-             c[i] = cp[i]
-         end
-     end
-\end{minted}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             c[i] = a[i] * b[i]
-         end
-     end
-\end{minted}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             ap[i] = a[i]
-         end
-     end
-     for i = _
-         c[i] = ap[i] * b[i]
-     end
-\end{minted}
-\end{wrapfigure}
-
-% \teo{
-% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
-% %
-% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
-% %
-% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
-% %
-% The Finch compiler uses the structure
-% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
-
-%% Many variations of this problem.
-%By combining many data structures and control flows into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
-%
-%We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
-%
-%(Note quite sure how much comes after this)
-%We should say the predicability is generating only the non-zeros!
-We leverage the intiial stuff of Finch to say we will in order generate the code that only grabs the non-zeros, modified in certain ways based on the fine grained ordering of computation)
-%(Use the below comment as an example - where the old paragraph talked about merging two lists.)
-
-%In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
-TACO did something similar?} In this model, performance engineers can more efficiently search the
-%complex landscape of programs and datastructures to find the best implementation.
-In this programming model, we can express certain concepts in programs, and others in data.
-For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
-%While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
-
-\subsection{Contributions}
-
-\begin{enumerate}
-\item 
-\teo{How about:
-A Looplet based level-format abstraction.
-%
-Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
-%
-The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
-}
-
-More complex array structures than ever before. A complete level-by-level
-structure-description language for expressing the structure of data
-hierarchically.\saman{level formats were done in TACO.} The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
-\item A rich structured array programming language with for-loops
-and complex control flow constructs at the same level of productivity
-of dense arrays. 
-%
-To our knowledge, the Finch programming language is the first 
-to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather. 
-%
-\saman{for beyond dense...don't want to sound like we are the fist to do for dense. } 
-\item A compiler that specializes programs to data structures 
-predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
-%
-%
-\teo{
-The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
-%
-By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
-%
-We observe that this strategy naturally extends the dense solution.
-}
-\teo{I think we can back it up in two ways:
-1. We need to modify the semantics to focus when on when things are converted into Looplets
-2.
-We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations.
-If we do triangle counting or something like that, we could maybe talk about that a bit.
-}
-\teo{
-\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
-}
-
-\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
-%We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
-Additionally, as a case study, a high-level array programming language and fusion interface for operations such as map, broadcast, or reduce that can be compiled to efficient code using the previous loop-level abstractions.
-%\item A complete set of level formats for expressing data patterns hierarchically in FiberTree-style decompositions.
-The first such set of formats to efficiently capture banded, triangular, run-length-encoded, or sparse-run-length-encoded datasets.
-The formats capture many use cases, from random updates to sequential construction.
-%\item The Finch array language, mirroring simple for-loops with imperative code blocks and if-conditions.
-The first array programming language for the above data formats to support multiple outputs, affine indexing, and imperfectly-nested loops.
-%\item Tensor lifecycles, a simple constraint on tensor reads and writes that elegantly restricts Finch programs to avoid complex data dependencies, and enables tensor polymorphism by providing implementers with well-defined functions to overload.
-%\item Wrapper Tensors which modify existing datastructures and recombine them to support new patterns, such as affine indexing, padding, transposition, and slicing.
-%\item Wrapper Levels which modify existing datastructures and enabling complex features such as atomic updates or contiguous versus separate allocation.
-%\item We define the first mappings from the existing pydata/sparse array api high-level operations to low level finch notation
-%\item <Performance Contributions>
-\end{enumerate}

### 2024-03-26 sections/intro.tex
Clean up intro after discussions

@@
 However, while dense arrays are the easiest way to program for performance, the world is not all dense.
-
-%Bring in structure
-% - old arrays are Dense, rectilinear, integer grids
-% - The world is not like that
-% - Sparsity, runs of repeated values, symmetry
-% - Lots of citations, perhaps a few figures from my powerpoint
-%Sparse arrays
-% - Networks (graphblas, graph frameworks)
-% - Simulations (BLAS, symblas, banded blas, etc)
-% - Databases (database engines)
-%Run-length encoding
-%Symmetry
-%Bands
-%Padding
-%Blocks
 
@@
 Prior implementations are incomplete because the abstractions they use are tightly coupled with the specific data structures that they support.
-%many of these special purpose compilers represent a program's control flow via a specialized representation that is tightly integrated with a narrow class of supported data structures.
 %
@@
 %
-Taichi enriches single static assignment form with specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
-% TODO: Make me shorter
-The reason these systems tightly couple their control flow to narrow classes of data structures is because producing efficient programs over structured data requires overcoming two challenges that occur when we intersect complex control flow with structured data:
-%Taichi enriches single static assignment form with specialized instruction for accessing a single sparse structure drawn from a particular class, which is why it supports the most control flow for a single sparse structure~\cite{hu_taichi_2019}.
-%
-%Furthermore, the limits and advantages of a DSL can be traced to the specific coupling of a program's control flow with the data structure: many new TACO features required modifying algorithms that worked with TACO's merge lattices and Tiramisu can express programs that Halide can't because Tiramisu uses a polyhedral model as opposed to a model built on iterating over regular grids~\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse, baghdadi2019tiramisu}.
-%
-%In this work, we will choose  a new representation of a program's data structures and control flows that covers a wide range of programs.
-%
-%However, simply combining a large variety of structured data with a large variety of programs does not necessarily yield a performance advantages.
-%
-%After all previous implementations choose to tightly couple to a specific data structure in part to ensure the data structures are iterated over efficiently.
-%
-%To find and produce efficient programs for structured data, we must overcome two challenges that occur when we intersect complex control flow with structured data:
-
-%before introducing this, we must discuss performance: if we want to also produce efficient code for structured data, we face two challenges.
-
-%build on some conception of an iteration space: TACO has merge lattices, Polyhedral compilers rely on the polyhedral model, Halide and Corra uses rectangular intervals, and so on.
-%
-%And we believe the limitations of these systems can be traced back to their specific conceptions.} \willow{I agree with what this sentence is saying, but I'm concerned that the sentiment is unclear, because here it may seem like we say "iteration space bad" and later say "iteration space good".
-Can we perhaps adjust here to say may of these special purpose compilers use overly restrictive models of the iteration space of the program, but that Finch is good because it uses general control flow to express a wider variety of iteration spaces.
-At a high level, we need to define "iteration space" or "control flow" or "Looplets" if we want to use them in the intro, and depending on how much we want to use these words we may want to avoid them here.}
-%
-% \teo{For example, each modification to expressions that TACO could support required modifying the merge lattice algorithms\cite{kjolstad_tensor_2017, henry_compilation_2021, senanayake2020sparse} and several programs could not be supported in Halide, but could be in Tiramisu due to the limitations of a rectangular iteration space~\cite{baghdadi2019tiramisu}.
-% %
-% And many iteration space concepts simply don't allow for skipping work that isn't simply the complement of an interval.
-% %
-% In our work, we will show how we can use Looplets as our main tool to manage the iteration spaces of our computations and data structures.
-% %
-% However, to support as many data structures and iteration spaces as is implied above, we face two challenges if we want to also produce efficient code over structured data.
-% }
-%We see two main challenges to writing efficient code over structured data.
+Taichi enriches single static assignment form with a specialized instruction for accessing only a single sparse structure, but it supports more control flow ~\cite{hu_taichi_2019}.
+These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data.
+The challenge is twofold:
+%writing efficient code over structured data
 
@@
 For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens2022autoscheduling, zhang2021gamma}.
-% TODO: Add Gamma
 %For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
@@
 
-\begin{wrapfigure}{r}{0.2\linewidth}
-\begin{minted}{julia}
-     for i = _
-         cp[i] = ap[i] * b[i]
-     end
-     for i = _
-         if i % 2 == 0
-             c[i] = cp[i]
-         end
-     end
-\end{minted}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             c[i] = a[i] * b[i]
-         end
-     end
-\end{minted}
-\begin{minted}{julia}
-     for i = _
-         if i % 2 == 0
-             ap[i] = a[i]
-         end
-     end
-     for i = _
-         c[i] = ap[i] * b[i]
-     end
-\end{minted}
-\end{wrapfigure}
-
-% \teo{
-% Due to this fine grained mixing of control flow and data structures, we dub a new programming model for Finch: ``Fine Grained Data Structure Driven Array Programming''.
-% %
-% Finch supports \textit{both} flexible programming constructs and diverse data structures. 
-% %
-% Finch supports a similar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
-% %
-% The Finch compiler uses the structure
-% of the data to generate efficient implementations of these programs by predictably converting the control flow into the same representation as the data, Looplets~\cite{ahrens_looplets_2023}.}
-
-%% Many variations of this problem.
-%By combining many data structures and control flows into a single language, a performance engineer could use a sufficiently predicable and fine grained compiler to find efficient programs.
-%
-%We dub this model, ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, but then can reasonably iterate on the two with a compiler that predicable combines them, leading to efficient code eventually.
-%
-%(Note quite sure how much comes after this)
-%We should say the predicability is generating only the non-zeros!
-We leverage the intiial stuff of Finch to say we will in order generate the code that only grabs the non-zeros, modified in certain ways based on the fine grained ordering of computation)
-%(Use the below comment as an example - where the old paragraph talked about merging two lists.)
-
-%In this work, we propose a new programming model we call ``Datastructure-driven Array Programming'' in which the programmer specifies the structure of the data separately from the program, and the compiler uses these two descriptions to generate efficient code.\saman{Is this new?
-TACO did something similar?} In this model, performance engineers can more efficiently search the
-%complex landscape of programs and datastructures to find the best implementation.
-In this programming model, we can express certain concepts in programs, and others in data.
-For example, if we wish to merge two sorted lists, we express this as two sparse vectors which are true whenever the list contains the key in question, and we iterate over the \textit{entire} space of keys, writing to the output list whenever either of the vectors are true.
-%While the order in which we iterate over the data is expressed in the program, which datapoints are of interest and how to find them is expressed in the data.
-
 \subsection{Contributions}

### 2024-03-26 sections/cut.tex
calling it for today

@@
 %We see two main challenges to writing efficient code over structured data.
+\teo{How about:
+A Looplet based level-format abstraction.
+%
+Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
+%
+The first such set of formats to efficiently capture banded,
+triangular, run-length-encoded, or sparse datasets, and any combination thereof.
+}
+
+with key level formats that completely capture the main structural properties of data. for expressing the structure of data
+hierarchically.\saman{level formats were done in TACO.} The first such set of formats to efficiently capture banded,
+triangular, run-length-encoded, or sparse datasets, and any combination thereof.

### 2024-03-26 sections/intro.tex
fix
calling it for today

@@
 These systems tightly couple their control flow to narrow classes of data structures to avoid the challenges that occur when we intersect complex control flow with structured data.
-The challenge is twofold:
+There are two challenges:
 %writing efficient code over structured data
@@
 \item
-\teo{How about:
-A Looplet based level-format abstraction.
-%
-Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} features a flexible data structure description language for array-like computations, we show that we can describe more level formats than any other by a level format abstraction designed to target Looplets in the context of an array programming language with fine-grained control flow.
-%
-The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
-}
 
-More complex array structures than ever before. A complete level-by-level
-structure-description language for expressing the structure of data
-hierarchically.\saman{level formats were done in TACO.} The first such set of formats to efficiently capture banded,
-triangular, run-length-encoded, or sparse datasets, and any combination thereof.
+More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture
+banded, triangular, run-length-encoded, or sparse datasets, and any combination
+thereof. We have chosen a set of level formats that completely captures all combinations of relevant
+structural properties (zeros, repeated values, and/or blocks).
+Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level.
+
 \item A rich structured array programming language with for-loops
@@
 to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather. 
+structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands. 
 %
-\saman{for beyond dense...don't want to sound like we are the fist to do for dense. } 
 \item A compiler that specializes programs to data structures 

### 2024-03-27 sections/tensors.tex
cool
okay, I'm calling it
something like that
Update on Overleaf.
fix

@@
 \subsection{The 8 Key Level Structures}
+    \begin{wrapfigure}{r}{.35\textwidth}
+        \centering
+        \scriptsize
+        \begin{tabular}{|c|c|c|c|l|}
+            \hline
+            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
+            \hline
+             &  &  &  & Dense \\
+            \hline
+             &  & \checkmark &  & DenseRLE \\
+            \hline
+            \checkmark &  &  &  & Sparse \\
+            \hline
+            \checkmark &  &  & \checkmark & SparsePinpoint \\
+            \hline
+            \checkmark &  & \checkmark &  & SparseRLE \\
+            \hline
+            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
+            \hline
+            \checkmark & \checkmark &  &  & SparseVBL \\
+            \hline
+            \checkmark & \checkmark &  & \checkmark & SparseBand \\
+            \hline
+        \end{tabular}
+        \caption{All combinations of relevant structural properties and their
+        corresponding formats.  Note that blocks and runs need not be considered
+        together because we must store a run length for each run, and so there
+        isn't a significant storage benefit to combining them. Blocks and
+        singletons only make sense in the contex of sparsity, so we don't
+        consider them together either. We omit such combinations
+        from the exhaustive table.}
+        \label{tab:TypesOfStructure}
+    \end{wrapfigure}
     The main benefits of specializing to structure come from the following properties of the data:
@@
     tree, as shown in Figure~\ref{fig:structuraldiversity}.
-    \begin{table}
-        \centering
-        \begin{tabular}{|c|c|c|c|l|}
-            \hline
-            \textbf{Sparse} & \textbf{Blocked} & \textbf{Runs} & \textbf{Singletons} & \textbf{Corresponding Format} \\
-            \hline
-             &  &  &  & Dense \\
-            \hline
-             &  & \checkmark &  & DenseRLE \\
-            \hline
-            \checkmark &  &  &  & Sparse \\
-            \hline
-            \checkmark &  &  & \checkmark & SparsePinpoint \\
-            \hline
-            \checkmark &  & \checkmark &  & SparseRLE \\
-            \hline
-            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
-            \hline
-            \checkmark & \checkmark &  &  & SparseVBL \\
-            \hline
-            \checkmark & \checkmark &  & \checkmark & SparseBand \\
-            \hline
-        \end{tabular}
-        \caption{All combinations of relevant structural properties and their
-        corresponding formats.  Note that blocks and runs need not be considered
-        together because we must store a run length for each run, and so there
-        isn't a significant storage benefit to combining them. Blocks and
-        singletons only make sense in the contex of sparsity, so we don't
-        consider them together either. We omit such combinations
-        from the exhaustive table.}
-        \label{tab:TypesOfStructure}
-    \end{table}
 
@@
 necessary for our case studies.
+
+\subsection{Scalars}
+
+\subsubsection{Sparse Scalars}
+\subsubsection{Early Break Scalars}

### 2024-03-27 sections/cut.tex
formatting

@@
 triangular, run-length-encoded, or sparse datasets, and any combination thereof.
+
+%
+\teo{
+The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
+%
+By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
+%
+We observe that this strategy naturally extends the dense solution.
+}
+\teo{I think we can back it up in two ways:
+1. We need to modify the semantics to focus when on when things are converted into Looplets
+2.
+We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations.
+If we do triangle counting or something like that, we could maybe talk about that a bit.
+}
+\teo{
+\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
+}

### 2024-03-27 sections/intro.tex
Update on Overleaf.
formatting

@@
 % EASY CUT
-These frameworks have been the
-subject of extensive industry investment to enable performant implementations,
-and often operate at the peak capacity of the hardware they run on
-\cite{lo_roofline_2015}.
+%These frameworks have been the
+%subject of extensive industry investment to enable performant implementations,
+%and often operate at the peak capacity of the hardware they run on
+%\cite{lo_roofline_2015}.
 
@@
 
+\newcommand*\rot{\rotatebox{90}}
+
 \begin{table}[h!]
-  \centering
+\noindent
+\flushleft
+\begin{minipage}[t]{.49\textwidth}
   \scriptsize
   \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
+  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
   \hline
@@
   \label{tab:features}
-  \end{table}
-
-  \begin{table}[h!]
-  \centering
+\end{minipage} 
+\begin{minipage}[t]{.49\textwidth}
+  \flushright
+  \scriptsize
   \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \textbf{Halide} & \textbf{Taco} & \textbf{Cora} & \textbf{Taichi} & \textbf{StructTensor} & \textbf{Finch} \\
+  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
   \hline
@@
   \label{tab:data_structures}
+\end{minipage}
 \end{table}
@@
 
-As an example, a programmer might explore the different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices).
+As an example, a programmer might explore different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices).
 The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
@@
 %
-\item A compiler that specializes programs to data structures 
-predictably,\saman{Interesting...can we back this claim up in the discription?} facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
+\item A compiler that specializes programs to data structures automatically,
+facilitating an expressive language that makes it easier to search the complex
+space of algorithms and data structures. 
 %
-%
-\teo{
-The Finch language offers extensible abstractions for elements of control flow (conditionals and indexing expressions) that the Finch compiler predictably converts into the same structured representation used to describe arrays.
-%
-By controlling how control flow is lowered into this representation, a programmer can control how structured data is combined with control flow to produce a single control flow, the program.
-%
-We observe that this strategy naturally extends the dense solution.
-}
-\teo{I think we can back it up in two ways:
-1. We need to modify the semantics to focus when on when things are converted into Looplets
-2.
-We need to highlight the specifics in the case studies: in particular, the blur and the symmetric operations.
-If we do triangle counting or something like that, we could maybe talk about that a bit.
-}
-\teo{
-\item Extensibility of the language and compiler with respect to data structures and control flow: (Maybe put stuff about the data structure extensibility here?)
-}
-
 \item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.

### 2024-03-27 sections/tensors.tex
formatting

@@
 \section{Bridging Looplets and Finch: The Tensor Interface}
-
 %
@@
 Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.
+
+\begin{wrapfigure}{R}{0.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
+    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
+    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
+    \label{fig:levelsvsfibers}
+\end{wrapfigure}
 
@@
 %2. (Done) We have identified 8 key level structures to represent most combinations of runs or pinpoints.
-%2. a. beautiful figure with datas as rows of larger structures
+%2. (Done) a. beautiful figure with datas as rows of larger structures
 %2. b. How do we write to each of these, in order? Randomly? random access? several important datastructures that show up along the way.
 %3. What is "Unfurling"? and how does it help us represent each structure? refer to looplet decompositions of each case in the earlier figure
-%4. What are lifecycles? How do lifecycles help us keep sane? Why are they necessary for correctness?
+%4. (Done) What are lifecycles? How do lifecycles help us keep sane? Why are they necessary for correctness?
 
 \subsection{Level Abstraction}
+
 Fiber-tree style tensor abstractions have been the subject of extensive study
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.45\linewidth]{LevelsVsFibers-matrix.png}\hfill%
-    \includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
-    \caption{Levels and fiber tree representations of a sparse matrix and a sparse tensor.
-On left, a matrix is represented in a fibertree corresponding to CSC format, with a dense outer level and a sparse inner level.
-On right, a tensor is represented in a fibertree with two sparse outer levels, and a dense inner level.
-Note that the element levels in this case form the leaves of the tree.}
-    \label{fig:levelsvsfibers}
-\end{figure}
-
 \subsection{The 8 Key Level Structures}
-    \begin{wrapfigure}{r}{.35\textwidth}
+    \begin{wrapfigure}{R}{.35\textwidth}
         \centering

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 Outer product computes the product of all pairs of elements from the input matrices.
-More detailed explanation is available in Section 2.2 of ~\cite{} \changwan{gamma}.
-\changwan{Inner, asym bad, }
+More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
+Figures ~\ref{spgemm_small} and ~\reef{spgemm_large} show
 
-Figure~\ref{spgemm_small} 
 TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
@@
         \hline
+    \end{tabular}
+\end{table}
+
+\begin{table}[htbp]
+    \centering
+    \caption{SpMV Sample Matrices}
+    \label{spmv_sample_matrices}
+    \begin{tabular}{|l|c|c|c|c|}
+     \textbf{Spy} &
+     \includegraphics[width=3cm]{spmv_matrices/saylr4.png} &
+     \includegraphics[width=3cm]{spmv_matrices/heart3.png} &
+     \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} &
+
     \end{tabular}
@@
 
-\subsubsection{The Benefits of Kernel Fusion}
+\subsubsection{Evaluating Finch Logic}
+To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations.
+Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 \begin{figure}
+\scriptsize
 \noindent % This ensures the minipages fill the page width
@@
     \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
-    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs \texttt{<<op>>=} rhs}} \\
+    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs <<op>>= rhs}} \\
     \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
@@
 
+We present a sketch of a small-step operational semantics, showing how to execute a Finch program in a host language.
 We break our description of the semantics of Finch into three parts, those
@@
 
+%What are values? (objects in the host language)
+%What is t really? (
+
 \begin{figure}
     \centering
-    \small
+    \footnotesize
 
@@
     \end{prooftree}
-
-    \caption{Basic evaluation semantics, roughly defining most of these language
-    constructs to function similarly to their classical definitions. Note that
-    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
-    environment and another state $t$ corresponding to the state in the
-    host language. Several Looplets introduce variables into the embedding
-    language, which may be read when evaluating the \finchvalue node.}
-    \label{fig:semantics_core}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \small
+    \vspace{12pt}
 
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer4[$Declare$]{\langle\finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle \finchdeclare(tns', init', dims'...), t \rangle)}
+    \hypo{\langle declare(tns', init', dims'...), t \rangle \rightarrow t'}
+    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], t')}
     \end{prooftree}
@@
     \hypo{e(tns) = tns'}
-    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \langle freeze(tns'), t \rangle)}
+    \hypo{\langle freeze(tns'), t \rangle \rightarrow t'}
+    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], t')}
     \end{prooftree}
@@
 
-    \caption{Lifecycle evaluation semantics. The state $s$ of the finch compiler
+    \caption{Basic evaluation semantics, roughly defining most of these language
+    constructs to function similarly to their classical definitions. Note that
+    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
+    environment and another state $t$ corresponding to the state in the
+    host language. Several Looplets introduce variables into the embedding
+    language, which may be read when evaluating the \finchvalue node.
+    %
+    The state $s$ of the finch compiler
     is a tuple $(e, t)$ of a variable value environment and another state $t$
-    corresponding to the state in the host language. All of the lifecycle
+    corresponding to the state in the host language.
+    This means that rules which modify $t$ are running in the host language. All of the lifecycle
     functions are designed to be implemented and executed in the host language,
     but these semantics enforce that each of these functions may update state in
-    the host language and flip the mode of the tensor between read and update.}
-    \label{fig:semantics_lifecycle}
+    the host language and flip the mode of the tensor between read and update}
+    \label{fig:semantics_core}
 \end{figure}
@@
     \centering
-    \small
+    \footnotesize
 
@@
     environment and another state $t$ corresponding to the state in the host
-    language.  Note that $E$ is an evaluation context that applies anywhere in
+    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
     the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to

### 2024-03-27 sections/eval.tex
cool
Update on Overleaf.

@@
 SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
-The inner product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
-Inner product involves multiplying two matrices by taking dot products of corresponding rows and columns.
-Outer product computes the product of all pairs of elements from the input matrices.
+The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
+Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns.
+Outer-product computes the product of all pairs of elements from the input matrices.
 More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
-Figures ~\ref{spgemm_small} and ~\reef{spgemm_large} show
+Figures ~\ref{spgemm_small} and ~\reef{spgemm_large} show SpGEMM results with Finch and TACO.
+Note that inner-product is pretty ineffi
 

### 2024-03-27 main.tex
Update on Overleaf.
move figures
fix

@@
 \usepackage{graphicx}
+\graphicspath{{figures/}}
 %\usepackage{amssymb}
@@
 \usepackage{multirow}
-%
+\usepackage{wrapfig}
+
+\makeatletter
+\DeclareFontFamily{OMX}{MnSymbolE}{}
+\DeclareSymbolFont{MnLargeSymbols}{OMX}{MnSymbolE}{m}{n}
+\SetSymbolFont{MnLargeSymbols}{bold}{OMX}{MnSymbolE}{b}{n}
+\DeclareFontShape{OMX}{MnSymbolE}{m}{n}{
+    <-6>  MnSymbolE5
+   <6-7>  MnSymbolE6
+   <7-8>  MnSymbolE7
+   <8-9>  MnSymbolE8
+   <9-10> MnSymbolE9
+  <10-12> MnSymbolE10
+  <12->   MnSymbolE12
+}{}
+\DeclareFontShape{OMX}{MnSymbolE}{b}{n}{
+    <-6>  MnSymbolE-Bold5
+   <6-7>  MnSymbolE-Bold6
+   <7-8>  MnSymbolE-Bold7
+   <8-9>  MnSymbolE-Bold8
+   <9-10> MnSymbolE-Bold9
+  <10-12> MnSymbolE-Bold10
+  <12->   MnSymbolE-Bold12
+}{}
+
+\let\llangle\@undefined
+\let\rrangle\@undefined
+\DeclareMathDelimiter{\llangle}{\mathopen}%
+                     {MnLargeSymbols}{'164}{MnLargeSymbols}{'164}
+\DeclareMathDelimiter{\rrangle}{\mathclose}%
+                     {MnLargeSymbols}{'171}{MnLargeSymbols}{'171}
+\makeatother
 
@@
 \newcommand{\HIDE}[1]{}
+
+\newcommand{\rothead}[1]{\rotatebox{90}{\textbf{#1}}}
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.
fix

@@
 More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
-Figures ~\ref{spgemm_small} and ~\reef{spgemm_large} show SpGEMM results with Finch and TACO.
-Note that inner-product is pretty ineffi
+Figures ~\ref{spgemm_small} and ~\ref{spgemm_large} show SpGEMM results with Finch and TACO.
+Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of inner-product is dominated by the cost of intersection
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 Figures ~\ref{spgemm_small} and ~\ref{spgemm_large} show SpGEMM results with Finch and TACO.
-Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of inner-product is dominated by the cost of intersection
+Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant comput
 

### 2024-03-27 sections/eval.tex

@@
 Figures ~\ref{spgemm_small} and ~\ref{spgemm_large} show SpGEMM results with Finch and TACO.
-Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant comput
-
+Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant computations.
 TACO does not support outer product due to its lack of support for scatter into a 2D workspace.

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant computations.
-TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
+For the output-product, TACO only support the dense output: TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
+So, due to memory constraints, we are unable to run TAC
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 For the output-product, TACO only support the dense output: TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
-So, due to memory constraints, we are unable to run TAC
+So, due to memory constraints, we are unable to run TACO's outer-product for large sparse matrices.
+So, for large sparse matrices, we only compared Gustavson's algorithms.
+As shown in Figure~\ref{spgemm_large}, Finch outperforms TACO with an average speed
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 So, for large sparse matrices, we only compared Gustavson's algorithms.
-As shown in Figure~\ref{spgemm_large}, Finch outperforms TACO with an average speed
+As shown in Figure~\ref{spgemm_small}, for inner-product, Finch achieves comparable performance with TACO, and for outer-product, 
+As shown in Figure~\ref{spgemm_large}, for large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

--- 2024-03-27 sections/eval.tex
@@
 So, for large sparse matrices, we only compared Gustavson's algorithms.
-As shown in Figure~\ref{spgemm_large}, Finch outperforms TACO with an average speed
+As shown in Figure~\ref{spgemm_small}, for inner-product, Finch achieves comparable performance with TACO, and for outer-product, teh
+As shown in Figure~\ref{spgemm_large}, for large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
 

### 2024-03-27 sections/eval.tex

--- 2024-03-27 sections/eval.tex
@@
 More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
-Figures ~\ref{spgemm_small} and ~\ref{spgemm_large} show SpGEMM results with Finch and TACO.
-Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant computations.
-For the output-product, TACO only support the dense output: TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
+Figures ~\ref{fig:spgemm_small} and ~\ref{fig:spgemm_large} show SpGEMM results with Finch and TACO.
+Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant traversals.
+For the output-product, TACO only supports the dense output: TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
 So, due to memory constraints, we are unable to run TACO's outer-product for large sparse matrices.
+On the other hand,
 So, for large sparse matrices, we only compared Gustavson's algorithms.
-As shown in Figure~\ref{spgemm_large}, Finch outperforms TACO with an average speed
+\changwan{mention TACO's gustavson's algorithm is hard-wired}
+As shown in Figure~\ref{fig:spgemm_small}, for inner-product, Finch achieves comparable performance with TACO, and for outer-product, the 
+As shown in Figure~\ref{fig:spgemm_large}, for large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
 
@@
 \begin{figure}
-    \label{spgemm_small}
+    \label{fig:spgemm_small}
 	\includegraphics[width=\linewidth]{spgemm_small_speedup_log_scale.png}
@@
 \begin{figure}
-\label{spgemm_large}
+\label{fig:spgemm_large}
 	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 %}
+
+Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics, focusing on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
+
+The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM.
+Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
+Further explanation is available in Section 2.2 of Zhang et al.
+[1].
+
+Figures \ref{fig:spgemm_small} and \ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO.
+Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals.
+Additionally, TACO does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
+In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.
+
+As depicted in Figure \ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product.
+For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
+It's worth mentioning that TACO's Gustavson's algorithm is hard-wired.
 
@@
 So, due to memory constraints, we are unable to run TACO's outer-product for large sparse matrices.
-On the other hand,
+On the other hand, Finch supports scatter into 2D workspace,
 So, for large sparse matrices, we only compared Gustavson's algorithms.

### 2024-03-27 sections/lang.tex
cool
fixed

@@
 
-%What are values? (objects in the host language)
-%What is t really? (
-
 \begin{figure}
@@
         \hypo{\langle f, (e, t) \rangle \Rightarrow g}
-        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \langle g(vals...), t \rangle}
+        \infer2[$Call$]{\langle\finchcall(f, args...), (e, t)\rangle \rightarrow \llangle g(vals...), t \rrangle}
     \end{prooftree}
-    \vspace{12pt}
+    \vspace{6pt}
 
@@
         \hypo{}
-        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \langle ex, t \rangle}
+        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \llangle ex, t \rrangle}
     \end{prooftree}
-    \vspace{12pt}
+    \vspace{6pt}
 
@@
     \end{prooftree}
-    \vspace{12pt}
+    \vspace{6pt}
 
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \hypo{\langle declare(tns', init', dims'...), t \rangle \rightarrow t'}
-    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], t')}
+    \infer4[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
     \end{prooftree}
-    \vspace{12pt}
+    \vspace{6pt}
 
@@
     \hypo{e(tns) = tns'}
-    \hypo{\langle freeze(tns'), t \rangle \rightarrow t'}
-    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], t')}
+    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle)}
     \end{prooftree}
-    \vspace{12pt}
+    \vspace{6pt}
 
@@
     \hypo{e(tns) = tns'}
-    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \langle thaw(tns'), t \rangle)}
+    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle)}
     \end{prooftree}
@@
 
-\vspace{12pt}
+\begin{multline*}
+E := [\cdot] | \finchloop(idx, ext, E | body) | \finchblock(E | head, E | tail) | \finchsieve(E | cond, E | body) |\\ 
+\finchdeclare(var, E|rhs, E|body) | \finchcall(E|f, E|args...) | \finchaccess(tns, E | idxs...)
+\end{multline*}
+
+\begin{prooftree}
+    \hypo{e(tns) \mapsto tns'}
+    \hypo{\llangle unfurl(tns', ext)\rrangle \mapsto tns''}
+    \infer2[$Unfurl$]{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), s\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), s\rangle}
+\end{prooftree}
+\vspace{6pt}
 
@@
     \hypo{e(i) = i'}
-    \hypo{\langle seek(i'), t \rangle \rightarrow t'}
+    \hypo{\llangle seek(i'), t \rrangle \rightarrow t'}
     \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
 \end{prooftree}
-\vspace{12pt}
+\vspace{6pt}
 
 \begin{prooftree}
-    \hypo{\langle cond, t \rangle \Rightarrow true}
+    \hypo{\llangle cond, t \rrangle \Rightarrow true}
     \infer1[]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(head, i...)], s\rangle}}
@@
 \begin{prooftree}
-    \hypo{\langle cond, t \rangle \Rightarrow false}
+    \hypo{\llangle cond, t \rrangle \Rightarrow false}
     \infer1[$Switch$]{\splitfrac{\langle E[\finchaccess(switch(cond, head, tail), i...)], s\rangle}{\rightarrow \langle E[\finchaccess(tail, i...)], s\rangle}}
 \end{prooftree}
-\vspace{12pt}
+\vspace{6pt}
 
@@
 \end{prooftree}
-\vspace{12pt}
+\vspace{6pt}
+
+\begin{prooftree}
+    \hypo{\llangle preamble, t \rrangle \rightarrow t'}
+    \hypo{\langle E[body], (e, t') \rangle \rightarrow (e', t'')}
+    \hypo{\llangle epilogue, t'' \rrangle \rightarrow t'''}
+    \infer3[$Thunk$]{\langle E[thunk(preamble, body, epilogue)], (e, t')\rangle \rightarrow (e', t''')}
+\end{prooftree}
+\vspace{6pt}
 
@@
 \end{prooftree}
-\vspace{12pt}
+\vspace{6pt}
 
 \begin{prooftree}
-    \hypo{\langle seek(a), t \rangle \rightarrow t'}
+    \hypo{\llangle seek(a), t \rrangle \rightarrow t'}
     \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
 \end{prooftree}
-\vspace{12pt}
+\vspace{6pt}
 
@@
     \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
-    \hypo{\langle next, t' \rangle \rightarrow t''}
+    \hypo{\llangle next, t' \rrangle \rightarrow t''}
     \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
@@
     the priority in which we apply looplet rules, which is as follows:
-    $Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup$
+    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
     Many looplets, most notably the thunk looplet, introduce variables into the

### 2024-03-27 sections/eval.tex

@@
 Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
-Further explanation is available in Section 2.2 of Zhang et al.
-[1].
+Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.
 
@@
 It's worth mentioning that TACO's Gustavson's algorithm is hard-wired.
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics.
-SpGEMM focuses on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
-The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM. 
-Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns.
-Outer-product computes the product of all pairs of elements from the input matrices.
-More detailed explanation is available in Section 2.2 of ~\cite{zhang2021gamma}.
-Figures ~\ref{fig:spgemm_small} and ~\ref{fig:spgemm_large} show SpGEMM results with Finch and TACO.
-Note that inner-product is very inefficient unless sparse matrices are nearly dense because the cost of intersections that incurs a lot of redundant traversals.
-For the output-product, TACO only supports the dense output: TACO does not support outer product due to its lack of support for scatter into a 2D workspace.
-So, due to memory constraints, we are unable to run TACO's outer-product for large sparse matrices.
-On the other hand, Finch supports scatter into 2D workspace,
-So, for large sparse matrices, we only compared Gustavson's algorithms.
-\changwan{mention TACO's gustavson's algorithm is hard-wired}
-As shown in Figure~\ref{fig:spgemm_small}, for inner-product, Finch achieves comparable performance with TACO, and for outer-product, the 
-As shown in Figure~\ref{fig:spgemm_large}, for large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
+Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
 Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.
@@
 Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals.
-Additionally, TACO does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
+Additionally, TACO only supports outer product when the output matrix format is dense.
+It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
 In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
-It's worth mentioning that TACO's Gustavson's algorithm is hard-wired.
+It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.
 

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-We used graphs in GraphBLAS~\cite{}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
 Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.

### 2024-03-27 sections/eval.tex
Update on Overleaf.

--- 2024-03-27 sections/eval.tex
@@
 To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-We used graphs in GraphBLAS~\cite{}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{beamer2012direction} is crucial to achieve high-performance.
 Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{triangle_count_speedup_over_duckdb.png}
-    \caption{Performance of Finch on triangle counting.}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{elementwise_speedup_over_duckdb.png}
-    \caption{Performance of Finch on elementwise (A+B)*C kernel as the sparsity of C varies.}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{sddmm_speedup_over_duckdb.png}
-    \caption{Performance of Finch on SDDMM.}
+	\includegraphics[width=.3\linewidth]{triangle_count_speedup_over_duckdb.png}
+	\includegraphics[width=.35\linewidth]{sddmm_speedup_over_duckdb.png}
+	\includegraphics[width=.3\linewidth]{elementwise_speedup_over_duckdb.png}
+    \caption{Performance of Finch Logic for common kernels.}
 \end{figure}

### 2024-03-27 main.tex
fix

@@
 
-Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes (from Fred's thesis!).
+Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
 So far this has failed in both scope of the ambition and execution of comparable features.
@@
 programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays.
-Our approach enables new loop optimizations across multiple domains, unifying techniques such as
-sparse tensors, databases, and lossless compression\kyle{nit: "unifying techniques in areas such as"}.
+Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
+sparse tensors, databases, and lossless compression.
 We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  

### 2024-03-27 sections/eval.tex
fix

--- 2024-03-27 sections/eval.tex
@@
 Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
-Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
+
 Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.
@@
 Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals.
-Additionally, TACO only supports outer product when the output matrix format is dense.
-It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
+Additionally, TACO does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
 In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.
@@
 For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
-It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.
+It's worth mentioning that TACO's Gustavson's algorithm is hard-wired.
 
@@
     \centering
-    \caption{SpMV Sample Matrices}
-    \label{spmv_sample_matrices}
-    \begin{tabular}{|m{3cm}|c|c|c|}
+    \caption{Transposed SpMV Sample Matrices}
+    \label{tab:transposed_spmv_sample_matrices}
+    \begin{tabular}{|
+        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
+        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
+        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
+        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
+        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
         \hline
-        \textbf{Spy} & \textbf{Group / Name} & \textbf{Dimensions} & \textbf{Best Finch Format}\\
+         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
         \hline
-        \includegraphics[width=3cm]{spmv_matrices/saylr4.png} & HB/saylr4 & 3,564 x 3,564 (22,316) & Symmetric SparseList \\
+        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
         \hline 
-        \includegraphics[width=3cm]{spmv_matrices/heart3.png} & Norris/heart3 & 2,339 x 2,339 (680,341) & SparseVBL \\
+        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
         \hline
-        \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} & toeplitz\_large\_band  & 10,000 x 10,000 (1,999,900) & SparseBand \\
+        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
         \hline
-        \includegraphics[width=3cm]{spmv_matrices/as-735.png} & SNAP/as-735 & 7,716 x 7,716 (26,467) & Symmetric SparseList-Pattern \\
+        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
         \hline
-    \end{tabular}
-\end{table}
-
-\begin{table}[htbp]
-    \centering
-    \caption{SpMV Sample Matrices}
-    \label{spmv_sample_matrices}
-    \begin{tabular}{|l|c|c|c|c|}
-     \textbf{Spy} &
-     \includegraphics[width=3cm]{spmv_matrices/saylr4.png} &
-     \includegraphics[width=3cm]{spmv_matrices/heart3.png} &
-     \includegraphics[width=3cm]{spmv_matrices/toeplitz_large_band.png} &
-
     \end{tabular}
@@
 \begin{figure}
-    \begin{minipage}[t]{0.5\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \includegraphics[width=\linewidth]{spmv_performance_sorted_(faster_than_taco).png}
-    \end{minipage}%
-    \begin{minipage}[t]{0.5\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \includegraphics[width=\linewidth]{spmv_performance_sorted_(slower_than_taco).png}
-    \end{minipage}
-    \caption{Performance of SpMV across various tools.}
-    \label{spmv_sorted}
-\end{figure}
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{spmv_performance_grouped.png}
+    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
     \caption{Performance of SpMV by Finch format.}
-    \label{spmv_grouped}
+    \label{fig:spmv_grouped}
     \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{erode2_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (2 iterations).}
+	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
+	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
+    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}
 \end{figure}
 
-\begin{figure}
-	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
-    \caption{Performance of Finch on erosion task (4 iterations).}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=\linewidth]{hist_speedup_over_opencv.png}
-    \caption{Performance of Finch on masked histogram task.}
-\end{figure}
+%\begin{figure}
+%	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
+%    \caption{Performance of Finch on erosion task (4 iterations).}
+%\end{figure}
 
@@
 To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
+We used graphs in GraphBLAS~\cite{}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{} is crucial to achiceve high-performance.
 Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.

### 2024-03-27 sections/eval.tex

@@
 Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
-
+Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
 Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.

### 2024-03-27 sections/eval.tex
Update on Overleaf.

@@
 Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can manipulate an index as a scalar to implement $argmin$.
+Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can access an index as a scalar to implement $argmin$.
 For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
@@
 Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.
-
 \begin{figure}
-	\includegraphics[width=.3\linewidth]{triangle_count_speedup_over_duckdb.png}
-	\includegraphics[width=.35\linewidth]{sddmm_speedup_over_duckdb.png}
-	\includegraphics[width=.3\linewidth]{elementwise_speedup_over_duckdb.png}
+\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
+  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
+  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
+  \vspace{0pt} \includegraphics[width=145pt, height=120pt]{figures/elementwise_speedup_over_duckdb.png}
+\end{tabular}
 \caption{Performance of Finch Logic for common kernels.}

### 2024-03-27 sections/eval.tex
shrink

@@
 \begin{figure}
-    \label{fig:spgemm_small}
-	\includegraphics[width=\linewidth]{spgemm_small_speedup_log_scale.png}
-    \caption{A comparison of several matrix multiplication algorithms between Finch and Taco smaller matrices, ordered from small to big dimension.
-Note that inner products necessarily requires $O(n^2)$ work and taco's outer products format is dense.
-Finch can use a sparse outer products format and thus has an asymptotic advantage that becomes evident as the output dimensions grow.}
-\end{figure}
-
-\begin{figure}
-\label{fig:spgemm_large}
-	\includegraphics[width=\linewidth]{spgemm_joel_speedup.png}
-    \caption{A comparison of gustavson's algorithm between Finch and Taco on some larger matrices}
+    \label{fig:spgemm}
+	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
+	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
+    \caption{A comparison of several matrix multiplication algorithms between
+    Finch and Taco. On left, we use smaller matrices, ordered from small to big
+    dimension. Note that inner products necessarily requires $O(n^2)$ work and
+    taco's outer products format is dense. Finch can use a sparse outer products
+    format and thus has an asymptotic advantage that becomes evident as the
+    output dimensions grow. On right, we use only gustavson's algorithm and compare on larger matrices.}
 \end{figure}

### 2024-03-27 sections/eval.tex
shrink more

@@
 
-Push-Pull BFS in Finch:
+\begin{figure}
+    \begin{minipage}{0.5\linewidth}
     \begin{minted}{julia}
@@
                     _F[k] |= true
-                P[k] <<choose(0)>>= j #Only set the parent for this vertex
+                    P[k] <<choose(0)>>= j
                 end
@@
 
+    \end{minted}
+\end{minipage}%
+\begin{minipage}{0.5\linewidth}
+    \begin{minted}{julia}
     function finch_bfs_pull_kernel(_F, F, AT, V, P)
@@
                         if F[follow(j)] && AT[j, k]
-                        p[] <<choose(0)>>= j #Only set the parent for this vertex
+                            p[] <<choose(0)>>= j
                         end
@@
     \end{minted}
+\end{minipage}
+\end{figure}
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 mirrors most imperative languages with for-loops and control flow. Notable
-inclusions to the language include \mintinline{julia}{for},
+inclusions to the language at the level of statements includes \mintinline{julia}{for},
 \mintinline{julia}{let}, blocks of code with \mintinline{julia}{if}, and 
 the lifecycle functions that let us declare, freeze, and thaw tensors.
+%
+At the level of expressions, we support a wide array of scalar operations through literals and calls to functions whose properties and definitions are defined externally.
+%
+The expressions can also interact with indices and extents.
+% FIXME: ???
+Finally, as detailed in the previous section, tensors are defined externally via an interface that supports the declare, freeze, thaw, and unfurl functions.
+%
+The first three are supported directly in the syntax whereas the four will be introduced through evaluation of \mintinline{julia}{for}-loops loops and accesses,
+in the next section.
+%
 
-Finch uses looplets to lower \mintinline{julia}{for}-loops, though a sparse
-program with complicated looplets should produce output tensors with the same
-semantic value as a dense program with ordinary for-loops.  We include the
-semantics of how looplets are lowered as it affects the structure of the results
-and the asymptotic complexity of the program. Some applications view the
-structure of tensors as semantically meaningful.
+Our syntax is highly permissive: by allowin blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
+%
+Our breadth-first seach also uses an
+\mintinline{julia}{if} statement to avoid operating on vertices outside the
+frontier. 
+%
+We have found that \mintinline{julia}{if} is particularly useful in
+our model for masking to avoid work.
+%
+However, to compile these constructs efficient, we must restrict this syntax in several ways, as we have alluded to in the previous section via life cycles.
+%
+In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.
 
-The \mintinline{julia}{let} statement (marked by the \finchdefine AST node)
-allows us to define tensor variables and reuse their value in multiple
-subexpressions without passing the variable through a tensor. Although it is too
-complex in this work to propagate constants through tensors, we can propagate
-constants (including 0) through a define statement. This allows the compiler to
-build up complicated functions of known operators and infer annihilation
-properties through them.It also allows use to eliminate common subexpressions,
-or tensor accesses, by reusing a tensor value in multiple places (such as the
-symmetric spmv in Figure \ref{fib:spmv_programs}). As another example, in the
-following program we don't need to define sparsity and annihilator properties of
-$f(a + b) * (a - b)$, as we can derive this from the more basic operators, and
-the code only reads from $A$ and $B$ once.
-\begin{minted}{julia}
-    let a = A[i, j]
-        let b = B[i, j]
-            C[i, j] = (a + b) * (a - b)
-        end
-    end
-\end{minted}
+% Finch uses looplets to lower \mintinline{julia}{for}-loops, though a sparse
+% program with complicated looplets should produce output tensors with the same
+% semantic value as a dense program with ordinary for-loops.  We include the
+% semantics of how looplets are lowered as it affects the structure of the results
+% and the asymptotic complexity of the program. Some applications view the
+% structure of tensors as semantically meaningful.
 
-We allow the use of blocks of code to describe multiple operations. This allows
-us to support both multiple outputs and temporary tensors, seen most prominently
-in our implementation of gustavson's algorithm for sparse-sparse matrix multiply
-and breadth-first-search. Our breadth-first seach also uses an
-\mintinline{julia}{if} statement to avoid operating on vertices outside the
-frontier. We have found that \mintinline{julia}{if} is particularly useful in
-our model for masking to avoid work.
+% The \mintinline{julia}{let} statement (marked by the \finchdefine AST node)
+% allows us to define tensor variables and reuse their value in multiple
+% subexpressions without passing the variable through a tensor. Although it is too
+% complex in this work to propagate constants through tensors, we can propagate
+% constants (including 0) through a define statement. This allows the compiler to
+% build up complicated functions of known operators and infer annihilation
+% properties through them.It also allows use to eliminate common subexpressions,
+% or tensor accesses, by reusing a tensor value in multiple places (such as the
+% symmetric spmv in Figure \ref{fib:spmv_programs}). As another example, in the
+% following program we don't need to define sparsity and annihilator properties of
+% $f(a + b) * (a - b)$, as we can derive this from the more basic operators, and
+% the code only reads from $A$ and $B$ once.
+% \begin{minted}{julia}
+%     let a = A[i, j]
+%         let b = B[i, j]
+%             C[i, j] = (a + b) * (a - b)
+%         end
+%     end
+% \end{minted}
 
-Finally, we add some explicit functions to advance the lifecycle of a tensor.
-Note that the user would not typically write \mintinline{julia}{@freeze} or
-\mintinline{julia}{@thaw} explicitly as those can be inferred through a separate
-pass.
+% We allow the use of blocks of code to describe multiple operations. This allows
+% us to support both multiple outputs and temporary tensors, seen most prominently
+% in our implementation of gustavson's algorithm for sparse-sparse matrix multiply
+% and breadth-first-search. Our breadth-first seach also uses an
+% \mintinline{julia}{if} statement to avoid operating on vertices outside the
+% frontier. We have found that \mintinline{julia}{if} is particularly useful in
+% our model for masking to avoid work.
+
+% Finally, we add some explicit functions to advance the lifecycle of a tensor.
+% Note that the user would not typically write \mintinline{julia}{@freeze} or
+% \mintinline{julia}{@thaw} explicitly as those can be inferred through a separate
+% pass.
 
@@
 We present a sketch of a small-step operational semantics, showing how to execute a Finch program in a host language.
-We break our description of the semantics of Finch into three parts, those
-relating to lowering the language itself in Figure~\ref{fig:semantics_core},
-those pertaining to the lifecycle of tensors in
-Figure~\ref{fig:semantics_lifecycle}, and the semantics of the looplets in
-Figure~\ref{fig:semantics_looplets}. Note that in addition to the rules seen
-here, we also apply extensive symbolic simplification of the program whenever we
-can. In the implementaiton of the language we have heuristics for when to run a
-simplification pass, but for the purposes of understanding the semantics of the
-language we should understand simplification rules as occuring whenever possible
-before continuing to execute the program as normal.
+%
+We distinguish evaluation of the language by evolving compiler state with $\langle \rangle$ and evaluation in the host language via $\llangle\rrangle$.
+%
+For example, in Figure~\ref{fig:semantics_core}, we declare the core semantics of the language and the Define rule (\mintinline{julia}{let}) adds a definition to the compiler state whereas the Call rule passes evaluation off to the host.
+%
+In Figure~\ref{fig:semantics_looplets}, we detail the Looplet evaluation semantics, which details how acceses within a loop can be replaced with Loopets, which are progressively rewritten to host code.
+%
+Due to the use of rewrite rules, our semantics here make use of an evaluation context.
 
-Note that both Looplets and Finch are designed to be embedded into a larger
-language, and a separate state variable is introduced for this language.
+% We break our description of the semantics of Finch into three parts, those
+% relating to lowering the language itself in Figure~\ref{fig:semantics_core},
+% those pertaining to the lifecycle of tensors in
+% Figure~\ref{fig:semantics_lifecycle}, and the semantics of the looplets in
+% Figure~\ref{fig:semantics_looplets}. Note that in addition to the rules seen
+% here, we also apply extensive symbolic simplification of the program whenever we
+% can. In the implementation of the language we have heuristics for when to run a
+% simplification pass, but for the purposes of understanding the semantics of the
+% language we should understand simplification rules as occurring whenever possible
+% before continuing to execute the program as normal.
+
+% Note that both Looplets and Finch are designed to be embedded into a larger
+% language, and a separate state variable is introduced for this language.
 
@@
     \rightarrow val'}
-    \infer1[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
+    \hypo{var\not\in e}
+    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
     \end{prooftree}
@@
     \hypo{s = (e, t)}
+    \hypo{\finchmode(tns)\not\in e}
     \hypo{e(tns) = tns'}
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer4[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
+    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
     \end{prooftree}

### 2024-03-27 sections/eval.tex
shrink stuff more
Update on Overleaf.

@@
 Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
-Further explanation is available in Section 2.2 of  ~\cite{zhang2021gamma}.
+Further explanation is available in Section 2.2 of~\cite{zhang2021gamma}.
 
-Figures \ref{fig:spgemm_small} and \ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO.
+Figures~\ref{fig:spgemm_small} and~\ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO.
 Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals.
@@
 It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
+So, we were unable to obtain numbers for inner-product and TACO's outer-product for large matrices.
 In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.
 
-As depicted in Figure \ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product.
-For large sparse matrices, as shown in Figure \ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
+As depicted in Figure~\ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product.
+For large sparse matrices, as shown in Figure~\ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
 It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.
@@
 
-To evaluate the capabilities of Finch in graph applications, we its performance for two fundamantal graph applications: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-We used graphs in GraphBLAS~\cite{mattson2019lagraph}, which include road networks (SNAP/roadNet-CA and GAP/GAP-road) where node degrees are bounded and a diametr is very long and scale-free graphs (the others) where the distribution of node degrees follows a power-law distribution and a diameter is very short, direction-optimization~\cite{beamer2012direction} is crucial to achieve high-performance.
-Direction-optimized BFS uses push and pull traversal to efficiently explore graphs.
-Push traversal starts from the source node, pushing neighboring nodes into the queue.
-Pull traversal starts from the target node, exploring predecessors (parents).
-This approach uses early-break in pull traversal, which is used to terminate the traversal prematurely once the target node is encountered, to avoid unnecessary edge traversal.
+To evaluate the capabilities of Finch in graph applications, we assessed its performance in two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
+The graphs used were obtained from GraphBLAS~\cite{mattson2019lagraph}, including road networks (SNAP/roadNet-CA and GAP/GAP-road), characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
+Direction-optimization~\cite{beamer2012direction} is crucial for achieving high performance in such scenarios.
 
-Figure~\cite{graph_result} (a) \changwan{add (a) in fig} shows BFS speedup over Graphs.jl, a BFS implementation with push traversal of Julia.
-`finch\_push\_only` only uses push traversal, whereas `finch\_push\_pull` applies direction-optimization used in GraphBLAS.
-Finch supports push/pull traversal, early break, so it supports direction-optimization.
-Finch additionaly supports the different loop orders and multiple outputs, which also improves the performance.
-As shown in Figure~\cite{graph_result} (a), direction-optimization significantly improves the performance for scale-free graphs.
-\changwan{how much average speedup?} GraphBLAS sometimes better than Finch since GraphBLAS incorporates all hardwired optimizations.
-We believe we will achieve a better performance if we incorporate those optimizations.
+Direction-optimized BFS utilizes push and pull traversal techniques to efficiently explore graphs.
+Push traversal originates from the source node, pushing neighboring nodes into the queue, while pull traversal starts from the target node, exploring predecessors (parents).
+Incorporating early-break in pull traversal enables premature termination upon encountering the target node, thereby avoiding unnecessary edge traversal.
+
+Figure \ref{graph_result} (a) illustrates BFS speedup over Graphs.jl, a BFS implementation with push traversal in Julia.
+finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
+Finch's support for push/pull traversal and early break facilitates direction-optimization.
+Additionally, Finch's support for different loop orders and multiple outputs further enhances performance.
+Direction-optimization notably enhances performance for scale-free graphs (average speedup to be included).
+Although GraphBLAS sometimes outperforms Finch due to its incorporation of all hardwired optimizations, we anticipate achieving better performance with Finch upon incorporating similar optimizations.
+\changwan{mark captions (a) and (b)}
+
+Figure \ref{graph_result} (b) presents Bellman-Ford speedup (with path lengths and shortest-path tree) over Graphs.jl.
+Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.
+Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05).
+Note that we did not include GAP-road as it takes too long time with Graph.jl.
+\changwan{(self)dataset name changed?}
 
 \begin{figure}
-	\includegraphics[width=\linewidth]{bfs_speedup_over_graphs.jl.png}
-	\includegraphics[width=\linewidth]{bellmanford_speedup_over_graphs.jl.png}
+	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
+	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
     \caption{Performance of graph apps across various tools.}
@@
 \subsubsection{Evaluating Finch Logic}
-To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) multiple elementwise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined execution and even takes advantage of vector instructions which are not yet present in Finch.
+To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) and element-wise operations.
+Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
+For the element-wise operations, we also provide an unfused Finch implementation and the native Julia implementation for reference
 \begin{figure}

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 
-Our syntax is highly permissive: by allowin blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
+Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
 %
@@
 
+We include the sketch of our semantics to highlight several key design choices.
+%
+\teo{Make this true:}
+First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
+%
+Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
+%
+We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
+%
+These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
+%
+Second, loops with tensor accesses enter into a looplet system via the Unfurl rule.
+%
+In this system, repeated structures  (Run) and constant informa
+
 % We break our description of the semantics of Finch into three parts, those
@@
 
+% \begin{multline*}
+%         Modes:=update \mid read \mid undefined \mid defined
+% \end{multline*}
     \begin{prooftree}
@@
     \begin{prooftree}
-        \hypo{\langle head, s_1 \rangle \rightarrow s_2}
+    \hypo{\langle head,s_{1} \rangle \rightarrow s_2}
     \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 %
-In this system, repeated structures  (Run) and constant informa
+In this system, repeated structures (Run) and constant information (Switch)  are slowly uncovered as accesses are lowered.
+%
+In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 %
-In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth
+In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth.
+%
+Since these occur within the context, the simplification system has full access to what is know
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

--- 2024-03-27 sections/lang.tex
@@
 %
-In this system, repeated structures  (Run) and constant informa
+In this system, repeated structures (Run) and constant information (Switch)  are slowly uncovered as accesses are lowered in various points in the program.
+%
+In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, de
+%
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 %
-In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, de
+In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definiti
 %

### 2024-03-27 sections/lang.tex
Update on Overleaf.

--- 2024-03-27 sections/lang.tex
@@
 %
-In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, de
+In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
 %
+Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow (complex index access expressions involving indirection and arithemtic)
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 %
-Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow (complex index access expressions involving indirection and arithemtic)
+Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be
 

### 2024-03-27 sections/lang.tex
Update on Overleaf.

@@
 %
-Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be
+Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be represented via looplets
 

### 2024-03-27 sections/lang.tex

--- 2024-03-27 sections/lang.tex
@@
 %
-Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be
+Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be represented via accesses to tensors
 

### 2024-03-28 sections/compiler.tex
Update on Overleaf.

@@
     \begin{table}[h]
+        \scriptsize
         \centering
-        \begin{tabular}{|>{\raggedright\arraybackslash}p{0.2\linewidth}|>{\raggedright\arraybackslash}p{0.8\linewidth}|}
+        \begin{tabular}{|>{\raggedright\arraybackslash}m{0.25\linewidth}|>{\raggedright\arraybackslash}m{0.75\linewidth}|}
         \hline
@@
     \begin{table}[h]
+        \scriptsize
         \centering
-        \begin{tabular}{|>{\raggedright\arraybackslash}p{0.5\linewidth}|>{\raggedright\arraybackslash}p{0.5\linewidth}|}
+        \begin{tabular}{|>{\raggedright\arraybackslash}m{0.3\linewidth}|>{\raggedright\arraybackslash}m{0.4\linewidth}|}
         \hline
@@
 \subsection{Dimensionalization}
-    Looplets typically require the dimension of the loop extent to match the
-    dimensions of the tensor.
+    Looplets typically require the dimension of the loop extent to match the dimensions of the tensor. 
+    %
+    However, it is cumbersome to write the dimensions
+    in loop programs, and most tensor compilers have a means of specifying the dimensions automatically.
+    %
+    In many pure Einsum languages, determining dimensions
+    is not needed because any tensor dimensions that share an index are assumed to be the same~\cite{kjolstad_tensor_2017}.
+    %
+    Other languages such as Halide perform bounds inference wherein
+    known bounds are symbolically propagated to fill in unknown bounds, often from the desired output/input sizes to intermediates via some conservative approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
+    %
+    We refer to the process of discovering suitable
+    dimensions as \textbf{dimensionalization}. 
+    %
 
-    However, it is cumbersome to write the dimensions
-    in loop programs, and most tensor compilers have a means of specifying the
-    dimensions automatically. We refer to the process of discovering suitable
-    dimensions as \textbf{dimensionalization}. \willow{Teo, please explain
-    alternative approaches}>
+    %\willow{Teo, please explain
+    %alternative approaches}>
 
-    In finch, we use a straightforward algorithm to automatically determine
-    suitable dimensions for loops and declaration statements (output tensors).
-    Finch determines the dimension of a loop index i from all of the tensors
-    using i in an access, as well as the bounds in the loop itself, and operates
-    similarly for declarations.
-
-    Our algorithm uses the following principles, assuming dimension types form a
-    lattice:
+    In Finch, we use a straightforward algorithm to automatically determine suitable dimensions for loops and declaration statements (output tensors).
+    %
+    Finch determines the dimension of a loop index i from all of the tensors using i in an access, as well as the bounds in the loop itself, and operates similarly for declarations.
+    %   
+    Our algorithm uses the following principles, assuming dimension types form a lattice:
     \begin{enumerate}
@@
 \subsection{Life Cycles and Scopes}
-
+\teo{Can we maybe skip this because we cover it in the semantics?}
 \subsection{Simplification and Algebraic Transformations}
@@
 \subsection{Bounds Analysis}
-
 \subsection{Performance Warnings}
+\teo{I also think this maybe belongs in an appendix.}

### 2024-03-28 sections/eval.tex
Update on Overleaf.

@@
 Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-For the element-wise operations, we also provide an unfused Finch implementation and the native Julia implementation for reference
+To do this, we express each of these kernels as a single select, join, groupby statement in SQL.
+For the element-wise operations, we also provide an unfused Finch implementation to show the impact of fusion.
+For triangle counting, we use the same set of graph matrices as in Fig.
+\ref{graph_result}.
+For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
+Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
+A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
+
+Across all three of these kernels, we see that the high level interface for Finch provides a significant improvement over DuckDB ranging from $1.2x-28x$.
+Generally, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
+This matches with findings in the database literature
+
 \begin{figure}
@@
   \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=145pt, height=120pt]{figures/elementwise_speedup_over_duckdb.png}
+  \vspace{0pt} \includegraphics[width=145pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
 \end{tabular}

### 2024-03-28 sections/intro.tex
Update on Overleaf.

@@
 As an example, a programmer might explore different ways to intersect only the even integers of two lists (represented as sparse vectors with sorted indices).
-The control flow here is only useful if the first example differs from the second in that it actually selects only even indices as the two integer lists are merged:
+The control flow here is only useful if the first example differs from the next two in that it actually selects only even indices as the two integer lists are merged and different from the last in that it does not require another tensor:
 %The Finch compiler uses the structure of the data to generate efficient implementations of these programs.
@@
 
-\begin{minipage}{0.333\linewidth}
+\begin{minipage}{0.25\linewidth}
 \begin{minted}{julia}
@@
          if i % 2 == 0
-             c[i] = a[i] * b[i]
+             c[i]=a[i]*b[i]
          end
@@
 \end{minipage}%
-\begin{minipage}{0.333\linewidth}
+\begin{minipage}{0.25\linewidth}
 \begin{minted}{julia}
@@
 \end{minipage}%
-\begin{minipage}{0.333\linewidth}
+\begin{minipage}{0.25\linewidth}
 \begin{minted}{julia}
      for i = _
-         cp[i] = ap[i] * b[i]
+         cp[i] = a[i] * b[i]
      end
@@
          end
+     end
+\end{minted}
+\end{minipage}%
+\begin{minipage}{0.25\linewidth}
+\begin{minted}{julia}
+     for i = _
+        if i % 2 == 0
+            f[i] = 1
+        end
+     end
+     for i = _
+        c[i] = a[i] * b[i] * f[i]
      end
@@
 facilitating an expressive language that makes it easier to search the complex
-space of algorithms and data structures. 
+space of algorithms and data structures. Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
 %

### 2024-03-28 sections/lang.tex
Update on Overleaf.

--- 2024-03-27 sections/lang.tex
@@
 %
-Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language so that more complex control flow can be
+
+Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
+%
+In the next section, we will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
+%
+Also in the next section, We will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
+\teo{More next sections teaserse here}
 

### 2024-03-28 sections/eval.tex
Update on Overleaf.

@@
 Across all three of these kernels, we see that the high level interface for Finch provides a significant improvement over DuckDB ranging from $1.2x-28x$.
-Generally, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
-This matches with findings in the database literature
+For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
+This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.
+For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$
+
+DuckDB's poor join order selection which attempts to perform $A+B$ before intersecting with the indices of $C$.
 

### 2024-03-28 sections/eval.tex
Update on Overleaf.

@@
 This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.
-For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$
-
-DuckDB's poor join order selection which attempts to perform $A+B$ before intersecting with the indices of $C$.
+For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$ before handling the disjunction on the indices implied by $A+B$.
+DuckDB handles the latter first which makes the computation linear in the size of $A+B$ rather than $C$.
 
@@
 \begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=145pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=145pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
+  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
+  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
+  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
 \end{tabular}

### 2024-03-28 sections/eval.tex
Update on Overleaf.
update

--- 2024-03-28 sections/eval.tex
@@
 
-\help{The main points of this section are:}
+%\help{The main points of this section are:}
 
-\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}
+%\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}
 
-\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}
+%\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}
 
-\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}
+%\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}
 
-\help{4.
+%\help{4.
 Take a look at the figure and point out that Finch is competitive with TACO on inner and gustavson, and that for outer product we see what we would expect from TACO: i.e.
@@
 
-\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.
+%\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.
 
-\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
+%\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
 \begin{figure}
@@
 \subsubsection{Graph Analytics}
-\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
+%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
 choose), and that Finch can do early break, and also the different loop orders and multiple outputs.

### 2024-03-28 sections/eval.tex

@@
 Across all three of these kernels, we see that the high level interface for Finch provides a significant improvement over DuckDB ranging from $1.2x-28x$.
-Generally, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
-This matches with findings in the database literature
+For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
+This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.
+For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$ before handling the disjunction on the indices implied by $A+B$.
+DuckDB handles the latter first which makes the computation linear in the size of $A+B$ rather than $C$.
 

### 2024-03-28 figures/export_looplet_figures.sh
make it nicer

@@ -17,3 +17,4 @@ file=LevelsVsFibers # File name excluding extension
 
 file=Structures # File name excluding extension
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file.png" "$file.drawio"
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 1 --output "$file-examples.png" "$file.drawio"

### 2024-03-28 sections/compiler.tex

@@
     %
-    In many pure Einsum languages, determining dimensions
+    In many pure Einsum languages like TACO, determining dimensions
     is not needed because any tensor dimensions that share an index are assumed to be the same~\cite{kjolstad_tensor_2017}.
     %
-    Other languages such as Halide perform bounds inference wherein
+    Other languages, such as Halide, perform bounds inference where
     known bounds are symbolically propagated to fill in unknown bounds, often from the desired output/input sizes to intermediates via some conservative approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
@@
     %
-
-    %\willow{Teo, please explain
-    %alternative approaches}>
 
@@
 \teo{Can we maybe skip this because we cover it in the semantics?}
+\willow{we should at least say here that we have a pass that adds freeze and thaw appropriately}
 \subsection{Simplification and Algebraic Transformations}

### 2024-03-29 sections/intro.tex
make it nicer
add jaeyeon

@@
 \begin{figure}
-	\includegraphics[width=\linewidth]{example_structures.png}
-    \caption{some example structures that might be nice to put into a figure}
+	\includegraphics[width=\linewidth]{Structures-examples.png}
+    \caption{A few examples of matrix structures arising in practice}
 \end{figure}
@@
 \begin{enumerate}
-\item
-
-More complex array structures than ever before. We are the first to extend level-by-level hierarchical descriptions to capture
-banded, triangular, run-length-encoded, or sparse datasets, and any combination
-thereof. We have chosen a set of level formats that completely captures all combinations of relevant
-structural properties (zeros, repeated values, and/or blocks).
+\item More complex array structures than ever before.
+We are the first to extend level-by-level hierarchical descriptions to capture banded, triangular, run-length-encoded, or sparse datasets, and any combination thereof.
+%
+We have chosen a set of level formats that completely captures all combinations of relevant structural properties (zeros, repeated values, and/or blocks).
+%
 Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level.
-
-\item A rich structured array programming language with for-loops
-and complex control flow constructs at the same level of productivity
-of dense arrays. 
 %
-To our knowledge, the Finch programming language is the first 
-to support if-conditions, early breaks, and multiple left hand sides over
-structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands. 
+\item A rich structured array programming language with for-loops and complex control flow constructs at the same level of productivity of dense arrays. 
 %
-\item A compiler that specializes programs to data structures automatically,
-facilitating an expressive language that makes it easier to search the complex
-space of algorithms and data structures. Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
+To our knowledge, the Finch programming language is the first to support if-conditions, early breaks, and multiple left hand sides over structured data, as well as complex accesses such as affine indexing or scatter/gather of sparse or structured operands.
+%
+\item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
+Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
+%
+\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs we implement in this work.
+%
+For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
 %

### 2024-03-29 figures/export_looplet_figures.sh
yay

@@ -18,3 +18,6 @@ file=LevelsVsFibers # File name excluding extension
 file=Structures # File name excluding extension
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file.png" "$file.drawio"
 /Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 1 --output "$file-examples.png" "$file.drawio"
+
+file=Spmv_annotated # File name excluding extension
+/Applications/draw.io.app/Contents/MacOS/draw.io --export --page-index 0 --output "$file.png" "$file.drawio"

### 2024-03-29 sections/tensors.tex
reformat

@@
 recommendations for additional level formats to be implemented and explain how
-users can also add their own custom formats. Note that all of the levels below store their shape and a sublevel.
+users can also add their own custom formats.
 
-\begin{enumerate}
-\item[Dense]
+A few challenges arise when implementing these levels. First, it is difficult to merge duplicate runs.
+%
+Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
+%
+Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
+%
+Instead, we merge the duplicates during the freeze phase.
+%
+We $freeze$ the sublevel, $declare$ a separate sublevel as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer.
+
+Second, we must be able to handle the case where a sublevel is not stored because a parent level is sparse.
+%
+Therefore, all of Finch's sparse formats use a dirty bit during writing to determine whether the sublevel has been modified from it's default fill value and thus, whether it needs to be stored.
+
+\begin{itemize}
+\item \textbf{Dense}
     The dense format is the simplest format, and simply maps $fiber(l, p)[i]
@@
     of its simplicity, freezing and thawing the level are no-ops.
-\item[DenseRLE]
-    The DenseRLE format is used to represent runs of repeated values. Notice
-    that a challenge arises when trying to write to a level with runs: it is
-    difficult to merge duplicate runs. Such a scenario might arise when merging
-    runs of subfibers of length 3, representing colors in an image.  Ideally, we
-    would be able to detect duplicate subfibers and merge them on the fly, but
-    we cannot determine which subfibers are equal because we cannot read them
-    while the sublevel is in update-only mode. Instead, we merge the duplicates
-    during the freeze phase. We $freeze$ the main sublevel, $declare$ a separate
-    sublevel as a buffer to store the deduplicated subfibers, and we can then
-    compare each of the subfibers in the main level, copying the deduplicated
-    subfibers into the buffer:
-    \begin{enumerate}
-        \item[$right$] A vector of indices corresponding to the end of each run in the level
-        \item[$ptr$] A vector of delimiters such that $right[ptr[p]:ptr[p+1] - 1]$ is the set of delimiters in the subfiber at position $p$.
-        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
-    \end{enumerate}
-\item[SparseList]
-    The sparse list format is the simplest sparse format, and is used to
-    construct the popular CSR, CSC, DCSR, DCSC, and CSF formats. The sparse list
-    format consists of the following fields:
-    \begin{enumerate}
-        \item[$idx$] A vector of nonzero indices in the level
-        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
-    \end{enumerate}
-    All of Finch's sparse formats use a dirty bit during writing to determine
-    whether the sublevel has been modified from it's default fill value and
-    thus, whether it needs to be stored.
-\item[SparsePinpoint]
-    Because the SparsePinpoint format will only ever store one nonzero in each subfiber,
-    we do not need the $ptr$ field.
-    \begin{enumerate}
-        \item[$idx$] A vector of nonzero indices in the level, one for each subfiber
-    \end{enumerate}
-\item[SparseRLE]
+\item \textbf{DenseRLE}
+    The DenseRLE format is used to represent runs of repeated values.
+    %
+    Stores two vectors, $right$ and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
+    %
+    Also stores a duplicate sublevel, $buf$, for deduplication.
+\item \textbf{SparseList}
+    The sparse list format is the simplest sparse format, and is used to construct the popular CSR, CSC, DCSR, DCSC, and CSF formats.
+    %
+    Stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
+\item \textbf{SparsePinpoint}
+    Because the SparsePinpoint format will only ever store one nonzero in each subfiber, we do not need the $ptr$ field.
+    %
+    Stores a vector $idx$ such that $idx[p]$ is the nonzero index in the subfiber at position $p$.
+\item \textbf{SparseRLE}
     Similar to the DenseRLE level, but because the runs are stored on top of a background of fill values, we must also store the start of each run.
-    \begin{enumerate}
-        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
-        \item[$right$] A vector of indices corresponding to the end of each run in the level
-        \item[$ptr$] A vector of delimiters such that $left[ptr[p]:ptr[p+1] - 1]$ and $right[ptr[p]:ptr[p+1] - 1]$ are the set of delimiters in the subfiber at position $p$.
-        \item[$buf$] A duplicate sublevel, used to store deduplicated subfibers during $freeze$.
-    \end{enumerate}
-\item[SparseInterval]
-    Similar to the SparseRLE level, but $ptr$ is redundant because we only store
-    one run. We also don't bother with deduplication here as we cannot store any
-    intermediate results with duplicates.
-    \begin{enumerate}
-        \item[$left$] A vector of indices corresponding to the beginning of each run in the level
-        \item[$right$] A vector of indices corresponding to the end of each run in the level
-    \end{enumerate}
-\item[SparseVBL]
-    The SparseVBL format is used to represent blocked data. The format consists
-    of the following fields:
-    \begin{enumerate}
-        \item[$idx$] A vector of indices at the end of each nonzero block in the level
-        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of block terminals in the subfiber at position $p$.
-        \item[$ofs$] A vector of delimiters such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ in subfiber $p$.
-    \end{enumerate}
-\item[SparseBand]
+    %
+    Stores three vectors, $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
+    %
+    Also stores a duplicate sublevel, $buf$, for deduplication.
+\item \textbf{SparseInterval}
+    Similar to the SparseRLE level, but $ptr$ is redundant because we only store one run.
+    %
+    We also don't bother with deduplication here as we cannot store any intermediate results with duplicates.
+    %
+    Stores two vectors, $left$ and $right$, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively.
+\item \textbf{SparseVBL}
+    The SparseVBL format is used to represent blocked data.
+    %
+    Stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$.
+\item \textbf{SparseBand}
     Like SparseVBL, but only one block is stored in each subfiber.
-        \item[$idx$] The index at the end of the block in the level
-        \item[$ofs$] A vector of delimiters such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block in subfiber $p$.
-\end{enumerate}
+    %
+    Stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$.
+\end{itemize}
+
+ Note that all of the levels below store their shape and a sublevel.
 

### 2024-03-29 sections/tensors.tex
okay?

@@
 
-\begin{itemize}
-\item \textbf{Dense}
-    The dense format is the simplest format, and simply maps $fiber(l, p)[i]
-    \rightarrow fiber(l.lvl, p * l.shape + i)$. This format is used to store dense data,
-    and is often a convenient format for the root level of a tensor. Because
-    of its simplicity, freezing and thawing the level are no-ops.
-\item \textbf{DenseRLE}
-    The DenseRLE format is used to represent runs of repeated values.
+\begin{table}[ht]
+    \centering
+    \begin{tabular}{p{14cm}}
+    \hline
+    \textbf{Sequential Reading and Updating} \\
+    \hline
+    \textbf{Dense}:
+    The dense format is the simplest format, mapping $fiber(l, p)[i] \rightarrow fiber(l.lvl, p \times l.shape + i)$.
     %
-    Stores two vectors, $right$ and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
+    This format is used to store dense data and is often a convenient format for the root level of a tensor.
     %
-    Also stores a duplicate sublevel, $buf$, for deduplication.
-\item \textbf{SparseList}
-    The sparse list format is the simplest sparse format, and is used to construct the popular CSR, CSC, DCSR, DCSC, and CSF formats.
+    Due to its simplicity, freezing and thawing the level are no-ops. \\
+    \textbf{DenseRLE}:
+    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
     %
-    Stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
-\item \textbf{SparsePinpoint}
-    Because the SparsePinpoint format will only ever store one nonzero in each subfiber, we do not need the $ptr$ field.
+    It also stores a duplicate sublevel, $buf$, for deduplication. \\
+    \textbf{SparseList}:
+    The simplest sparse format, used to construct popular formats like CSR, CSC, DCSR, DCSC, and CSF.
     %
-    Stores a vector $idx$ such that $idx[p]$ is the nonzero index in the subfiber at position $p$.
-\item \textbf{SparseRLE}
-    Similar to the DenseRLE level, but because the runs are stored on top of a background of fill values, we must also store the start of each run.
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
+    \textbf{SparsePinpoint}:
+    Similar to SparseList, but only one nonzero in each subfiber, eliminating the need for the $ptr$ field.
     %
-    Stores three vectors, $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
+    It stores a vector $idx$, such that $idx[p]$ is the nonzero index in the subfiber at position $p$. \\
+    \textbf{SparseRLE}:
+    Similar to DenseRLE level, but because runs are sparse, we must also store the start of each run.
     %
-    Also stores a duplicate sublevel, $buf$, for deduplication.
-\item \textbf{SparseInterval}
-    Similar to the SparseRLE level, but $ptr$ is redundant because we only store one run.
+    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
     %
-    We also don't bother with deduplication here as we cannot store any intermediate results with duplicates.
+    It also stores a duplicate sublevel, $buf$, for deduplication. \\
+    \textbf{SparseInterval}:
+    Similar to SparseRLE, but only stores one run per subfiber, eliminating the need for the $ptr$ field.
     %
-    Stores two vectors, $left$ and $right$, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively.
-\item \textbf{SparseVBL}
-    The SparseVBL format is used to represent blocked data.
+    This level does not deduplicate as it cannot store intermediate results with more than one run.
     %
-    Stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$.
-\item \textbf{SparseBand}
-    Like SparseVBL, but only one block is stored in each subfiber.
+    It stores two vectors, such that the run in subfiber $p$ begins and ends at $left[p]$ and $right[p]$ respectively. \\
+    \textbf{SparseVBL}: Used to represent blocked data.
+It stores three vectors, $idx$, $ptr$, and $ofs$, such that $ofs[ptr[p] + q]:ofs[ptr[p] + q + 1] - 1$ are the subpositions of block $q$ ending at index $idx[ptr[p] + q]$ in the subfiber at position $p$.
+\\
+    \textbf{SparseBand}: Similar to SparseVBL, but stores only one block per subfiber, eliminating the need for the $ptr$ field.
+It stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$.
+\\
+    \hline
+    \textbf{Random Access Reading and Updating} \\
+    \hline
+    \textbf{SparseHash}:
+    The sparse hash format uses a hash table to store the locations of nonzeros, and sorts the unique indices for iteration during the freeze phase.
     %
-    Stores two vectors $idx$ and $ofs$, such that $ofs[p]:ofs[p + 1] - 1$ are the subpositions of the block ending at $idx[p]$ in subfiber $p$.
-\end{itemize}
+    This allows for efficient random access, but not incremental construction, as the freeze phase runs in time proportional to the number of nonzeros in the entire level.
+    %
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
+    %
+    Also stores a hash table $tbl$ for construction and random access in the level. \\
+    \textbf{SparseBytemap}
+    The SparseBytemap format uses a bytemap to store which locations have been written to.
+    %
+    Unlike the SparseHash format, the bytemap assembles the entire space of possible subfibers.
+    %
+    This accelerates random access in the format, but requires a high memory overhead.
+    %
+    Because we don't want to reallocate all of the memory in each iteration, the declaration of this format instead re-assembles only the dirty locations in the tensor.
+    %
+    This format is analogous to the default workspace format used by TACO.
+    %
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
+    %
+    These vectors are used to collect dirty locations.
+    %
+    It also stores $tbl$, a dense array of booleans such that $tbl[shape * p + i]$ is true when there is a nonzero at index $i$ in the subfiber at position $p$. \\
+    \hline
+    \end{tabular}
+    \caption{Description of Data Storage Formats}
+    \label{table:formats}
+\end{table}
 

### 2024-03-29 sections/tensors.tex
more fixing

@@
 
+\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
+
+Our simplified view of a level is enabled by our use of looplets to represent
+the structure within each fiber. In fact, our level interface requires only 5
+highly general operations, described below.
+
+The first three of these functions, declare, freeze, and thaw, have to do with
+managing when tensors can be assumed mutable or immutable. As we use Looplets to
+represent iteration over a tensor, we must restrict the mutability of tensors
+while we iterate over them. For example, if a tensor declares it has a constant
+region from $i = 2:5$, but some other part of the computation modifies the
+tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
+write correct Looplet code if we can assume that the tensor is immutable while
+we are reading from it. Thus, we introduce the notion that a tensor can be in
+read-only mode or update-only mode.  In read-only mode, the tensor may only
+appear in the right-hand side of assignments. In update-only mode, the tensor
+may only appear in the left-hand side of an assignment, either being overwritten
+or incremented by some operator.  We can switch between these modes using freeze
+and thaw functions. The declare function is used to allocate a tensor,
+initialize it to some specified size and value, and leave it in update-only
+mode.
+
+The unfurl function is used to manage iteration over a subfiber. At the point
+when it comes time to iterate over a tensor, be in on the left or right hand
+side of an assignment, we call the unfurl function to precompute whatever state
+and datastructures are necessary to return a looplet nest that would iterate
+over that level of a tensor. The unfurl function is called directly before
+iterating over the corresponding loop, so it has access to any state variables introduced
+by freezing or thawing the tensor.
+
+\begin{figure}[htbp]
+    \scriptsize
+    \raggedright
+\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
+of size $dims$ and an initial value of $init$. Requires the level to be in
+read-only mode.
+\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
+level for reading. Requires the level to be in update-only mode.
+\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
+to be in read-only mode.
+\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
+$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
+returns a looplet nest over the values in the read-only fiber.  When $mode =
+\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
+fiber. Often, skipping over mutable locations allows the level to know which
+locations must be stored. Additionally, a dirty bit may be used to communicate
+whether the mutable subfiber has been written to, which allows the parent fiber
+to know whether the subfiber must be stored explicitly.
+\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
+level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
+only ever called on unassembled positions, but some levels (such as dense levels
+or bytemaps) may support reassembly.
+\caption{The five functions that define a level.}
+\end{figure}
+
+Our view of a level as a fiber allocator implies an allocation function
+$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
+$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
+function, instead relying on initialization to reset the fiber if it needs to be
+reused. While all of the previous functions are used to manage the lifecycle and
+iteration over a general tensor, the assemble function is quite specific to the
+level abstraction, and the notion of positions within sublevels. Note: it was an
+intentional choice to hold the parent level responsible for managing the
+data of the sublevels, which positions they allocate, etc. This allows the parent
+level to reuse allocation logic from internal index datastructures. For example,
+a sparse level might use a list of indices to store which nonzeros are present,
+and when it comes time to resize that list, it could also call assemble to resize the
+sublevel, reducing the number of branches in the code. The assemble function
+lends itself particularly to a "vector doubling" allocation approach, which we
+have found to be effective and flexible when managing the allocation
+of sparse left hand sides. This benefit is made clear in our evaluation section,
+where we see that prior systems like TACO do not support all possible loop
+orderings and format combinations for sparse matrix multiply because they do
+not have a flexible enough allocation strategy, instead using a two-phase approach
+which requires computing a complicated closed-form kernel to iterate over the
+data twice to determine the number of required output nonzeros.
+
 \subsection{The 8 Key Level Structures}
-    \begin{wrapfigure}{R}{.35\textwidth}
+
+    The main benefits of specializing to structure come from the following properties of the data:
+    \begin{itemize}
+        \item \textbf{Sparsity} Sparse data is data that is mostly zero, or some other
+        fill value. When we specialize on this data, we can use annihilation ($x
+        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
+        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
+        redundant work.
+
+        \item \textbf{Blocks} Blocked data is a subset of sparse data where the nonzeros
+        are clustered and occur adjacent to one another. This provides us with
+        two opportunities: We can avoid storing the locations of the nonzeros
+        individually, and we can use more efficient randomly accessible
+        iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.
+
+        \item \textbf{Runs} Runs of repeated values may occur in dense or sparse code,
+        cutting down on storage and allowing us to use integration rules such as 
+        \mintinline{julia}{for i = 1:n; s += x end} $\rightarrow$
+        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops \cite{donenfeld_unified_2022,ahrens_looplets_2023}.
+
+        \item \textbf{Singletons} When we have only one non-fill region in sparse data,
+        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
+    \end{itemize}
+
+    \begin{wrapfigure}{r}{.35\textwidth}
         \centering
@@
     \end{wrapfigure}
-    The main benefits of specializing to structure come from the following properties of the data:
-    \begin{enumerate}
-        \item[Sparsity] Sparse data is data that is mostly zero, or some other
-        fill value. When we specialize on this data, we can use annihilation ($x
-        * 0 = 0$), identity ($x * 1 = 1$), or other constant propagation
-        properties ($ifelse(false, x, y) = y$) to simplify the computation and avoid
-        redundant work.
 
-        \item[Blocks] Blocked data is a subset of sparse data where the nonzeros
-        are clustered and occur adjacent to one another. This provides us with
-        two opportunities: We can avoid storing the locations of the nonzeros
-        individually, and we can use more efficient randomly accessible
-        iterators within the block. \cite{im_optimizing_2001, vuduc_performance_2002, ahrens_looplets_2023}.
-
-        \item[Runs] Runs of repeated values may occur in dense or sparse code,
-        cutting down on storage and allowing us to use integration rules such as 
-        \mintinline{julia}{for i = 1:n; s += x end} $\rightarrow$
-        \mintinline{julia}{s += n * x} or code motion to lift operations out of loops \cite{donenfeld_unified_2022,ahrens_looplets_2023}.
-
-        \item[Singletons] When we have only one non-fill region in sparse data,
-        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
-    \end{enumerate}
-
-    In Finch, we have identified 8 key level structures that can represent all
-    of the relevant combinations of these properties, summarized in Table
-    \ref{tab:TypesOfStructure}. We examine each structure in turn, describing
-    some of the key properties and potential use cases of each. In this sense,
-    the structures we consider are exhaustive. We can represent a wide variety
-    of hierarchical tensor structures by combining these level structures in a
-    tree, as shown in Figure~\ref{fig:structuraldiversity}.
+    We consider all combinations of these structural properties in Table
+    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
+    the the 8 resulting situations. While it is impossible to write code which
+    precisely addresses every possible structure, the combination of our level
+    formats can express a wide variety of hierarchical structures to a
+    sufficient granularity that we can generate efficient code for them.  For
+    example, though banded arrays are a superset of ragged arrays, representing
+    a ragged array using a banded format does not add much overhead.  The
+    structures we consider are exhaustive in the sense that they address all 
+    combinations of sparsity, blocks, runs, and singletons in each level. We can
+    represent a wide variety of hierarchical tensor structures by combining
+    these level structures in a tree, as shown in
+    Figure~\ref{fig:structuraldiversity}.
 
@@
 
-\subsection{Tensor Lifecycle, Declare, Freeze, Thaw, Unfurl}
-
-Our simplified view of a level is enabled by our use of looplets to represent
-the structure within each fiber. In fact, our level interface requires only 5
-highly general operations, described below.
-
-The first three of these functions, declare, freeze, and thaw, have to do with
-managing when tensors can be assumed mutable or immutable. As we use Looplets to
-represent iteration over a tensor, we must restrict the mutability of tensors
-while we iterate over them. For example, if a tensor declares it has a constant
-region from $i = 2:5$, but some other part of the computation modifies the
-tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
-write correct Looplet code if we can assume that the tensor is immutable while
-we are reading from it. Thus, we introduce the notion that a tensor can be in
-read-only mode or update-only mode.  In read-only mode, the tensor may only
-appear in the right-hand side of assignments. In update-only mode, the tensor
-may only appear in the left-hand side of an assignment, either being overwritten
-or incremented by some operator.  We can switch between these modes using freeze
-and thaw functions. The declare function is used to allocate a tensor,
-initialize it to some specified size and value, and leave it in update-only
-mode.
-
-The unfurl function is used to manage iteration over a subfiber. At the point
-when it comes time to iterate over a tensor, be in on the left or right hand
-side of an assignment, we call the unfurl function to precompute whatever state
-and datastructures are necessary to return a looplet nest that would iterate
-over that level of a tensor. The unfurl function is called directly before
-iterating over the corresponding loop, so it has access to any state variables introduced
-by freezing or thawing the tensor.
-
-Our view of a level as a fiber allocator implies an allocation function
-$assemble(tns, pos_{start}:pos_{stop})$, which allocates fibers at positions
-$pos_{start}:pos_{stop}$ in the level. We don't specify a de-allocation
-function, instead relying on initialization to reset the fiber if it needs to be
-reused. While all of the previous functions are used to manage the lifecycle and
-iteration over a general tensor, the assemble function is quite specific to the
-level abstraction, and the notion of positions within sublevels. Note: it was an
-intentional choice to hold the parent level responsible for managing the
-data of the sublevels, which positions they allocate, etc. This allows the parent
-level to reuse allocation logic from internal index datastructures. For example,
-a sparse level might use a list of indices to store which nonzeros are present,
-and when it comes time to resize that list, it could also call assemble to resize the
-sublevel, reducing the number of branches in the code. The assemble function
-lends itself particularly to a "vector doubling" allocation approach, which we
-have found to be effective and flexible when managing the allocation
-of sparse left hand sides. This benefit is made clear in our evaluation section,
-where we see that prior systems like TACO do not support all possible loop
-orderings and format combinations for sparse matrix multiply because they do
-not have a flexible enough allocation strategy, instead using a two-phase approach
-which requires computing a complicated closed-form kernel to iterate over the
-data twice to determine the number of required output nonzeros.
-
-\begin{figure}
-    \raggedright
-\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
-of size $dims$ and an initial value of $init$. Requires the level to be in
-read-only mode.
-\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
-level for reading. Requires the level to be in update-only mode.
-\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
-to be in read-only mode.
-\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
-$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
-returns a looplet nest over the values in the read-only fiber.  When $mode =
-\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
-fiber. Often, skipping over mutable locations allows the level to know which
-locations must be stored. Additionally, a dirty bit may be used to communicate
-whether the mutable subfiber has been written to, which allows the parent fiber
-to know whether the subfiber must be stored explicitly.
-\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
-level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
-only ever called on unassembled positions, but some levels (such as dense levels
-or bytemaps) may support reassembly.
-\caption{The five functions that define a level.}
-\end{figure}
-
-\subsection{Level Formats}
     Having described the 8 basic level structures and the functions that define a
@@
 
-A few challenges arise when implementing these levels. First, it is difficult to merge duplicate runs.
-%
-Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
-%
-Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
-%
-Instead, we merge the duplicates during the freeze phase.
-%
-We $freeze$ the sublevel, $declare$ a separate sublevel as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer.
+A few challenges arise when implementing these levels. First,
 
-Second, we must be able to handle the case where a sublevel is not stored because a parent level is sparse.
-%
-Therefore, all of Finch's sparse formats use a dirty bit during writing to determine whether the sublevel has been modified from it's default fill value and thus, whether it needs to be stored.
+Second,
+
+Note that all of the levels below store their shape and a sublevel.
+
+\help{we need also to talk about pattern and element}
+
+\subsection{Flexible Level Formats}
+The levels we described in the previous section are all oriented towards bulk,
+sequential creation of formats. However, when users want to be able to write out
+of order (which is a common requirement arising from loop order or from the
+problem itself, it occurs in our spgemm algorithms and our histogram example in
+the evaluation section), we must use more complicated datastructures like hash
+tables and trees to support the random access. Because these datastructures are
+more complex and have a higher implementation burden and performance overhead,
+we only support random access construction of sparse or dense structures.  We
+can use these two more general structures as intermediates to convert to our
+more specialized structures later.
+
+As future work, we also recommend the implementation of a tree-based format, as
+this would enable incremental construction. Currently, the cost of freezing and
+thawing is usually proportional to the number of stored fibers, so an operation
+that sets a single value in a tensor would be prohibitively expensive (i.e.
+$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
+intermediate randomly accessible structure, because (unlike blocks or
+singletons), runs have asymptotic benefits. However, these cases were not
+necessary for our case studies.
+
+\subsection{Scalars}
+
+\subsubsection{Sparse Scalars}
+\subsubsection{Early Break Scalars}
 
@@
     \centering
+    \scriptsize
     \begin{tabular}{p{14cm}}
@@
     %
-    It also stores a duplicate sublevel, $buf$, for deduplication. \\
+    A challenge arises for this level: it is difficult to merge duplicate runs.
+    %
+    Such a scenario might arise when merging runs of subfibers of length 3, representing colors in an image.
+    %
+    Ideally, we would be able to detect duplicate subfibers and merge them on the fly, but we cannot determine which subfibers are equal because we cannot read them while the sublevel is in update-only mode.
+    %
+    Instead, we merge the duplicates during the freeze phase.
+    %
+    We $freeze$ the sublevel, $declare$ a separate sublevel $buf$ as a buffer to store the deduplicated subfibers, and then compare each of the subfibers in the main level, copying the deduplicated subfibers into the buffer.
+\\
+    %
+    %It also stores a duplicate sublevel, $buf$, for deduplication. \\
     \textbf{SparseList}:
@@
     %
-    It also stores a duplicate sublevel, $buf$, for deduplication. \\
+    Like DenseRLE, it also stores a duplicate sublevel, $buf$, for deduplication. \\
     \textbf{SparseInterval}:
@@
     \end{tabular}
-    \caption{Description of Data Storage Formats}
+    \caption{The main level formats supported by Finch. Note that all non-leaf
+    levels store a the dimension of the subfibers and a child level. Since we
+    must be able to handle the case where a sublevel is not stored because a
+    parent level is sparse, all of Finch's sparse formats use a dirty bit during
+    writing to determine whether the sublevel has been modified from it's
+    default fill value and thus, whether it needs to be stored.}
     \label{table:formats}
 \end{table}
-
- Note that all of the levels below store their shape and a sublevel.
-
-\help{we need also to talk about pattern and element}
-
-\subsection{Flexible Level Formats}
-The levels we described in the previous section are all oriented towards bulk,
-sequential creation of formats. However, when users want to be able to write out
-of order (which is a common requirement arising from loop order or from the
-problem itself, it occurs in our spgemm algorithms and our histogram example in
-the evaluation section), we must use more complicated datastructures like hash
-tables and trees to support the random access. Because these datastructures are
-more complex and have a higher implementation burden and performance overhead,
-we only support random access construction of sparse or dense structures.  We
-can use these two more general structures as intermediates to convert to our
-more specialized structures later.
-
-\begin{enumerate}
-\item[SparseHash]
-    The sparse hash format uses a hash table to store the locations of nonzeros,
-    and sorts the unique indices for iteration during the freeze phase. This
-    allows for efficient random access, but not incremental construction, as the
-    freeze phase runs in time proportional to the number of nonzeros in the
-    entire level.
-    \begin{enumerate}
-        \item[$idx$] A vector of nonzero indices in the level
-        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
-        \item[$tbl$] A hash table used for construction of the level.
-    \end{enumerate}
-\item[SparseBytemap]
-    The SparseBytemap format uses a bytemap to store which locations have been
-    written to. Unlike the SparseHash format, the bytemap assembles the entire
-    space of possible subfibers. This accelerates random access in the format,
-    but requires a high memory overhead. Because we don't want to reallocate all
-    of the memory in each iteration, the declaration of this format instead
-    re-assembles only the dirty locations in the tensor. This format is
-    analogous to the default workspace format used by TACO.
-    \begin{enumerate}
-        \item[$idx$] A vector of nonzero indices in the level
-        \item[$ptr$] A vector of delimiters such that $idx[ptr[p]:ptr[p+1] - 1]$ is the set of nonzero indices in the subfiber at position $p$.
-        \item[$tbl$] A dense array of booleans used to mark dirty locations in the tensor.
-    \end{enumerate}
-\end{enumerate}
-
-As future work, we also recommend the implementation of a tree-based format, as
-this would enable incremental construction. Currently, the cost of freezing and
-thawing is usually proportional to the number of stored fibers, so an operation
-that sets a single value in a tensor would be prohibitively expensive (i.e.
-$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
-intermediate randomly accessible structure, because (unlike blocks or
-singletons), runs have asymptotic benefits. However, these cases were not
-necessary for our case studies.
-
-\subsection{Scalars}
-
-\subsubsection{Sparse Scalars}
-\subsubsection{Early Break Scalars}

### 2024-03-29 sections/tensors.tex
cool

@@
     formats can express a wide variety of hierarchical structures to a
-    sufficient granularity that we can generate efficient code for them.  For
-    example, though banded arrays are a superset of ragged arrays, representing
-    a ragged array using a banded format does not add much overhead.  The
-    structures we consider are exhaustive in the sense that they address all 
-    combinations of sparsity, blocks, runs, and singletons in each level. We can
-    represent a wide variety of hierarchical tensor structures by combining
-    these level structures in a tree, as shown in
+    sufficient granularity that we can generate code which utilizes the four
+    properties.  For example, though banded arrays are a superset of ragged
+    arrays, representing a ragged array using a banded format does not add much
+    overhead.  The structures we consider are exhaustive in the sense that they
+    address all combinations of sparsity, blocks, runs, and singletons in each
+    level. We can represent a wide variety of hierarchical tensor structures by
+    combining these level structures in a tree, as shown in
     Figure~\ref{fig:structuraldiversity}.
@@
 
-    Having described the 8 basic level structures and the functions that define a
-level format, we can describe the concrete level formats supported by Finch,
-some of the challenges in their implementation, and how we overcome them. Note
-that these formats are not meant to be exhaustive; we will later give
-recommendations for additional level formats to be implemented and explain how
-users can also add their own custom formats.
-
-A few challenges arise when implementing these levels. First,
-
-Second,
-
-Note that all of the levels below store their shape and a sublevel.
-
-\help{we need also to talk about pattern and element}
-
-\subsection{Flexible Level Formats}
-The levels we described in the previous section are all oriented towards bulk,
+\subsection{Randomly Accessible Temporary Levels}
+To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
 sequential creation of formats. However, when users want to be able to write out
@@
 necessary for our case studies.
+
+\help{we need also to talk about pattern and element}
+
+\help{this section fails to address freeze or thaw and how it relates to the concrete levels}
+
+\subsection{Leaf Levels}
+
+The leaf level stores the actual entries of the tensor. In most cases, it is
+sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
+values are the same, an additional optimization can be made by storing the
+identical value only once. In this work, we introduce the concept of a \textbf{pattern}
+level to handle the boolean case, returning $true$ for all ``stored'' values.
 

### 2024-03-29 sections/tensors.tex
fix

@@
 identical value only once. In this work, we introduce the concept of a \textbf{pattern}
-level to handle the boolean case, returning $true$ for all ``stored'' values.
+level to handle this binary case.
+Pattern has a fill value of $false$, and returning $true$ for all ``stored'' values.
+The pattern level allows us to easily represent unweighted graphs or other boolean matrices.
 

### 2024-03-29 sections/intro.tex
Update on Overleaf.

@@
   Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
-  High-Level API           &            & \checkmark &            &            &            & \checkmark \\
-  Automatic API Fusion     &            &            &            &            &            & \checkmark \\
+  % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
+  % Automatic API Fusion     &            &            &            &            &            & \checkmark \\
 %  Parallelism             & \checkmark & \checkmark & \checkmark & \checkmark&           & \checkmark \\
@@
   Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
+    Arbitrary Read/Write              &    \checkmark        &  &            &  &            &\checkmark \\
   \end{tabular}
@@
 
-In this work, we propose a new programming language, Finch, which supports \textit{both} flexible control flow and diverse data structures.
+\textbf{In this work, we propose a new programming language, Finch, which supports \textit{both} flexible control flow and diverse data structures.}
 %
-Finch facilitates a programming model which resolves the challenges of computing over structured arrays by combining control flow and data structures into a common representation where they can be co-optimized.
+Finch facilitates a programming model which resolves the challenges of computing over structured arrays by \textbf{combining control flow and data structures into a common representation where they can be co-optimized}.
 %

### 2024-03-29 main.tex
Update on Overleaf.

@@
 %% article.
-\begin{abstract}
-\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.
-Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.
-However, arrays can only handle dense rectilinear integer grids.
-Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
-We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs.
-Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
-Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
-}
-
-Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
-Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
-Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler
-technology.  Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
-However, a large portion of the data, either obtained from nature or generated by humans, have structure.
-
-Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
-So far this has failed in both scope of the ambition and execution of comparable features.
-Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
-Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data.
-Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.
-For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-
-In this paper, we describe a compiler, Finch,
-which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
-programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays.
-Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
-sparse tensors, databases, and lossless compression.
-We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.  
-\end{abstract}
+\input{sections/abstract}
 
@@
 
-\section Frequently Requested Finch kernels
+\section{Frequently Requested Finch kernels}
 

### 2024-03-29 sections/abstract.tex
Update on Overleaf.

\begin{abstract}
\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.  Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.  However, arrays can only handle dense rectilinear integer grids. Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry. We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression. }

\HIDE{
Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler technology.
Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
However, a large portion of the data, either obtained from nature or generated by humans, have structure.

Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes. So far this has failed in both scope of the ambition and execution of comparable features.  Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry. 
Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data. Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers. For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.

In this paper, we describe a compiler, Finch,
which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays. Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
sparse tensors, databases, and lossless compression.
We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.}

From FORTRAN to Numpy, arrays have revolutionized how we express computation. However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
Support for structured data is fragmented and incomplete.
Existing frameworks limit the array structures and program control flow they support to better simplify the problem.

In this work, we propose a new programming language, Finch, which supports \textit{both}
flexible control flow and diverse data structures. Finch facilitates a programming model which
resolves the challenges of computing over structured arrays by combining control flow and data
structures into a common representation where they can be co-optimized. In particular,
Finch automatically specializes the control flow to the data so that performance engineers can
focus on experimenting with many algorithms. Finch supports a familiar programming language
of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as
sparsity, run-length-encoding, symmetry, triangles, padding, or blocks. Finch reliably utilizes the key
properties of structure, such as structural zeros, repeated values, or clustered non-zeros. We show that this leads to dramatic speedups in operations such as spmv and spgemm, image processing, graph analytics, and a high-level tensor operator fusion interface.
\end{abstract}

### 2024-03-29 sections/background.tex
formatting
Update on Overleaf.

@@
-\section{Background}
-\subsection{Looplets}
+\section{Background On Looplets}
 Finch represents iteration patterns using Looplets, a language that decomposes
@@
 \begin{figure}[ht]
+    \footnotesize
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchlookup(seek, body)$} The Lookup Looplet represents a
+        $\finchlookup(seek, body)$: The Lookup Looplet represents a
         randomly accessible region of an iterator. The body of the lookup is
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchrun(body)$} The Run Looplet represents a constant
+        $\finchrun(body)$: The Run Looplet represents a constant
         region of an iterator. The body of the run is understood to have one
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchphase(c:d, body)$} The Phase Looplet represents a
+        $\finchphase(c:d, body)$: The Phase Looplet represents a
         restriction of the range on which a loop should execute, and allows us
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchswitch(cond, head, tail)$} The Switch Looplet allows
+        $\finchswitch(cond, head, tail)$: The Switch Looplet allows
         us to specialize the body of a looplet based on a condition, evaluated
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchthunk(preamble, body, epilogue)$} The Thunk Looplet
+        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
         allows us to cache certain computations in the state under which the
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchsequence(head, tail)$} The Sequence looplet represents the
+        $\finchsequence(head, tail)$: The Sequence looplet represents the
         concatenation of two looplets. Both arguments must be phase looplets, and
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchspike(body, tail)$} The Spike Looplet represents a run
+        $\finchspike(body, tail)$ The Spike Looplet represents a run
         followed by a single value. In this paper, Spike will be considered a
@@
     \begin{minipage}[c]{0.6\linewidth}
-        \paragraph{$\finchstepper([seek], next, body)$} The stepper looplet
+        $\finchstepper([seek], next, body)$ The stepper looplet
         represents a variable number of looplets, concatenated. Since our
@@
 \end{figure}
-
-\subsection{FiberTrees}
-
-\subsection{Concordant Iteration}
-
-\subsection{Protocols}

### 2024-03-29 sections/intro.tex
Update on Overleaf.

@@
   \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
+  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
   \hline
-  Einsums and Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
+  Einsums/Contractions & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\
   % High-Level API           &            & \checkmark &            &            &            & \checkmark \\
@@
   Early Break              &            & \checkmark &            & \checkmark &            &\checkmark \\
-    Arbitrary Read/Write              &    \checkmark        &  &            &  &            &\checkmark \\
+  Unrestricted Read/Write              &    \checkmark        &  &            &  &            &  \\
   \end{tabular}
@@
   \begin{tabular}{l|cccccc}
-  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{StructTensor} & \rothead{Finch} \\
+  \textbf{Feature / Tool} & \rothead{Halide} & \rothead{Taco} & \rothead{Cora} & \rothead{Taichi} & \rothead{Stur} & \rothead{Finch} \\
   \hline
@@
   Padded                   & \checkmark &            &            &            &            & \checkmark \\
-  One Sparse               &            & \checkmark &            & \checkmark &            &\checkmark \\
-  Sparse                   &            & \checkmark &            &            &            &\checkmark \\
+  One Sparse Operand              &            & \checkmark &            & \checkmark &            &\checkmark \\
+  Multiple Sparse  Operands                 &            & \checkmark &            &            &            &\checkmark \\
   Run-length               &            &            &            &            &            & \checkmark \\
@@
 \item A compiler that specializes programs to data structures automatically, facilitating an expressive language that makes it easier to search the complex space of algorithms and data structures.
-Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
+Finch reliably utilizes four key properties of structure, such as structural zeros, repeated values, clustered non-zeros, and singletons.
 %
@@
 %\item <Performance Contributions>
+
 \end{enumerate}
+
+This advances the state of the art over the Looplets work, as the Looplets presented a way to merge iterators over multiple single dimensional structures, whereas we present a way to combine single-dimensional looplet structures into multi-dimensional tensors and operate on them with a programming language with fully-featured control flow.

### 2024-03-29 sections/tensors.tex
Update on Overleaf.

@@
 \begin{figure}[htbp]
-    \scriptsize
+    \footnotesize
     \raggedright
-\paragraph{$declare(lvl, init, dims...)$} Declares the level to hold subtensors
+$declare(lvl, init, dims...)$: Declares the level to hold subtensors
 of size $dims$ and an initial value of $init$. Requires the level to be in
 read-only mode.
-\paragraph{$freeze(lvl)$} Finalizes the updates in the level, and readies the
+
+$freeze(lvl)$: Finalizes the updates in the level, and readies the
 level for reading. Requires the level to be in update-only mode.
-\paragraph{$thaw(lvl)$} Prepares the level to accept updates. Requires the level
+
+$thaw(lvl)$: Prepares the level to accept updates. Requires the level
 to be in read-only mode.
-\paragraph{$unfurl(fiber(lvl, pos), ext, mode)$} Unfurls the fiber at position
+
+$unfurl(fiber(lvl, pos), ext, mode)$: Unfurls the fiber at position
 $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
@@
 to know whether the subfiber must be stored explicitly.
-\paragraph{$assemble(lvl, pos_{start}, pos_{stop})$} Allocates subfibers in the
+
+$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
 level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
@@
         \centering
-        \scriptsize
+        \footnotesize
         \begin{tabular}{|c|c|c|c|l|}
@@
     \centering
-    \scriptsize
+    \footnotesize
     \begin{tabular}{p{14cm}}
+
     \hline
@@
 \\
+
     \hline
@@
     \hline
+    \textbf{Leaf Levels} \\
+    \hline
+    \textbf{Element}:
+    The element level uses an array $val$ to store a value for each position $p$. The zero (fill) value is configurable.\\
+    \textbf{Pattern}:
+    The pattern level statically represents a leaf level with a fill value of $false$ and whose stored values are all $true$. \\
+    \hline
     \end{tabular}

### 2024-03-29 main.tex
shrink to fit

@@
     fontsize=\scriptsize,
-    linenos, % Enable line numbers
+    linenos=false, % Enable line numbers
     breaklines, % Enable line breaks

### 2024-03-29 sections/compiler.tex
shrink to fit

@@
 \subsection{Wrapperization}
+    \begin{wrapfigure}{R}{0.2\textwidth}
+        \vspace{-1\intextsep}
+        \begin{minted}{julia}
+            for i=_, j=_
+                if i <= j
+                    s[] += A[i, j]
+                end
+            end
+        \end{minted}
+        $\downarrow$
+        \begin{minted}{julia}
+            for i=_, j=_
+                if UpTriMask()[i, j]
+                    s[] += A[i, j]
+                end
+            end
+        \end{minted}
+        $\downarrow$
+        \begin{minted}{julia}
+            for i = 1:n
+                for j = 1:i
+                    s[] += A[i, j]
+                end
+            end
+        \end{minted}
+        \caption{Wrapperization}\label{fig:wrapperization}
+    \end{wrapfigure}
 
@@
     or by introducing \textbf{mask arrays} which replace index expressions like
-    \mintinline{julia}{i == j} with their equivalent masks (in this case, a
-    diagonal mask tensor).
+    \mintinline{julia}{i <= j} with their equivalent masks (in this case, a
+    triangular mask tensor).
 
@@
     allow us to lift computations with masks to the level of the loop, without modifying the loop directly.
-
-    For example, \begin{minted}{julia}
-    for i=_, j=_
-        if i <= j
-            s[] += A[i, j]
-        end
-    end
-    \end{minted}
-
-    would likely compile to something like
-
-    \begin{minted}{julia}
-    for i = 1:n
-        for j = 1:i
-            s[] += A[i, j]
-        end
-    end
-    \end{minted}
 
@@
     Our algorithm uses the following principles, assuming dimension types form a lattice:
+    \begin{wrapfigure}{r}{0.3\textwidth}
+        \begin{minted}{julia}
+            #A is 3 x 4
+            #B is 4 x 5
+            C .= 0
+            for i = 1:3
+                for j = _
+                    for k = _
+                        C[i, j] += A[i, k] * B[k, j]
+                    end
+                end
+            end
+        \end{minted}
+        $\downarrow$
+        \begin{minted}{julia}
+            C .= 0
+            for i = 1:3
+                for j = 1:5
+                    for k = 1:4
+                        C[i, j] += A[i, k] * B[k, j]
+                    end
+                end
+            end
+        \end{minted}
+        \caption{Dimensionalization}\label{fig:dimensionalization}
+    \end{wrapfigure}
     \begin{enumerate}
@@
 
-    For example, consider the following code
-
-    \begin{minted}{julia}
-    A = fsprand(3, 4, 0.5)
-    B = fsprand(4, 5, 0.5)
-    C = Tensor(Dense(SparseList(Element(0.0))))
-    @finch begin
-    C .= 0
-    for i = 1:3
-        for j = _
-            for k = _
-                C[i, j] += A[i, k] * B[k, j]
-            end
-        end
-    end
-    \end{minted}
-
-    In the above code, the second dimension of A must match the first dimension of B.
+    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B.
 Also, the first dimension of A must match the i loop dimension, 1:3.
@@
     Dimensionalization occurs after wrapper arrays are de-sugared. You can
-    therefore exempt a tensor from dimensionalization by wrapping the
-    corresponding index in ~ to produce a "PermissiveArray". For example,
-
-    \begin{minted}{julia}
-    @finch begin
-    y .= 0
-    for i = 1:3
-        y[~i] += x[i]
-    end
-    \end{minted}
-
-    does not set the dimension of y, and y does not participate in dimensionalization.
+    therefore exempt a mode in an access from dimensionalization by wrapping the
+    corresponding index in $~$ to produce a "PermissiveArray" (e.g. $A[~i]$).
 
@@
     We make expressions concordant by inserting loops with one iteration. For example, the following expression can be made concordant by adding a loop for the out of order access to \texttt{i}.
-    \begin{minted}{julia}
-        @finch begin
+    \begin{figure}[htbp]
+        \centering % Center the figure
+        % First pair of examples
+        \begin{minipage}{0.3\textwidth}
+        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
         for i = _
@@
         \end{minted}
-
-    \begin{minted}{julia}
-        @finch begin
+        \end{minipage}
+        $\rightarrow$
+        \begin{minipage}{0.4\textwidth}
+        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
             for i = _
@@
                         for l = _
-                        if i == l #diagonal mask
+                            if i == l
                                 s[] += A[l, k, j]
@@
         \end{minted}
-
-    We can concordize unrecognized index expressions using a similar approach:
-
-    \begin{minted}{julia}
+        \end{minipage}\hfill%
+        \begin{minipage}{0.2\textwidth}
+            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
                 for i = _
@@
                 \end{minted}
-
-    \begin{minted}{julia}
+            $\downarrow$
+            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
             for i = _
@@
             \end{minted}
+        \end{minipage}
 
-\subsection{Life Cycles and Scopes}
-\teo{Can we maybe skip this because we cover it in the semantics?}
-\willow{we should at least say here that we have a pass that adds freeze and thaw appropriately}
-\subsection{Simplification and Algebraic Transformations}
-
-\subsection{Bounds Analysis}
-\subsection{Performance Warnings}
-\teo{I also think this maybe belongs in an appendix.}
+        \caption{Transformations demonstrating concordization of unrecognized index expressions.}
+        \end{figure}
+\begin{wrapfigure}{r}{0.45\textwidth}
+    \vspace{-1\intextsep}
+    \begin{minipage}{0.2\textwidth}
+    \begin{minted}{julia}
+        y .= 0
+        for i = _
+            y[i] .= x[i] + 1
+        end
+        for i = _
+            x[i] += 1
+            y[i] += 1
+        end 
+        for i = _
+            x[i] += y[i]
+        end
+    \end{minted}
+    \end{minipage}
+    $\rightarrow$
+    \begin{minipage}{0.2\textwidth}
+    \begin{minted}{julia}
+        y .= 0
+        for i = _
+            y[i] .= x[i] + 1
+        end
+        @thaw(x)
+        @freeze(y)
+        for i = _
+            x[i] += 1
+            y[i] += 1
+        end
+        @freeze(x)
+        for i = _
+            x[i] += y[i]
+        end
+    \end{minted}
+    \end{minipage}
+    \caption{Life Cycle Automation.}\label{fig:lifecycles}
+\end{wrapfigure}
+\subsection{Life Cycle Automation}
+Finally, we introduce a pass to avoid needed to manually insert the \mintinline{julia}{@freeze} or \mintinline{julia}{@thaw} macros.
+The pass will insert these statements at the appropriate place in the program if they have not been inserted already, easing the burden on the programmer and bridging between structured and dense languages.
+An example is given in Figure \ref{fig:lifecycles}.

### 2024-03-31 sections/background.tex
Update on Overleaf.

@@
 semantics, rather than precise implementation. Since Finch is designed to be
-embedded into another language, we allow for several of the looplets to utilize
+embedded into another language\saman{not clear}, we allow for several of the looplets to utilize
 functions that modify the state of that language. It is assumed that if a
 looplet introduces a variable to be used in a child looplet, the child looplet
-will not modify that variaable.
+will not modify that variable.\saman{No context yet to understand this. Need to introduce what looplets are (Fig 2)!}
 

### 2024-03-31 sections/eval.tex
Update on Overleaf.
yay

@@
     \end{minipage}
-    \caption{Finch SpMV Programs}
+    \caption{Finch \saman{row-major, column-major and symmetric} SpMV Programs}
     \label{spmv_programs}
@@
 \begin{table}[htbp]
+    \footnotesize
     \centering
@@
     \centering
+    \footnotesize
     \caption{Transposed SpMV Sample Matrices}
@@
 \begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
+    \includegraphics[width=\linewidth]{Spmv_annotated.png}
     \caption{Performance of SpMV by Finch format.}
@@
 First, the output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the aggregate operator and to convert the pointwise expression into a Finch expression.
+Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to convert the pointwise expression into a Finch expression.
 At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor.

### 2024-03-31 sections/intro.tex
Update on Overleaf.

@@
 moler_history_2020, abadi_tensorflow_2016,
-paszke_pytorch_2019,ragan-kelley_halide_2013}.
+paszke_pytorch_2019,ragan-kelley_halide_2013}\saman{Put these citations next to each language - otherwise associated with the breakthroughs...}.
 % EASY CUT
@@
 caching, prefetching, multiple outputs, scatters, gathers, vectorization,
-loop-carry-dependencies, and more. Several optimizations have been developed for
+loop-carry-dependencies, and more. Several\saman{myriad of} optimizations have been developed for
 dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
@@
 constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
+contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. \saman{when you cite multiple projects, put the citations next to each project.}
 %
@@
 \end{minipage}
+\vspace{-24pt}
 \end{table}

### 2024-03-31 sections/lang.tex
update
Update on Overleaf.

@@
 mirrors most imperative languages with for-loops and control flow. Notable
-inclusions to the language at the level of statements includes \mintinline{julia}{for},
+statements that have been added to the language include \mintinline{julia}{for},
 \mintinline{julia}{let}, blocks of code with \mintinline{julia}{if}, and 
@@
 %
-However, to compile these constructs efficient, we must restrict this syntax in several ways, as we have alluded to in the previous section via life cycles.
+However, to compile these constructs efficiently, we must restrict this syntax in several ways, as we have alluded to in the previous section, via life cycles.
 %
@@
 %
-\teo{Make this true:}
 First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
@@
 %
-Also in the next section, We will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
-\teo{More next sections teaserse here}
+Also in the next section, we will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
 
@@
     \begin{prooftree}
-    \hypo{\langle val, (e, t) \rangle
+    \hypo{\langle val, (e, t, d) \rangle
     \rightarrow val'}
-    \hypo{var\not\in e}
-    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t) \rangle}}
+    \hypo{var\not\in d}
+    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t, d)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t, d\cup\{var\}) \rangle}}
     \end{prooftree}
@@
         \hypo{}
-        \infer1[$Literal$]{\langle\finchliteral(val), (e, t)\rangle \rightarrow val}
+        \infer1[$Literal$]{\langle\finchliteral(val), (e, t, d)\rangle \rightarrow val}
     \end{prooftree}
@@
         \hypo{}
-        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t)\rangle \Rightarrow \llangle ex, t \rrangle}
+        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t, d)\rangle \Rightarrow \llangle ex, t \rrangle}
     \end{prooftree}
@@
     \begin{prooftree}
-        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t)\rangle}{\rightarrow e(\finchvar(name))}}
+        \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t, d)\rangle}{\rightarrow e(\finchvar(name))}}
     \end{prooftree}
@@
     \begin{prooftree}
-        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t)\rangle}{\rightarrow e(\finchindex(name))}}
+        \infer0[$Index$]{\splitfrac{\langle\finchindex(name), (e, t, d)\rangle}{\rightarrow e(\finchindex(name))}}
     \end{prooftree}
@@
     \begin{prooftree}
-    \hypo{\langle head,s_{1} \rangle \rightarrow s_2}
-    \infer1[$BlockHead$]{\langle\finchblock(head, tail...), s_1 \rangle \rightarrow \langle \finchblock(tail...), s_2 \rangle}
+    \hypo{s = (e, t, d)}
+    \hypo{\langle head, (e, t, \{\}) \rangle \rightarrow s^{\prime}}
+    \hypo{\langle \finchblock(tail...), s^{\prime} \rangle \rightarrow (e^{\prime}, t^{\prime}, d^{\prime})})
+        \infer3[$BlockHead$]{\langle\finchblock(head, tail...), s \rangle \rightarrow (e^{\prime}, t^{\prime}, d) }
     \end{prooftree}
     \hfill
-    \begin{prooftree}
-        \infer0[$EmptyBlock$]{\langle\finchblock(), s\rangle \rightarrow s}
-    \end{prooftree}
+
     \vspace{6pt}
@@
     \begin{prooftree}
-        \hypo{\langle cond, s \rangle \Rightarrow true}
-        \infer1[$SieveTrue$]{\langle\finchsieve(cond, body), s\rangle \rightarrow \langle body, s \rangle}
+    \hypo{\langle cond, (e, t, d) \rangle \Rightarrow true}
+    \hypo{\langle body, (e, t, \{\}) \rangle \rightarrow (e^{\prime}, t^{\prime}, d^{\prime})}
+        \infer2[$SieveTrue$]{\langle\finchsieve(cond, body), (e, t, d)\rangle \rightarrow (e^{\prime}, t^{\prime}, d)}
     \end{prooftree}
@@
     \begin{prooftree}  
-    \hypo{s = (e, t)}
-    \hypo{\finchmode(tns)\not\in e}
+    \hypo{s = (e, t, d)}
+    \hypo{tns\not\in d}
     \hypo{e(tns) = tns'}
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle)}
+    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
     \end{prooftree}
@@
     \begin{prooftree}  
-    \hypo{s = (e, t)}
+    \hypo{s = (e, t,d )}
     \hypo{e(\finchmode(tns)) = update}
+    \hypo{tns\in d}
     \hypo{e(tns) = tns'}
-    \infer3[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle)}
+    \infer4[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle, d)}
     \end{prooftree}
@@
     \begin{prooftree}  
-    \hypo{s = (e, t)}
+    \hypo{s = (e, t, d)}
     \hypo{e(\finchmode(tns)) = read}
+    \hypo{tns\in d}
     \hypo{e(tns) = tns'}
-    \infer3[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle)}
+    \infer4[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle, d)}
     \end{prooftree}
@@
     \caption{Basic evaluation semantics, roughly defining most of these language
-    constructs to function similarly to their classical definitions. Note that
-    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
-    environment and another state $t$ corresponding to the state in the
-    host language. Several Looplets introduce variables into the embedding
+    constructs to function similarly to their classical definitions.
+    %
+    The state $s$ of the finch compiler is a tuple $(e, t, d)$ of a variable value
+    environment, another state $t$ corresponding to the state in the
+    host language, and finally the set of tensors defined within the current scope, $d$. 
+    %
+    This means that rules which modify $t$ are running in the host language.
+    %
+    Several Looplets introduce variables into the embedding
     language, which may be read when evaluating the \finchvalue node.
 %
-    The state $s$ of the finch compiler
-    is a tuple $(e, t)$ of a variable value environment and another state $t$
-    corresponding to the state in the host language.
-    This means that rules which modify $t$ are running in the host language. All of the lifecycle
+     All of the lifecycle
     functions are designed to be implemented and executed in the host language,
@@
     \caption{Looplet evaluation semantics. 
-    The state $s$ of the finch compiler is a tuple $(e, t)$ of a variable value
-    environment and another state $t$ corresponding to the state in the host
-    language. This means that rules which modify $t$ are running in the host language. Note that $E$ is an evaluation context that applies anywhere in
-    the syntax tree. The nonlocal evaluations of Looplets are what allow looplets to
-    hoist conditions and subranges out of loops. However, this also means we must specify
+    %
+    Basic evaluation semantics, roughly defining most of these language
+    constructs to function similarly to their classical definitions.
+    %
+    The state of the compiler is described in the previous figure.
+    %
+     Note that $E$ is an evaluation context that applies anywhere in
+    the syntax tree. 
+    %
+    The nonlocal evaluations of Looplets are what allow looplets to
+    hoist conditions and subranges out of loops.
+    %
+    However, this also means we must specify
     the priority in which we apply looplet rules, which is as follows:
     $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
+    %
     Many looplets, most notably the thunk looplet, introduce variables into the
-    host language environment.  While variables introduced by a looplet may be
+    host language environment. 
+    %
+    While variables introduced by a looplet may be
     modified by the looplet itself (steppers often increment some state
     variables), we forbid child looplets from modifying any state variables that
-    they didn't introduce. This allows us to treat the \finchvalue node as a
+    they didn't introduce. 
+    %
+    This allows us to treat the \finchvalue node as a
     constant.
@@
     Note: the $Simplify$ rule references $algebra$, which is our variable
-    defining a set of straightforward simplification rules. These rules include
+    defining a set of straightforward simplification rules. 
+    %
+    These rules include
     simple properties like $x * 0 \rightarrow 0$ to more complicated ones such
-    as constant propagation. We omit the full set of rules for brevity, but
+    as constant propagation. 
+    %
+    We omit the full set of rules for brevity, but
     point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}

### 2024-03-31 sections/tensors.tex
Update on Overleaf.

@@
 %
-At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible.
+At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible\saman{?????}.
 %
@@
 
-The first three of these functions, declare, freeze, and thaw, have to do with
-managing when tensors can be assumed mutable or immutable. As we use Looplets to
+The first three of these functions, declare, freeze, and thaw, \saman{Italics for keywords?}have to do with
+managing when tensors can be assumed mutable or immutable. \saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.} As we use Looplets to
 represent iteration over a tensor, we must restrict the mutability of tensors
@@
 
+\saman{Can we not put future work into the submission!}
 As future work, we also recommend the implementation of a tree-based format, as

### 2024-04-01 sections/tensors.tex
Update on Overleaf.

@@
 %
-At this point, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators, but Finch programs and data structures are sufficiently flexible that this impossible\saman{?????}.
+Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
+%
+However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
 %
@@
 Our simplified view of a level is enabled by our use of looplets to represent
-the structure within each fiber. In fact, our level interface requires only 5
+the structure within each fiber.
+%
+In fact, our level interface requires only 5
 highly general operations, described below.
@@
 The first three of these functions, declare, freeze, and thaw, \saman{Italics for keywords?}have to do with
-managing when tensors can be assumed mutable or immutable. \saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.} As we use Looplets to
+managing when tensors can be assumed mutable or immutable.
+%
+\saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.}
+%
+As we use Looplets to
 represent iteration over a tensor, we must restrict the mutability of tensors
-while we iterate over them. For example, if a tensor declares it has a constant
+while we iterate over them. 
+%
+For example, if a tensor declares it has a constant
 region from $i = 2:5$, but some other part of the computation modifies the
-tensor at $i = 3$, this would result in incorrect behavior. It is much easier to
+tensor at $i = 3$, this would result in incorrect behavior.
+%
+It is much easier to
 write correct Looplet code if we can assume that the tensor is immutable while
-we are reading from it. Thus, we introduce the notion that a tensor can be in
-read-only mode or update-only mode.  In read-only mode, the tensor may only
-appear in the right-hand side of assignments. In update-only mode, the tensor
+we are reading from it.
+%
+Thus, we introduce the notion that a tensor can be in
+read-only mode or update-only mode.  
+%
+In read-only mode, the tensor may only
+appear in the right-hand side of assignments
+%
+. In update-only mode, the tensor
 may only appear in the left-hand side of an assignment, either being overwritten
-or incremented by some operator.  We can switch between these modes using freeze
-and thaw functions. The declare function is used to allocate a tensor,
+or incremented by some operator. 
+%
+We can switch between these modes using freeze
+and thaw functions.
+%
+The declare function is used to allocate a tensor,
 initialize it to some specified size and value, and leave it in update-only
@@
 
-The unfurl function is used to manage iteration over a subfiber. At the point
-when it comes time to iterate over a tensor, be in on the left or right hand
+The unfurl function is used to manage iteration over a subfiber. 
+%
+At the point when it comes time to iterate over a tensor, be in on the left or right hand
 side of an assignment, we call the unfurl function to precompute whatever state
 and datastructures are necessary to return a looplet nest that would iterate
-over that level of a tensor. The unfurl function is called directly before
+over that level of a tensor.
+%
+The unfurl function is called directly before
 iterating over the corresponding loop, so it has access to any state variables introduced

### 2024-04-01 sections/tensors.tex
rewrite the intro

@@
 %
-The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
+%The Finch language provides descriptions of computations that iterate over a subset of a regular grid that is lexicographically ordered.
 %
-Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
+%Based on our description of Looplets, the reader might believe that compilation of a Finch program simply involves simply replacing for loops over a range with for loops over iterators.
 %
-However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
+%However, Finch's data structures and control flow are sufficiently complex that several challenges exist.
 %
-First, the Finch language interacts with multi-dimensional tensors whereas the Looplet abstraction is best suited towards iterators over a single dimension.
+Arrays use multiple dimensions to organize data with respect to orthogonal concepts.
 %
-We require a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
+Thus, the Finch language supports multi-dimensional arrays.
 %
-Second, since the iteration order of a Finch program might not match that of a data structure (a discordant traversal), different iterators need to be requested for the same data depending on the traversal order of the program.
+Unfortunately, the Looplet abstraction is best suited towards iterators over a single dimension.
 %
-So we require a bridge that can provide different iteration orders depending on the context.
+Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
 % 
-Third, since Finch programs can read and write to the same data, multi-dimensional tensors need to provide iterators for reading and writing as well as machinery to manage transition between these states.
+This bridge must address three challenges.
+%
+First, looplets represent an instance of an iterator over a tensor, but we may access the same tensor twice with different indices.
+%
+Thus, the $unfurl$ function creates separate looplet nests for each iterator.
+%
+Next, since Finch programs are not just single einsums, they may read and write to the same data at different times.
+%
+The $declare$, $freeze$, and $thaw$ functions provide machinery to manage transition between these states.
+%
+Finally, we must be able to write looplet nests that modify tensors, as well as reading them.
+%
+The $assemble$ function manages the allocation of new data in the tensor.
 

### 2024-04-01 sections/tensors.tex
comment

@@
 
+\willow{After making changes to the introduction of this section, some of the following text feels redundant}
 To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.

### 2024-04-01 sections/eval.tex
add intro text to evaluation.

@@
 \section{Evaluation}
+
+We evaluate Finch on a broad set of applications to showcase it's efficiency,
+flexibility, and expressibility. All of our implementations highlight the
+benefits of data structure and algorithm co-design.  Our implementation of
+sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
+performance engineering into the language of Finch, using temporaries and
+randomly-accessible workspace formats to efficiently implement the three main
+approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
+the benefits of precise structural specialization. Our studies of image
+morphology and graph applications show how Finch's programming model can express more complex
+real-world kernels. Finally, we explain
+how flexible operators, formats, and indexing expressions in Finch have
+supported a flexible implementation of the Python Array API, supporting fused exection.
 

### 2024-04-01 sections/tensors.tex
add intro text to evaluation.
Update on Overleaf.

--- 2024-04-01 sections/tensors.tex
@@
 
-To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
+In the rest of this section, we describe this functionality in detail after we review the basics of level abstractions~\cite{sze2017efficient,chou2022compilation, chou2018format}.
 %
-Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors~\cite{sze2017efficient,chou2022compilation, chou2018format}.
-%
-We recapitulate the essential details here.
-%
-Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
-%
-We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
-%
-These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
-%
-Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
+Then we
+
+% \willow{After making changes to the introduction of this section, some of the following text feels redundant}
+% To build our bridge, we embrace a set of abstractions: level formats/Fiber Trees, iteration context dependent instantiation of iterators, and tensor life cycles.
+% %
+% Our first abstraction mostly already exists in the literature: a manner of specifying a data structure for a multi-dimensional tensor out of data structures for single dimensional tensors
+% %
+% We recapitulate the essential details here.
+% %
+% Our next two abstractions add to to the first by providing a mechanism to use data structures generated by the first abstraction in a greater variety of contexts while maintaining per-dimension encapsulation of array data structures.
+% %
+% We introduce an interface to instatiate iterators in a variety of contexts in our programs and we introduce the lifecycle interface to manage when we read and write to multi-dimensional iterators.
+% %
+% These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
+% %
+% Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}.
 

### 2024-04-01 sections/tensors.tex
Update on Overleaf.

@@
 
+\teo{Here is a new paragraph.}
 In the rest of this section, we describe this functionality in detail after we review the basics of level abstractions~\cite{sze2017efficient,chou2022compilation, chou2018format}.
 %
-Then we
+Then we discuss $declare$, $freeze$, $thaw$, and $unfurl$ functions as part of a life cycle abstraction as well as the $assemble$ function.
+%
+These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to Looplets and expanding the contexts in which they can be used.
+%
+Finally, we describe several levels that we implement in this abstraction, which allow us to implement our taxonomy of known structures.
+
+% Previous efforts to compile a greater variety of sparse array programs left these bridges untouched ~\cite{henry_compilation_2021, won2023unified, senanayake2020sparse}
 

### 2024-04-01 sections/tensors.tex
Update on Overleaf.

@@
 %
-Finally, we describe several levels that we implement in this abstraction, which allow us to implement our taxonomy of known structures.
+Finally, we describe several levels that we implement in this abstraction, which allow us to implement a taxonomy of known structures that appear in tensor computations.
 

### 2024-04-01 sections/related.tex
Update on Overleaf.

\section{Related Work}

The related work on array languages, compilers, libraries, and frameworks, both for dense and structured computation, spans several areas.

%% Libraries: Numpy, blas, graphblas, SplAtt, Scipy
%% Languages: Halide, Lift, TVM, OptiML
%% Languages: Lgen/Spiral, LAMapping system,
%% languages sparse: TVM Sparse, Tiachi, SPF, TACO, Stut, various extensiosn to TACO,
%% SQL: HyPer, SDQL + Formats, Vectorwise, Morphesu,

%% Most similar to us: SDQL

### 2024-04-01 sections/related.tex
Update on Overleaf.

@@
 
-The related work on array languages, compilers, libraries, and frameworks, both for dense and structured computation, spans several areas.
+The related work on array languages and libraries, both for dense and structured computation, spans several areas.
+%
+We highlight these works based on if they are a library or language vs. if they are dense or more general.
 
-%% Libraries: Numpy, blas, graphblas, SplAtt, Scipy
+\paragraph{Dense Libraries:}
+
+\paragraph{Structured Libraries:}
+
+\paragraph{Dense Tensor Compilers:}
+
+\paragraph{Tensor Compilers for Structured Data:}
+
+%% Libraries: Numpy, LAPACK, blas, graphblas, SplAtt, Scipy, Petsc, 
 %% Languages: Halide, Lift, TVM, OptiML
@@
 
-%% Most similar to us: SDQL
+%% Most similar to us: SDQL, SPF Leader Follower, Taichi

### 2024-04-01 main.tex
Update on Overleaf.
close reading of evaluation up to SPGEMM

--- 2024-03-29 main.tex
@@
     style=colorful, % Set code style to colorful
-    fontsize=\scriptsize,
-    linenos, % Enable line numbers
+    fontsize=\tiny,
+    linenos=false, % Enable line numbers
     breaklines, % Enable line breaks

### 2024-04-01 sections/eval.tex
add reproducibility paragraph
close reading of evaluation up to SPGEMM

@@
-\section{Evaluation}
+\section{Case Studies}
 
@@
 
-\subsection{Data-Driven Performance Engineering}
-\subsubsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-%\changwan{
-%Explain what is SpGEMM. why SpGEMM hard? (not sure)
-%three approaches. Asym time complexity?
-%}
+All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
+2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
+through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
+whichever happens first.
 
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics, focusing on efficiently computing the product of two sparse matrices to produce an output sparse matrix.
+\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
+\begin{figure}
+    \begin{minipage}{0.333\linewidth}
+    \begin{minted}{julia}
+    @finch begin
+      C .= 0
+      for j=_
+        for i=_
+          for k=_
+            C[i, j] += AT[k, i] * B[k, j]
+          end
+        end
+      end
+      return C
+    end
+    \end{minted}
+    \end{minipage}%
+    \begin{minipage}{0.333\linewidth}
+    \begin{minted}{julia}
+    w = Tensor(SparseByteMap(Element(0)))
+    @finch begin
+      C .= 0
+      for j=_
+        w .= 0
+        for k=_
+          for i=_
+            w[i] += A[i, k] * B[k, j]
+          end
+        end
+        for i=_
+          C[i, j] = w[i]
+        end
+      end
+    end
+    \end{minted}
+    \end{minipage}%
+    \begin{minipage}{0.333\linewidth}
+    \begin{minted}{julia}
+    w = Tensor(SparseHash(SparseHash(Element(0))))
+    @finch begin
+      w .= 0
+      for k=_
+        for j=_
+          for i=_
+            w[i, j] += A[i, k] * BT[j, k]
+          end
+        end
+      end
+      C .= 0
+      for j=_, i=_
+        C[i, j] = w[i, j]
+      end
+    end
+    \end{minted}
+    \end{minipage}
+    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
+\end{figure}
 
-The inner-product, outer product, and Gustavson's algorithm are three different approaches to performing SpGEMM.
-Inner-product involves multiplying two matrices by taking dot products of corresponding rows and columns, while outer-product computes the product of all pairs of elements from the input matrices.
-Gustavson's algorithm computes the output matrix one row at a time, by scaling and merging rows of the second sparse matrix corresponding to the nonzero columns of a row of the first sparse matrix.
-Further explanation is available in Section 2.2 of~\cite{zhang2021gamma}.
+Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
+We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
 
-Figures~\ref{fig:spgemm_small} and~\ref{fig:spgemm_large} illustrate SpGEMM results with Finch and TACO.
-Notably, inner-product can be very inefficient unless sparse matrices are nearly dense due to the cost of intersections incurring redundant traversals.
-Additionally, TACO only supports outer product when the output matrix format is dense.
-It does not support outer product due to its lack of support for scatter into a 2D workspace, limiting its use for large sparse matrices.
-So, we were unable to obtain numbers for inner-product and TACO's outer-product for large matrices.
-In contrast, Finch supports scatter into 2D workspace, allowing for comparison with Gustavson's algorithm for large sparse matrices.
+There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
+%
+The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
+%
+Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
+%
+Inner products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
 
-As depicted in Figure~\ref{fig:spgemm_small}, Finch achieves comparable performance with TACO for inner-product.
-For large sparse matrices, as shown in Figure~\ref{fig:spgemm_large}, Finch outperforms TACO with an average speedup of 1.05.
-It's worth mentioning that TACO's Gustavson's algorithm is hard-wired whereas Finch's data structure-driving programming enables to systematically generate SpGEMM using explicit temporary, bytemap format, and matrix transposition.
+Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
+%
+Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
+%
+Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
+%
+In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
+%
+Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
+%
+It's also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
 
-%\help{The main points of this section are:}
+As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
+For large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
 
-%\help{1. summarize the difference between inner, gustavson, and outer, cite gamma but this also needs to stand on it's own}
-
-%\help{2. Explain why bytemap temporary makes sense for gustavson, but not for outer or inner}
-
-%\help{3. Point out that TACO does not even support outer product because it does not support scatter into 2D workspace}
-
-%\help{4.
-Take a look at the figure and point out that Finch is competitive with TACO on inner and gustavson, and that for outer product we see what we would expect from TACO: i.e.
-Finch Outer with dense performs similarly to TACO with dense, and Finch Outer with sparse performs much better as the matrix size increases.}
-
-%\help{5. point out that for the large matrices finch is competitive with TACO for gustavson}.
-
-%\help{6. explain the features of "datastructure-driven programming" that are used here in this problem. i.e. explicit temporary, bytemap format, transpositions}.
 \begin{figure}
-    \label{fig:spgemm}
 	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
@@
     dimension. Note that inner products necessarily requires $O(n^2)$ work and
-    taco's outer products format is dense. Finch can use a sparse outer products
+    TACO's outer products format is dense. Finch can use a sparse outer products
     format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and compare on larger matrices.}
+    output dimensions grow. On right, we use only gustavson's algorithm and
+    compare on larger matrices.}
+    \label{fig:spgemm}
 \end{figure}
 
-Examples that demonstrate performance engineering in a datastructure-driven model
-
-\subsubsection{SpMV}
+\subsection{SpMV}
 The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
@@
 
-\subsection{Programming over flexible data}
-
-\subsubsection{Image Morphology}
+\subsection{Image Morphology}
 
@@
 
-\subsubsection{Graph Analytics}
+\subsection{Graph Analytics}
 %\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.

### 2024-04-01 sections/tensors.tex

@@
 
-\teo{Here is a new paragraph.}
 In the rest of this section, we describe this functionality in detail after we review the basics of level abstractions~\cite{sze2017efficient,chou2022compilation, chou2018format}.

### 2024-04-01 sections/eval.tex
Update on Overleaf.

@@
 As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-For large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
+For large sparse matrices, Finch outperforms TACO with an average speedup of 1.05 \changwan{(self)may remove this number for consistency with other exps.
+finch\_outer\_bytemap not explained.}.
 
@@
 The graphs used were obtained from GraphBLAS~\cite{mattson2019lagraph}, including road networks (SNAP/roadNet-CA and GAP/GAP-road), characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high performance in such scenarios.
+Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios.
 

### 2024-04-01 sections/eval.tex
packing

--- 2024-04-01 sections/eval.tex
@@
 As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-For large sparse matrices, Finch outperforms TACO with an average speedup of 1.05.
+Finch outperforms TACO on larger matrices, with an average speedup of 1.05.
 
@@
 
-\subsection{SpMV}
-The sparse matrix-vector multiplication kernel (SpMV) is a common operation in sparse linear algebra with many applications including conjugate gradients, graph algorithms, numerical analysis, and neural networks [cite].
-SpMV is a bandwidth bound kernel, and thus historically has been subject to many modifications to improve throughput and reduce the associated control flow [cite].
-The wide range of applications unsurprisingly results in a wide range of types of datasets making it an effective kernel to demonstrate the utility of having flexible data formats.
+\subsection{Sparse Matrix-Vector Multiply (SpMV)}
+Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
+kernel, with a wide range of applications \cite{liu_csr5_2015,
+zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
+been proposed to reduce the footprint with a minimal impact on complexity
+\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
+results in a wide range of structures, making it an effective kernel to
+demonstrate the utility of flexible data formats.
 
-In this case study, we highlight a few different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
+In this case study, we highlight different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
 Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g.
@@
 
-% TODO: make this be 3 columns
 \begin{figure}
@@
 
+Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
+In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
+We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
+
+Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
+
+Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
+%
+All of these have their advantages.
+%
+The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
+
+The erosion operation turns off a pixel unless all of it's neighbors are on.
+%
+This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
+%
+Note that two instances of structure occur here, both in the mask and in the convolutional filter itself.
+%
+We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
+Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise or
+run-length-encoded algorithms.
+
+\begin{figure}
+    \scriptsize
+    \begin{minipage}{0.5\linewidth}
+    Wordwise Erosion:
+    \begin{minted}{julia}
+        output .= false
+        for y = _
+          tmp .= false
+          for x = _
+            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
+          end
+          for x = _
+            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
+          end
+        end
+    \end{minted}
+    \vspace{12pt}
+    Masked Histogram:
+    \begin{minted}{julia}
+        bins .= 0 
+        for x=_
+          for y=_
+            if mask[y, x]
+              bins[div(img[y, x], 16) + 1] += 1
+            end
+          end
+        end
+    \end{minted}
+    \end{minipage}%
+    \begin{minipage}{0.5\linewidth}
+    Bitwise Erosion:
+    \begin{minted}{julia}
+        output .= 0
+        for y = _
+          tmp .= 0
+          for x = _
+            if mask[x, y]
+              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
+            end
+          end
+          for x = _
+            if mask[x, y]
+              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
+                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
+                  output[x, y] = res
+                end
+              end
+            end
+          end
+        end
+    \end{minted}
+    \end{minipage}
+    \caption{Two approaches to erosion in Finch.
+The $coalesce$ function defines the out of bounds value.
+On left, the naive approach.
+On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
+On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
+    \label{fig:morphology_listing}
+\end{figure}
+
 \help{In this section, explain what erosion is, link a few cool images that show erosion, and then explain what the finch kernel looks like.
@@
 Talk about hist and also the fact that Finch supports scatter, here and in spgemm.}
+
+\help{Strength of flexible operators on bitwise ops}
 

### 2024-04-01 sections/abstract.tex

@@
 focus on experimenting with many algorithms. Finch supports a familiar programming language
-of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as
+of loops, statements, if conditions, breaks, etc., over a wide variety of array structures, such as
 sparsity, run-length-encoding, symmetry, triangles, padding, or blocks. Finch reliably utilizes the key

### 2024-04-01 sections/related.tex
Update on Overleaf.

@@
 
-\paragraph{Dense Libraries:}
+\paragraph{Dense Libraries:} There are many well known libraries that specialize in dense computations, exemplified by BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}.
+%
+This pattern is carried over into well known dense array libraries that use BLAS, exemplified by Numpy
+
+%
 
@@
 
-%% Libraries: Numpy, LAPACK, blas, graphblas, SplAtt, Scipy, Petsc, 
+%% Libraries: Numpy, LAPACK, blas, graphblas, SplAtt, Scipy, Petsc, Graph libraries (ligraph, lagraph,...)
 %% Languages: Halide, Lift, TVM, OptiML
 %% Languages: Lgen/Spiral, LAMapping system,
-%% languages sparse: TVM Sparse, Tiachi, SPF, TACO, Stut, various extensiosn to TACO,
+%% languages sparse: TVM Sparse, Tiachi, SPF, TACO, Stut, various extensiosn to TACO, Graphit
 %% SQL: HyPer, SDQL + Formats, Vectorwise, Morphesu,

### 2024-04-01 sections/related.tex
Update on Overleaf.

@@
 %
-This pattern is carried over into well known dense array libraries that use BLAS, exemplified by Numpy
-
-%
+This pattern is carried over into well known dense array libraries that use BLAS, exemplified by Numpy.
 
 \paragraph{Structured Libraries:}
+
+Many libraries support BLAS plus a few sparse array types, typically CSR, CSC, BCSR, and COO.
+%
+Examples include Scipy, PETSc, Armadillo, and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, eigenweb}.
 

### 2024-04-01 sections/related.tex
Update on Overleaf.

@@
 Examples include Scipy, PETSc, Armadillo, and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, eigenweb}.
+%
+Several of these libraries also feature some graph or mesh algorithms built on sparse matrices, which represent other sparse structures.
+%
+GraphBlas perhaps exemplifies this pattern with support for many sparse matrix operations on a variety of formats~\cite{kepner2016mathematical} although a variety of other graph libraries adopt a similar approach such as LAgraph~\cite{mattson2019lagraph}
 

### 2024-04-02 sections/eval.tex
cool

@@
 Finch outperforms TACO on larger matrices, with an average speedup of 1.05.
+\changwan{(self)may remove this number for consistency with other exps.
+finch\_outer\_bytemap not explained.}
 
@@
 This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
+\help{link a few cool images that show erosion}
 %
-Note that two instances of structure occur here, both in the mask and in the convolutional filter itself.
+Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
 %
 We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise or
-run-length-encoded algorithms.
+Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
+or run-length-encoded algorithms, and a masked histogram kernel.
+
+We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
+in the masked kernel and get better performance.
+%
+We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
+
+We used four datasets.
+We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
+We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
+All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
+IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
+The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
+mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (51.6x on MNIST and 20.3x on Omniglot).
+The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
+A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
 
@@
 
-\help{In this section, explain what erosion is, link a few cool images that show erosion, and then explain what the finch kernel looks like.
-in particular, be sure to point out that we're doing unrolled convolution, and how that maps to the generated code (merging shifted sparse iterators).}
-
-\help{Explain the role of formats in this kernel, how sparseRLE(PAttern) is really the right thing to use here}
-
-\help{Explain that finch can support a bitwise version, and that finch can mask the bitwise kernel too to get performance with a small change to the code}
-
-\help{Point out that masks with SparseRLE can sometimes perform better because they lift the masking out of the inner loop.
-Talk about hist and also the fact that Finch supports scatter, here and in spgemm.}
-
-\help{Strength of flexible operators on bitwise ops}
-
 \begin{figure}
@@
 	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}
+    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
 \end{figure}
-
-%\begin{figure}
-%	\includegraphics[width=\linewidth]{erode4_speedup_over_opencv.png}
-%    \caption{Performance of Finch on erosion task (4 iterations).}
-%\end{figure}
 

### 2024-04-02 sections/related.tex

--- 2024-04-01 sections/related.tex
@@
 Examples include Scipy, PETSc, Armadillo, and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, eigenweb}.
+%
+Several of these libraries also feature some graph or mesh algorithms built on sparse matrices.
+%
+GraphBlas exemplifies this pattern with support for many sparse matrix operations on a variety of formats~\cite{kepner2016mathematical} although a variety of other graph libraries adopt a similar approach such as LAgraph or GBase~\cite{mattson2019lagraph, kang2011gbase} among many others.
 

### 2024-04-02 sections/related.tex
Update on Overleaf.

--- 2024-04-01 sections/related.tex
@@
 
-\paragraph{Dense Libraries:} There are many well known libraries that specialize in dense computations, exemplified by BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}.
+\paragraph{Libraries for Dense Data:} There are many well known libraries that specialize in dense computations, exemplified by BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}.
 %
 This pattern is carried over into well known dense array libraries that use BLAS, exemplified by Numpy.
+%
+Many research projects have advanced on BLAS, such as BatchedBlas and BLIS~\cite{dongarra2017design, van2015blis}.
 
-\paragraph{Structured Libraries:}
+\paragraph{Libraries for Structured Data:}
 
-Many libraries support BLAS plus a few sparse array types, typically CSR, CSC, BCSR, and COO.
+Many libraries support BLAS plus a few sparse array types, typically CSR, CSC, BCSR, Banded, and COO.
 %
-Examples include Scipy, PETSc, Armadillo, and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, eigenweb}.
+Examples include Scipy, PETSc, Armadillo, OSKI, Cyclops, MKL (add cite), and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, vuduc2005oski,solomonik2013cyclops, eigenweb}.
+%
+There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt}.
+%
+Several of these libraries also feature some graph or mesh algorithms built on sparse matrices.
+%
+GraphBlas exemplifies this pattern with support for many sparse matrix operations on a variety of formats~\cite{kepner2016mathematical} although a variety of other graph libraries adopt a similar approach such as LAgraph or GBase~\cite{mattson2019lagraph, kang2011gbase} among many others~\cite{ashari2014fast,huang2020ge}.
+%
+In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays an few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
+%
 
-\paragraph{Dense Tensor Compilers:}
+\paragraph{Compilers for Dense Data:}
+%% Languages: Halide, Lift, TVM, OptiML
+Outside of general purpose compilers, many compilers have been developed for optimizing dense data on a variety of control flow.
+%
+Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM, Exo, and Elevate~\cite{chen2018tvm, ikarashi2022exocompilation, hagedorn2020achieving}.
+%
+These languages typically support most control flow except for an early break though some don't support arbitrary reading/writing or even indirect accesses.
+%
+Several polyhedral languages, such as Polly, Tiramisu, CHiLL, Pluto, AlphaZ ~\cite{grosser2012polly, baghdadi2019tiramisu,chen2008framework, bondhugula2008pluto, yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
+%
+These are based on ISL~\cite{verdoolaege2010isl}.
+%
+The density of this research represents the density of support for dense computation.
 
-\paragraph{Tensor Compilers for Structured Data:}
+\paragraph{ Compilers for Structured Data:}
+Several compilers exist for several types of structured data, often featuring separate languages for the storage of the structured data and the computation.
+%
+The TACO compiler originally supported just plain einsum computations, but has been extended several times to support (single dimensional) local tensors, breaks via semi-rings, windowing, tiling, and convolution~\cite{kjolstad_tensor_2017, kjolstad_tensor_2019, senanayake2020sparse, henry_compilation_2021,won2023unified}.
+%
+Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures and tree like structures~\cite{kjolstad_tensor_2017, chou2018format, chou2022compilation}.
+%
+The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}.
+%
+Similarly, SDQL originally supported only hash tables, but has been extended with one format language that allows one to specify tensors as queries on several base storage types~\cite{schleich2023optimizing} and via another system that optimizing programs by taking advantage of predicates describing symmetries and other structures on a tensor~\cite{ghorbani2023compiling}.
+%
+Taichi is another language with a separate program and data structure specification, but Taichi focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
+%
+Outside of these works, the window of comparison fogs up.
+%
+The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
+%
+Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
+%
+However, many SQL or dataframe systems rely on BTrees, columnar, or Hashtables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
+%
+However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
+%
+Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a recursive FFT. 
+%
+SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
+%
+
+%Stut and SDQL - similar but more workspaces and rings. Mention formats paper of theirs - similar to looplets, but to our knowledge doesn't support all of oru structures.
+%Tiachi - 
+%SPF - ????
+%Spiral
+%SQL - Vectorwise/Morpheus/HyPer and others?
+
+\paragraph{Other Architectures:} We would be remiss if we didn't mention another axis: architectural support. 
+%
+We saved this angle for future work, but we must note that much of this work varies in support for other architectures, which is arguably another element of control flow.
+%%TACO
+%%Tiachi
+%%SPF
+%%Stut
+%%Graphit
+%%Spiral
+%%SQL
 
 %% Libraries: Numpy, LAPACK, blas, graphblas, SplAtt, Scipy, Petsc, Graph libraries (ligraph, lagraph,...)
-%% Languages: Halide, Lift, TVM, OptiML
-%% Languages: Lgen/Spiral, LAMapping system,
+
+%% Languages: Lgen/Spiral
 %% languages sparse: TVM Sparse, Tiachi, SPF, TACO, Stut, various extensiosn to TACO, Graphit

### 2024-04-02 sections/eval.tex
cleanin up graphs

@@
 The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (51.6x on MNIST and 20.3x on Omniglot).
+mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
+We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
 The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
@@
 
-To evaluate the capabilities of Finch in graph applications, we assessed its performance in two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
-The graphs used were obtained from GraphBLAS~\cite{mattson2019lagraph}, including road networks (SNAP/roadNet-CA and GAP/GAP-road), characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios.
+We used finch to implement two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
+Our BFS implementation and graphs datasets are taken from Yang et.
+al \cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
 
-Direction-optimized BFS utilizes push and pull traversal techniques to efficiently explore graphs.
-Push traversal originates from the source node, pushing neighboring nodes into the queue, while pull traversal starts from the target node, exploring predecessors (parents).
-Incorporating early-break in pull traversal enables premature termination upon encountering the target node, thereby avoiding unnecessary edge traversal.
+Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
+Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
+The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
+Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
 
-Figure \ref{graph_result} (a) illustrates BFS speedup over Graphs.jl, a BFS implementation with push traversal in Julia.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Additionally, Finch's support for different loop orders and multiple outputs further enhances performance.
-Direction-optimization notably enhances performance for scale-free graphs (average speedup to be included).
-Although GraphBLAS sometimes outperforms Finch due to its incorporation of all hardwired optimizations, we anticipate achieving better performance with Finch upon incorporating similar optimizations.
-\changwan{mark captions (a) and (b)}
-
-Figure \ref{graph_result} (b) presents Bellman-Ford speedup (with path lengths and shortest-path tree) over Graphs.jl.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.
-Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05).
-Note that we did not include GAP-road as it takes too long time with Graph.jl.
-\changwan{(self)dataset name changed?}
+Figure \ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS\cite{mattson2019lagraph}.
+For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
+Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
+On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05).
+Note that we did not include GAP-road as it took too long to run.
 
@@
 	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \caption{Performance of graph apps across various tools.}
-     \label{graph_result}
+    \caption{Performance of graph apps across various tools.
+finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
+Finch's support for push/pull traversal and early break facilitates direction-optimization.
+Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
+     \label{fig:graph_result}
 \end{figure}
@@
 \begin{figure}
-    \begin{minipage}{0.5\linewidth}
+    \begin{minipage}{0.33\linewidth}
     \begin{minted}{julia}
@@
 
-    function finch_bfs_push_kernel(_F, F, A, V, P)
+    function bfs_push(_F, F, A, V, P)
       @finch begin
@@
 \end{minipage}%
-\begin{minipage}{0.5\linewidth}
+\begin{minipage}{0.33\linewidth}
     \begin{minted}{julia}
-    function finch_bfs_pull_kernel(_F, F, AT, V, P)
+    function bfs_pull(_F, F, AT, V, P)
       p = ShortCircuitScalar{0}()
@@
     \end{minted}
+\end{minipage}%
+\begin{minipage}{0.33\linewidth}
+  \begin{minted}{julia}
+  _D = Tensor(Dense(Element(Inf)), n)
+  D = Tensor(Dense(Element(Inf)), n)
+  function bellmanford(A, _D, D, _F, F)
+    @finch begin
+    F .= false
+    for j = _
+      if _F[j]
+        for i = _
+          let d = _D[j] + A[i, j]
+            D[i] <<min>>= d
+            F[i] |= d < _D[i]
+          end
+        end
+      end
+    end
+  end
+\end{minted}
 \end{minipage}
+\caption{Graph Applications written in Finch.
+Note that parents are calculated separately for Bellman-Ford.
+The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
 \end{figure}

### 2024-04-02 sections/compiler.tex
Update on Overleaf.
shrinking

@@
 The properties of such a program are as follows:
-\begin{enumerate}
-    \item[Access with Indices] Though Finch allows general expressions in an
+\begin{itemize}
+    \item \textbf{Access with Indices:} Though Finch allows general expressions in an
     access (i.e. `A[i + j]` or `A[I[i]]`), the normal form restricts to allow only indices in 
     accesses (i.e. `A[i]`), rather than more general expressions.
-    \item[Evaluable Dimensions] Loop dimensions and declaration dimensions must
+    \item \textbf{Evaluable Dimensions:} Loop dimensions and declaration dimensions must
     be evaluable at the time we compile them, so we restrict the normal form to
@@
     statement.
-    \item[Concordant] Finch is column-major by default, and the normal form
+    \item \textbf{Concordant:} Finch is column-major by default, and the normal form
     requires the order of indices in an access to match the order in which loops
@@
     \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end} is not.
-    \item[Lifecycle Constraints] Tensors in read mode may appear on the right
+    \item \textbf{Lifecycle Constraints:} Tensors in read mode may appear on the right
     hand side only. Tensors in update mode may appear on the left hand side
@@
     defined.
-\end{enumerate}
+\end{itemize}
 
@@
 \subsection{Wrapperization}
-    \begin{wrapfigure}{R}{0.2\textwidth}
+    \begin{wrapfigure}{R}{0.17\textwidth}
         \vspace{-1\intextsep}
@@
     \mintinline{julia}{i <= j} with their equivalent masks (in this case, a
-    triangular mask tensor).
+    triangular mask tensor).  Wrappers and masks are summarized in \ref{tab:wrappers}.
 
@@
     implement the looplet functions of \ref{fig:semantics_looplets} with some minor modifications
-    to shift dimensions, for example. Wrapper arrays are summarized in Table \ref{tab:wrappers}
+    to shift dimensions, for example.
 
     Mask arrays have a more straightforward implementation using static looplets
-    that are constructed during the unfurl step. They are summarized in table \ref{tab:masks}. Mask tensors
+    that are constructed during the unfurl step. Mask tensors
     allow us to lift computations with masks to the level of the loop, without modifying the loop directly.
@@
         \hline
-        \end{tabular}
-        \caption{Wrapper arrays and some example indexing sugar they enable.}
-    \label{table:wrappers}
-    \end{table}
-
-    \begin{table}[h]
-        \scriptsize
-        \centering
-        \begin{tabular}{|>{\raggedright\arraybackslash}m{0.3\linewidth}|>{\raggedright\arraybackslash}m{0.4\linewidth}|}
-        \hline
-        \textbf{Transformation Example} & \textbf{Description} \\
-        \hline
         $i < j \rightarrow$ UpTriMask()[i, j - 1] & Upper triangular mask, true if $i < j$. \\
@@
         \end{tabular}
-        \caption{Mask tensors and some example cases they are inserted.}
-        \label{table:masks}
+        \caption{Wrapper arrays, masks, and some example indexing sugar they enable.}
+    \label{tab:wrappers}
     \end{table}
@@
     Our algorithm uses the following principles, assuming dimension types form a lattice:
-    \begin{wrapfigure}{r}{0.3\textwidth}
+    \begin{wrapfigure}{r}{0.25\textwidth}
         \begin{minted}{julia}
@@
         \end{minted}
-        \caption{Dimensionalization}\label{fig:dimensionalization}
+        \caption{Dimensionalization.}\label{fig:dimensionalization}
     \end{wrapfigure}
@@
 
-    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B.
-Also, the first dimension of A must match the i loop dimension, 1:3.
-Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5).
-If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.
-
-    Dimensionalization occurs after wrapper arrays are de-sugared. You can
-    therefore exempt a mode in an access from dimensionalization by wrapping the
-    corresponding index in $~$ to produce a "PermissiveArray" (e.g. $A[~i]$).
-
-\subsection{Concordization}
-    After wrapperization, all arrays are normalized to column-major ordering and
-    the unrecognized expressions are left in the access expressions. At that
-    point, we run a pass over the code to make the program concordant.
-
-    We make expressions concordant by inserting loops with one iteration. For example, the following expression can be made concordant by adding a loop for the out of order access to \texttt{i}.
-    \begin{figure}[htbp]
+    \begin{wrapfigure}{r}{0.4\linewidth}
         \centering % Center the figure
         % First pair of examples
-        \begin{minipage}{0.3\textwidth}
-        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
+        \begin{minipage}{0.16\textwidth}
+        \begin{minted}{julia}
+          for i = _
+            for j = _
+                s[] += A[i, j]
+            end
+          end
+        \end{minted}
+        $\downarrow$
+        \begin{minted}{julia}
           for i = _
@@
               for k = _
-                    s[] += A[i, k, j]
-                end
-            end
-        end
-        \end{minted}
-        \end{minipage}
-        $\rightarrow$
-        \begin{minipage}{0.4\textwidth}
-        \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
-            for i = _
-                for j = _
-                    for k = _
-                        for l = _
-                            if i == l
-                                s[] += A[l, k, j]
-                            end
+                if i == k
+                  s[] += A[k, j]
                 end
@@
         \end{minipage}\hfill%
-        \begin{minipage}{0.2\textwidth}
-            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
+        \begin{minipage}{0.12\textwidth}
+            \begin{minted}{julia}
               for i = _
@@
             $\downarrow$
-            \begin{minted}[fontsize=\scriptsize, breaklines]{julia}
+            \begin{minted}{julia}
               for i = _
@@
             \end{minted}
-        \end{minipage}
+        \end{minipage}\hfill
+        \caption{Examples of concordization, transforming accesses to column major with bare indices.}\label{fig:concordization}
+    \end{wrapfigure}
 
-        \caption{Transformations demonstrating concordization of unrecognized index expressions.}
-        \end{figure}
-\begin{wrapfigure}{r}{0.45\textwidth}
-    \vspace{-1\intextsep}
-    \begin{minipage}{0.2\textwidth}
+    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B.
+Also, the first dimension of A must match the i loop dimension, 1:3.
+Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5).
+If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.
+
+    Dimensionalization occurs after wrapper arrays are de-sugared. You can
+    therefore exempt a mode in an access from dimensionalization by wrapping the
+    corresponding index in \mintinline{julia}{~} to produce a "PermissiveArray" (e.g. \mintinline{julia}{A[~i]}).
+
+\subsection{Concordization}
+    After wrapperization, all arrays are normalized to column-major ordering and
+    the unrecognized expressions are left in the access expressions. At that
+    point, we run a pass over the code to make the program concordant. We make
+    expressions concordant by inserting loops with one iteration. Examples are given in Figure~\ref{fig:concordization}.
+
+    \begin{wrapfigure}{R}{0.34\linewidth}
+        \begin{minipage}{0.15\textwidth}
         \begin{minted}{julia}
@@
         $\rightarrow$
-    \begin{minipage}{0.2\textwidth}
+        \begin{minipage}{0.15\textwidth}
         \begin{minted}{julia}
@@
         \end{minipage}
-    \caption{Life Cycle Automation.}\label{fig:lifecycles}
+        \caption{Life cycle automation.} \label{fig:lifecycles}
     \end{wrapfigure}
@@
 Finally, we introduce a pass to avoid needed to manually insert the \mintinline{julia}{@freeze} or \mintinline{julia}{@thaw} macros.
+%
 The pass will insert these statements at the appropriate place in the program if they have not been inserted already, easing the burden on the programmer and bridging between structured and dense languages.
-An example is given in Figure \ref{fig:lifecycles}.
+% An example is given in Figure \ref{fig:lifecycles}.

### 2024-04-02 sections/eval.tex
packing

@@
 
-\begin{figure}
-    \begin{minipage}[t]{0.315\textwidth}
+\begin{wrapfigure}{r}{0.41\linewidth}
+    \begin{minipage}[t]{0.18\textwidth}
         \vspace{0pt} % Add this to ensure top alignment within minipage
         \begin{minted}{julia}
-            @finch begin
             y .= 0
@@
             end
-                return y
-            end
         \end{minted}
-    \end{minipage}%
-    \begin{minipage}[t]{0.315\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
+        \vspace{24pt} % Add this to ensure top alignment within minipage
         \begin{minted}{julia}
-            @finch begin
             y .= 0
@@
             end
-                return y
-            end
         \end{minted}
-    \end{minipage}
-    \begin{minipage}[t]{0.36\textwidth}
+    \end{minipage}\hfill%
+    \begin{minipage}[t]{0.22\textwidth}
         \vspace{0pt} % Add this to ensure top alignment within minipage
         \begin{minted}{julia}
-            @finch begin
             y .= 0
@@
                 end
-                        y[j] += y_j[] + diag[j] * x_j
+                #D is the diagonal
+                y[j] += y_j[] + D[j] * x_j
               end
-                end
-                return y
             end
@@
     \end{minipage}
-    \caption{Finch \saman{row-major, column-major and symmetric} SpMV Programs}
+    \caption{Finch row-major, column-major and symmetric SpMV Programs}
     \label{spmv_programs}
-\end{figure}
+\end{wrapfigure}
 

### 2024-04-02 sections/tensors.tex
packing

@@
 
-\saman{Can we not put future work into the submission!}
-As future work, we also recommend the implementation of a tree-based format, as
-this would enable incremental construction. Currently, the cost of freezing and
-thawing is usually proportional to the number of stored fibers, so an operation
-that sets a single value in a tensor would be prohibitively expensive (i.e.
-$A[3, 4] = 3.14$). We would also recommend supporting SparseRLE as an
-intermediate randomly accessible structure, because (unlike blocks or
-singletons), runs have asymptotic benefits. However, these cases were not
-necessary for our case studies.
-
-\help{we need also to talk about pattern and element}
+\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction.
+Currently, the cost of freezing and thawing is usually proportional to the number of stored fibers, so an operation that sets a single value in a tensor would be prohibitively expensive (i.e.
+$A[3, 4] = 3.14$).
+We would also recommend supporting SparseRLE as an intermediate randomly accessible structure, because (unlike blocks or singletons), runs have asymptotic benefits.
+However, these cases were not necessary for our case studies.}
 

### 2024-04-02 sections/background.tex
Update on Overleaf.
cool

@@
 Finch represents iteration patterns using Looplets, a language that decomposes
-datastructure iterators hierarchically. Looplets represent the control-flow
+datastructure iterators hierarchically. 
+%
+Looplets represent the control-flow
 structures needed to iterate over any given datastructure, or multiple
-datastructures simultaneously. In particular, looplets are good at lifting code
+datastructures simultaneously. 
+%
+In particular, looplets are good at lifting code
 to the highest loop level that it's needed and subdividing iteration
-hierarchically in coordinate space. Because looplets are compiled with
+hierarchically in coordinate space.
+%
+Because looplets are compiled with
 progressive lowering, structure-specific mathematical optimizations such as
@@
     \footnotesize
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchlookup(seek, body)$: The Lookup Looplet represents a
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-lookup.png}
+        \includegraphics[scale=0.20]{Looplets-lookup.png}
     \end{minipage}
+    \vspace{3pt}
 
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchrun(body)$: The Run Looplet represents a constant
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-run.png}
+        \includegraphics[scale=0.20]{Looplets-run.png}
     \end{minipage}
+    \vspace{3pt}
 
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchphase(c:d, body)$: The Phase Looplet represents a
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-phase.png}
+        \includegraphics[scale=0.20]{Looplets-phase.png}
     \end{minipage}
-    \begin{minipage}[c]{0.6\linewidth}
+    \vspace{3pt}
+
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchswitch(cond, head, tail)$: The Switch Looplet allows
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-switch.png}
+        \includegraphics[scale=0.20]{Looplets-switch.png}
     \end{minipage}
-    \begin{minipage}[c]{0.6\linewidth}
+    \vspace{3pt}
+
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-thunk.png}
+        \includegraphics[scale=0.20]{Looplets-thunk.png}
     \end{minipage}
+    \vspace{3pt}
 
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchsequence(head, tail)$: The Sequence looplet represents the
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-sequence.png}
+        \includegraphics[scale=0.20]{Looplets-sequence.png}
     \end{minipage}
+    \vspace{3pt}
 
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchspike(body, tail)$ The Spike Looplet represents a run
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-spike.png}
+        \includegraphics[scale=0.20]{Looplets-spike.png}
     \end{minipage}
+    \vspace{3pt}
 
-    \begin{minipage}[c]{0.6\linewidth}
+    \begin{minipage}[c]{0.65\linewidth}
         $\finchstepper([seek], next, body)$ The stepper looplet
@@
     \end{minipage}%
-    \begin{minipage}[c]{0.4\linewidth}
+    \begin{minipage}[c]{0.35\linewidth}
         \centering
-        \includegraphics[scale=0.25]{Looplets-stepper.png}
+        \includegraphics[scale=0.20]{Looplets-stepper.png}
     \end{minipage}

### 2024-04-02 sections/cut.tex
cool

@@
 }
+
+\scriptsize
+\noindent % This ensures the minipages fill the page width
+\begin{minipage}{0.4\linewidth}
+\begin{align*}
+    \finchliteral(val \in \mathbb{V}) &:= \text{\mintinline{julia}{val}} \\
+    \finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) &:= \text{\mintinline{julia}{ex :: type}} \\
+    \finchtensor(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchindex(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchvar(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
+    \finchextent(a \in E, b \in E) &:= \text{\mintinline{julia}{a : b}} \\
+    \finchcall(f \in E, args\ldots \in E) &:= \text{\mintinline{julia}{f(args...)}} \\
+    \finchaccess(tns \in T, idxs\ldots \in E) &:= \text{\mintinline{julia}{tns[idxs...]}} \\
+    \finchfreeze(tns \in T) &:= \text{\mintinline{julia}{@freeze(tns)}} \\
+    \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
+    \finchmode(tns \in T) &:= \text{\mintinline{julia}{@mode(tns)}} \\
+\end{align*}
+\end{minipage}%
+\begin{minipage}{0.5\linewidth}
+\begin{align*}
+    \mathbb{V} &:= \text{the set of all values.} \\
+    \mathbb{S} &:= \text{the set of all symbols.} \\
+    \mathbb{T} &:= \text{the set of all types.} \\
+    I &:= \finchindex \\
+    A &:= \finchaccess \\
+    V &:= \finchvar \\
+    T &:= \finchtensor \\
+    E &:= \finchliteral \mid \finchvalue \mid \finchindex \\
+        &\quad \mid \finchvar \mid \finchextent \mid \finchcall \mid \finchaccess \\
+    S &:= \finchassign \mid \finchloop \mid \finchdefine \\
+        &\quad \mid \finchsieve \mid \finchblock \\
+\end{align*}
+\end{minipage}%
+\begin{align*}
+    \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
+    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs <<op>>= rhs}} \\
+    \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
+    \finchdefine(var \in V, val \in E, body \in S) &:= \text{\mintinline{julia}{let var = val; body end}} \\
+    \finchsieve(cond \in E, body \in S) &:= \text{\mintinline{julia}{if cond; body end}} \\
+    \finchblock(bodies\ldots \in S) &:= \text{\mintinline{julia}{begin; bodies... end}}
+\end{align*}

### 2024-04-02 sections/lang.tex
cool

@@
 \section{The Finch Language}
+
+\begin{wrapfigure}{r}{0.51\linewidth}
+    \vspace{-26pt}
+    \begin{minted}{julia}
+     EXPR := LITERAL | VALUE | INDEX | VARIABLE | EXTENT | CALL | ACCESSS
+     STMT := ASSIGN | LOOP | DEFINE | SIEVE | BLOCK
+
+      DECLARE := TENSOR .= EXPR(EXPR...)   #V is the set of all values
+       FREEZE := @freeze(TENSOR)           #S is the set of all Symbols
+         THAW := @thaw(TENSOR)             #T is the set of all types                                      
+       ASSIGN := ACCESS <<EXPR>>= EXPR    
+         LOOP := for INDEX = EXPR              LITERAL := V                           
+                   STMT                          VALUE := S::T                        
+                 end                            TENSOR := S                            
+       DEFINE := let VARIABLE = EXPR             INDEX := S                            
+                   STMT                       VARIABLE := S                            
+                 end                            EXTENT := EXPR : EXPR                  
+        SIEVE := if EXPR                          CALL := EXPR(EXPR...)                
+                   STMT                         ACCESS := TENSOR[EXPR...]              
+                 end                              MODE := @mode(TENSOR)                
+        BLOCK := begin                     
+                   STMT...                 
+                 end                        
+    \end{minted}                                                   
+    \vspace{-12pt}
+\caption{The syntax of the finch language. Compare this grammar to the Concrete
+Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
+addition of multiple left hand sides through code blocks, access with expressions, and explicit initialiation.}\label{fig:syntax}
+\vspace{-12pt}
+\end{wrapfigure}
 
@@
 % pass.
-
-\begin{figure}
-\scriptsize
-\noindent % This ensures the minipages fill the page width
-\begin{minipage}{0.4\linewidth}
-\begin{align*}
-    \finchliteral(val \in \mathbb{V}) &:= \text{\mintinline{julia}{val}} \\
-    \finchvalue(ex \in \mathbb{S}, type \in \mathbb{T}) &:= \text{\mintinline{julia}{ex :: type}} \\
-    \finchtensor(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
-    \finchindex(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
-    \finchvar(name \in \mathbb{S}) &:= \text{\mintinline{julia}{name}} \\
-    \finchextent(a \in E, b \in E) &:= \text{\mintinline{julia}{a : b}} \\
-    \finchcall(f \in E, args\ldots \in E) &:= \text{\mintinline{julia}{f(args...)}} \\
-    \finchaccess(tns \in T, idxs\ldots \in E) &:= \text{\mintinline{julia}{tns[idxs...]}} \\
-    \finchfreeze(tns \in T) &:= \text{\mintinline{julia}{@freeze(tns)}} \\
-    \finchthaw(tns \in T) &:= \text{\mintinline{julia}{@thaw(tns)}} \\
-    \finchmode(tns \in T) &:= \text{\mintinline{julia}{@mode(tns)}} \\
-\end{align*}
-\end{minipage}%
-\begin{minipage}{0.5\linewidth}
-\begin{align*}
-    \mathbb{V} &:= \text{the set of all values.} \\
-    \mathbb{S} &:= \text{the set of all symbols.} \\
-    \mathbb{T} &:= \text{the set of all types.} \\
-    I &:= \finchindex \\
-    A &:= \finchaccess \\
-    V &:= \finchvar \\
-    T &:= \finchtensor \\
-    E &:= \finchliteral \mid \finchvalue \mid \finchindex \\
-        &\quad \mid \finchvar \mid \finchextent \mid \finchcall \mid \finchaccess \\
-    S &:= \finchassign \mid \finchloop \mid \finchdefine \\
-        &\quad \mid \finchsieve \mid \finchblock \\
-\end{align*}
-\end{minipage}%
-\begin{align*}
-    \finchdeclare(tns \in T, init \in E, dims\ldots \in E) &:= \text{\mintinline{julia}{tns .= init(dims...)}} \\
-    \finchassign(lhs \in A, op \in E, rhs \in E) &:= \text{\mintinline{julia}{lhs <<op>>= rhs}} \\
-    \finchloop(idx \in I, range \in E, body \in S) &:= \text{\mintinline{julia}{for idx = range; body end}} \\
-    \finchdefine(var \in V, val \in E, body \in S) &:= \text{\mintinline{julia}{let var = val; body end}} \\
-    \finchsieve(cond \in E, body \in S) &:= \text{\mintinline{julia}{if cond; body end}} \\
-    \finchblock(bodies\ldots \in S) &:= \text{\mintinline{julia}{begin; bodies... end}}
-\end{align*}
-\caption{The syntax of the finch language. Note that instead of a classic EBNF
-form, we have constrained the classes of subterms on the left hand side to allow
-for descriptive names of the arguments. Compare this grammar to the Concrete
-Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
-addition of multiple left hand sides through code blocks, and the explicit
-initialization and finalization of tensors.}\label{fig:syntax}
-\end{figure}
 

### 2024-04-02 sections/related.tex
Update on Overleaf.

@@
 %
-However, many SQL or dataframe systems rely on BTrees, columnar, or Hashtables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
+However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
 %

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 \begin{table}[htbp]
-    \footnotesize
+    \scriptsize
     \centering
@@
     \centering
-    \footnotesize
+    \scriptsize
     \caption{Transposed SpMV Sample Matrices}
@@
     \includegraphics[width=\linewidth]{Spmv_annotated.png}
+    \vspace{-18pt}
     \caption{Performance of SpMV by Finch format.}
@@
     \end{minipage}
+    \vspace{-12pt}
     \caption{Two approaches to erosion in Finch.
@@
 	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
+ \vspace{-12pt}
     \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
@@
 	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
+    \vspace{-12pt}
     \caption{Performance of graph apps across various tools.
@@
 \begin{align*}
+\footnotesize
     \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
@@
 \begin{align*}
+\footnotesize
 &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
@@
 \end{tabular}
+\vspace{-12pt}
 \caption{Performance of Finch Logic for common kernels.}

### 2024-04-02 sections/related.tex
Update on Overleaf.

@@
 %
-In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays an few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
+In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays and a few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
 %

### 2024-04-02 sections/abstract.tex
updates

@@
 \begin{abstract}
-\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.
-Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.
-However, arrays can only handle dense rectilinear integer grids.
-Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
-We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs.
-Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
-Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
-}
-
-\HIDE{
-Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
-Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
-Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler technology.
-Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
-However, a large portion of the data, either obtained from nature or generated by humans, have structure.
-
-Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
-So far this has failed in both scope of the ambition and execution of comparable features.
-Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
-Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data.
-Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.
-For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
-
-In this paper, we describe a compiler, Finch,
-which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
-programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays.
-Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
-sparse tensors, databases, and lossless compression.
-We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.}
-
 From FORTRAN to Numpy, arrays have revolutionized how we express computation. However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.

### 2024-04-02 sections/background.tex
Update on Overleaf.
updates

@@
 \section{Background On Looplets}
-Finch represents iteration patterns using Looplets, a language that decomposes
-datastructure iterators hierarchically. 
+Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
 %
-Looplets represent the control-flow
-structures needed to iterate over any given datastructure, or multiple
-datastructures simultaneously. 
+Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
 %
-In particular, looplets are good at lifting code
-to the highest loop level that it's needed and subdividing iteration
-hierarchically in coordinate space.
+In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space.
 %
-Because looplets are compiled with
-progressive lowering, structure-specific mathematical optimizations such as
-integrals, multiply by zero, etc. can be implemented using simple compiler
-passes like term rewriting and constant propagation during the intermediate
-lowering stages. \cite{ahrens_looplets_2023}
+Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
+can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages \cite{ahrens_looplets_2023}.
 
-In this work, we partially simplify the presentation of Looplets to focus on the
-semantics, rather than precise implementation. Since Finch is designed to be
-embedded into another language\saman{not clear}, we allow for several of the looplets to utilize
-functions that modify the state of that language. It is assumed that if a
-looplet introduces a variable to be used in a child looplet, the child looplet
-will not modify that variable.\saman{No context yet to understand this. Need to introduce what looplets are (Fig 2)!}
+The looplets are described in Figure~\ref{fig:looplets}.
+We simplify the presentation to focus on the semantics, rather than precise implementation.
+Several looplets introduce or modify variables in the scope of the target language.
+It is assumed that if a looplet introduces a variable to be used in a child looplet, the child looplet will not modify that variable.
 
@@
     \end{minipage}
-    \caption{The looplet language, as understood in a correct execution of a Finch program.}
+    \caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
 \end{figure}

### 2024-04-02 sections/cut.tex
updates

@@
+\HIDE{From FORTRAN to Numpy, arrays have revolutionized how we express computation.
+Arrays are the highest-performing datastructure with a long history of investment and innovation, from hardware support to compiler technology.
+However, arrays can only handle dense rectilinear integer grids.
+Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
+We describe a compiler, Finch, which adapts existing programs and interfaces to the structure and sparsity of the inputs.
+Finch enables programmers to capture complex, real-world data scenarios with the same productivity they expect from dense arrays.
+Our approach enables new loop optimizations across multiple domains, unifying techniques such as sparse tensors, databases, and lossless compression.
+}
+
+\HIDE{
+Of the nearly 44 zettabytes of data humans have gathered to date, most of it is collected, processed, and stored as multi-dimensional arrays.
+Multi-dimensional arrays are the most prominent data type from the first programming language FORTRAN to modern Numpy.
+Furthermore, multi-dimensional arrays are the highest-performing data structure with a long history of investment and innovation, from hardware support to compiler technology.
+Arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
+However, a large portion of the data, either obtained from nature or generated by humans, have structure.
+
+Over the years compiler researchers have valiantly attempted to put sparse tensor algebra on the same compiler transformation and code generation footing as dense tensor algebra and array codes.
+So far this has failed in both scope of the ambition and execution of comparable features.
+Real world data have structure beyond sparsity including runs of repeated values, dense blocks or diagonals, ragged edges, or symmetry.
+Thus, focusing only on sparsity limits the scope needed to handle the staggering variety of data.
+Second, the arrays and loops abstraction introduced in FORTRAN supports lot more complex control flow not supported by einsum based modern sparse array compilers.
+For example, complex control flow such as conditional branches and early exits are needed for many important computations such as graph analytics.
+
+In this paper, we describe a compiler, Finch,
+which adapts existing programs and interfaces to the structure and sparsity of the inputs. Finch enables
+programmers to capture real-world scenarios with structured data and complex control flow with the same productivity they expect from dense arrays.
+Our approach enables new loop optimizations across multiple domains, unifying techniques in areas such as
+sparse tensors, databases, and lossless compression.
+We hope to truly bring tensor algebra on structured data to the same compiler transformation and code generation footing as dense tensor algebra and array codes.}
 % \teo{

### 2024-04-02 sections/intro.tex
Update on Overleaf.
updates

--- 2024-03-29 sections/intro.tex
@@
 %
-Modern
-array programming languages such as NumPy, SciPy, MatLab, TensorFlow, PyTorch,
-and Halide have pushed the limits of productive data processing with arrays,
-fueling breakthroughs in machine learning, scientific computing, image
-processing, and more  \cite{harris_array_2020, virtanen_scipy_2020,
-moler_history_2020, abadi_tensorflow_2016,
-paszke_pytorch_2019,ragan-kelley_halide_2013}.
+Modern array programming languages such as NumPy~\cite{harris_array_2020},
+SciPy~\cite{virtanen_scipy_2020}, MatLab~\cite{moler_history_2020},
+TensorFlow~\cite{abadi_tensorflow_2016}, PyTorch~\cite{paszke_pytorch_2019}, and
+Halide~\cite{ragan-kelley_halide_2013} have pushed the limits of productive data
+processing with arrays, fueling breakthroughs in machine learning, scientific
+computing, image processing, and more.
 % EASY CUT
@@
 caching, prefetching, multiple outputs, scatters, gathers, vectorization,
-loop-carry-dependencies, and more. Several optimizations have been developed for
+loop-carry-dependencies, and more. A myriad of optimizations have been developed for
 dense arrays, such as loop fusion, loop tiling, loop unrolling, and loop
@@
 Special-purpose
-compilers like TACO, Taichi, StructTensor, or CoRa which support a select subset of structured data
+compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani2023compiling}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
 structures (only sparse, or only ragged arrays) must compromise by greatly
 constraining the classes of programs which they support, such as tensor
-contractions \cite{kjolstad_tensor_2019, hu_taichi_2019, ghorbani2023compiling, fegade_cora_2022}. 
+contractions.
 %
@@
   \end{tabular}
-  \caption{Feature support across various tools.}
+  \caption{Control flow support across various tools.}
   \label{tab:features}
@@
   \end{tabular}
-  \caption{Support for various data structures across tools. Finch supports \textbf{both} complex programs and complex data structures.}
+  \caption{Data structure support across various tools.  Finch supports \textbf{both} complex programs and complex data structures.}
   \label{tab:data_structures}
 \end{minipage}
+\vspace{-24pt}
 \end{table}

### 2024-04-02 sections/tensors.tex
Update on Overleaf.
updates

@@
 The $assemble$ function manages the allocation of new data in the tensor.
-
-\begin{wrapfigure}{R}{0.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
-    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
-    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
-    \label{fig:levelsvsfibers}
-\end{wrapfigure}
 
@@
 %postype(lvl)
+
+\begin{wrapfigure}{R}{0.5\linewidth}
+    \centering
+    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
+    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
+    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
+    \label{fig:levelsvsfibers}
+\end{wrapfigure}
 
@@
 \subsection{Scalars}
-
-\subsubsection{Sparse Scalars}
-\subsubsection{Early Break Scalars}
+Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
+%
+Scalars can also affect how the leaves are lowered.
+%
+Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
+%
+Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
+%
+We also provide early break scalars, which modify the stepper looplets to
+re-specialize the loop whenever a reduction into an early break scalar hits an
+annihilator.
+%
+We assume that we don't need to re-specialize other looplets because the stepper
+is the only one which repeats a non-constant number of times.
 
@@
     \hline
-    \textbf{Sequential Reading and Updating} \\
+    \textbf{Sequential Access Levels} \\
     \hline
@@
     \hline
-    \textbf{Random Access Reading and Updating} \\
+    \textbf{Random Access Levels} \\
     \hline
@@
     \hline
+    \textbf{Scalars} \\
+    \hline
+    \textbf{Scalar}:
+    A dense scalar that, unlike a variable, supports reduction.  \\
+    \textbf{SparseScalar}:
+    A scalar with a dirty bit which specializes on the fill value when it occurs.\\
+    \textbf{EarlyBreakScalar}:
+    A scalar which triggers early breaks in reductions whenever an annihilator is encountered.\\
+    \hline
     \end{tabular}

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
     \end{minipage}
+    \vspace{-12pt}
     \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
+    \vspace{-12pt}
 \end{figure}
@@
 	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
+    \vspace{-8pt}
     \caption{A comparison of several matrix multiplication algorithms between
@@
     \label{fig:spgemm}
+    \vspace{-12pt}
 \end{figure}
@@
     \end{minipage}
+    \vspace{-8pt}
     \caption{Finch row-major, column-major and symmetric SpMV Programs}
     \label{spmv_programs}
+    \vspace{-12pt}
 \end{wrapfigure}
@@
     \centering
+    \vspace{-12pt}
     \caption{SpMV Tensor Formats}
+    \vspace{-12pt}
     \label{spmv_tensor_formats}
@@
     \end{tabular}
+    \vspace{-8pt}
 \end{table}
@@
     \caption{Transposed SpMV Sample Matrices}
+    \vspace{-12pt}
     \label{tab:transposed_spmv_sample_matrices}
@@
 In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They generally either limit the dimensionality to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support the fusion of arbitrary operators which can have a drastic impact on performance as we show in Fig.
-(\kyle{PUT FIG HERE}).
-We believe that a flexible, online compiler like Finch which can manage the complexity of arbitrary operations between inputs with a wide variety of formats is the secret ingredient needed to make the array API performant for structured data.
-
-There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both the output and the algorithm.
-The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
-This approach results in a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
-To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user.
-This is implemented through 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which lowers Finch Logic programs to one or more Finch programs while making heuristic decisions about output format, loop ordering, and protocols.
+They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
+Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Fig.
+\ref{fig:fusion}.
+We believe that a flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the secret ingredient needed to make the array API performant for structured data.
+To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
+This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
 
@@
 The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that every operator (with the exception of $\finchtable$ and $\finchalias$ which are analogous to the scan operator in RA) takes in an indexed tensor and outputs an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate the order.
-This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.
-
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define queries and plans.
-A query is the assignment of the output of an expression to a name.
-We use this assignment to denote that we materialize an expression in memory.
-Later queries can then access the result of earlier queries through the alias operator which allows multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and it produces a set of tensors as output.
+This means that operators take in a set of indexed tensors and output an indexed tensor.
+Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
+To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
+A $\finchquery$ assigns the output of an expression to a name.
+We use this to denote that we materialize an expression.
+Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
+A plan is a sequence of queries and outputs a set of tensors.
 
 \begin{align*}
-\footnotesize
-    \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
+    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
+     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
+    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
 \end{align*}
@@
 \begin{align*}
-\footnotesize
-&\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
+\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
 \end{align*}
@@
 \vspace{-12pt}
+\label{fig:fusion}
 \caption{Performance of Finch Logic for common kernels.}

### 2024-04-02 main.tex
Update on Overleaf.
DO NOT USE THE CODE USE THE CORRECT CODE FOR YOU PAPER

@@
 
-\ifthenelse{1=1}{
+\ifthenelse{1=0}{
 \newcommand{\TODO}[1]{\textcolor{red}{TODO: #1}}
@@
   <concept>
-  <concept_id>00000000.0000000.0000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_id>10011007.10011006.10011008.10011024.10011027</concept_id>
+  <concept_desc>Software and its engineering~Control structures</concept_desc>
   <concept_significance>500</concept_significance>
@@
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
+  <concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
+  <concept_desc>Software and its engineering~Data types and structures</concept_desc>
+  <concept_significance>500</concept_significance>
+  </concept>
+  <concept>
+  <concept_id>10011007.10011006.10011008.10011009.10011010</concept_id>
+  <concept_desc>Software and its engineering~Imperative languages</concept_desc>
   <concept_significance>300</concept_significance>
@@
   <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
- </concept>
- <concept>
-  <concept_id>00000000.00000000.00000000</concept_id>
-  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-  <concept_significance>100</concept_significance>
+  <concept_id>10002950.10003705</concept_id>
+  <concept_desc>Mathematics of computing~Mathematical software</concept_desc>
+  <concept_significance>500</concept_significance>
   </concept>
@@
 
-\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-
+\ccsdesc[500]{Software and its engineering~Control structures}
+\ccsdesc[500]{Software and its engineering~Data types and structures}
+\ccsdesc[300]{Software and its engineering~Imperative languages}
+\ccsdesc[500]{Mathematics of computing~Mathematical software}
 %%
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Do, Not, Us, This, Code, Put, the, Correct, Terms, for,
-  Your, Paper}
+\keywords{Array, Sparse, Structure, Control Flow, Language, Compiler}
 
@@
 \input{sections/eval}
+\input{sections/related}
+\input{sections/conclusion}
 %%

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 Finch outperforms TACO on larger matrices, with an average speedup of 1.05.
-\changwan{(self)may remove this number for consistency with other exps.
-finch\_outer\_bytemap not explained.}
+
+%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
 
@@
 
-Given this language, we now describe how to define a few example functions from the API as plans in the Finch Logic language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can access an index as a scalar to implement $argmin$.
+We now describe how to define a few example functions from the API in this language.
+Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
 For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 \subsubsection{Finch Logic}
-The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
+The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
 This means that operators take in a set of indexed tensors and output an indexed tensor.
@@
 Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
+For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
 \begin{align*}
-\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
+&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
 \end{align*}

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 \end{align*}
-Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
-In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
-However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
 
 \subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting.
-This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e.
+Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
+The semantically important requirements of this standard form are 1) all inputs (i.e.
 tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
@@
 
-During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first.
-Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias.
-At this point, the first semantic requirement is satisfied.
+During the standardization process, a concordization pass is performed which examines each query, heuristically selects a loop order, and transposes inputs which don't match that order.
 Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required.

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 The last one guarantees that the innermost expression can be computed as a single kernel.
-
-During the standardization process, a concordization pass is performed which examines each query, heuristically selects a loop order, and transposes inputs which don't match that order.
-Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required.
-The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.
+To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
+In the future,
 

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In the future,
+In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
 
 \subsubsection{Finch Interpreter} 
-Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process.
-First, the output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to convert the pointwise expression into a Finch expression.
+Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
+The output format is identified by unpacking the outer $\finchreformat$ statement.
+Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to process the pointwise expression into a Finch expression.
 At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor.

### 2024-04-02 sections/eval.tex
Update on Overleaf.

@@
 \ref{fig:fusion}.
-We believe that a flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the secret ingredient needed to make the array API performant for structured data.
+A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the array API performant for structured data.
 To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
@@
 The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to process the pointwise expression into a Finch expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor.
+Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
+At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
 Lastly, the concordant loop order is identified and instantiated.
-The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
+The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
 
@@
 \begin{figure}
+\label{fig:fusion}
 \begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
@@
 \vspace{-12pt}
-\label{fig:fusion}
 \caption{Performance of Finch Logic for common kernels.}

### 2024-04-02 sections/appendix.tex
fixes

--- 2024-04-02 sections/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It's also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.05.
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner products necessarily requires $O(n^2)$ work and
-    TACO's outer products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of structures, making it an effective kernel to
-demonstrate the utility of flexible data formats.
-
-In this case study, we highlight different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
-Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g.
-sparsity \textit{and} symmetry).
-
-We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset in Figure \ref{spmv_grouped} is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-\subsubsection{Tensor Formats}
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
+\section{SpMV Datasets}
 
@@
   \scriptsize
-    \caption{Transposed SpMV Sample Matrices}
+  \caption{SpMV Sample Matrices}
   \vspace{-12pt}
-    \label{tab:transposed_spmv_sample_matrices}
+  \label{tab:spmv_sample_matrices}
   \begin{tabular}{|
@@
 \end{table}
-
-\subsubsection{Symmetric SpMV}
-Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix.
-We can apply this strategy with any level format.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used finch to implement two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et.
-al \cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure \ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05).
-Note that we did not include GAP-road as it took too long to run.
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Fig.
-\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluating Finch Logic}
-To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby statement in SQL.
-For the element-wise operations, we also provide an unfused Finch implementation to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Fig.
-\ref{graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-Across all three of these kernels, we see that the high level interface for Finch provides a significant improvement over DuckDB ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
-This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.
-For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$ before handling the disjunction on the indices implied by $A+B$.
-DuckDB handles the latter first which makes the computation linear in the size of $A+B$ rather than $C$.
-
-\begin{figure}
-\label{fig:fusion}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-02 sections/eval.tex
fixes

--- 2024-04-02 sections/eval.tex
@@
 \cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of structures, making it an effective kernel to
-demonstrate the utility of flexible data formats.
+results in a wide range of array structures, making it an effective kernel to
+demonstrate the utility of our programming model.
 
-In this case study, we highlight different Finch formats as specified in Table \ref{spmv_tensor_formats}, and the performance effects of conforming a dataset’s structure with its storage format, which Finch's datastructure-driven model enables us to do.
-Finch also provides the control flow necessary to manipulate data reads and writes, enabling exploitation of multiple structural patterns concurrently (e.g.
-sparsity \textit{and} symmetry).
+We varied both the data formats and the programs we used them with. Our formats
+are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
+Figure~\ref{spmv_programs}.
+
+In varying the program, we consider both row and column-major SpMV programs, as
+well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
+effectively. Our program restricts our attention to the canonical trial (using
+masks), reuses the reads to the canonical triangle of the symmetric matrix
+(using a $\finchdefine$ statement), and writes the results to both relevant
+locations (using multiple outputs).
+
+All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
 
@@
 Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created several synthetic matrices containing bands or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
+We also created some synthetic matrices containing bands of varying sizes.
+% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
 The dense vector is randomly generated.
@@
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
+
+We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
+Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
+Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
+There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
+For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
+Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
+Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
+Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
+
+%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
+
+%\subsubsection{Tensor Formats}
+
+%\subsubsection{Symmetric SpMV}
+% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
+%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
 
@@
 \end{table}
-
-\subsubsection{Tensor Formats}
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic Toeplitz banded matrices we constructed performed the best with the SparseBand matrix, in particular with the toeplitz\_large\_band and the toeplitz\_medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-
-\begin{table}[htbp]
-    \centering
-    \scriptsize
-    \caption{Transposed SpMV Sample Matrices}
-    \vspace{-12pt}
-    \label{tab:transposed_spmv_sample_matrices}
-    \begin{tabular}{|
-        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
-        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
-        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
-        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|
-        >{\centering\arraybackslash}m{0.2\linewidth-2\tabcolsep-1.2\arrayrulewidth}|}
-        \hline
-         & \textbf{HB/saylr4} & \textbf{Norris/heart3} & \textbf{large\_band} & \textbf{SNAP/as-735} \\
-        \hline
-        \textbf{Spy} & \includegraphics[width=\linewidth]{spmv_matrices/saylr4.png} & \includegraphics[width=\linewidth]{spmv_matrices/heart3.png} & \includegraphics[width=\linewidth]{spmv_matrices/toeplitz_large_band.png} & \includegraphics[width=\linewidth]{spmv_matrices/as-735.png} \\
-        \hline 
-        \textbf{Group / Name} & HB/saylr4 & Norris/heart3 & large\_band & SNAP/as-735 \\
-        \hline
-        \textbf{Dimensions} & 3,564 x 3,564 (22,316) & 2,339 x 2,339 (680,341) & 10,000 x 10,000 (1,999,900) & 7,716 x 7,716 (26,467) \\
-        \hline
-        \textbf{Best Finch Format} & Symmetric SparseList & SparseVBL & SparseBand & Symmetric SparseList-Pattern \\
-        \hline
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-\subsubsection{Symmetric SpMV}
-Finch enables us to exploit symmetry in the sparse matrix of the SpMV kernel by providing the capabilities to reuse memory reads and insert control flow logic to restrict iterations to either the lower or upper triangle of the sparse matrix.
-We can apply this strategy with any level format.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
 
@@
 \subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
+The expression fragment of the logical language takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
 This means that operators take in a set of indexed tensors and output an indexed tensor.
@@
 Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
+For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
 \begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
+\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
 \end{align*}
+Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
+In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
+However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
 
 \subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
+Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting.
+This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e.
 tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
@@
 The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In the future,
+
+During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first.
+Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias.
+At this point, the first semantic requirement is satisfied.
+Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required.
+The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.
 

### 2024-04-02 sections/eval.tex

@@
 Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer wrapping of $\finchplan(\finchquery(out,...),out)$.
+For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
 \begin{align*}
-\footnotesize &\text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-\footnotesize &\text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-\footnotesize &\text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
+&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
 \end{align*}
-Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
-In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
-However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
 
 \subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves the above questions about loop ordering and output formatting.
-This standard form has a few syntactic requirements, but the semantically important requirements are 1) all inputs (i.e.
+Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
+The semantically important requirements of this standard form are 1) all inputs (i.e.
 tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
@@
 The last one guarantees that the innermost expression can be computed as a single kernel.
-
-During the standardization process, a concordization pass is performed which examines each query in order and selects a loop order based on a heuristic which loops over intersecting variables first.
-Then, each query is examined again and a transposition query is inserted for each input which doesn't match the loop order and then the input is replaced with an alias.
-At this point, the first semantic requirement is satisfied.
-Next, a formatting pass is performed which examines each query and selects a level format for each output index based on the formats of the inputs and whether random writes are required.
-The procedure for this is a simple set of rules which attempts to aggressively preserve the structure present in the input tensors.
+To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
+In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
 
 \subsubsection{Finch Interpreter} 
-Once the Finch Logic program has been converted to the standard form, the Finch Interpreter can execute each query, in order, through a straightforward lowering process.
-First, the output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and to convert the pointwise expression into a Finch expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual tensor.
+Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
+The output format is identified by unpacking the outer $\finchreformat$ statement.
+Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
+At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
 Lastly, the concordant loop order is identified and instantiated.
-The lowered query can then be compiled and executed with the Finch compiler, and the result is assigned to $name$ in the plan's scope before proceeding to the next query in the plan.
+The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
 

### 2024-04-03 sections/eval.tex
Update on Overleaf.

@@
 \ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the array API performant for structured data.
+A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
 To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
@@
 
-\subsubsection{Evaluating Finch Logic}
-To demonstrate the performance of Finch Logic, we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies; 1) triangle counting on graphs 2) SDDMM 3) and element-wise operations.
+\subsubsection{Evaluation}
+To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
 Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby statement in SQL.
-For the element-wise operations, we also provide an unfused Finch implementation to show the impact of fusion.
+To do this, we express each of these kernels as a single select, join, groupby query.
+For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
 For triangle counting, we use the same set of graph matrices as in Fig.
@@
 
-Across all three of these kernels, we see that the high level interface for Finch provides a significant improvement over DuckDB ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries.
-This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.
-For the element-wise operations, this stems from Finch's ability to aggressively intersect with the indices of $C$ before handling the disjunction on the indices implied by $A+B$.
-DuckDB handles the latter first which makes the computation linear in the size of $A+B$ rather than $C$.
+Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
+For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
+For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
 

### 2024-04-03 sections/related.tex
updates
Update on Overleaf.

@@
 %
-Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a recursive FFT. 
+Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT. 
 %
-SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
+SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller FFT matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
 %
+%RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these
 

### 2024-04-03 sections/eval.tex
Update on Overleaf.

@@
 We used finch to implement two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et.
-al \cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
+Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
 

### 2024-04-03 sections/related.tex
update

@@
 
-The related work on array languages and libraries, both for dense and structured computation, spans several areas.
+The related work on array languages and libraries spans several areas, from libraries to languages, from dense to structured computation.
+
+\paragraph{Libraries for Dense Data:}
+Many libraries specialize in dense computations.
 %
-We highlight these works based on if they are a library or language vs. if they are dense or more general.
-
-\paragraph{Libraries for Dense Data:} There are many well known libraries that specialize in dense computations, exemplified by BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}.
-%
-This pattern is carried over into well known dense array libraries that use BLAS, exemplified by Numpy.
+Perhaps the most well-known example is Numpy~\cite{harris_array_2020},
+and a classic example is the BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}. 
 %
@@
 %
-Examples include Scipy, PETSc, Armadillo, OSKI, Cyclops, MKL (add cite), and Eigen~\cite{virtanen2020scipy, abhyankarpetsc, Rumengan2021, vuduc2005oski,solomonik2013cyclops, eigenweb}.
+Examples include Scipy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
 %
-There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt}.
+There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt} (MTTKRP on CSF).
 %
@@
 %
-GraphBlas exemplifies this pattern with support for many sparse matrix operations on a variety of formats~\cite{kepner2016mathematical} although a variety of other graph libraries adopt a similar approach such as LAgraph or GBase~\cite{mattson2019lagraph, kang2011gbase} among many others~\cite{ashari2014fast,huang2020ge}.
+The GraphBLAS~\cite{kepner2016mathematical} supports primitive semiring operations (operations beyond $(+, *)$, such as $(min, +)$ multiplication) which can be composed to enable graph algorithms, some of which are collected in LAGraph~\cite{mattson2019lagraph}.
+%
+Similarly, the Mapreduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
+algorithms in the GBASE library\cite{kang2011gbase}.
 %
@@
 %
-
 \paragraph{Compilers for Dense Data:}
@@
 %
-Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM, Exo, and Elevate~\cite{chen2018tvm, ikarashi2022exocompilation, hagedorn2020achieving}.
+Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
 %
@@
 %
-Several polyhedral languages, such as Polly, Tiramisu, CHiLL, Pluto, AlphaZ ~\cite{grosser2012polly, baghdadi2019tiramisu,chen2008framework, bondhugula2008pluto, yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
+Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu\cite{baghdadi2019tiramisu}, CHiLL\cite{chen2008framework}, Pluto\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
 %
@@
 %
-The TACO compiler originally supported just plain einsum computations, but has been extended several times to support (single dimensional) local tensors, breaks via semi-rings, windowing, tiling, and convolution~\cite{kjolstad_tensor_2017, kjolstad_tensor_2019, senanayake2020sparse, henry_compilation_2021,won2023unified}.
+The TACO compiler originally supported just plain einsum computations\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
 %
-Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures and tree like structures~\cite{kjolstad_tensor_2017, chou2018format, chou2022compilation}.
+Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions.
 %
-The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}.
+The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
 %
-Similarly, SDQL originally supported only hash tables, but has been extended with one format language that allows one to specify tensors as queries on several base storage types~\cite{schleich2023optimizing} and via another system that optimizing programs by taking advantage of predicates describing symmetries and other structures on a tensor~\cite{ghorbani2023compiling}.
+Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
 %
-Taichi is another language with a separate program and data structure specification, but Taichi focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
+%Taichi is another language with a separate program and data structure specification, but
+The Taichi language focuses on a single sparse data structure made from dense blocks, bit-masks, and pointers~\cite{hu_taichi_2019}.
 %
-Outside of these works, the window of comparison fogs up.
+%Outside of these works, the window of comparison fogs up.
 %
@@
 %
-However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectowrise, LaraDB, GMAP, or SciDB building physical layouts with other constructs based in array programming~\cite{boncz2012vectorwise,hutchison2017laradb,  tsatalos1996gmap, stonebraker2013scidb}.
+However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
 %
@@
 %
-Lastly, we mention SPIRAL, as a system that can express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT. 
-%
-SPIRAL focuses on recursively defined linear algebra and computations on the same (e.g. FFT matrices as tensor products of smaller FFT matrices)~\cite{franchetti2018spiral,franchetti2009operator}.
+Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT \cite{franchetti2018spiral,franchetti2009operator}.
 %
 %RECUMA AND UNISPARSE are in OOPSLA R1, we should probably address these
-
-%Stut and SDQL - similar but more workspaces and rings. Mention formats paper of theirs - similar to looplets, but to our knowledge doesn't support all of oru structures.
-%Tiachi - 
-%SPF - ????
-%Spiral
-%SQL - Vectorwise/Morpheus/HyPer and others?
 

### 2024-04-03 sections/related.tex
cool

@@
 
-\paragraph{Other Architectures:} We would be remiss if we didn't mention another axis: architectural support. 
+\paragraph{Other Architectures:}
+
+Both structured and dense libraries support a multitude of architectures, as do dense compilers.
+
+Sparse compilers have been extended to many architectures.
+An extension of TACO supports GPU \cite{senanayake2020sparse}, Cyclops\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine\cite{hsu_sparse_2022} supports custom hardware.
+
+We believe that supporting control flow is the first step towards architectural support for complex structures.
+
+%Sparse and structured com We would be remiss if we didn't mention another axis: architectural support. 
 %
-We saved this angle for future work, but we must note that much of this work varies in support for other architectures, which is arguably another element of control flow.
+%We saved this angle for future work, but we must note that much of this work varies in support for other architectures, which is arguably another element of control flow.
 %%TACO

### 2024-04-03 sections/eval.tex
Update on Overleaf.

@@
 %
-Inner products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
+Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
 
@@
 %
-It's also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
+It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
 
 As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.05.
+Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
+%arith-mean
 
@@
     Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner products necessarily requires $O(n^2)$ work and
-    TACO's outer products format is dense. Finch can use a sparse outer products
+    dimension. Note that inner-products necessarily requires $O(MN)$ work and
+    TACO's outer-products format is dense. Finch can use a sparse outer products
     format and thus has an asymptotic advantage that becomes evident as the
@@
 
-We used finch to implement two fundamental graph algorithms: Breadth-First Search (BFS) and Bellman-Ford single-source shortest path.
+We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
 Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
@@
 
-Figure \ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS\cite{mattson2019lagraph}.
+Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
 For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
 Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 1.05).
+On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
 Note that we did not include GAP-road as it took too long to run.
+%speedup is arith-mean
 

### 2024-04-03 sections/related.tex
Update on Overleaf.

@@
 %
-Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu\cite{baghdadi2019tiramisu}, CHiLL\cite{chen2008framework}, Pluto\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
+Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu~\cite{baghdadi2019tiramisu}, CHiLL~\cite{chen2008framework}, Pluto~\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
 %
@@
 %
-The TACO compiler originally supported just plain einsum computations\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
+The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
 %
@@
 %
-However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
+However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
 %
@@
 %
-Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT \cite{franchetti2018spiral,franchetti2009operator}.
+Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti2018spiral,franchetti2009operator}.
 %
@@
 Sparse compilers have been extended to many architectures.
-An extension of TACO supports GPU \cite{senanayake2020sparse}, Cyclops\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine\cite{hsu_sparse_2022} supports custom hardware.
+An extension of TACO supports GPU~\cite{senanayake2020sparse}, Cyclops~\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal~\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine~\cite{hsu_sparse_2022} supports custom hardware.
 

### 2024-04-04 sections/appendix.tex
fixes

@@
 \end{table}
+
+\section{Graph Algorithms}
+
+\subsection{Finch Breadth-First Search}
+\begin{minted}{julia}
+    function bfs_finch_kernel(edges, edgesT, source=5, alpha = 0.01)
+    (n, m) = size(edges)
+    edges = pattern!(edges)
+    @assert n == m
+    F = Tensor(SparseByteMap(Pattern()), n)
+    _F = Tensor(SparseByteMap(Pattern()), n)
+    @finch F[source] = true
+    F_nnz = 1 
+    V = Tensor(Dense(Element(false)), n)
+    @finch V[source] = true
+    P = Tensor(Dense(Element(0)), n)
+    @finch P[source] = source
+    while F_nnz > 0 
+        if F_nnz/m > alpha # pull
+            p = ShortCircuitScalar{0}()
+            _F .= false
+            for k=_ 
+                if !V[k]
+                    p .= 0
+                    for j=_ 
+                        if F[follow(j)] && AT[j, k]
+                            p[] <<choose(0)>>= j
+                        end 
+                    end 
+                    if p[] != 0
+                        _F[k] |= true
+                        P[k] = p[] 
+                    end 
+                end 
+            end
+        else # push
+            _F .= false
+            for j=_, k=_ 
+                if F[j] && A[k, j] && !(V[k])
+                    _F[k] |= true
+                    P[k] <<choose(0)>>= j
+                end 
+            end 
+        end 
+        c = Scalar(0)
+        @finch begin
+            for k=_ 
+                let _f = _F[k]
+                    V[k] |= _f
+                    c[] += _f
+                end 
+            end 
+        end 
+        (F, _F) = (_F, F)
+        F_nnz = c[] 
+    end 
+    return P
+end
+\end{minted}
+\subsection{GraphBLAS Breadth-First Search}
+\begin{minted}{c}
+//------------------------------------------------------------------------------
+// LG_BreadthFirstSearch_SSGrB:  BFS using Suitesparse extensions
+//------------------------------------------------------------------------------
+
+// LAGraph, (c) 2019-2022 by The LAGraph Contributors, All Rights Reserved.
+// SPDX-License-Identifier: BSD-2-Clause
+//
+// For additional details (including references to third party source code and
+// other files) see the LICENSE file or contact permission@sei.cmu.edu. See
+// Contributors.txt for a full list of contributors. Created, in part, with
+// funding and support from the U.S. Government (see Acknowledgments.txt file).
+// DM22-0790
+
+// Contributed by Timothy A. Davis, Texas A&M University
+
+//------------------------------------------------------------------------------
+
+// This is an Advanced algorithm.  G->AT and G->out_degree are required for
+// this method to use push-pull optimization.  If not provided, this method
+// defaults to a push-only algorithm, which can be slower.  This is not
+// user-callable (see LAGr_BreadthFirstSearch instead).  G->AT and
+// G->out_degree are not computed if not present.
+
+// References:
+//
+// Carl Yang, Aydin Buluc, and John D. Owens. 2018. Implementing Push-Pull
+// Efficiently in GraphBLAS. In Proceedings of the 47th International
+// Conference on Parallel Processing (ICPP 2018). ACM, New York, NY, USA,
+// Article 89, 11 pages. DOI: https://doi.org/10.1145/3225058.3225122
+//
+// Scott Beamer, Krste Asanovic and David A. Patterson, The GAP Benchmark
+// Suite, http://arxiv.org/abs/1508.03619, 2015.  http://gap.cs.berkeley.edu/
+
+// revised by Tim Davis (davis@tamu.edu), Texas A&M University
+
+#define LG_FREE_WORK        \
+{                           \
+    GrB_free (&w) ;         \
+    GrB_free (&q) ;         \
+}
+
+#define LG_FREE_ALL         \
+{                           \
+    LG_FREE_WORK ;          \
+    GrB_free (&pi) ;        \
+    GrB_free (&v) ;         \
+}
+
+#include "LG_internal.h"
+
+int LG_BreadthFirstSearch_SSGrB
+(
+    GrB_Vector *level,
+    GrB_Vector *parent,
+    const LAGraph_Graph G,
+    GrB_Index src,
+    char *msg
+)
+{
+
+    //--------------------------------------------------------------------------
+    // check inputs
+    //--------------------------------------------------------------------------
+
+    LG_CLEAR_MSG ;
+    GrB_Vector q = NULL ;           // the current frontier
+    GrB_Vector w = NULL ;           // to compute work remaining
+    GrB_Vector pi = NULL ;          // parent vector
+    GrB_Vector v = NULL ;           // level vector
+
+#if !LAGRAPH_SUITESPARSE
+    LG_ASSERT (false, GrB_NOT_IMPLEMENTED) ;
+#else
+
+    bool compute_level  = (level != NULL) ;
+    bool compute_parent = (parent != NULL) ;
+    if (compute_level ) (*level ) = NULL ;
+    if (compute_parent) (*parent) = NULL ;
+    LG_ASSERT_MSG (compute_level || compute_parent, GrB_NULL_POINTER,
+        "either level or parent must be non-NULL") ;
+
+    LG_TRY (LAGraph_CheckGraph (G, msg)) ;
+
+    //--------------------------------------------------------------------------
+    // get the problem size and cached properties
+    //--------------------------------------------------------------------------
+
+    GrB_Matrix A = G->A ;
+
+    GrB_Index n, nvals ;
+    GRB_TRY (GrB_Matrix_nrows (&n, A)) ;
+    LG_ASSERT_MSG (src < n, GrB_INVALID_INDEX, "invalid source node") ;
+
+    GRB_TRY (GrB_Matrix_nvals (&nvals, A)) ;
+
+    GrB_Matrix AT = NULL ;
+    GrB_Vector Degree = G->out_degree ;
+    if (G->kind == LAGraph_ADJACENCY_UNDIRECTED ||
+       (G->kind == LAGraph_ADJACENCY_DIRECTED &&
+        G->is_symmetric_structure == LAGraph_TRUE))
+    {
+        // AT and A have the same structure and can be used in both directions
+        AT = G->A ;
+    }
+    else
+    {
+        // AT = A' is different from A.  If G->AT is NULL, then a push-only
+        // method is used.
+        AT = G->AT ;
+    }
+
+    // direction-optimization requires G->AT (if G is directed) and
+    // G->out_degree (for both undirected and directed cases)
+    bool push_pull = (Degree != NULL && AT != NULL) ;
+
+    // determine the semiring type
+    GrB_Type int_type = (n > INT32_MAX) ? GrB_INT64 : GrB_INT32 ;
+    GrB_Semiring semiring ;
+
+    if (compute_parent)
+    {
+        // use the ANY_SECONDI_INT* semiring: either 32 or 64-bit depending on
+        // the # of nodes in the graph.
+        semiring = (n > INT32_MAX) ?
+            GxB_ANY_SECONDI_INT64 : GxB_ANY_SECONDI_INT32 ;
+
+        // create the parent vector.  pi(i) is the parent id of node i
+        GRB_TRY (GrB_Vector_new (&pi, int_type, n)) ;
+        GRB_TRY (GxB_set (pi, GxB_SPARSITY_CONTROL, GxB_BITMAP + GxB_FULL)) ;
+        // pi (src) = src denotes the root of the BFS tree
+        GRB_TRY (GrB_Vector_setElement (pi, src, src)) ;
+
+        // create a sparse integer vector q, and set q(src) = src
+        GRB_TRY (GrB_Vector_new (&q, int_type, n)) ;
+        GRB_TRY (GrB_Vector_setElement (q, src, src)) ;
+    }
+    else
+    {
+        // only the level is needed, use the LAGraph_any_one_bool semiring
+        semiring = LAGraph_any_one_bool ;
+
+        // create a sparse boolean vector q, and set q(src) = true
+        GRB_TRY (GrB_Vector_new (&q, GrB_BOOL, n)) ;
+        GRB_TRY (GrB_Vector_setElement (q, true, src)) ;
+    }
+
+    if (compute_level)
+    {
+        // create the level vector. v(i) is the level of node i
+        // v (src) = 0 denotes the source node
+        GRB_TRY (GrB_Vector_new (&v, int_type, n)) ;
+        GRB_TRY (GxB_set (v, GxB_SPARSITY_CONTROL, GxB_BITMAP + GxB_FULL)) ;
+        GRB_TRY (GrB_Vector_setElement (v, 0, src)) ;
+    }
+
+    // workspace for computing work remaining
+    GRB_TRY (GrB_Vector_new (&w, GrB_INT64, n)) ;
+
+    GrB_Index nq = 1 ;          // number of nodes in the current level
+    double alpha = 8.0 ;
+    double beta1 = 8.0 ;
+    double beta2 = 512.0 ;
+    int64_t n_over_beta1 = (int64_t) (((double) n) / beta1) ;
+    int64_t n_over_beta2 = (int64_t) (((double) n) / beta2) ;
+
+    //--------------------------------------------------------------------------
+    // BFS traversal and label the nodes
+    //--------------------------------------------------------------------------
+
+    bool do_push = true ;       // start with push
+    GrB_Index last_nq = 0 ;
+    int64_t edges_unexplored = nvals ;
+    bool any_pull = false ;     // true if any pull phase has been done
+
+    // {!mask} is the set of unvisited nodes
+    GrB_Vector mask = (compute_parent) ? pi : v ;
+
+    for (int64_t nvisited = 1, k = 1 ; nvisited < n ; nvisited += nq, k++)
+    {
+
+        //----------------------------------------------------------------------
+        // select push vs pull
+        //----------------------------------------------------------------------
+
+        if (push_pull)
+        {
+            if (do_push)
+            {
+                // check for switch from push to pull
+                bool growing = nq > last_nq ;
+                bool switch_to_pull = false ;
+                if (edges_unexplored < n)
+                {
+                    // very little of the graph is left; disable the pull
+                    push_pull = false ;
+                }
+                else if (any_pull)
+                {
+                    // once any pull phase has been done, the # of edges in the
+                    // frontier has no longer been tracked.  But now the BFS
+                    // has switched back to push, and we're checking for yet
+                    // another switch to pull.  This switch is unlikely, so
+                    // just keep track of the size of the frontier, and switch
+                    // if it starts growing again and is getting big.
+                    switch_to_pull = (growing && nq > n_over_beta1) ;
+                }
+                else
+                {
+                    // update the # of unexplored edges
+                    // w<q>=Degree
+                    // w(i) = outdegree of node i if node i is in the queue
+                    GRB_TRY (GrB_assign (w, q, NULL, Degree, GrB_ALL, n,
+                        GrB_DESC_RS)) ;
+                    // edges_in_frontier = sum (w) = # of edges incident on all
+                    // nodes in the current frontier
+                    int64_t edges_in_frontier = 0 ;
+                    GRB_TRY (GrB_reduce (&edges_in_frontier, NULL,
+                        GrB_PLUS_MONOID_INT64, w, NULL)) ;
+                    edges_unexplored -= edges_in_frontier ;
+                    switch_to_pull = growing &&
+                        (edges_in_frontier > (edges_unexplored / alpha)) ;
+                }
+                if (switch_to_pull)
+                {
+                    // switch from push to pull
+                    do_push = false ;
+                }
+            }
+            else
+            {
+                // check for switch from pull to push
+                bool shrinking = nq < last_nq ;
+                if (shrinking && (nq <= n_over_beta2))
+                {
+                    // switch from pull to push
+                    do_push = true ;
+                }
+            }
+            any_pull = any_pull || (!do_push) ;
+        }
+
+        //----------------------------------------------------------------------
+        // q = kth level of the BFS
+        //----------------------------------------------------------------------
+
+        int sparsity = do_push ? GxB_SPARSE : GxB_BITMAP ;
+        GRB_TRY (GxB_set (q, GxB_SPARSITY_CONTROL, sparsity)) ;
+
+        // mask is pi if computing parent, v if computing just level
+        if (do_push)
+        {
+            // push (saxpy-based vxm):  q'{!mask} = q'*A
+            GRB_TRY (GrB_vxm (q, mask, NULL, semiring, q, A, GrB_DESC_RSC)) ;
+        }
+        else
+        {
+            // pull (dot-product-based mxv):  q{!mask} = AT*q
+            GRB_TRY (GrB_mxv (q, mask, NULL, semiring, AT, q, GrB_DESC_RSC)) ;
+        }
+
+        //----------------------------------------------------------------------
+        // done if q is empty
+        //----------------------------------------------------------------------
+
+        last_nq = nq ;
+        GRB_TRY (GrB_Vector_nvals (&nq, q)) ;
+        if (nq == 0)
+        {
+            break ;
+        }
+
+        //----------------------------------------------------------------------
+        // assign parents/levels
+        //----------------------------------------------------------------------
+
+        if (compute_parent)
+        {
+            // q(i) currently contains the parent id of node i in tree.
+            // pi{q} = q
+            GRB_TRY (GrB_assign (pi, q, NULL, q, GrB_ALL, n, GrB_DESC_S)) ;
+        }
+        if (compute_level)
+        {
+            // v{q} = k, the kth level of the BFS
+            GRB_TRY (GrB_assign (v, q, NULL, k, GrB_ALL, n, GrB_DESC_S)) ;
+        }
+    }
+
+    //--------------------------------------------------------------------------
+    // free workspace and return result
+    //--------------------------------------------------------------------------
+
+    if (compute_parent) (*parent) = pi ;
+    if (compute_level ) (*level ) = v ;
+    LG_FREE_WORK ;
+    return (GrB_SUCCESS) ;
+#endif
+}
+\end{minted}
+\subsection{Finch Bellman-Ford}
+\begin{minted}{julia}
+function bellmanford_finch_kernel(edges, source=1)
+    (n, m) = size(edges)
+    @assert n == m
+    dists_prev = Tensor(Dense(Element(Inf)), n)
+    dists_prev[source] = 0 
+    dists = Tensor(Dense(Element(Inf)), n)
+    active_prev = Tensor(SparseByteMap(Pattern()), n)
+    active_prev[source] = true
+    active = Tensor(SparseByteMap(Pattern()), n)
+    parents = Tensor(Dense(Element(0)), n)
+    for iter = 1:n  
+        @finch begin
+            for j=_ 
+                if active_prev[j]
+                    dists[j] <<min>>= dists_prev[j]
+                end 
+            end 
+        end 
+        @finch begin
+            active .= false
+            for j = _ 
+                if active_prev[j]
+                    for i = _ 
+                        let d = dists_prev[j] + edges[i, j]
+                            dists[i] <<min>>= d
+                            active[i] |= d < dists_prev[i]
+                        end 
+                    end 
+                end 
+            end 
+        end 
+        if countstored(active) == 0
+            break
+        end 
+        dists_prev, dists = dists, dists_prev
+        active_prev, active = active, active_prev
+    end 
+    @finch begin
+        for j = _ 
+            for i = _ 
+                let d = edges[i, j]
+                    if d < Inf && dists[j] + d <= dists[i]
+                        parents[i] <<choose(0)>>= j
+                    end 
+                end 
+            end 
+        end 
+    end 
+    return (dists=dists, parents=parents)
+end
+\end{minted}
+
+\subsection{GraphBLAS Bellman-Ford}
+\begin{minted}{c}
+//------------------------------------------------------------------------------
+// LAGraph_BF_full1a.c: Bellman-Ford single-source shortest paths, returns tree,
+// while diagonal of input matrix A needs not to be explicit 0
+//------------------------------------------------------------------------------
+
+// LAGraph, (c) 2019-2022 by The LAGraph Contributors, All Rights Reserved.
+// SPDX-License-Identifier: BSD-2-Clause
+//
+// For additional details (including references to third party source code and
+// other files) see the LICENSE file or contact permission@sei.cmu.edu. See
+// Contributors.txt for a full list of contributors. Created, in part, with
+// funding and support from the U.S. Government (see Acknowledgments.txt file).
+// DM22-0790
+
+// Contributed by Jinhao Chen and Timothy A. Davis, Texas A&M University
+
+//------------------------------------------------------------------------------
+
+// This is the fastest variant that computes both the parent & the path length.
+
+// LAGraph_BF_full1a: Bellman-Ford single source shortest paths, returning both
+// the path lengths and the shortest-path tree.
+
+// LAGraph_BF_full performs a Bellman-Ford to find out shortest path, parent
+// nodes along the path and the hops (number of edges) in the path from given
+// source vertex s in the range of [0, n) on graph given as matrix A with size
+// n*n. The sparse matrix A has entry A(i, j) if there is an edge from vertex i
+// to vertex j with weight w, then A(i, j) = w.
+
+// LAGraph_BF_full1a returns GrB_SUCCESS if it succeeds.  In this case, there
+// are no negative-weight cycles in the graph, and d, pi, and h are returned.
+// The vector d has d(k) as the shortest distance from s to k. pi(k) = p+1,
+// where p is the parent node of k-th node in the shortest path. In particular,
+// pi(s) = 0. h(k) = hop(s, k), the number of edges from s to k in the shortest
+// path.
+
+// If the graph has a negative-weight cycle, GrB_NO_VALUE is returned, and the
+// GrB_Vectors d(k), pi(k) and h(k)  (i.e., *pd_output, *ppi_output and
+// *ph_output respectively) will be NULL when negative-weight cycle detected.
+
+// Otherwise, other errors such as GrB_OUT_OF_MEMORY, GrB_INVALID_OBJECT, and
+// so on, can be returned, if these errors are found by the underlying
+// GrB_* functions.
+
+//------------------------------------------------------------------------------
+
+#define LG_FREE_WORK                   \
+{                                      \
+    GrB_free(&d);                      \
+    GrB_free(&dmasked);                \
+    GrB_free(&dless);                  \
+    GrB_free(&Atmp);                   \
+    GrB_free(&BF_Tuple3);              \
+    GrB_free(&BF_lMIN_Tuple3);         \
+    GrB_free(&BF_PLUSrhs_Tuple3);      \
+    GrB_free(&BF_LT_Tuple3);           \
+    GrB_free(&BF_lMIN_Tuple3_Monoid);  \
+    GrB_free(&BF_lMIN_PLUSrhs_Tuple3); \
+    LAGraph_Free ((void**)&I, NULL);   \
+    LAGraph_Free ((void**)&J, NULL);   \
+    LAGraph_Free ((void**)&w, NULL);   \
+    LAGraph_Free ((void**)&W, NULL);   \
+    LAGraph_Free ((void**)&h, NULL);   \
+    LAGraph_Free ((void**)&pi, NULL);  \
+}
+
+#define LG_FREE_ALL                    \
+{                                      \
+    LG_FREE_WORK ;                     \
+    GrB_free (pd_output);              \
+    GrB_free (ppi_output);             \
+    GrB_free (ph_output);              \
+}
+
+#include <LAGraph.h>
+#include <LAGraphX.h>
+#include <LG_internal.h>  // from src/utility
+
+typedef void (*LAGraph_binary_function) (void *, const void *, const void *) ;
+
+//------------------------------------------------------------------------------
+// data type for each entry of the adjacent matrix A and "distance" vector d;
+// <INFINITY,INFINITY,INFINITY> corresponds to nonexistence of a path, and
+// the value  <0, 0, NULL> corresponds to a path from a vertex to itself
+//------------------------------------------------------------------------------
+
+typedef struct
+{
+    double w;    // w  corresponds to a path weight.
+    GrB_Index h; // h  corresponds to a path size or number of hops.
+    GrB_Index pi;// pi corresponds to the penultimate vertex along a path.
+                 // vertex indexed as 1, 2, 3, ... , V, and pi = 0 (as nil)
+                 // for u=v, and pi = UINT64_MAX (as inf) for (u,v) not in E
+}
+BF_Tuple3_struct;
+
+//------------------------------------------------------------------------------
+// binary functions, z=f(x,y), where Tuple3xTuple3 -> Tuple3
+//------------------------------------------------------------------------------
+
+void BF_lMIN3
+(
+    BF_Tuple3_struct *z,
+    const BF_Tuple3_struct *x,
+    const BF_Tuple3_struct *y
+)
+{
+    if (x->w < y->w
+        || (x->w == y->w && x->h < y->h)
+        || (x->w == y->w && x->h == y->h && x->pi < y->pi))
+    {
+        if (z != x) { *z = *x; }
+    }
+    else
+    {
+        *z = *y;
+    }
+}
+
+void BF_PLUSrhs3
+(
+    BF_Tuple3_struct *z,
+    const BF_Tuple3_struct *x,
+    const BF_Tuple3_struct *y
+)
+{
+    z->w = x->w + y->w ;
+    z->h = x->h + y->h ;
+    z->pi = (x->pi != UINT64_MAX && y->pi != 0) ?  y->pi : x->pi ;
+}
+
+void BF_LT3
+(
+    bool *z,
+    const BF_Tuple3_struct *x,
+    const BF_Tuple3_struct *y
+)
+{
+    (*z) = (x->w < y->w
+        || (x->w == y->w && x->h < y->h)
+        || (x->w == y->w && x->h == y->h && x->pi < y->pi)) ;
+}
+
+// Given a n-by-n adjacency matrix A and a source vertex s.
+// If there is no negative-weight cycle reachable from s, return the distances
+// of shortest paths from s and parents along the paths as vector d. Otherwise,
+// returns d=NULL if there is a negtive-weight cycle.
+// pd_output is pointer to a GrB_Vector, where the i-th entry is d(s,i), the
+//   sum of edges length in the shortest path
+// ppi_output is pointer to a GrB_Vector, where the i-th entry is pi(i), the
+//   parent of i-th vertex in the shortest path
+// ph_output is pointer to a GrB_Vector, where the i-th entry is h(s,i), the
+//   number of edges from s to i in the shortest path
+// A has weights on corresponding entries of edges
+// s is given index for source vertex
+GrB_Info LAGraph_BF_full1a
+(
+    GrB_Vector *pd_output,      //the pointer to the vector of distance
+    GrB_Vector *ppi_output,     //the pointer to the vector of parent
+    GrB_Vector *ph_output,      //the pointer to the vector of hops
+    const GrB_Matrix A,         //matrix for the graph
+    const GrB_Index s           //given index of the source
+)
+{
+    GrB_Info info;
+    char *msg = NULL ;
+    // tmp vector to store distance vector after n (i.e., V) loops
+    GrB_Vector d = NULL, dmasked = NULL, dless = NULL;
+    GrB_Matrix Atmp = NULL;
+    GrB_Type BF_Tuple3;
+
+    GrB_BinaryOp BF_lMIN_Tuple3;
+    GrB_BinaryOp BF_PLUSrhs_Tuple3;
+    GrB_BinaryOp BF_LT_Tuple3;
+
+    GrB_Monoid BF_lMIN_Tuple3_Monoid;
+    GrB_Semiring BF_lMIN_PLUSrhs_Tuple3;
+
+    GrB_Index nrows, ncols, n, nz;  // n = # of row/col, nz = # of nnz in graph
+    GrB_Index *I = NULL, *J = NULL; // for col/row indices of entries from A
+    GrB_Index *h = NULL, *pi = NULL;
+    double *w = NULL;
+    BF_Tuple3_struct *W = NULL;
+
+    if (pd_output  != NULL) *pd_output  = NULL;
+    if (ppi_output != NULL) *ppi_output = NULL;
+    if (ph_output  != NULL) *ph_output  = NULL;
+
+    LG_ASSERT (A != NULL && pd_output != NULL &&
+        ppi_output != NULL && ph_output != NULL, GrB_NULL_POINTER) ;
+
+    GRB_TRY (GrB_Matrix_nrows (&nrows, A)) ;
+    GRB_TRY (GrB_Matrix_ncols (&ncols, A)) ;
+    GRB_TRY (GrB_Matrix_nvals (&nz, A));
+    LG_ASSERT_MSG (nrows == ncols, -1002, "A must be square") ;
+    n = nrows;
+    LG_ASSERT_MSG (s < n, GrB_INVALID_INDEX, "invalid source node") ;
+
+    //--------------------------------------------------------------------------
+    // create all GrB_Type GrB_BinaryOp GrB_Monoid and GrB_Semiring
+    //--------------------------------------------------------------------------
+    // GrB_Type
+    GRB_TRY (GrB_Type_new(&BF_Tuple3, sizeof(BF_Tuple3_struct)));
+
+    // GrB_BinaryOp
+    GRB_TRY (GrB_BinaryOp_new(&BF_LT_Tuple3,
+        (LAGraph_binary_function) (&BF_LT3), GrB_BOOL, BF_Tuple3, BF_Tuple3));
+    GRB_TRY (GrB_BinaryOp_new(&BF_lMIN_Tuple3,
+        (LAGraph_binary_function) (&BF_lMIN3), BF_Tuple3, BF_Tuple3,BF_Tuple3));
+    GRB_TRY (GrB_BinaryOp_new(&BF_PLUSrhs_Tuple3,
+        (LAGraph_binary_function)(&BF_PLUSrhs3),
+        BF_Tuple3, BF_Tuple3, BF_Tuple3));
+
+    // GrB_Monoid
+    BF_Tuple3_struct BF_identity = (BF_Tuple3_struct) { .w = INFINITY,
+        .h = UINT64_MAX, .pi = UINT64_MAX };
+    GRB_TRY (GrB_Monoid_new_UDT(&BF_lMIN_Tuple3_Monoid, BF_lMIN_Tuple3,
+        &BF_identity));
+
+    //GrB_Semiring
+    GRB_TRY (GrB_Semiring_new(&BF_lMIN_PLUSrhs_Tuple3,
+        BF_lMIN_Tuple3_Monoid, BF_PLUSrhs_Tuple3));
+
+    //--------------------------------------------------------------------------
+    // allocate arrays used for tuplets
+    //--------------------------------------------------------------------------
+#if 1
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &I, nz, sizeof(GrB_Index), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &J, nz, sizeof(GrB_Index), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &w, nz, sizeof(double), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &W, nz, sizeof(BF_Tuple3_struct),
+        msg)) ;
+
+    //--------------------------------------------------------------------------
+    // create matrix Atmp based on A, while its entries become BF_Tuple3 type
+    //--------------------------------------------------------------------------
+
+    GRB_TRY (GrB_Matrix_extractTuples_FP64(I, J, w, &nz, A));
+    int nthreads, nthreads_outer, nthreads_inner ;
+    LG_TRY (LAGraph_GetNumThreads (&nthreads_outer, &nthreads_inner, msg)) ;
+    nthreads = nthreads_outer * nthreads_inner ;
+    printf ("nthreads %d\n", nthreads) ;
+    int64_t k;
+    #pragma omp parallel for num_threads(nthreads) schedule(static)
+    for (k = 0; k < nz; k++)
+    {
+        W[k] = (BF_Tuple3_struct) { .w = w[k], .h = 1, .pi = I[k] + 1 };
+    }
+    GRB_TRY (GrB_Matrix_new(&Atmp, BF_Tuple3, n, n));
+    GRB_TRY (GrB_Matrix_build_UDT(Atmp, I, J, W, nz, BF_lMIN_Tuple3));
+    LAGraph_Free ((void**)&I, NULL);
+    LAGraph_Free ((void**)&J, NULL);
+    LAGraph_Free ((void**)&W, NULL);
+    LAGraph_Free ((void**)&w, NULL);
+
+#else
+
+    todo: GraphBLAS could use a new kind of unary operator, not z=f(x), but
+
+    [z,flag] = f (aij, i, j, k, nrows, ncols, nvals, etc, ...)
+    flag: keep or discard.  Combines GrB_apply and GxB_select.
+
+    builtins:
+        f(...) =
+            i, bool is true
+            j, bool is true
+            i+j*nrows, etc.
+            k
+            tril, triu (like GxB_select): return aij, and true/false boolean
+
+        z=f(x,i).  x: double, z:tuple3, i:GrB_Index with the row index of x
+        // z = (BF_Tuple3_struct) { .w = x, .h = 1, .pi = i + 1 };
+
+    GrB_apply (Atmp, op, A, ...)
+
+    in the BFS, this is used:
+        op:  z = f ( .... ) = i
+        to replace x(i) with i
+
+#endif
+
+    //--------------------------------------------------------------------------
+    // create and initialize "distance" vector d, dmasked and dless
+    //--------------------------------------------------------------------------
+    GRB_TRY (GrB_Vector_new(&d, BF_Tuple3, n));
+    // make d dense
+    GRB_TRY (GrB_Vector_assign_UDT(d, NULL, NULL, (void*)&BF_identity,
+        GrB_ALL, n, NULL));
+    // initial distance from s to itself
+    BF_Tuple3_struct d0 = (BF_Tuple3_struct) { .w = 0, .h = 0, .pi = 0 };
+    GRB_TRY (GrB_Vector_setElement_UDT(d, &d0, s));
+
+    // creat dmasked as a sparse vector with only one entry at s
+    GRB_TRY (GrB_Vector_new(&dmasked, BF_Tuple3, n));
+    GRB_TRY (GrB_Vector_setElement_UDT(dmasked, &d0, s));
+
+    // create dless
+    GRB_TRY (GrB_Vector_new(&dless, GrB_BOOL, n));
+
+    //--------------------------------------------------------------------------
+    // start the Bellman Ford process
+    //--------------------------------------------------------------------------
+    bool any_dless= true;      // if there is any newly found shortest path
+    int64_t iter = 0;          // number of iterations
+
+    // terminate when no new path is found or more than V-1 loops
+    while (any_dless && iter < n - 1)
+    {
+        // execute semiring on dmasked and A, and save the result to dmasked
+        GRB_TRY (GrB_vxm(dmasked, GrB_NULL, GrB_NULL,
+            BF_lMIN_PLUSrhs_Tuple3, dmasked, Atmp, GrB_NULL));
+
+        // dless = d .< dtmp
+        GRB_TRY (GrB_eWiseMult(dless, NULL, NULL, BF_LT_Tuple3, dmasked, d,
+            NULL));
+
+        // if there is no entry with smaller distance then all shortest paths
+        // are found
+        GRB_TRY (GrB_reduce (&any_dless, NULL, GrB_LOR_MONOID_BOOL, dless,
+            NULL)) ;
+        if(any_dless)
+        {
+            // update all entries with smaller distances
+            //GRB_TRY (GrB_apply(d, dless, NULL, BF_Identity_Tuple3,
+            //    dmasked, NULL));
+            GRB_TRY (GrB_assign(d, dless, NULL, dmasked, GrB_ALL, n, NULL));
+
+            // only use entries that were just updated
+            //GRB_TRY (GrB_Vector_clear(dmasked));
+            //GRB_TRY (GrB_apply(dmasked, dless, NULL, BF_Identity_Tuple3,
+            //    d, NULL));
+            //try:
+            GRB_TRY (GrB_assign(dmasked, dless, NULL, d, GrB_ALL, n, GrB_DESC_R));
+        }
+        iter ++;
+    }
+
+    // check for negative-weight cycle only when there was a new path in the
+    // last loop, otherwise, there can't be a negative-weight cycle.
+    if (any_dless)
+    {
+        // execute semiring again to check for negative-weight cycle
+        GRB_TRY (GrB_vxm(dmasked, GrB_NULL, GrB_NULL,
+            BF_lMIN_PLUSrhs_Tuple3, dmasked, Atmp, GrB_NULL));
+
+        // dless = d .< dtmp
+        GRB_TRY (GrB_eWiseMult(dless, NULL, NULL, BF_LT_Tuple3, dmasked, d,
+            NULL));
+
+        // if there is no entry with smaller distance then all shortest paths
+        // are found
+        GRB_TRY (GrB_reduce (&any_dless, NULL, GrB_LOR_MONOID_BOOL, dless,
+            NULL)) ;
+        if(any_dless)
+        {
+            // printf("A negative-weight cycle found. \n");
+            LG_FREE_ALL;
+            return (GrB_NO_VALUE) ;
+        }
+    }
+
+    //--------------------------------------------------------------------------
+    // extract tuple from "distance" vector d and create GrB_Vectors for output
+    //--------------------------------------------------------------------------
+
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &I, n, sizeof(GrB_Index), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &W, n, sizeof(BF_Tuple3_struct),
+        msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &w, n, sizeof(double), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &h, n, sizeof(GrB_Index), msg)) ;
+    LAGRAPH_TRY (LAGraph_Malloc ((void **) &pi, n, sizeof(GrB_Index), msg)) ;
+
+    // todo: create 3 unary ops, and use GrB_apply?
+
+    GRB_TRY (GrB_Vector_extractTuples_UDT (I, (void *) W, &n, d));
+
+    for (k = 0; k < n; k++)
+    {
+        w [k] = W[k].w ;
+        h [k] = W[k].h ;
+        pi[k] = W[k].pi;
+    }
+    GRB_TRY (GrB_Vector_new(pd_output,  GrB_FP64,   n));
+    GRB_TRY (GrB_Vector_new(ppi_output, GrB_UINT64, n));
+    GRB_TRY (GrB_Vector_new(ph_output,  GrB_UINT64, n));
+    GRB_TRY (GrB_Vector_build (*pd_output , I, w , n, GrB_MIN_FP64  ));
+    GRB_TRY (GrB_Vector_build (*ppi_output, I, pi, n, GrB_MIN_UINT64));
+    GRB_TRY (GrB_Vector_build (*ph_output , I, h , n, GrB_MIN_UINT64));
+    LG_FREE_WORK;
+    return (GrB_SUCCESS) ;
+}
+
+\end{minted}

### 2024-04-04 sections/eval.tex
Update on Overleaf.
fixes

@@
 
+\begin{wrapfigure}{r}{0.41\linewidth}
+  \vspace{-28pt}
+  \begin{minipage}[t]{0.18\textwidth}
+      \begin{minted}{julia}
+          y .= 0
+          for j = _, i = _
+            y[i] += A[i, j] * x[j]
+          end
+      \end{minted}
+      \vspace{24pt} % Add this to ensure top alignment within minipage
+      \begin{minted}{julia}
+          y .= 0
+          for j = _, i = _
+            y[j] += A[i, j] * x[i]
+          end
+      \end{minted}
+  \end{minipage}\hfill%
+  \begin{minipage}[t]{0.22\textwidth}
+      \vspace{0pt} % Add this to ensure top alignment within minipage
+      \begin{minted}{julia}
+          y .= 0
+          for j = _
+            let x_j = x[j]
+              y_j .= 0
+              for i = _
+                let A_ij = A[i, j]
+                  y[i] += x_j * A_ij
+                  y_j[] += A_ij * x[i]
+                end
+              end
+              #D is the diagonal
+              y[j] += y_j[] + D[j] * x_j
+            end
+          end
+      \end{minted}
+  \end{minipage}
+  \vspace{-8pt}
+  \caption{Finch row-major, column-major and symmetric SpMV Programs}
+  \label{spmv_programs}
+  \vspace{-12pt}
+\end{wrapfigure}
 \subsection{Sparse Matrix-Vector Multiply (SpMV)}
@@
 
-We display speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library, as depicted in Figures \ref{spmv_sorted} and \ref{spmv_grouped}.
+Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
 We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
@@
 The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset in Figure \ref{spmv_grouped} is the fastest among the formats and programs we tested.
+We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
 Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
@@
 They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Fig.
-\ref{fig:fusion}.
+Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
 A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
@@
 For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Fig.
-\ref{graph_result}.
+For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
 For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
@@
 \begin{figure}
-\label{fig:fusion}
 \begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
@@
 \vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.}
+\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
 \end{figure}

### 2024-04-04 sections/related.tex
fixes

@@
 
-Both structured and dense libraries support a multitude of architectures, as do dense compilers.
-
 Sparse compilers have been extended to many architectures.
 An extension of TACO supports GPU~\cite{senanayake2020sparse}, Cyclops~\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal~\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine~\cite{hsu_sparse_2022} supports custom hardware.
-
-We believe that supporting control flow is the first step towards architectural support for complex structures.
+We believe that supporting control flow is the first step towards architectural support beyond unstructured sparsity.
 

### 2024-04-04 sections/eval/eval.tex
reorg

--- 2024-04-04 sections/eval.tex
@@
 
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-  \vspace{-28pt}
-  \begin{minipage}[t]{0.18\textwidth}
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[i] += A[i, j] * x[j]
-          end
-      \end{minted}
-      \vspace{24pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[j] += A[i, j] * x[i]
-          end
-      \end{minted}
-  \end{minipage}\hfill%
-  \begin{minipage}[t]{0.22\textwidth}
-      \vspace{0pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _
-            let x_j = x[j]
-              y_j .= 0
-              for i = _
-                let A_ij = A[i, j]
-                  y[i] += x_j * A_ij
-                  y_j[] += A_ij * x[i]
-                end
-              end
-              #D is the diagonal
-              y[j] += y_j[] + D[j] * x_j
-            end
-          end
-      \end{minted}
-  \end{minipage}
-  \vspace{-8pt}
-  \caption{Finch row-major, column-major and symmetric SpMV Programs}
-  \label{spmv_programs}
-  \vspace{-12pt}
-\end{wrapfigure}
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of array structures, making it an effective kernel to
-demonstrate the utility of our programming model.
-
-We varied both the data formats and the programs we used them with. Our formats
-are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
-Figure~\ref{spmv_programs}.
-
-In varying the program, we consider both row and column-major SpMV programs, as
-well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-effectively. Our program restricts our attention to the canonical trial (using
-masks), reuses the reads to the canonical triangle of the symmetric matrix
-(using a $\finchdefine$ statement), and writes the results to both relevant
-locations (using multiple outputs).
-
-All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
-
-Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created some synthetic matrices containing bands of varying sizes.
-% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-%\subsubsection{Tensor Formats}
-
-%\subsubsection{Symmetric SpMV}
-% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
+\include{spmv}
+\include{spgemm}
+\include{graphs}
+\include{morphology}
 

### 2024-04-04 sections/eval/graphs.tex
reorg

--- 2024-04-04 sections/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-  \vspace{-28pt}
-  \begin{minipage}[t]{0.18\textwidth}
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[i] += A[i, j] * x[j]
-          end
-      \end{minted}
-      \vspace{24pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[j] += A[i, j] * x[i]
-          end
-      \end{minted}
-  \end{minipage}\hfill%
-  \begin{minipage}[t]{0.22\textwidth}
-      \vspace{0pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _
-            let x_j = x[j]
-              y_j .= 0
-              for i = _
-                let A_ij = A[i, j]
-                  y[i] += x_j * A_ij
-                  y_j[] += A_ij * x[i]
-                end
-              end
-              #D is the diagonal
-              y[j] += y_j[] + D[j] * x_j
-            end
-          end
-      \end{minted}
-  \end{minipage}
-  \vspace{-8pt}
-  \caption{Finch row-major, column-major and symmetric SpMV Programs}
-  \label{spmv_programs}
-  \vspace{-12pt}
-\end{wrapfigure}
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of array structures, making it an effective kernel to
-demonstrate the utility of our programming model.
-
-We varied both the data formats and the programs we used them with. Our formats
-are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
-Figure~\ref{spmv_programs}.
-
-In varying the program, we consider both row and column-major SpMV programs, as
-well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-effectively. Our program restricts our attention to the canonical trial (using
-masks), reuses the reads to the canonical triangle of the symmetric matrix
-(using a $\finchdefine$ statement), and writes the results to both relevant
-locations (using multiple outputs).
-
-All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
-
-Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created some synthetic matrices containing bands of varying sizes.
-% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-%\subsubsection{Tensor Formats}
-
-%\subsubsection{Symmetric SpMV}
-% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
 \subsection{Graph Analytics}
@@
 \end{figure}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluation}
-To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby query.
-For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
-For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
-
-\begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-04 sections/eval/morphology.tex
reorg

--- 2024-04-04 sections/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-  \vspace{-28pt}
-  \begin{minipage}[t]{0.18\textwidth}
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[i] += A[i, j] * x[j]
-          end
-      \end{minted}
-      \vspace{24pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[j] += A[i, j] * x[i]
-          end
-      \end{minted}
-  \end{minipage}\hfill%
-  \begin{minipage}[t]{0.22\textwidth}
-      \vspace{0pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _
-            let x_j = x[j]
-              y_j .= 0
-              for i = _
-                let A_ij = A[i, j]
-                  y[i] += x_j * A_ij
-                  y_j[] += A_ij * x[i]
-                end
-              end
-              #D is the diagonal
-              y[j] += y_j[] + D[j] * x_j
-            end
-          end
-      \end{minted}
-  \end{minipage}
-  \vspace{-8pt}
-  \caption{Finch row-major, column-major and symmetric SpMV Programs}
-  \label{spmv_programs}
-  \vspace{-12pt}
-\end{wrapfigure}
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of array structures, making it an effective kernel to
-demonstrate the utility of our programming model.
-
-We varied both the data formats and the programs we used them with. Our formats
-are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
-Figure~\ref{spmv_programs}.
-
-In varying the program, we consider both row and column-major SpMV programs, as
-well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-effectively. Our program restricts our attention to the canonical trial (using
-masks), reuses the reads to the canonical triangle of the symmetric matrix
-(using a $\finchdefine$ statement), and writes the results to both relevant
-locations (using multiple outputs).
-
-All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
-
-Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created some synthetic matrices containing bands of varying sizes.
-% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-%\subsubsection{Tensor Formats}
-
-%\subsubsection{Symmetric SpMV}
-% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
 \subsection{Image Morphology}
@@
 \end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluation}
-To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby query.
-For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
-For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
-
-\begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-04 sections/eval/spgemm.tex
reorg

--- 2024-04-04 sections/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
 \subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
@@
 %
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
+Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(N)$ intermediate.
 %
@@
 \end{figure}
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-  \vspace{-28pt}
-  \begin{minipage}[t]{0.18\textwidth}
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[i] += A[i, j] * x[j]
-          end
-      \end{minted}
-      \vspace{24pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[j] += A[i, j] * x[i]
-          end
-      \end{minted}
-  \end{minipage}\hfill%
-  \begin{minipage}[t]{0.22\textwidth}
-      \vspace{0pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _
-            let x_j = x[j]
-              y_j .= 0
-              for i = _
-                let A_ij = A[i, j]
-                  y[i] += x_j * A_ij
-                  y_j[] += A_ij * x[i]
-                end
-              end
-              #D is the diagonal
-              y[j] += y_j[] + D[j] * x_j
-            end
-          end
-      \end{minted}
-  \end{minipage}
-  \vspace{-8pt}
-  \caption{Finch row-major, column-major and symmetric SpMV Programs}
-  \label{spmv_programs}
-  \vspace{-12pt}
-\end{wrapfigure}
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of array structures, making it an effective kernel to
-demonstrate the utility of our programming model.
-
-We varied both the data formats and the programs we used them with. Our formats
-are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
-Figure~\ref{spmv_programs}.
-
-In varying the program, we consider both row and column-major SpMV programs, as
-well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-effectively. Our program restricts our attention to the canonical trial (using
-masks), reuses the reads to the canonical triangle of the symmetric matrix
-(using a $\finchdefine$ statement), and writes the results to both relevant
-locations (using multiple outputs).
-
-All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
-
-Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created some synthetic matrices containing bands of varying sizes.
-% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-%\subsubsection{Tensor Formats}
-
-%\subsubsection{Symmetric SpMV}
-% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluation}
-To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby query.
-For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
-For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
-
-\begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-04 sections/eval/spmv.tex
reorg

--- 2024-04-04 sections/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
 \begin{wrapfigure}{r}{0.41\linewidth}
@@
   \end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluation}
-To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby query.
-For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
-For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
-
-\begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-04 sections/eval.tex
allset

@@
 
-\subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    @finch begin
-      C .= 0
-      for j=_
-        for i=_
-          for k=_
-            C[i, j] += AT[k, i] * B[k, j]
-          end
-        end
-      end
-      return C
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseByteMap(Element(0)))
-    @finch begin
-      C .= 0
-      for j=_
-        w .= 0
-        for k=_
-          for i=_
-            w[i] += A[i, k] * B[k, j]
-          end
-        end
-        for i=_
-          C[i, j] = w[i]
-        end
-      end
-    end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
-    \begin{minted}{julia}
-    w = Tensor(SparseHash(SparseHash(Element(0))))
-    @finch begin
-      w .= 0
-      for k=_
-        for j=_
-          for i=_
-            w[i, j] += A[i, k] * BT[j, k]
-          end
-        end
-      end
-      C .= 0
-      for j=_, i=_
-        C[i, j] = w[i, j]
-      end
-    end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
-    \vspace{-12pt}
-\end{figure}
-
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
-We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
-%
-The inner-products algorithm takes dot products of corresponding rows and columns, while the outer-products algorithm sums the outer products of corresponding columns and rows.
-%
-Gustavson's algorithm sums the rows of $B$ scaled by the corresponding nonzero columns in each row of $A$.
-%
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
-
-Figure~\ref{fig:spgemm_listing} implements all three approaches in Finch, and Figure~\ref{fig:spgemm} compares the performance of Finch to TACO.
-%
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(M)$ intermediate.
-%
-In our Finch implementation of outer products, we use a sparse hash table, as it is fully-sparse and randomly accessible.
-%
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
-%
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
-As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
-
-%\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-  \vspace{-28pt}
-  \begin{minipage}[t]{0.18\textwidth}
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[i] += A[i, j] * x[j]
-          end
-      \end{minted}
-      \vspace{24pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _, i = _
-            y[j] += A[i, j] * x[i]
-          end
-      \end{minted}
-  \end{minipage}\hfill%
-  \begin{minipage}[t]{0.22\textwidth}
-      \vspace{0pt} % Add this to ensure top alignment within minipage
-      \begin{minted}{julia}
-          y .= 0
-          for j = _
-            let x_j = x[j]
-              y_j .= 0
-              for i = _
-                let A_ij = A[i, j]
-                  y[i] += x_j * A_ij
-                  y_j[] += A_ij * x[i]
-                end
-              end
-              #D is the diagonal
-              y[j] += y_j[] + D[j] * x_j
-            end
-          end
-      \end{minted}
-  \end{minipage}
-  \vspace{-8pt}
-  \caption{Finch row-major, column-major and symmetric SpMV Programs}
-  \label{spmv_programs}
-  \vspace{-12pt}
-\end{wrapfigure}
-\subsection{Sparse Matrix-Vector Multiply (SpMV)}
-Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-kernel, with a wide range of applications \cite{liu_csr5_2015,
-zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-been proposed to reduce the footprint with a minimal impact on complexity
-\cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
-results in a wide range of array structures, making it an effective kernel to
-demonstrate the utility of our programming model.
-
-We varied both the data formats and the programs we used them with. Our formats
-are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
-Figure~\ref{spmv_programs}.
-
-In varying the program, we consider both row and column-major SpMV programs, as
-well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-effectively. Our program restricts our attention to the canonical trial (using
-masks), reuses the reads to the canonical triangle of the symmetric matrix
-(using a $\finchdefine$ statement), and writes the results to both relevant
-locations (using multiple outputs).
-
-All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
-
-Figure~\ref{fig:spmv_grouped} displays speedup relative to TACO, SuiteSparseGraphBLAS, and Julia’s standard library.
-We test using sparse matrices from a large selection of datasets spanning several previous papers: the datasets used by Vuduc et al.
-to test the OSKI interface \cite{vuduc2005oski}, Ahrens et al.
-to test a variable block row format partitioning strategy \cite{ahrens_optimal_2021}, and Kjolstad et al.
-to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
-We also created some synthetic matrices containing bands of varying sizes.
-% or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-The dense vector is randomly generated.
-We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
-
-We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
-Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
-Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
-There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
-For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
-Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
-Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
-
-%We consider the Dense(SparseList(Element)) format with the column-major SpMV program to be the Finch baseline as it is the closet analog to the sparse matrix format and SpMV program in other libraries.
-
-%\subsubsection{Tensor Formats}
-
-%\subsubsection{Symmetric SpMV}
-% Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
-%However, the regular row- or column-major Finch SpMV programs have better performance for symmetric matrices than the symmetric Finch SpMV program for the other more specialized formats, likely because we need in-order accesses to fully capitalize on the specialized storage.
-
-\begin{wrapfigure}{r}{0.41\linewidth}
-    \begin{minipage}[t]{0.18\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[i] += A[i, j] * x[j]
-            end
-        \end{minted}
-        \vspace{24pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _, i = _
-              y[j] += A[i, j] * x[i]
-            end
-        \end{minted}
-    \end{minipage}\hfill%
-    \begin{minipage}[t]{0.22\textwidth}
-        \vspace{0pt} % Add this to ensure top alignment within minipage
-        \begin{minted}{julia}
-            y .= 0
-            for j = _
-              let x_j = x[j]
-                y_j .= 0
-                for i = _
-                  let A_ij = A[i, j]
-                    y[i] += x_j * A_ij
-                    y_j[] += A_ij * x[i]
-                  end
-                end
-                #D is the diagonal
-                y[j] += y_j[] + D[j] * x_j
-              end
-            end
-        \end{minted}
-    \end{minipage}
-    \vspace{-8pt}
-    \caption{Finch row-major, column-major and symmetric SpMV Programs}
-    \label{spmv_programs}
-    \vspace{-12pt}
-\end{wrapfigure}
-
-\begin{table}[htbp]
-    \scriptsize
-    \centering
-    \vspace{-12pt}
-    \caption{SpMV Tensor Formats}
-    \vspace{-12pt}
-    \label{spmv_tensor_formats}
-    \begin{tabular}{|l|l|l|l|}
-        \hline
-        \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-        \hline
-        \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-        \cline{3-4} 
-        & & Pattern & sparse, boolean-valued matrices \\
-        \cline{2-4} 
-        & SparseVBL & Element & real-valued matrices with blocked structure \\
-        \cline{2-4}
-        & SparseBand & Element & real-valued matrices with diagonal band \\
-        \cline{2-4}
-        & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-        \cline{3-4} 
-        & & Pattern & matrix with runs of true or false \\
-        \hline 
-    \end{tabular}
-    \vspace{-8pt}
-\end{table}
-
-% Perhaps add the following section again in a future iteration of paper
-% \subsubsection{4D Blocked SpMV}
-% Finch also provides us the capability of writing an SpMV kernel [point to figure] that computes the output in blocks.
-Specifically, we rewrite an n x n matrix as a n/b x n/b x b x b 4-dimensional tensor and the vector as an n x n/b matrix where b is the block-size.
-We represent the blocked matrix with a SparseList as its second level (i.e.
-of format Dense(SparseList(Dense(Dense(Element(0.0)))))) so that only the non-zero blocks are stored.
-Then, we perform SpMV on each b x b block individually.
-Note that although SparseVBL already stores consecutive nonzeros in blocks, the benefit of a 4D-blocked kernel is that it additionally computes the output block by block.
-This method enables us to take advantage of spatial and temporal locality via register reuse [cite].
-
-% We evaluated the 4D-blocked kernel on the Kronecker product of a neural network matrix (erdos-renyi, with sparsity p = ?) with a blocked matrix of size 10x10 and a dense and randomly generated vector.
-We found a 1.04x speedup to TACO, indicating comparable performance, and a 1.35x speedup to computing a non-blocked SpMV with the SparseVBL format, the fastest performing 2D Finch format for this matrix.
-
-% Structured matrices with induced dense blocks of equivalent size commonly arise in machine learning applications.
-In feature extraction, the Kronecker product of image data and a smaller matrix is computed to extract relevant features [cite].
-Graph adjacency matrices or Laplacian matrices may also exhibit dense blocks when certain subsets of nodes or edges are densely interconnected.
-For instance, the adjacency matrix of the Erdös-Renyi random graph has a real symmetric bxb random block at each non-vanishing entry [cite].
-
-%Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-\begin{figure}
-    \includegraphics[width=\linewidth]{Spmv_annotated.png}
-    \vspace{-18pt}
-    \caption{Performance of SpMV by Finch format.}
-    \label{fig:spmv_grouped}
-    \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
-\end{figure}
-
-\subsection{Image Morphology}
-
-Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
-We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
-
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
-%
-All of these have their advantages.
-%
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
-
-The erosion operation turns off a pixel unless all of it's neighbors are on.
-%
-This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
-%
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
-%
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
-
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
-in the masked kernel and get better performance.
-%
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
-
-\begin{figure}
-    \scriptsize
-    \begin{minipage}{0.5\linewidth}
-    Wordwise Erosion:
-    \begin{minted}{julia}
-        output .= false
-        for y = _
-          tmp .= false
-          for x = _
-            tmp[x] = coalesce(input[x, ~(y-1)], true) & input[x, y] & coalesce(input[x, ~(y+1)], true)
-          end
-          for x = _
-            output[x, y] = coalesce(tmp[~(x-1)], true) & tmp[x] & coalesce(tmp[~(x+1)], true)
-          end
-        end
-    \end{minted}
-    \vspace{12pt}
-    Masked Histogram:
-    \begin{minted}{julia}
-        bins .= 0 
-        for x=_
-          for y=_
-            if mask[y, x]
-              bins[div(img[y, x], 16) + 1] += 1
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}%
-    \begin{minipage}{0.5\linewidth}
-    Bitwise Erosion:
-    \begin{minted}{julia}
-        output .= 0
-        for y = _
-          tmp .= 0
-          for x = _
-            if mask[x, y]
-              tmp[x] = coalesce(input[x, ~(y-1)], 0xFFFFFFFF) & input[x, y] & coalesce(input[x, ~(y+1)], 0xFFFFFFFF)
-            end
-          end
-          for x = _
-            if mask[x, y]
-              let tl = coalesce(tmp[~(x-1)], 0xFFFFFFFF), t = tmp[x], tr = coalesce(tmp[~(x+1)], 0xFFFFFFFF)
-                let res = ((tr << (8 * sizeof(UInt) - 1)) | (t >> 1)) & t & ((t << 1) | (tl >> (8 * sizeof(UInt) - 1)))
-                  output[x, y] = res
-                end
-              end
-            end
-          end
-        end
-    \end{minted}
-    \end{minipage}
-    \vspace{-12pt}
-    \caption{Two approaches to erosion in Finch.
-The $coalesce$ function defines the out of bounds value.
-On left, the naive approach.
-On $SparseRLE(Pattern())$ inputs, this only performs operations at the boundaries of constant regions.
-On right, a bitwise approach, using a mask to limit work to nonzero blocks of bits.}
-    \label{fig:morphology_listing}
-\end{figure}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{erode2_speedup_over_opencv.png}%
-	\includegraphics[width=0.5\linewidth]{hist_speedup_over_opencv.png}
- \vspace{-12pt}
-    \caption{Performance of Finch on image morphology tasks. On left, we run 2 iterations of erosion. On right, we run a masked histogram.}\label{fig:morphology}
-\end{figure}
-
-\subsection{Graph Analytics}
-%\help{In this case, the two main highlights are that Finch can do arbitrary operators (i.e.
-choose), and that Finch can do early break, and also the different loop orders and multiple outputs.
-We may need to explain a little bit about what push pull is.
-For bellman, the main point is that we need multiple outputs, sparse inputs, masks, and sparse outputs with differing formats at differing points.}
-
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
-
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
-Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier.
-The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway.
-Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
-
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
-For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs.
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
-On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53).
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
-    \begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    V = Tensor(Dense(Element(false)))
-    P = Tensor(Dense(Element(0)))
-    F = Tensor(SparseByteMap(Pattern()))
-    _F = Tensor(SparseByteMap(Pattern()))
-    A = Tensor(Dense(SparseList(Pattern())))
-    AT = Tensor(Dense(SparseList(Pattern())))
-
-    function bfs_push(_F, F, A, V, P)
-      @finch begin
-        _F .= false
-        for j=_, k=_
-          if F[j] && A[k, j] && !(V[k])
-            _F[k] |= true
-            P[k] <<choose(0)>>= j
-          end
-        end
-        return _F
-      end
-    end
-
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-    \begin{minted}{julia}
-    function bfs_pull(_F, F, AT, V, P)
-      p = ShortCircuitScalar{0}()
-      @finch begin
-        _F .= false
-        for k=_
-          if !V[k]
-            p .= 0
-            for j=_
-              if F[follow(j)] && AT[j, k]
-                p[] <<choose(0)>>= j
-              end
-            end
-            if p[] != 0
-              _F[k] |= true
-              P[k] = p[]
-            end
-          end
-        end
-        return _F
-      end
-    end
-    \end{minted}
-\end{minipage}%
-\begin{minipage}{0.33\linewidth}
-  \begin{minted}{julia}
-  _D = Tensor(Dense(Element(Inf)), n)
-  D = Tensor(Dense(Element(Inf)), n)
-  function bellmanford(A, _D, D, _F, F)
-    @finch begin
-    F .= false
-    for j = _
-      if _F[j]
-        for i = _
-          let d = _D[j] + A[i, j]
-            D[i] <<min>>= d
-            F[i] |= d < _D[i]
-          end
-        end
-      end
-    end
-  end
-\end{minted}
-\end{minipage}
-\caption{Graph Applications written in Finch.
-Note that parents are calculated separately for Bellman-Ford.
-The $choose(z)$ operator is a GraphBLAS concept which returns any argument that is not $z$.}\label{fig:graph_listing}
-\end{figure}
+\include{eval/spmv}
+\include{eval/spgemm}
+\include{eval/graphs}
+\include{eval/morphology}
 
@@
 
+%There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both the output and the algorithm.
+The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
+This approach results in a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
+To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user.
 \subsubsection{Finch Logic}
@@
 A plan is a sequence of queries and outputs a set of tensors.
+%This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.
 
@@
 
+%Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
+In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
+However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
+
 \subsubsection{Standardizing \& Heuristic Optimization}
@@
 In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
+
+% based on a heuristic which loops over intersecting variables first.
 
@@
 A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
+
+%we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies
 

### 2024-04-04 sections/appendix.tex
reorg
need to edit contributions

@@
 
-\section{Graph Algorithms}
+\section{Graph Algorithm Listings}\label{sec:graph_listings}
 
@@
 \begin{minted}{c}
+//------------------------------------------------------------------------------
+// LAGr_BreadthFirstSearch:  breadth-first search dispatch
+//------------------------------------------------------------------------------
+
+// LAGraph, (c) 2019-2022 by The LAGraph Contributors, All Rights Reserved.
+// SPDX-License-Identifier: BSD-2-Clause
+//
+// For additional details (including references to third party source code and
+// other files) see the LICENSE file or contact permission@sei.cmu.edu. See
+// Contributors.txt for a full list of contributors. Created, in part, with
+// funding and support from the U.S. Government (see Acknowledgments.txt file).
+// DM22-0790
+
+// Contributed by Scott McMillan, SEI Carnegie Mellon University
+
+//------------------------------------------------------------------------------
+
+// Breadth-first-search via push/pull method if using SuiteSparse:GraphBLAS
+// and its GxB extensions, or a push-only method otherwise.  The former is
+// much faster.
+
+// This is an Advanced algorithm.  SuiteSparse can use a push/pull method if
+// G->AT and G->out_degree are provided.  G->AT is not required if G is
+// undirected.  The vanilla method is always push-only.
+
+#include "LG_alg_internal.h"
+
+int LAGr_BreadthFirstSearch
+(
+    // output:
+    GrB_Vector *level,
+    GrB_Vector *parent,
+    // input:
+    const LAGraph_Graph G,
+    GrB_Index src,
+    char *msg
+)
+{
+
+#if LAGRAPH_SUITESPARSE
+    return LG_BreadthFirstSearch_SSGrB   (level, parent, G, src, msg) ;
+#else
+    return LG_BreadthFirstSearch_vanilla (level, parent, G, src, msg) ;
+#endif
+}
+
 //------------------------------------------------------------------------------

### 2024-04-04 sections/eval.tex
oops
ugh
reorg

@@
 
-\include{eval/spmv}
-\include{eval/spgemm}
-\include{eval/graphs}
-\include{eval/morphology}
-
-\subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
-To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
-
-%There is a gap between the expressiveness of the purely declarative Array API and the Finch language which defines both the output and the algorithm.
-The naive approach to solving this would be to explicitly define a mapping from each function in the API to a Finch program which eagerly produces the output.
-This approach results in a simple translation procedure but misses crucial opportunities for optimization across a chain of function calls and needlessly materializes large intermediate results.
-To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user.
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
-This means that operators take in a set of indexed tensors and output an indexed tensor.
-Conceptually, an indexed tensor can be thought of as a relation with an ordering on the index attributes and a separate value attribute, and we include the $\finchreorder$ operator to manipulate this order.
-To express materialization, reuse of common sub-expressions, and multiple outputs, we define $\finchquery$ and $\finchplan$.
-A $\finchquery$ assigns the output of an expression to a name.
-We use this to denote that we materialize an expression.
-Later queries can access the result of earlier queries through the $\finchalias$ operator, allowing multiple queries to benefit from a shared computation.
-A plan is a sequence of queries and outputs a set of tensors.
-%This ordering is necessary for two reasons 1) the outputs of a function in the Array API must match a defined order on the dimensions based on the order of the inputs' dimensions 2) to benefit from concordant iteration we must ensure that the operands to a mapjoin have compatible index orders.
-
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
-
-We now describe how to define a few example functions from the API in this language.
-Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
-For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
-
-%Notably, these plans do not specify important details about the computation such as the format of intermediates and the order of the loops.
-In the following discussion, we provide sensible heuristics and show that they provide good performance on important kernels.
-However, for larger or more complex programs, it would be important to apply a cost-based optimization strategy which we leave for future work.
-
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
-The semantically important requirements of this standard form are 1) all inputs (i.e.
-tables and alias operators) in a query's RHS must conform to a common ordering of the indices 2) the outermost operator of each query's RHS must be a reformat 3) the expression within the reformat must be a pointwise expression, optionally wrapped in an aggregate operator.
-The former allows the interpreter to identify the loop order for each kernel.
-The second determines the output format for each intermediate.
-The last one guarantees that the innermost expression can be computed as a single kernel.
-To achieve this: 1) a concordization pass examines each query, heuristically selects a loop order, and transposes inputs which don't match that order 2) a formatting pass selects a level format for each output index based on the formats of the inputs and whether the loop order requires random writes.
-In future work, these heuristics could be improved with cost-based optimization similar to query optimization in the database setting.
-
-% based on a heuristic which loops over intersecting variables first.
-
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
-Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
-At this point, any aliases to the result of previous queries are replaced with an access to the actual result.
-Lastly, the concordant loop order is identified and instantiated.
-The lowered query is then compiled and executed with the Finch compiler, and the result is assigned to $name$ before proceeding to the next query.
-
-\subsubsection{Evaluation}
-To demonstrate the performance of our array implementation, we evaluate it on 1) triangle counting 2) SDDMM 3) and element-wise operations.
-Further, we compare against DuckDB as a state of the art system which implements a form of kernel fusion through pipelined query execution.
-To do this, we express each of these kernels as a single select, join, groupby query.
-For the element-wise case, we provide an unfused Finch method to show the impact of fusion.
-For triangle counting, we use the same set of graph matrices as in Figure~\ref{fig:graph_result}.
-For SDDMM, we use this set of graph matrices for the sparse matrix, and we produce random dense matrices with embedding dimension 25.
-Lastly, for the elementwise operations, we use uniformly sparse matrices with dimension 10000 by 10000.
-A/B have sparsity $.1$, and we vary the sparsity of C in the X axis.
-
-%we evaluate it on a series of kernels which benefit from the kind of kernel fusion that it automatically applies
-
-Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
-For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.
-
-\begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
-\end{figure}
-
-%matmul, mttkrp, repeated ttm, triangle counting, multiple pointwise,
-%in-place.
-%dot((v^t .* u), w)) vs. 
-%(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}
+\input{sections/eval/spmv}
+\input{sections/eval/spgemm}
+\input{sections/eval/graphs}
+\input{sections/eval/morphology}
+\input{sections/eval/api}

### 2024-04-04 sections/eval/api.tex
reorg

--- 2024-04-04 sections/eval/eval.tex
@@
-\section{Case Studies}
-
-We evaluate Finch on a broad set of applications to showcase it's efficiency,
-flexibility, and expressibility. All of our implementations highlight the
-benefits of data structure and algorithm co-design.  Our implementation of
-sparse-sparse-matrix multiply (SpGEMM) translates classical lessons from sparse
-performance engineering into the language of Finch, using temporaries and
-randomly-accessible workspace formats to efficiently implement the three main
-approaches. Our study of sparse-matrix-dense-vector multiply (SpMV) highlights
-the benefits of precise structural specialization. Our studies of image
-morphology and graph applications show how Finch's programming model can express more complex
-real-world kernels. Finally, we explain
-how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
-
-All experiments were run on a single core of a 12-core 2-socket Intel Xeon E5-2695 v2 running at
-2.40GHz with 128GB of memory. Finch is implemented in Julia v1.9, targeting LLVM
-through Julia. All timings are the minimum of 10,000 runs or 5s of measurement,
-whichever happens first.
-
-\include{spmv}
-\include{spgemm}
-\include{graphs}
-\include{morphology}
-
 \subsection{Implementing Numpy's Array API in Finch}

### 2024-04-04 sections/background.tex
Update on Overleaf.

@@
     \end{minipage}
-    \caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
+    \caption{The looplet language, as understood in a correct execution of a Finch program.}
+    \vspace{-8pt}
 \end{figure}
+
+\begin{figure}[ht]
+\footnotesize
+\begin{tabular} {|l|c|}
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchlookup(seek, body)$: The Lookup Looplet represents a
+        randomly accessible region of an iterator. The body of the lookup is
+        understood to have one less dimension than the lookup itself, as we have
+        already ``looked up'' that index in the tensor by the time we reach the
+        body. \texttt{seek(i)} is a function that updates state to the given
+        index.
+    \end{minipage}% &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-lookup.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchrun(body)$: The Run Looplet represents a constant
+        region of an iterator. The body of the run is understood to have one
+        less dimension than the lookup itself, as all of the bodies are
+        identical.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-run.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchphase(c:d, body)$: The Phase Looplet represents a
+        restriction of the range on which a loop should execute, and allows us
+        to succinctly express the ranges on which children of compound looplets
+        are defined.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-phase.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchswitch(cond, head, tail)$: The Switch Looplet allows
+        us to specialize the body of a looplet based on a condition, evaluated
+        in the embedding context. If the condition is true, we use `head`,
+        otherwise we use `tail`. Switch has a high lowering priority so we can
+        see what's inside of it and lower that appropriately. This also lifts
+        the condition as high as possible into the loop nest. The condition is
+        assumed to evaluate to a boolean.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-switch.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
+        allows us to cache certain computations in the state under which the
+        body will execute. This is useful for computing and caching the results
+        of expensive computations.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-thunk.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchsequence(head, tail)$: The Sequence looplet represents the
+        concatenation of two looplets. Both arguments must be phase looplets, and
+        are assumed to be nonoverlapping, covering, and in order.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-sequence.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchspike(body, tail)$ The Spike Looplet represents a run
+        followed by a single value. In this paper, Spike will be considered a
+        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
+        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
+        handled with special care, since they are an opportunity to align the
+        final run to the end of the root loop extent, without using any special
+        bounds inference.
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-spike.png}
+    \end{minipage} \\
+    \begin{minipage}[c]{0.65\linewidth}
+        $\finchstepper([seek], next, body)$ The stepper looplet
+        represents a variable number of looplets, concatenated. Since our
+        looplets may be skipped over due to conditions or various rewrites, the
+        $seek$ function allows us to fast-forward the state to the start of the
+        root loop extent when it comes time to lower the stepper. The $next$
+        function advances the state to the next iteration of the stepper. 
+    \end{minipage} &
+    \begin{minipage}[c]{0.35\linewidth}
+        \centering
+        \includegraphics[scale=0.20]{Looplets-stepper.png}
+    \end{minipage}
+    \end{tabular}
+    \caption{The looplet language, as understood in a correct execution of a Finch program.}
+\end{figure}

### 2024-04-04 sections/background.tex
Update on Overleaf.

@@
 
+% \begin{figure}[ht]
+%     \footnotesize
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchlookup(seek, body)$: The Lookup Looplet represents a
+%         randomly accessible region of an iterator. The body of the lookup is
+%         understood to have one less dimension than the lookup itself, as we have
+%         already ``looked up'' that index in the tensor by the time we reach the
+%         body. \texttt{seek(i)} is a function that updates state to the given
+%         index.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-lookup.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchrun(body)$: The Run Looplet represents a constant
+%         region of an iterator. The body of the run is understood to have one
+%         less dimension than the lookup itself, as all of the bodies are
+%         identical.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-run.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchphase(c:d, body)$: The Phase Looplet represents a
+%         restriction of the range on which a loop should execute, and allows us
+%         to succinctly express the ranges on which children of compound looplets
+%         are defined.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-phase.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchswitch(cond, head, tail)$: The Switch Looplet allows
+%         us to specialize the body of a looplet based on a condition, evaluated
+%         in the embedding context. If the condition is true, we use `head`,
+%         otherwise we use `tail`. Switch has a high lowering priority so we can
+%         see what's inside of it and lower that appropriately. This also lifts
+%         the condition as high as possible into the loop nest. The condition is
+%         assumed to evaluate to a boolean.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-switch.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
+%         allows us to cache certain computations in the state under which the
+%         body will execute. This is useful for computing and caching the results
+%         of expensive computations.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-thunk.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchsequence(head, tail)$: The Sequence looplet represents the
+%         concatenation of two looplets. Both arguments must be phase looplets, and
+%         are assumed to be nonoverlapping, covering, and in order.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-sequence.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchspike(body, tail)$ The Spike Looplet represents a run
+%         followed by a single value. In this paper, Spike will be considered a
+%         shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
+%         \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
+%         handled with special care, since they are an opportunity to align the
+%         final run to the end of the root loop extent, without using any special
+%         bounds inference.
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-spike.png}
+%     \end{minipage}
+%     \vspace{3pt}
+
+%     \begin{minipage}[c]{0.65\linewidth}
+%         $\finchstepper([seek], next, body)$ The stepper looplet
+%         represents a variable number of looplets, concatenated. Since our
+%         looplets may be skipped over due to conditions or various rewrites, the
+%         $seek$ function allows us to fast-forward the state to the start of the
+%         root loop extent when it comes time to lower the stepper. The $next$
+%         function advances the state to the next iteration of the stepper. 
+%     \end{minipage}%
+%     \begin{minipage}[c]{0.35\linewidth}
+%         \centering
+%         \includegraphics[scale=0.20]{Looplets-stepper.png}
+%     \end{minipage}
+%     \caption{The looplet language, as understood in a correct execution of a Finch program.}
+%     \vspace{-8pt}
+% \end{figure}
+
 \begin{figure}[ht]
+\begingroup
 \footnotesize
+\setlength{\tabcolsep}{10pt} 
+\renewcommand{\arraystretch}{1.5}
+\begin{tabular} {|l|c|} 
+\hline
     \begin{minipage}[c]{0.65\linewidth}
@@
         index.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
+    \end{minipage} &
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-lookup.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchrun(body)$: The Run Looplet represents a constant
-        region of an iterator. The body of the run is understood to have one
-        less dimension than the lookup itself, as all of the bodies are
-        identical.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-run.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchphase(c:d, body)$: The Phase Looplet represents a
-        restriction of the range on which a loop should execute, and allows us
-        to succinctly express the ranges on which children of compound looplets
-        are defined.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-phase.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchswitch(cond, head, tail)$: The Switch Looplet allows
-        us to specialize the body of a looplet based on a condition, evaluated
-        in the embedding context. If the condition is true, we use `head`,
-        otherwise we use `tail`. Switch has a high lowering priority so we can
-        see what's inside of it and lower that appropriately. This also lifts
-        the condition as high as possible into the loop nest. The condition is
-        assumed to evaluate to a boolean.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-switch.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
-        allows us to cache certain computations in the state under which the
-        body will execute. This is useful for computing and caching the results
-        of expensive computations.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-thunk.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchsequence(head, tail)$: The Sequence looplet represents the
-        concatenation of two looplets. Both arguments must be phase looplets, and
-        are assumed to be nonoverlapping, covering, and in order.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-sequence.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchspike(body, tail)$ The Spike Looplet represents a run
-        followed by a single value. In this paper, Spike will be considered a
-        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
-        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
-        handled with special care, since they are an opportunity to align the
-        final run to the end of the root loop extent, without using any special
-        bounds inference.
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-spike.png}
-    \end{minipage}
-    \vspace{3pt}
-
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchstepper([seek], next, body)$ The stepper looplet
-        represents a variable number of looplets, concatenated. Since our
-        looplets may be skipped over due to conditions or various rewrites, the
-        $seek$ function allows us to fast-forward the state to the start of the
-        root loop extent when it comes time to lower the stepper. The $next$
-        function advances the state to the next iteration of the stepper. 
-    \end{minipage}%
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-stepper.png}
-    \end{minipage}
-    \caption{The looplet language, as understood in a correct execution of a Finch program.}
-    \vspace{-8pt}
-\end{figure}
-
-\begin{figure}[ht]
-\footnotesize
-\begin{tabular} {|l|c|}
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchlookup(seek, body)$: The Lookup Looplet represents a
-        randomly accessible region of an iterator. The body of the lookup is
-        understood to have one less dimension than the lookup itself, as we have
-        already ``looked up'' that index in the tensor by the time we reach the
-        body. \texttt{seek(i)} is a function that updates state to the given
-        index.
-    \end{minipage}% &
-    \begin{minipage}[c]{0.35\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-lookup.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-run.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-phase.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-switch.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-thunk.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-sequence.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-spike.png}
-    \end{minipage} \\
+    \end{minipage} \\ \hline
     \begin{minipage}[c]{0.65\linewidth}
@@
     \end{minipage} &
-    \begin{minipage}[c]{0.35\linewidth}
+    \begin{minipage}[c]{0.25\linewidth}
         \centering
         \includegraphics[scale=0.20]{Looplets-stepper.png}
-    \end{minipage}
+    \end{minipage} \\\hline
 \end{tabular}
+\endgroup
+\vspace{-8pt}
 \caption{The looplet language, as understood in a correct execution of a Finch program.}

### 2024-04-04 sections/eval/api.tex
Update on Overleaf.

@@
 In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-They're either limited to vectors and matrices or only support tabular representations in order to reduce the complexity of interactions between different formats.
+In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
 Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
@@
 To avoid this, we have implemented a lazy evaluation approach that collects API calls and then computes output when requested by the user.
+
 \subsubsection{Finch Logic}

### 2024-04-04 sections/intro.tex
Update on Overleaf.
need to edit contributions

@@
 	\includegraphics[width=\linewidth]{Structures-examples.png}
+    \vspace{-12pt}
     \caption{A few examples of matrix structures arising in practice}
+    \vspace{-8pt}
 \end{figure}
@@
 
-\begin{table}[h!]
+\begin{table}[b!]
+\vspace{-12pt}
 \noindent
@@
 \item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
+\willow{add speedups and loc}
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
@@
 
-This advances the state of the art over the Looplets work, as the Looplets presented a way to merge iterators over multiple single dimensional structures, whereas we present a way to combine single-dimensional looplet structures into multi-dimensional tensors and operate on them with a programming language with fully-featured control flow.
+This advances the state-of-the-art over the Looplets work, as the Looplets presented a way to merge iterators over multiple single dimensional structures, whereas we present a way to combine single-dimensional looplet structures into multi-dimensional tensors and operate on them with a programming language with fully-featured control flow.

### 2024-04-04 sections/intro.tex
Update on Overleaf.

@@
 \end{figure}
+\kyle{In this paragraph, I wonder if it should be more "our world is full of structured data" which can be described by structured arrays.
+The "X describes Y" framing feels less fundamental than "Y is Z" like "images, masks, geometry, and transactions consist of long runs of identical values that can be compressed via run length encoding".
+This might be overly clunky though.}
 Our world is full of structured arrays.
@@
 %
-Operators, such as convolution, can be expressed as structured arrays.
+Even complex operators can be expressed as structured arrays.
 %
@@
 %
-\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs we implement in this work.
+\item Our compiler is highly extensible, evidenced by the variety of level formats and control flow constructs that we implement in this work.
 %
@@
 %
-\item We evaluate the efficiency, flexibility, and expressability of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and a high-level tensor operator fusion interface.
+\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and the Python Array API.
 \willow{add speedups and loc}
@@
 
-This advances the state-of-the-art over the Looplets work, as the Looplets presented a way to merge iterators over multiple single dimensional structures, whereas we present a way to combine single-dimensional looplet structures into multi-dimensional tensors and operate on them with a programming language with fully-featured control flow.
+This advances the state-of-the-art over the Looplets work.
+While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further combine single-dimensional looplet structures into multi-dimensional tensors and operate on them in a programming language with fully-featured control flow.

### 2024-04-04 sections/lang.tex
Update on Overleaf.
fixes

@@
 \begin{wrapfigure}{r}{0.51\linewidth}
-    \vspace{-26pt}
+    %\vspace{-26pt}
     \begin{minted}{julia}
@@
 addition of multiple left hand sides through code blocks, access with expressions, and explicit initialiation.}\label{fig:syntax}
-\vspace{-12pt}
+
 \end{wrapfigure}
@@
     \hypo{var\not\in d}
-    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t, d)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t, d\cup\{var\}) \rangle}}
+    \infer2[$Define$]{\splitfrac{\langle\finchdefine(var, val, body), (e, t, d)\rangle}{\rightarrow \langle body, (e[var \mapsto val'], t, \{\}) \rangle}}
     \end{prooftree}
@@
     \hypo{s = (e, t, d)}
-    \hypo{\langle head, (e, t, \{\}) \rangle \rightarrow s^{\prime}}
-    \hypo{\langle \finchblock(tail...), s^{\prime} \rangle \rightarrow (e^{\prime}, t^{\prime}, d^{\prime})})
-        \infer3[$BlockHead$]{\langle\finchblock(head, tail...), s \rangle \rightarrow (e^{\prime}, t^{\prime}, d) }
+    \hypo{\langle stmt_{0}, (e, t, \{\}) \rangle \rightarrow s_{1}}
+    \hypo{\langle stmt_{1}, s_{1} \rangle \rightarrow s_{2}}
+    \hypo{\dots}
+    \hypo{\langle stmt_{n}, s_{n} \rangle \rightarrow s_{n+1}}
+    \hypo{s_{n+1}= (e^{\prime}, t^{\prime}, d^{\prime})}
+        \infer6[$Block$]{\langle\finchblock(stmt_{0}, \dots, stmt_{n}), s \rangle \rightarrow (e^{\prime}, t^{\prime}, d) }
     \end{prooftree}
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle declare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
+    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchupdate], \llangle declare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
     \end{prooftree}
@@
     \hypo{s = (e, t,d )}
-    \hypo{e(\finchmode(tns)) = update}
+    \hypo{e(\finchmode(tns)) = \finchupdate}
     \hypo{tns\in d}
     \hypo{e(tns) = tns'}
-    \infer4[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto read], \llangle freeze(tns'), t \rrangle, d)}
+    \infer4[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchread], \llangle freeze(tns'), t \rrangle, d)}
     \end{prooftree}
@@
     \hypo{s = (e, t, d)}
-    \hypo{e(\finchmode(tns)) = read}
+    \hypo{e(\finchmode(tns)) = \finchread}
     \hypo{tns\in d}
@@
     \infer4[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle, d)}
+    \end{prooftree}
+
+\vspace{6pt}
+\begin{prooftree}
+    \hypo{e(tns) = tns'}
+    \hypo{\langle op, s\rangle \rightarrow op'}
+    \hypo{\langle rhs, s\rangle \rightarrow rhs'}
+    \infer[no rule]3{
+        \begin{prooftree}
+    \hypo{e(\finchmode(tns)) = \finchupdate}
+    \hypo{\llangle unfurl(tns', \finchupdate, op', rhs'), t\rrangle \rightarrow  t'}
+    \infer[simple]2[$Assign$]{\langle E[\finchassign(\finchaccess(tns), op, rhs)], (e, t, d)\rangle \rightarrow (e, t', d)}
+    \end{prooftree}}
 \end{prooftree}
@@
 \begin{multline*}
-E := [\cdot] | \finchloop(idx, ext, E | body) | \finchblock(E | head, E | tail) | \finchsieve(E | cond, E | body) |\\ 
+E := [\cdot] | \finchloop(idx, ext, E | body) | \finchblock(E | head, E | tail) | \finchsieve(E | cond, E | body) | \finchassign(E | lhs, op, E | rhs)\\ 
 \finchdeclare(var, E|rhs, E|body) | \finchcall(E|f, E|args...) | \finchaccess(tns, E | idxs...)
@@
     \hypo{e(tns) \mapsto tns'}
-    \hypo{\llangle unfurl(tns', ext)\rrangle \mapsto tns''}
-    \infer2[$Unfurl$]{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), s\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), s\rangle}
+    \hypo{e(\finchmode(tns)) \mapsto m}
+    \hypo{\llangle unfurl(tns', ext, m), t\rrangle \Rightarrow tns''}
+    \infer3[$Unfurl$]{\langle \finchloop(i, ext, E[\finchaccess(tns, j..., i)]), s\rangle \rightarrow \langle \finchloop(i, ext, E[\finchaccess(tns'', j..., i)]), s\rangle}
 \end{prooftree}
@@
 \begin{prooftree}
-    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(run(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
+    \infer0[$Run$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(\finchrun(body), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(body, j...)]), s\rangle}}
 \end{prooftree}
@@
     \hypo{\llangle seek(i'), t \rrangle \rightarrow t'}
-    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(lookup(seek, body), j..., i)], (e, t)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t')\rangle}}
+    \infer2[$Lookup$]{\splitfrac{\langle E[\finchaccess(\finchlookup(seek, body), j..., i)], (e, t, d)\rangle}{\rightarrow \langle E[\finchaccess(body, j...)], (e, t', d)\rangle}}
+\end{prooftree}
+\vspace{6pt}
+
+\begin{prooftree}
+    \infer0[$AcceptRun$]{\splitfrac{\langle \finchloop(i, \finchextent(a, b), E[\finchassign(\finchaccess(\finchrun(body), j..., i), op, rhs)]), s\rangle}{\rightarrow \langle \finchloop(i, \finchextent(a, b), E[\finchsieve(i == a, \finchassign(\finchaccess(body, j...), op, rhs))]), s\rangle}}
 \end{prooftree}
@@
 \begin{prooftree}
-    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
+    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, \finchextent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
 \end{prooftree}
@@
     \hypo{\llangle preamble, t \rrangle \rightarrow t'}
-    \hypo{\langle E[body], (e, t') \rangle \rightarrow (e', t'')}
+    \hypo{\langle E[body], (e, t', d) \rangle \rightarrow (e', t'', d)}
     \hypo{\llangle epilogue, t'' \rrangle \rightarrow t'''}
-    \infer3[$Thunk$]{\langle E[thunk(preamble, body, epilogue)], (e, t')\rangle \rightarrow (e', t''')}
+    \infer3[$Thunk$]{\langle E[thunk(preamble, body, epilogue)], (e, t', d)\rangle \rightarrow (e', t''', d)}
 \end{prooftree}
@@
     \hypo{\llangle seek(a), t \rrangle \rightarrow t'}
-    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t')\rangle}}
+    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t, d)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t', d)\rangle}}
 \end{prooftree}
@@
 \begin{prooftree}
-    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t) \rangle \rightarrow (e', t')}
+    \hypo{\langle \finchloop(i, ext, E[\finchaccess(body, j..., i)]), (e, t, d) \rangle \rightarrow (e', t', d)}
     \hypo{\llangle next, t' \rrangle \rightarrow t''}
-    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'')\rangle}}
+    \infer2[$StepperNext$]{\splitfrac{\langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), s\rangle}{\rightarrow \langle \finchloop(i, ext, E[\finchaccess(stepper(body, next), j..., i)]), (e', t'', d)\rangle}}
+\end{prooftree}
+
+    \vspace{6pt}
+
+%    \begin{prooftree}
+%    \hypo{s = (e, t, d)}
+%    \hypo{\langle a, s \rangle \Rightarrow a'}
+%    \hypo{\langle b, s \rangle \Rightarrow b'}
+%    \hypo{a' < b'}
+%    \hypo{\langle body, (e[i \mapsto a'], t, \{\}) \rangle \rightarrow s_2}
+%    \infer5[$Loop$]{\langle\finchloop(i, \finchextent(a, b), body), s_1\rangle \rightarrow \finchloop(i, \finchextent(a + 1, b), body), s_2\rangle}
+%    \end{prooftree}
+%    \hfill
+%    \vspace{6pt}
+
+    \begin{prooftree}
+    \infer0[$Loop$]{\splitfrac{\langle\finchloop(i, \finchextent(a, b), body), s\rangle}{\rightarrow \finchblock(\finchdeclare(i, a, body), \finchsieve(a < b, \finchloop(i, \finchextent(a + 1, b), body)), s\rangle}}
+    \end{prooftree}
+
+\vspace{6pt}
+    \begin{prooftree}
+    \hypo{e(tns) \mapsto tns'}
+    \hypo{e(\finchmode(tns)) \mapsto \finchread}
+    \hypo{\llangle unfurl(tns', \finchread), t\rrangle \rightarrow tns''}
+    \infer3[$Access$]{\langle E[\finchaccess(tns)], s\rangle \rightarrow \langle E[tns''], s\rangle} 
 \end{prooftree}
@@
     the priority in which we apply looplet rules, which is as follows:
-    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > Unfurl$
+    $Thunk > Phase > Switch > Simplify > Run > Spike > Sequence > StepperSeek > StepperNext > Lookup > AcceptRun > Unfurl > Loop > Access$.
     %

### 2024-04-04 sections/intro.tex
Update on Overleaf.

@@
 This advances the state-of-the-art over the Looplets work.
-While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further combine single-dimensional looplet structures into multi-dimensional tensors and operate on them in a programming language with fully-featured control flow.
+While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further, combining single-dimensional looplet structures into multi-dimensional tensors and operate on them in a programming language with fully-featured control flow.

### 2024-04-04 sections/intro.tex
Update on Overleaf.

--- 2024-04-04 sections/intro.tex
@@
 This advances the state-of-the-art over the Looplets work.
-While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further combine single-dimensional looplet structures into multi-dimensional tensors and operate on them in a programming language with fully-featured control flow.
+While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further by combining single-dimensional looplets into multi-dimensional tensors and operating on them in a programming language with fully-featured control flow.

### 2024-04-04 sections/background.tex
fix

@@
 \footnotesize
-\setlength{\tabcolsep}{10pt} 
-\renewcommand{\arraystretch}{1.5}
-\begin{tabular} {|l|c|} 
+%\setlength{\tabcolsep}{10pt} 
+\begin{tabular} {|p{0.65\linewidth}|c|} 
 \hline
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchlookup(seek, body)$: The Lookup Looplet represents a
-        randomly accessible region of an iterator. The body of the lookup is
-        understood to have one less dimension than the lookup itself, as we have
-        already ``looked up'' that index in the tensor by the time we reach the
-        body. \texttt{seek(i)} is a function that updates state to the given
-        index.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-lookup.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchrun(body)$: The Run Looplet represents a constant
-        region of an iterator. The body of the run is understood to have one
-        less dimension than the lookup itself, as all of the bodies are
-        identical.
-    \end{minipage} &
+    $\finchlookup(seek, body)$: The Lookup Looplet represents a randomly accessible region of an iterator.
+The body of the lookup is understood to have one less dimension than the lookup itself, as we have already ``looked up'' that index in the tensor by the time we reach the body.
+\texttt{seek(i)} is a function that updates state to the given index.
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-lookup.png}}
+    \\
+    \hline
+    $\finchrun(body)$: The Run Looplet represents a constant region of an iterator.
+The body of the run is understood to have one less dimension than the lookup itself, as all of the bodies are identical.
+     &
     \begin{minipage}[c]{0.25\linewidth}
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchphase(c:d, body)$: The Phase Looplet represents a
-        restriction of the range on which a loop should execute, and allows us
-        to succinctly express the ranges on which children of compound looplets
-        are defined.
+        $\finchphase(c:d, body)$: The Phase Looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
     \end{minipage} &
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchswitch(cond, head, tail)$: The Switch Looplet allows
-        us to specialize the body of a looplet based on a condition, evaluated
-        in the embedding context. If the condition is true, we use `head`,
-        otherwise we use `tail`. Switch has a high lowering priority so we can
-        see what's inside of it and lower that appropriately. This also lifts
-        the condition as high as possible into the loop nest. The condition is
-        assumed to evaluate to a boolean.
+        $\finchswitch(cond, head, tail)$: The Switch Looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context.
+If the condition is true, we use `head`, otherwise we use `tail`.
+Switch has a high lowering priority so we can see what's inside of it and lower that appropriately.
+This also lifts the condition as high as possible into the loop nest.
+The condition is assumed to evaluate to a boolean.
     \end{minipage} &
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
-        allows us to cache certain computations in the state under which the
-        body will execute. This is useful for computing and caching the results
-        of expensive computations.
+        $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet allows us to cache certain computations in the state under which the body will execute.
+This is useful for computing and caching the results of expensive computations.
     \end{minipage} &
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchsequence(head, tail)$: The Sequence looplet represents the
-        concatenation of two looplets. Both arguments must be phase looplets, and
-        are assumed to be nonoverlapping, covering, and in order.
+        $\finchsequence(head, tail)$: The Sequence looplet represents the concatenation of two looplets.
+Both arguments must be phase looplets, and are assumed to be nonoverlapping, covering, and in order.
     \end{minipage} &
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchspike(body, tail)$ The Spike Looplet represents a run
-        followed by a single value. In this paper, Spike will be considered a
-        shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
-        \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
-        handled with special care, since they are an opportunity to align the
-        final run to the end of the root loop extent, without using any special
-        bounds inference.
+        $\finchspike(body, tail)$ The Spike Looplet represents a run followed by a single value.
+In this paper, Spike will be considered a shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)), \finchphase(j:j, \finchrun(tail)))$.
+In the Finch compiler, spikes are handled with special care, since they are an opportunity to align the final run to the end of the root loop extent, without using any special bounds inference.
     \end{minipage} &
@@
     \begin{minipage}[c]{0.65\linewidth}
-        $\finchstepper([seek], next, body)$ The stepper looplet
-        represents a variable number of looplets, concatenated. Since our
-        looplets may be skipped over due to conditions or various rewrites, the
-        $seek$ function allows us to fast-forward the state to the start of the
-        root loop extent when it comes time to lower the stepper. The $next$
-        function advances the state to the next iteration of the stepper. 
+        $\finchstepper([seek], next, body)$ The stepper looplet represents a variable number of looplets, concatenated.
+Since our looplets may be skipped over due to conditions or various rewrites, the $seek$ function allows us to fast-forward the state to the start of the root loop extent when it comes time to lower the stepper.
+The $next$ function advances the state to the next iteration of the stepper.
     \end{minipage} &

### 2024-04-04 sections/background.tex
fix

@@
 \begin{figure}[ht]
-\begingroup
 \footnotesize
-%\setlength{\tabcolsep}{10pt} 
 \begin{tabular} {|p{0.65\linewidth}|c|} 
@@
         &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-run.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-run.png}}
+    \\ 
+    \hline
     $\finchphase(c:d, body)$: The Phase Looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-phase.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-phase.png}}
+    \\ 
+    \hline
     $\finchswitch(cond, head, tail)$: The Switch Looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context.
@@
 The condition is assumed to evaluate to a boolean.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-switch.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-switch.png}}
+    \\ 
+    \hline
     $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet allows us to cache certain computations in the state under which the body will execute.
 This is useful for computing and caching the results of expensive computations.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-thunk.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-thunk.png}}
+    \\ 
+    \hline
     $\finchsequence(head, tail)$: The Sequence looplet represents the concatenation of two looplets.
 Both arguments must be phase looplets, and are assumed to be nonoverlapping, covering, and in order.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-sequence.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchspike(body, tail)$ The Spike Looplet represents a run followed by a single value.
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-sequence.png}}
+    \\ 
+    \hline
+    $\finchspike(body, tail)$: The Spike Looplet represents a run followed by a single value.
 In this paper, Spike will be considered a shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)), \finchphase(j:j, \finchrun(tail)))$.
 In the Finch compiler, spikes are handled with special care, since they are an opportunity to align the final run to the end of the root loop extent, without using any special bounds inference.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-spike.png}
-    \end{minipage} \\ \hline
-    \begin{minipage}[c]{0.65\linewidth}
-        $\finchstepper([seek], next, body)$ The stepper looplet represents a variable number of looplets, concatenated.
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-spike.png}}
+    \\ 
+    \hline
+    $\finchstepper([seek], next, body)$: The stepper looplet represents a variable number of looplets, concatenated.
 Since our looplets may be skipped over due to conditions or various rewrites, the $seek$ function allows us to fast-forward the state to the start of the root loop extent when it comes time to lower the stepper.
 The $next$ function advances the state to the next iteration of the stepper.
-    \end{minipage} &
-    \begin{minipage}[c]{0.25\linewidth}
-        \centering
-        \includegraphics[scale=0.20]{Looplets-stepper.png}
-    \end{minipage} \\\hline
+    &
+    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-stepper.png}}
+    \\
+    \hline
     \end{tabular}
-\endgroup
 \vspace{-8pt}

### 2024-04-04 sections/eval/api.tex

@@
 Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from DuckDB's use of binary join plans which, while not materializing intermediates, don't optimally intersect the nonzero indices for cyclic queries \footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
+For triangle counting and SDDMM, this improvement stems from Finch's efficient intersection of the nonzero indices as opposed to DuckDB's use of binary join plans\footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
 For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.

### 2024-04-04 sections/background.tex
fixes for completeness

@@
 Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
-can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages \cite{ahrens_looplets_2023}.
+can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.
 
@@
 It is assumed that if a looplet introduces a variable to be used in a child looplet, the child looplet will not modify that variable.
-
-% \begin{figure}[ht]
-%     \footnotesize
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchlookup(seek, body)$: The Lookup Looplet represents a
-%         randomly accessible region of an iterator. The body of the lookup is
-%         understood to have one less dimension than the lookup itself, as we have
-%         already ``looked up'' that index in the tensor by the time we reach the
-%         body. \texttt{seek(i)} is a function that updates state to the given
-%         index.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-lookup.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchrun(body)$: The Run Looplet represents a constant
-%         region of an iterator. The body of the run is understood to have one
-%         less dimension than the lookup itself, as all of the bodies are
-%         identical.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-run.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchphase(c:d, body)$: The Phase Looplet represents a
-%         restriction of the range on which a loop should execute, and allows us
-%         to succinctly express the ranges on which children of compound looplets
-%         are defined.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-phase.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchswitch(cond, head, tail)$: The Switch Looplet allows
-%         us to specialize the body of a looplet based on a condition, evaluated
-%         in the embedding context. If the condition is true, we use `head`,
-%         otherwise we use `tail`. Switch has a high lowering priority so we can
-%         see what's inside of it and lower that appropriately. This also lifts
-%         the condition as high as possible into the loop nest. The condition is
-%         assumed to evaluate to a boolean.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-switch.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet
-%         allows us to cache certain computations in the state under which the
-%         body will execute. This is useful for computing and caching the results
-%         of expensive computations.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-thunk.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchsequence(head, tail)$: The Sequence looplet represents the
-%         concatenation of two looplets. Both arguments must be phase looplets, and
-%         are assumed to be nonoverlapping, covering, and in order.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-sequence.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchspike(body, tail)$ The Spike Looplet represents a run
-%         followed by a single value. In this paper, Spike will be considered a
-%         shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)),
-%         \finchphase(j:j, \finchrun(tail)))$.  In the Finch compiler, spikes are
-%         handled with special care, since they are an opportunity to align the
-%         final run to the end of the root loop extent, without using any special
-%         bounds inference.
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-spike.png}
-%     \end{minipage}
-%     \vspace{3pt}
-
-%     \begin{minipage}[c]{0.65\linewidth}
-%         $\finchstepper([seek], next, body)$ The stepper looplet
-%         represents a variable number of looplets, concatenated. Since our
-%         looplets may be skipped over due to conditions or various rewrites, the
-%         $seek$ function allows us to fast-forward the state to the start of the
-%         root loop extent when it comes time to lower the stepper. The $next$
-%         function advances the state to the next iteration of the stepper. 
-%     \end{minipage}%
-%     \begin{minipage}[c]{0.35\linewidth}
-%         \centering
-%         \includegraphics[scale=0.20]{Looplets-stepper.png}
-%     \end{minipage}
-%     \caption{The looplet language, as understood in a correct execution of a Finch program.}
-%     \vspace{-8pt}
-% \end{figure}
+For more background on Looplets, we recommend the original work \cite{ahrens_looplets_2023}.
 
@@
 \vspace{-8pt}
-\caption{The looplet language, as understood in a correct execution of a Finch program.}
+\caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
 \end{figure}

### 2024-04-04 sections/intro.tex
fixes for completeness

@@
 %
-\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies, showing that Finch can be used to accelerate a wide range of applications,  from classic operations such as spmv and spgemm, to more complex applications such as image processing, graph analytics, and the Python Array API.
-\willow{add speedups and loc}
+\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as spmv ($2.51\times$) and spgemm ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $0.25\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
@@
 
-This advances the state-of-the-art over the Looplets work.
+This advances the state-of-the-art over the looplets work \cite{ahrens_looplets_2023}.
 While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further by combining single-dimensional looplets into multi-dimensional tensors and operating on them in a programming language with fully-featured control flow.

### 2024-04-04 sections/tensors.tex
improvements
Update on Overleaf.
fixes for completeness

@@
 The $assemble$ function manages the allocation of new data in the tensor.
+
+\begin{wrapfigure}{R}{0.5\linewidth}
+    \centering
+    \vspace{-8pt}
+    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
+    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
+    \vspace{-8pt}
+    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
+    \label{fig:levelsvsfibers}
+    \vspace{-8pt}
+\end{wrapfigure}
 
@@
 
-\begin{wrapfigure}{R}{0.5\linewidth}
-    \centering
-    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
-    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
-    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
-    \label{fig:levelsvsfibers}
-\end{wrapfigure}
-
 Instead of storing the data for each subfiber separately, most sparse tensor
@@
 \begin{figure}[htbp]
-    \footnotesize
+    \scriptsize
     \raggedright
@@
 
+\vspace{3pt}
+
 $freeze(lvl)$: Finalizes the updates in the level, and readies the
 level for reading. Requires the level to be in update-only mode.
+
+\vspace{3pt}
 
@@
 
-$unfurl(fiber(lvl, pos), ext, mode)$: Unfurls the fiber at position
+\vspace{3pt}
+
+$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
 $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
@@
 \finchupdate$, returns a looplet nest over mutable subfibers in the update-only
-fiber. Often, skipping over mutable locations allows the level to know which
-locations must be stored. Additionally, a dirty bit may be used to communicate
+fiber. When applied to a scalar or leaf node, $ext$ is omitted and we may also use $op$ and $rhs$ to update the scalar value. Often, skipping over mutable locations allows the level to know which
+locations must be stored. Often, a dirty bit may be used in $tns$ to communicate
 whether the mutable subfiber has been written to, which allows the parent fiber
 to know whether the subfiber must be stored explicitly.
+
+\vspace{3pt}
 
@@
 or bytemaps) may support reassembly.
+\vspace{-8pt}
 \caption{The five functions that define a level.}
+\vspace{-12pt}
 \end{figure}
@@
 
+   We consider all combinations of these structural properties in Table
+    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
+    the the 8 resulting situations. While it is impossible to write code which
+    precisely addresses every possible structure, the combination of our level
+    formats can express a wide variety of hierarchical structures to a
+    sufficient granularity that we can generate code which utilizes the four
+    properties.  For example, though banded arrays are a superset of ragged
+    arrays, representing a ragged array using a banded format does not add much
+    overhead.  The structures we consider are exhaustive in the sense that they
+    address all combinations of sparsity, blocks, runs, and singletons in each
+    level. We can represent a wide variety of hierarchical tensor structures by
+    combining these level structures in a tree, as shown in
+    Figure~\ref{fig:structuraldiversity}.
+
     \begin{wrapfigure}{r}{.35\textwidth}
         \centering
-        \footnotesize
+        \vspace{-8pt}
+        \scriptsize
         \begin{tabular}{|c|c|c|c|l|}
@@
         \end{tabular}
+        \vspace{-8pt}
         \caption{All combinations of relevant structural properties and their
@@
         \label{tab:TypesOfStructure}
+        \vspace{-12pt}
     \end{wrapfigure}
-
-    We consider all combinations of these structural properties in Table
-    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
-    the the 8 resulting situations. While it is impossible to write code which
-    precisely addresses every possible structure, the combination of our level
-    formats can express a wide variety of hierarchical structures to a
-    sufficient granularity that we can generate code which utilizes the four
-    properties.  For example, though banded arrays are a superset of ragged
-    arrays, representing a ragged array using a banded format does not add much
-    overhead.  The structures we consider are exhaustive in the sense that they
-    address all combinations of sparsity, blocks, runs, and singletons in each
-    level. We can represent a wide variety of hierarchical tensor structures by
-    combining these level structures in a tree, as shown in
-    Figure~\ref{fig:structuraldiversity}.
 
@@
         \includegraphics[width=\linewidth]{Structures.png}\hfill%
+        \vspace{-8pt}
         \caption{Several examples of matrix structures represented using the
@@
         \label{fig:structuraldiversity}
+        \vspace{-18pt}
     \end{figure}
@@
     \centering
-    \footnotesize
+    \scriptsize
+    \renewcommand{\arraystretch}{1.3}
     \begin{tabular}{p{14cm}}

### 2024-04-04 sections/lang.tex
done

@@
 % FIXME: ???
-Finally, as detailed in the previous section, tensors are defined externally via an interface that supports the declare, freeze, thaw, and unfurl functions.
+Finally, as detailed in the previous section, tensors are defined externally via an interface that supports the $declare$, $freeze$, $thaw$, and $unfurl$ functions.
 %
@@
 %
-For example, in Figure~\ref{fig:semantics_core}, we declare the core semantics of the language and the Define rule (\mintinline{julia}{let}) adds a definition to the compiler state whereas the Call rule passes evaluation off to the host.
+For example, in Figure~\ref{fig:semantics_core}, we declare the core semantics of the language and the $Define$ rule (\mintinline{julia}{let}) adds a definition to the compiler state whereas the $Call$ rule passes evaluation off to the host.
 %
@@
 %
-Second, loops with tensor accesses enter into a looplet system via the Unfurl rule.
+Second, loops with tensor accesses enter into a looplet system via the $Unfurl$ rule.
 %
-In this system, repeated structures (Run) and constant information (Switch)  are slowly uncovered as accesses are lowered in various points in the program.
+In this system, repeated structures and constants are slowly uncovered as accesses are lowered in various points in the program (e.g. $Run$  and $Switch$, respectively).
 %
-In this process, we are able to use rewrite rules (Simplify) to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
+In this process, we are able to use rewrite rules in $Simplify$ to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
 %
 
-Though the user may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
+Though the reader may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
 %
@@
 
-% \begin{multline*}
-%         Modes:=update \mid read \mid undefined \mid defined
-% \end{multline*}
     \begin{prooftree}
@@
     \begin{prooftree}
-        \hypo{}
-        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t, d)\rangle \Rightarrow \llangle ex, t \rrangle}
-    \end{prooftree}
-    \vspace{6pt}
-
-    \begin{prooftree}
         \infer0[$Variable$]{\splitfrac{\langle\finchvar(name), (e, t, d)\rangle}{\rightarrow e(\finchvar(name))}}
@@
     \begin{prooftree}
-    \hypo{s = (e, t, d)}
-    \hypo{\langle stmt_{0}, (e, t, \{\}) \rangle \rightarrow s_{1}}
-    \hypo{\langle stmt_{1}, s_{1} \rangle \rightarrow s_{2}}
-    \hypo{\dots}
-    \hypo{\langle stmt_{n}, s_{n} \rangle \rightarrow s_{n+1}}
-    \hypo{s_{n+1}= (e^{\prime}, t^{\prime}, d^{\prime})}
-        \infer6[$Block$]{\langle\finchblock(stmt_{0}, \dots, stmt_{n}), s \rangle \rightarrow (e^{\prime}, t^{\prime}, d) }
+    \hypo{\langle body, s \rangle \rightarrow s'}
+    \infer1[$Block$]{\langle\finchblock(body, tail...), s \rangle \rightarrow \langle \finchblock(tail...), s' \rangle }
     \end{prooftree}
     \hfill
+    \begin{prooftree}
+        \hypo{}
+        \infer1[$Value$]{\langle\finchvalue(ex, type), (e, t, d)\rangle \Rightarrow \llangle ex, t \rrangle}
+    \end{prooftree}
 
@@
     \hypo{\forall i \langle init, dims_i \rangle \Rightarrow dims'_i}
-    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchupdate], \llangle declare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
+    \infer5[$Declare$]{\langle \finchdeclare(tns, init, dims), s\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchupdate], \llangle \finchdeclare(tns', init', dims'...), t \rrangle, d\cup\{ tns\})}
     \end{prooftree}
@@
     \hypo{e(tns) = tns'}
-    \infer4[$Freeze$]{\langle\finchfreeze(tns), (e, t)\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchread], \llangle freeze(tns'), t \rrangle, d)}
+    \infer4[$Freeze$]{\langle\finchfreeze(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchread], \llangle freeze(tns'), t \rrangle, d)}
     \end{prooftree}
@@
     \hypo{e(tns) = tns'}
-    \infer4[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto update], \llangle thaw(tns'), t \rrangle, d)}
+    \infer4[$Thaw$]{\langle\finchthaw(tns), s\rangle \rightarrow (e [\finchmode(tns) \mapsto \finchupdate], \llangle thaw(tns'), t \rrangle, d)}
     \end{prooftree}
@@
     Several Looplets introduce variables into the embedding
-    language, which may be read when evaluating the \finchvalue node.
+    language, which may be read when evaluating the {\finchvalue} node.
 %
@@
     but these semantics enforce that each of these functions may update state in
-    the host language and flip the mode of the tensor between read and update}
+    the host language and flip the mode of the tensor between {\finchread} and {\finchupdate}}
     \label{fig:semantics_core}
@@
 \begin{prooftree}
-    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, \finchextent(a, b), E[\finchaccess(phase(extent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, extent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
+    \infer0[$Phase$]{\splitfrac{\langle \finchloop(i, \finchextent(a, b), E[\finchaccess(\finchphase(\finchextent(c, d), body), j..., i)]), s\rangle}{\rightarrow \langle \text{\finchloop}(i, \finchextent(max(a, c), min(b, d)), E[\text{\finchaccess}(body, j..., i)]), s\rangle}}
 \end{prooftree}
@@
     \hypo{\llangle seek(a), t \rrangle \rightarrow t'}
-    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t, d)\rangle}{\rightarrow \langle \finchloop(i, extent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t', d)\rangle}}
+    \infer1[$StepperSeek$]{\splitfrac{\langle \finchloop(i, \finchextent(a, b), E[\finchaccess(stepper(seek, body, next), j..., i)]), (e, t, d)\rangle}{\rightarrow \langle \finchloop(i, \finchextent(a, b), E[\finchaccess(stepper(body, next), j..., i)]), (e, t', d)\rangle}}
 \end{prooftree}
@@
     \begin{prooftree}
-    \infer0[$Loop$]{\splitfrac{\langle\finchloop(i, \finchextent(a, b), body), s\rangle}{\rightarrow \finchblock(\finchdeclare(i, a, body), \finchsieve(a < b, \finchloop(i, \finchextent(a + 1, b), body)), s\rangle}}
+    \infer0[$Loop$]{\splitfrac{\langle\finchloop(i, \finchextent(a, b), body), s\rangle}{\rightarrow \langle \finchblock(\finchdefine(i, a, body), \finchsieve(a < b, \finchloop(i, \finchextent(a + 1, b), body))), s\rangle}}
     \end{prooftree}
@@
     %
-    This allows us to treat the \finchvalue node as a
+    This allows us to treat the {\finchvalue} node as a
     constant.

### 2024-04-05 sections/eval/graphs.tex
need to edit contributions
Update on Overleaf.

@@
 
-We used Finch to implement two fundamental graph algorithms: Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
-Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
+We used Finch to implement both Breadth-first search (BFS) and Bellman-Ford single-source shortest path.
+%
+Our BFS implementation and graphs datasets are taken from Yang et al.~\cite{yang_implementing_2018}, including both road networks and scale-free graphs (bounded node degree vs. power law node degree).
+
+%road networks, characterized by bounded node degrees and long diameters, and scale-free graphs, where node degree distribution follows a power-law distribution and diameters are short.
 
 Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
+%
 Push traversal visits the neighbors of each frontier node, while pull traversal visits every node and checks to see if it has a neighbor in the frontier. 
+%
 The advantage of pull traversal is that we may terminate our search once we find a node in the frontier, saving time in the event the push traversal were to visit most of the graph anyway. 
+%
 Early break is the critical part of control flow in this algorithm, though the algorithms also require different loop orders, multiple outputs, and custom operators.
 
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, an open-source Julia library, and the LAGraph Library, which implements graph algorithms in the language of linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
+Figure~\ref{fig:graph_result} compares performance to Graphs.jl, a Julia library, and the LAGraph Library, which implements graph algorithms with sparse linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
+%
 For the BFS algorithm, direction-optimization notably enhances performance for scale-free graphs. 
-Although GraphBLAS uses hardwired optimizations, Finch is competitive on these benchmarks.
+%
+Although GraphBLAS uses hardwired optimizations, Finch is competitive. 
+%
 On Bellman-Ford (with path lengths and shortest-path tree), Finch's support for multiple outputs, sparse inputs, and masks leads to superior performance over GraphBLAS (average speedup of 3.53). 
-Note that we did not include GAP-road as it took too long to run.
-%speedup is arith-mean
+%
+We did not include GAP-road as it timed out. %speedup is arith-mean
+%
+In Appendix~\ref{sec:graph_listings}, we display the code for BFS and Bellman-Ford in Finch (57 and 50 LOC) and LAGraph (215 and 227 LOC), and invite readers to compare the clarity of the algorithms.
 

### 2024-04-05 sections/eval/morphology.tex
Update on Overleaf.

@@
 Some image processing pipelines stand to benefit from structured data processing \cite{donenfeld_unified_2022}.
-In this case study, we focus on binary image morpology and the logical transformation of binary images and masks.
+%
+We focus on binary image morpology and the logical transformation of binary images and masks.
+%
 We consider two operations: binary erosion (computing a mask), and a masked histogram (using a mask to avoid work).
+%
+We use images are all binary, either by design or having been thresholded.
 
-Samples from our datasets are shown in Figure~\ref{fig:image_datasets}. Note that these images are all binary, either by design or having been thresholded.
-
-Finch allows us to modify both our algorithm and our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
+Finch allows us choose our datastructure, so we may choose to use either a dense representation with bytes ($Dense(Element(0x00))$), a bit-packed representation ($Dense(Element(UInt64))$), or a run-length encoded representation that represents runs of true or false ($SparseRLE(Pattern())$).
 %
@@
 %
-The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes between true and false.
+The dense representation induces the least overhead, the bit-packed representation can take advantage of bitwise binary ops, and the run-length encoded version only uses memory and compute when the pattern changes.
+% between true and false.
 
+Similarly, since Finch let's us choose our algorithm we can implement erosion in a few ways.
+%
 The erosion operation turns off a pixel unless all of it's neighbors are on.
@@
 This can be used to shrink the boundaries of a mask, and remove point instances of noise \cite{fisher_hypermedia_1996}.
-\help{link a few cool images that show erosion}
+%\help{link a few cool images that show erosion}
 %
-Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
+This introduces three instances of structure in the control flow: the mask, the padding of inputs, and the convolutional filter.
 %
-We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor.
-Figure \ref{fig:erosion_listing} displays example erosion algorithms for bitwise
-or run-length-encoded algorithms, and a masked histogram kernel.
+We focused on the filter.
+%Note that three instances of structured data are used here, both in the mask, in the padding of the inputs, and in the convolutional filter itself.
+%
+We might understand the filter as a structured tensor of circular shifts, or we might understand each shifted view of the data in an unrolled stencil computation as a structured tensor, or a two part stencil where we compute the horizontal then vertical part of the stencil.
+%
+We experimented with these options and found that the last approach performed best, due to fitting the storage formats while reducing the amount of work with intermediate temporaries.
+%
+Figure \ref{fig:morphology_listing} displays example erosion algorithms for bitwise
+or run-length-encoded algorithms.
+%
+% and a masked histogram kernel.
 
-We compare to OpenCV, and chose the histogram since the OpenCV histogram function also accepts a mask. If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
+We compared against OpenCV.
+%
+We used four datasets.
+We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
+%
+We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
+%
+All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions. 
+%
+IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
+%
+We believe the 51.6x on MNIST is due to calling overhead in OpenCV). 
+%
+The bitwise kernels were effective as well, and would be more effective on datasets with less structure. 
+%
+A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
+
+We also implemented a histogram kernel.
+%
+We used an indirect access into the output to implement this (Figure \ref{fig:morphology_listing}), something not many sparse frameworks support.
+%
+We compare to OpenCV since the OpenCV histogram function also accepts a mask. 
+%
+If we use $SparseRLE(Pattern())$ for our mask, we can reduce the branching
 in the masked kernel and get better performance.
 %
-We this is also an instance where we use a computed index into the output, something not many sparse and structured programming frameworks support.
-
-We used four datasets.
-We randomly selected 100 images from the MNIST \cite{lecun_gradient-based_1998} and Omniglot \cite{lake_human-level_2015} character recognition datasets, as well as a dataset of human line drawings \cite{eitz_how_2012}.
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}.
-All images were thresholded, and we also include versions of the images that have been magnified before thresholding, to induce larger constant regions.
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
-The improvements with SparseRLE are seen in the histogram task as well, as it allows us to
-mask off contiguous regions of computation, instead of individual pixels, reducing the branches in the mask and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches.
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV).
-The bitwise kernels were effective as well, and would be more effective on datasets with less structure.
-A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
+The improvements with SparseRLE are seen in the histogram task too, as it allows us to mask off contiguous regions of computation, instead of individual pixels, reducing the branches and leading to a significant speedup (20.3x on Omniglot and 20.8x on sketches).
+%
+%
+%A strength of Finch is that it supports structured datasets, even over bitwise operations, allowing us to implement the bitwise kernel and then mask it.
 

### 2024-04-05 sections/eval/spgemm.tex
Update on Overleaf.

@@
 
-Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
+%Sparse Matrix-Matrix Multiplication (SpGEMM) is a fundamental operation in scientific computing and data analytics. 
 We compute the $M \times N$ sparse matrix $C$ as the product of $M \times K$ and $K \times N$ sparse matrices $A$ and $B$.
-
+%
 There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
@@
 Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
+%
+We will show that our ability to implement these latter methods exceeds that of TACO, translating to asymptotic benefits. 
+%
 
@@
 %
-Note that these algorithms mainly differ in their loop order, but that different datastructures can be used to support the various access patterns induced.
-%
-Although a sparse bytemap has a dense memory footprint, we use it in our Finch implemenation of Gustavson's for the smaller $O(N)$ intermediate.
+Note that these algorithms mainly differ in their loop order, but that different data structures can be used to support the various access patterns induced.
 %
@@
 %
-Unfortunately, TACO does not support multidimensional sparse workspaces, only supporting outer-products with a dense output, a limitation with asymptotic consequences.
+Since, TACO does not support multidimensional sparse workspaces, its outer products uses a dense intermediate, which leads to an asymptotic slow down shown in Figure~\ref{fig:spgemm}.
 %
-It is also worth mentioning that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporary formats and transpositions.
-
+Similarly, although a sparse bytemap has a dense memory footprint, we use it in our Finch implementation of Gustavson's for the smaller $O(N)$ intermediate.
+%
+We note that the bytemap format in TACO's Gustavson's implementation is hard-wired, whereas Finch's programming model allows us to write algorithms with explicit temporaries and transpositions.
+%
+Without such hard wiring, TACO would have to use a dense intermediate to support random writes.
+%
+But without if conditions, TACO would then be forced to propagate all values in the dense vector to the sparse output matrix, turning the output dense and leading to the same asymptotic results as in the case of outer products.
+%
 As depicted in Figure~\ref{fig:spgemm}, Finch achieves comparable performance with TACO on smaller matrices when we use the same datastructures, and significant improvements when we use better datastructures.
-Finch outperforms TACO on larger matrices, with an average speedup of 1.25.
-%arith-mean
+%
+Finch outperforms TACO on larger matrices, with an average speedup of 1.25. %arith-mean
 

### 2024-04-05 sections/eval/spmv.tex
Update on Overleaf.

@@
   \subsection{Sparse Matrix-Vector Multiply (SpMV)}
-  Sparse matrix-vector multiplication (SpMV) is perhaps the most studied sparse
-  kernel, with a wide range of applications \cite{liu_csr5_2015,
-  zhou_enabling_2020}. Because SpMV is a bandwidth bound kernel, many formats have
-  been proposed to reduce the footprint with a minimal impact on complexity
-  \cite{langr_evaluation_2016}. The wide range of applications unsurprisingly
+  Sparse matrix-vector multiplication (SpMV) has a wide range of applications and has been thoroughly studied~\cite{liu_csr5_2015,
+  zhou_enabling_2020}. 
+  %
+  Because SpMV is bandwidth bound, many formats have
+  been proposed to reduce the footprint~\cite{langr_evaluation_2016}. 
+  %
+  The wide range of applications unsurprisingly
   results in a wide range of array structures, making it an effective kernel to
   demonstrate the utility of our programming model. 
-
-  We varied both the data formats and the programs we used them with. Our formats
+  %
+  We varied both the data formats and the SpMV algorithm. 
+  %
+  Our formats
   are shown in Table~\ref{spmv_tensor_formats}, and our programs are listed in
   Figure~\ref{spmv_programs}.
-
+  %
   In varying the program, we consider both row and column-major SpMV programs, as
-  well as a symmetric variant of SpMV. Finch enables us to exploit symmetry
-  effectively. Our program restricts our attention to the canonical trial (using
+  well as a symmetric SpMV. 
+  %
+  Finch enables us to exploit symmetry
+  effectively.
+  %
+  Our program restricts our attention to the canonical triangle (using
   masks), reuses the reads to the canonical triangle of the symmetric matrix
@@
   locations (using multiple outputs).
-
+  %
   All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
@@
   % or blocks of varying sizes as well as a permutation matrix to encapsulate a few additional use cases.
-  The dense vector is randomly generated.
+  %The dense vector is randomly generated. 
+  %
   We tested using the row-major and column-major Finch programs in Figure \ref{spmv_programs} as well as the symmetric program where applicable; the performance displayed for Finch on each dataset is the fastest among the formats and programs we tested.
-Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) in TACO so we use column-major SpMV in TACO as our baseline.
+For TACO, Column-major SpMV consistently performs better than row-major SpMV (an average of 1.36x better) so we use column-major SpMV in TACO as our baseline.
 
   We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
+  %
   Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
+  %
   Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
+  %
   There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
+  %
   For example, the SparseList-Pattern for email-Eu-core resulted in a speedup of 2.51, while the SparseList-Element format resulted in a slowdown of 1.84 over TACO.
+  %
   Every symmetric matrix in the SparseList and SparseList-Pattern formats has better performance when we use a Finch SpMV program that takes advantage of this symmetry.
+  %
   Symmetric SpMV with the SparseList level format in Finch results in an average of 1.27x speedup over TACO and symmetric SpMV with the SparseList-Pattern format in Finch results in an average speedup of 1.21x over TACO.
+  %
   Notably, there is a 1.91x speedup for the HB/saylr4 matrix over TACO.
@@
 
-  \begin{wrapfigure}{r}{0.41\linewidth}
-      \begin{minipage}[t]{0.18\textwidth}
-          \vspace{0pt} % Add this to ensure top alignment within minipage
-          \begin{minted}{julia}
-              y .= 0
-              for j = _, i = _
-                y[i] += A[i, j] * x[j]
-              end
-          \end{minted}
-          \vspace{24pt} % Add this to ensure top alignment within minipage
-          \begin{minted}{julia}
-              y .= 0
-              for j = _, i = _
-                y[j] += A[i, j] * x[i]
-              end
-          \end{minted}
-      \end{minipage}\hfill%
-      \begin{minipage}[t]{0.22\textwidth}
-          \vspace{0pt} % Add this to ensure top alignment within minipage
-          \begin{minted}{julia}
-              y .= 0
-              for j = _
-                let x_j = x[j]
-                  y_j .= 0
-                  for i = _
-                    let A_ij = A[i, j]
-                      y[i] += x_j * A_ij
-                      y_j[] += A_ij * x[i]
-                    end
-                  end
-                  #D is the diagonal
-                  y[j] += y_j[] + D[j] * x_j
-                end
-              end
-          \end{minted}
-      \end{minipage}
-      \vspace{-8pt}
-      \caption{Finch row-major, column-major and symmetric SpMV Programs}
-      \label{spmv_programs}
-      \vspace{-12pt}
-  \end{wrapfigure}
+  % \begin{wrapfigure}{r}{0.41\linewidth}
+  %     \begin{minipage}[t]{0.18\textwidth}
+  %         \vspace{0pt} % Add this to ensure top alignment within minipage
+  %         \begin{minted}{julia}
+  %             y .= 0
+  %             for j = _, i = _
+  %               y[i] += A[i, j] * x[j]
+  %             end
+  %         \end{minted}
+  %         \vspace{24pt} % Add this to ensure top alignment within minipage
+  %         \begin{minted}{julia}
+  %             y .= 0
+  %             for j = _, i = _
+  %               y[j] += A[i, j] * x[i]
+  %             end
+  %         \end{minted}
+  %     \end{minipage}\hfill%
+  %     \begin{minipage}[t]{0.22\textwidth}
+  %         \vspace{0pt} % Add this to ensure top alignment within minipage
+  %         \begin{minted}{julia}
+  %             y .= 0
+  %             for j = _
+  %               let x_j = x[j]
+  %                 y_j .= 0
+  %                 for i = _
+  %                   let A_ij = A[i, j]
+  %                     y[i] += x_j * A_ij
+  %                     y_j[] += A_ij * x[i]
+  %                   end
+  %                 end
+  %                 #D is the diagonal
+  %                 y[j] += y_j[] + D[j] * x_j
+  %               end
+  %             end
+  %         \end{minted}
+  %     \end{minipage}
+  %     \vspace{-8pt}
+  %     \caption{Finch row-major, column-major and symmetric SpMV Programs}
+  %     \label{spmv_programs}
+  %     \vspace{-12pt}
+  % \end{wrapfigure}
 

### 2024-04-05 sections/tensors.tex
Update on Overleaf.

@@
     \raggedright
-$declare(lvl, init, dims...)$: Declares the level to hold subtensors
-of size $dims$ and an initial value of $init$. Requires the level to be in
-read-only mode.
+\fbox{\parbox{\textwidth}{$declare(lvl, init, dims...)$: Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
+read-only mode.}}
 
@@
 
-$freeze(lvl)$: Finalizes the updates in the level, and readies the
-level for reading. Requires the level to be in update-only mode.
+\fbox{\parbox{\textwidth}{$freeze(lvl)$: Finalizes the updates in the level, and readies the
+level for reading. Requires the level to be in update-only mode.}}
 
@@
 
-$thaw(lvl)$: Prepares the level to accept updates. Requires the level
-to be in read-only mode.
+\fbox{\parbox{\textwidth}{$thaw(lvl)$: Prepares the level to accept updates. Requires the level
+to be in read-only mode.}}
 
@@
 
-$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
+\fbox{\parbox{\textwidth}{$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
 $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
@@
 whether the mutable subfiber has been written to, which allows the parent fiber
-to know whether the subfiber must be stored explicitly.
+to know whether the subfiber must be stored explicitly.}}
 
@@
 
-$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
+\fbox{\parbox{\textwidth}{$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
 level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
 only ever called on unassembled positions, but some levels (such as dense levels
-or bytemaps) may support reassembly.
+or bytemaps) may support reassembly.}}
 \vspace{-8pt}
@@
 \end{figure}
+
+\begin{figure}[ht]
+\footnotesize
+\begin{tabular} {|c|p{0.65\linewidth}|} 
+    \hline
+    $declare(lvl, init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
+read-only mode.
+\hline
+
+    \end{tabular}
+
+\vspace{3pt}
+
+\fbox{\parbox{\textwidth}{$freeze(lvl)$: Finalizes the updates in the level, and readies the
+level for reading. Requires the level to be in update-only mode.}}
+
+\vspace{3pt}
+
+\fbox{\parbox{\textwidth}{$thaw(lvl)$: Prepares the level to accept updates. Requires the level
+to be in read-only mode.}}
+
+\vspace{3pt}
+
+\fbox{\parbox{\textwidth}{$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
+$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
+returns a looplet nest over the values in the read-only fiber.  When $mode =
+\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
+fiber. When applied to a scalar or leaf node, $ext$ is omitted and we may also use $op$ and $rhs$ to update the scalar value. Often, skipping over mutable locations allows the level to know which
+locations must be stored. Often, a dirty bit may be used in $tns$ to communicate
+whether the mutable subfiber has been written to, which allows the parent fiber
+to know whether the subfiber must be stored explicitly.}}
+
+\vspace{3pt}
+
+\fbox{\parbox{\textwidth}{$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
+level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
+only ever called on unassembled positions, but some levels (such as dense levels
+or bytemaps) may support reassembly.}}
+\vspace{-8pt}
+\caption{The five functions that define a level.}
+\vspace{-12pt}
+\end{figure}
+
+% \begin{figure}[htbp]
+%     \scriptsize
+% \begin{tabular}{|p|p|} 
+% \hline
+%      Function & Description  \\
+%      \hline
+%      $declare(lvl, init, dims...)$ & Declares the level to hold subtensors
+% of size $dims$ and an initial value of $init$. Requires the level to be in
+% read-only mode.\\
+% \hline
+% $freeze(lvl)$ & Finalizes the updates in the level, and readies the
+% level for reading. Requires the level to be in update-only mode.\\
+% \hline
+% $thaw(lvl)$ & Prepares the level to accept updates. Requires the level
+% to be in read-only mode.\\
+% \hline
+% $unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$ & When $ext$ is given, unfurls the fiber at position $pos$ in the level $lvl$ over the extent $ext$.
+When $mode = \finchread$, returns a looplet nest over the values in the read-only fiber.
+When $mode = \finchupdate$, returns a looplet nest over mutable subfibers in the update-only fiber.
+When applied to a scalar or leaf node, $ext$ is omitted and we may also use $op$ and $rhs$ to update the scalar value.
+Often, skipping over mutable locations allows the level to know which
+% locations must be stored. Often, a dirty bit may be used in $tns$ to communicate
+% whether the mutable subfiber has been written to, which allows the parent fiber
+% to know whether the subfiber must be stored explicitly.\\
+% \hline
+
+% \end{tabular}
+% \caption{The five functions that define a level.}
+% \vspace{-12pt}
+% \end{figure}
 

### 2024-04-05 sections/tensors.tex
Update on Overleaf.

@@
     $declare(lvl, init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
-read-only mode.
+read-only mode.\\
 \hline
+$freeze(lvl)$ & Finalizes the updates in the level, and readies the
+level for reading. Requires the level to be in update-only mode.\\
+\hline
+$thaw(lvl)$ & Prepares the level to accept updates. Requires the level
+to be in read-only mode.\\
+\hline
+$unfurl(fiber(lvl, pos),$
 
-    \end{tabular}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$freeze(lvl)$: Finalizes the updates in the level, and readies the
-level for reading. Requires the level to be in update-only mode.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$thaw(lvl)$: Prepares the level to accept updates. Requires the level
-to be in read-only mode.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
+$ [ext], mode, [op], [rhs])$ & When $ext$ is given, unfurls the fiber at position
 $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
@@
 whether the mutable subfiber has been written to, which allows the parent fiber
-to know whether the subfiber must be stored explicitly.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
+to know whether the subfiber must be stored explicitly.\\
+\hline
+$assemble(lvl, pos_{start}, pos_{stop})$& Allocates subfibers in the
 level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
 only ever called on unassembled positions, but some levels (such as dense levels
-or bytemaps) may support reassembly.}}
-\vspace{-8pt}
+or bytemaps) may support reassembly.\\
+\hline
+    \end{tabular}
+
 \caption{The five functions that define a level.}
-\vspace{-12pt}
+
 \end{figure}

### 2024-04-05 sections/tensors.tex
Update on Overleaf.

@@
 \footnotesize
-\begin{tabular} {|c|p{0.65\linewidth}|} 
+\begin{tabular} {|m{5em}|p{0.65\linewidth}|} 
     \hline
-    $declare(lvl, init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
+    $declare(lvl,$\newline{}$ init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
 read-only mode.\\
@@
 \hline
-$unfurl(fiber(lvl, pos),$
-
-$ [ext], mode, [op], [rhs])$ & When $ext$ is given, unfurls the fiber at position
+$unfurl($\newline{}$fiber(lvl, pos),$\newline{}$ [ext], mode,$ \newline{} $[op], [rhs])$ & When $ext$ is given, unfurls the fiber at position
 $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,

### 2024-04-05 sections/eval/spmv.tex
cool

@@
   All of our programs apply to all of our level formats, enabling exploitation of multiple structural patterns concurrently (e.g. sparsity \textit{and} symmetry).
+
+  \begin{wrapfigure}{r}{0.5\linewidth}
+    \scriptsize
+    \centering
+    \vspace{-12pt}
+    \vspace{-12pt}
+    \begin{tabular}{|r|l|l|}
+        \hline
+        \textbf{Format} & \textbf{Style of Matrix}\\
+        \hline
+        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
+        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & boolean, sparse \\
+        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
+        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseBand(Element(0.0)))} & real, banded \\
+        %\mintinline{julia}{Dense(SparsePoint(Element(0.0)))} & real-valued matrices with one value per row \\
+        %\mintinline{julia}{Dense(SparseRLE(Pattern(0.0)))} & matrix with runs of true or false \\
+        \hline 
+    \end{tabular}
+    \caption{SpMV Tensor Formats}
+    \label{spmv_tensor_formats}
+    \vspace{-8pt}
+  \end{wrapfigure}
 
@@
 
-  \begin{table}[htbp]
-      \scriptsize
-      \centering
-      \vspace{-12pt}
-      \caption{SpMV Tensor Formats}
-      \vspace{-12pt}
-      \label{spmv_tensor_formats}
-      \begin{tabular}{|l|l|l|l|}
-          \hline
-          \textbf{Outer Level} & \textbf{Inner Level} & \textbf{Scalar Values} & \textbf{Style of Matrix}\\
-          \hline
-          \multirow{6}{*}{Dense} & \multirow{2}{*}{SparseList} & Element & sparse, real-valued matrices \\
-          \cline{3-4} 
-          & & Pattern & sparse, boolean-valued matrices \\
-          \cline{2-4} 
-          & SparseVBL & Element & real-valued matrices with blocked structure \\
-          \cline{2-4}
-          & SparseBand & Element & real-valued matrices with diagonal band \\
-          \cline{2-4}
-          & \multirow{2}{*}{SparsePoint} & Element & real-valued matrices with one value per row \\
-          \cline{3-4} 
-          & & Pattern & matrix with runs of true or false \\
-          \hline 
-      \end{tabular}
-      \vspace{-8pt}
-  \end{table}
-
   % Perhaps add the following section again in a future iteration of paper
@@
   \begin{figure}
-      \includegraphics[width=\linewidth]{Spmv_annotated.png}
+      \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
       \vspace{-18pt}
@@
       \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-"R" indicates row-major implementation and "C" indicates column-major implementation in Finch.
-The baseline Finch format is unsymmetric Dense(SparseList(Element)).
+Programs are from Figure~\ref{spmv_programs} and formats are from Figure~\ref{spmv_tensor_formats}.
+``finch\_baseline'' is the faster of row major or column major ``SparseList''.
   \end{figure}

### 2024-04-05 sections/eval/api.tex
fixes

@@
 \subsection{Implementing Numpy's Array API in Finch}
-In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this API for structured data suffer from either incompleteness or inefficiency.
-In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
-Further, existing work doesn't support operator fusion which can have a drastic impact on performance as we show in Figure~\ref{fig:fusion}.
-A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to make the full array API performant for structured data.
+In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
+%
+These APIs have hundreds of required functions, from mapreduce to slicing and windowing. 
+%
+Compilers with simpler interfaces (such as TACOs simple einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
+%
+As the API's were designed for only one structure (Dense), the burden only grows with the introduction of as additional structures such as sparsity, as these structures interact with the complex interface and with each other.
+%
+%In order to reduce the complexity of interactions between different formats, they're either limited to vectors and matrices or only support tabular representations.
+%
+Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degredation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
+%
+A flexible compiler like Finch, which can produce efficient code for arbitrary operations between inputs in a wide variety of formats, is the missing piece needed to implement the full array API for structured data with competitive performance.
+%
 To handle the expansiveness of the Array API while preserving opportunities for fusion and whole workflow optimization, we pursued a lazy evaluation strategy mediated by a high-level query language.
-This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Finch Interpreter, which executes Finch Logic as a sequence of Finch programs.
+%
+This is implemented by 1) Finch Logic, a minimal, high-level language for expressing array operations and 2) the Physical Interpreter, which executes Finch Logic as a sequence of Finch programs.
+
+%Add a list of all of the functions we currently support, and claim that other frameworks cannot do the same.
+%Masks and wrapper arrays -> needed to implement indexing and slicing
+%Custom operators -> needed to implement argmin, norm, ifelse, and other "weird" operations
+
+%Add back the nice stuff that kyle cut
+%Languages like TACO that are very simple can be trivially incorporated as a library function interface into python.
+%However, Finch provides many other features and control flow than TACO, that are more challenging to integrate.
+%Instead of pushing Finch literally into the hands of python users, we are using their language to describe functions, similar to pytorch.
 
@@
 
-\subsubsection{Finch Logic}
-The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
+\begin{wrapfigure}{R}{0.6\linewidth}
+  \scriptsize
+\begin{minted}{julia}
+  EXPR := table(TENSOR, FIELDS...) |         #TENSOR is an immediate tensor
+    ALIAS | FIELD | mapjoin(OP, EXPRS...) |  #OP is a immediate function
+    aggregate(OP, INIT, EXPR, FIELDS...) |   #INIT is an immediate init value
+    reorder(EXPR, FIELDS...) |               #FIELD is a name for an index
+    relabel(EXPR, FIELDS...) |               #ALIAS is a name for a tensor
+    reformat(TENSOR, EXPR)                     PLAN := plan(QUERIES..., PRODUCES)
+                                              QUERY := query(ALIAS, EXPR)
+\end{minted}
+\caption{\scriptsize Finch Logic Syntax. }\label{fig:finch_logic_syntax}
+\begin{minted}{python}
+  sum(A, dims=2) = aggregate(+, relabel(A, i_1, ..., i_d), i_2)
+  matmul(A, dims=2) = aggregate(+, mapjoin(*, relabel(A, i, j), relabel(B, j, k)))
+  argmin(A) = aggregate(minby, (Inf, len(A)), mapjoin(tuple, relabel(A, i), i), i)
+  norm2(A) = aggregate(scaled_plus, (0, 0), mapjoin(tuple, mapjoin(sign, A), mapjoin(abs, A)), fields(A)...)
+  where(C, A, B) = mapjoin(ifelse, relabel(C, i), relabel(A, i), relabel(B, i))
+  slice(A, i:j) = table(WindowedArray(A, i:j), k)
+  cat(A, B) = mapjoin(coalesce, table(PermissiveArray(A), i), table(OffsetArray(B, length(A)), i))
+\end{minted}
+\caption{\scriptsize Implementations of a few common functions in Finch Logic.
+Here, \texttt{minby(x, y)} compares \texttt{x[1]} and \texttt{y[1]} and returns the smaller \texttt{x} or \texttt{y}.
+\texttt{scaled\_plus} rescales whichever argument is smaller, then adds the values with the same scale.
+\texttt{ifelse(c, a, b)} returns \texttt{a} when \texttt{c} is true,  \texttt{b} otherwise.
+Finch can understand declared properties and rewrite rules for all of these functions and types, extending these algorithms to operate on structured arguments.} \label{fig:api_extensible}
+\end{wrapfigure}
+\subsubsection{Finch Logic and the Physical Interpreter}
+Finch Logic is used as a common representation to implement and fuse the operations in the Array API.
+
+\paragraph{Finch Logic}
+The syntax of Finch Logic is described in Figure~\ref{fig:finch_logic_syntax}.
+%
+Note that the syntax is designed to be minimal and expressive, taking inspiration from relational algebra with the goal of being able to express most of the operations in the Array API.
+%
+Our \texttt{mapjoin} represents pointwise application, while \texttt{aggregate} represents reductions.
+%
+We name the result of each computation with \texttt{query}, and join several expressions in a \texttt{plan}.
+%
+\texttt{reorder} reorders the indices while \texttt{relabel} relabels them.
+%
+\texttt{reformat} hints at materialization into its argument.
+
+Finch's support for arbitrary user-defined functions allows us to support complex operations such as \mintinline{python}{argmin}, \mintinline{python}{norm}, and \mintinline{python}{where}, all over structured data.
+Finch's support for wrapper arrays allows us to express indexing operations.
+These situations are shown in Figure~\ref{fig:api_extensible}.
+\begin{figure}
+
+\end{figure}
+
+\paragraph{Heuristic Optimization}
+Once we express our operation in Finch Logic, we use a quick heuristic to fuse the expressions. 
+%
+We heuristically split nested \texttt{aggregate} queries into separate queries, and fuse all of the \texttt{mapjoin}s into corresponding \texttt{aggregate}s.
+%
+We then heuristically choose a loop order for each query with \texttt{reorder} nodes, and introduce separate \texttt{query} nodes to transpose so that the resulting expressions are concordant.
+%
+Our heuristic assigns a format to each output by recursively evaluating the properties of structure (sparsity, repetition, etc.) and checking whether they are preserved under the operation in question.
+
+%The expression fragment takes inspiration from relational algebra while incorporating an ordering on the dimensions of a tensor.
 This means that operators take in a set of indexed tensors and output an indexed tensor.
@@
 
-\begin{align*}
-    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
-     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
-    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
-     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
-\end{align*}
+%\begin{align*}
+%    \footnotesize \finchplan(queries..., names...) \quad\quad\quad \finchquery(name, expr) \quad\quad\quad \finchreorder(expr, idxs...)\\
+%     \footnotesize \finchrelabel(expr, idxs...) \quad\quad\quad \finchreformat(expr, format) \quad\quad\quad \finchmapjoin(op, exprs...) \\
+%    \footnotesize\finchaggregate(op, expr)\quad\quad\quad\quad \finchtable(tns, idxs...)  \quad\quad\quad\quad\quad\quad \finchalias(name) \quad\quad\\
+%     \footnotesize expr:= \finchreorder | \finchrelabel | \finchreformat |\finchmapjoin | \finchaggregate | \finchtable | \finchalias \quad\quad
+%\end{align*}
 
-We now describe how to define a few example functions from the API in this language.
+%We now describe how to define a few example functions from the API in this language.
 Due to the flexibility of Finch, we can use the custom operators $minby(x,y)$ (which compares $x[1]$ and $y[1]$ and returns the smaller $x$ or $y$) and $tuple(x, y)$ (which returns the tuple $(x,y)$), and we can treat an index as a scalar to implement $argmin$.
 For conciseness, we omit the outer $\finchplan(\finchquery(out,...),out)$.
-\begin{align*}
-&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
-&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
-&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
-\end{align*}
+%\begin{align*}
+%&\footnotesize \text{sum}(M, dims=[2]) \rightarrow \finchaggregate(+,\finchrelabel(M, i_1,\ldots,i_d), i_2)\\
+%&\footnotesize \text{matmul}(A, B) \rightarrow \finchaggregate(+,\finchmapjoin(*, \finchrelabel(A, i, j), \finchrelabel(B, j, k)), j)\\
+%&\footnotesize \text{argmin}(A, dims=[2]) \rightarrow \finchaggregate(minby,\finchmapjoin(tuple, \finchrelabel(A, i_1,\ldots,i_d), \finchtable(i_2)), i_2)
+%\end{align*}
 
@@
 
-\subsubsection{Standardizing \& Heuristic Optimization}
-Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
+%\paragraph{Execution}
+%\subsubsection{Standardizing \& Heuristic Optimization}
+%Before a plan in Finch Logic can be interpreted, it must be converted to a standard form which resolves ambiguity about loop ordering and output formatting.
 The semantically important requirements of this standard form are 1) all inputs (i.e.
@@
 
-\subsubsection{Finch Interpreter} 
-Once the program is in a standard form, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
-The output format is identified by unpacking the outer $\finchreformat$ statement.
+\paragraph{Finch Interpreter} 
+Once the program is in a standard form where each \texttt{query} has a specified format with \texttt{reformat} and at most one \texttt{aggregate}, the Finch Interpreter executes each query, in order, through a straightforward lowering process.
+%
+%Each result is assigned to the corresponding alias the .
+% The output format is identified by unpacking the outer $\finchreformat$ statement.
 Next, the inner expression of the is unpacked to identify the $\finchaggregate$ operator and extract the pointwise expression.
@@
 \begin{figure}
-\begin{tabular}{p{0.33\textwidth} p{0.33\textwidth} p{0.33\textwidth}}
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/triangle_count_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=135pt]{figures/sddmm_speedup_over_duckdb.png} &
-  \vspace{0pt} \includegraphics[width=140pt, height=123pt]{figures/elementwise_speedup_over_duckdb.png}
-\end{tabular}
-\vspace{-12pt}
-\caption{Performance of Finch Logic for common kernels.} \label{fig:fusion}
+  \centering % Centers the figure block in the text area
+  % First minipage for the first image
+  \begin{minipage}[t]{0.32\linewidth}
+    \vspace{0pt}
+    \centering
+    \includegraphics[width=\linewidth]{figures/triangle_count_speedup_over_duckdb.png}
+    \vspace{0pt}
+  \end{minipage}%
+  % Second minipage for the second image
+  \begin{minipage}[t]{0.32\linewidth}
+    \vspace{0pt}
+    \centering
+    \includegraphics[width=\linewidth]{figures/sddmm_speedup_over_duckdb.png}
+    \vspace{0pt}
+  \end{minipage}%
+  % Third minipage for the third image
+  \begin{minipage}[t]{0.36\linewidth}
+    \vspace{0pt}
+    \centering
+    \includegraphics[width=\linewidth]{figures/elementwise_speedup_over_duckdb.png}
+    \vspace{0pt}
+  \end{minipage}
+
+  \vspace{-24pt}
+  \caption{Performance of Finch Logic for common kernels.}
+  \label{fig:fusion}
 \end{figure}
@@
 %(v^t .* dot(u, w))
-
-\willow{Note: I may want to explain format inference somewhere in here, but I'll
-have to get to it a little later, perhaps after the paper deadline}
-\willow{Note: I think it would be cool to include something about how we support
-numerically stable norms and argmin in this model}

### 2024-04-05 sections/abstract.tex
fixes for completeness
packing

@@
 \begin{abstract}
-From FORTRAN to Numpy, arrays have revolutionized how we express computation. However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
+From FORTRAN to Numpy, arrays have revolutionized how we express computation.
+However, arrays in these, and almost all prominent systems, can only handle dense rectilinear integer grids.
 Real world arrays often contain underlying structure, such as sparsity, runs of repeated values, or symmetry.
@@
 
-In this work, we propose a new programming language, Finch, which supports \textit{both}
-flexible control flow and diverse data structures. Finch facilitates a programming model which
-resolves the challenges of computing over structured arrays by combining control flow and data
-structures into a common representation where they can be co-optimized. In particular,
-Finch automatically specializes the control flow to the data so that performance engineers can
-focus on experimenting with many algorithms. Finch supports a familiar programming language
-of loops, statements, if conditions, breaks, etc., over a wide variety of array structures, such as
-sparsity, run-length-encoding, symmetry, triangles, padding, or blocks. Finch reliably utilizes the key
-properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
+In this work, we propose a new programming language, Finch, which supports \textit{both} flexible control flow and diverse data structures.
+Finch facilitates a programming model which resolves the challenges of computing over structured arrays by combining control flow and data structures into a common representation where they can be co-optimized.
+Finch automatically specializes control flow to data so that performance engineers can focus on experimenting with many algorithms.
+Finch supports a familiar programming language of loops, statements, ifs, breaks, etc., over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
 We show that this leads to dramatic speedups in operations such as spmv and spgemm, image processing, graph analytics, and a high-level tensor operator fusion interface.

### 2024-04-05 sections/intro.tex
squish

@@
 %
-Compilers understand dense array computations across many
+Compilers understand dense computations across many
 programming constructs, such as for and while loops, breaks, parallelism,
@@
 \end{figure}
-\kyle{In this paragraph, I wonder if it should be more "our world is full of structured data" which can be described by structured arrays.
-The "X describes Y" framing feels less fundamental than "Y is Z" like "images, masks, geometry, and transactions consist of long runs of identical values that can be compressed via run length encoding".
-This might be overly clunky though.}
-Our world is full of structured arrays.
+
+Our world is full of structured data.
 %

### 2024-04-05 sections/abstract.tex
Update on Overleaf.

@@
 Finch reliably utilizes the key properties of structure, such as structural zeros, repeated values, or clustered non-zeros.
-We show that this leads to dramatic speedups in operations such as spmv and spgemm, image processing, graph analytics, and a high-level tensor operator fusion interface.
+We show that this leads to dramatic speedups in operations such as SpMV and SpGEMM, image processing, graph analytics, and a high-level tensor operator fusion interface.
 \end{abstract}

### 2024-04-05 sections/intro.tex
Update on Overleaf.

@@
 %
-Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g.
-higher order FEMs) as well as in intermediate structures in many linear solvers (e.g.
-GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
+Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
 %

### 2024-04-05 sections/eval/spmv.tex
Update on Overleaf.

@@
 \begin{wrapfigure}{r}{0.41\linewidth}
-    \vspace{-28pt}
+    \vspace{-36pt}
     \begin{minipage}[t]{0.18\textwidth}
         \begin{minted}{julia}
+
             y .= 0
@@
     \end{minipage}
-    \vspace{-8pt}
+    \vspace{-12pt}
     \caption{Finch row-major, column-major and symmetric SpMV Programs}
     \label{spmv_programs}
-    \vspace{-12pt}
+    \vspace{-36pt}
   \end{wrapfigure}
   \subsection{Sparse Matrix-Vector Multiply (SpMV)}
-  Sparse matrix-vector multiplication (SpMV) has a wide range of applications and has been thoroughly studied~\cite{liu_csr5_2015,
+  Sparse matrix-vector multiply (SpMV) has a wide range of applications and has been thoroughly studied~\cite{liu_csr5_2015,
   zhou_enabling_2020}. 
@@
     \centering
-    \vspace{-12pt}
-    \vspace{-12pt}
+    \vspace{-16pt}
     \begin{tabular}{|r|l|l|}
@@
     \end{tabular}
+      \vspace{-8pt}
     \caption{SpMV Tensor Formats}

### 2024-04-05 sections/intro.tex
Update on Overleaf.

@@
 %
-Finch supports a familiar programming language of loops, statements, if conditions, breaks, etc, over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
+Finch supports a familiar programming language of loops, statements, if conditions, breaks, etc., over a wide variety of array structures, such as sparsity, run-length-encoding, symmetry, triangles, padding, or blocks.
 %

### 2024-04-05 sections/intro.tex
fix

--- 2024-04-05 sections/intro.tex
@@
 %
-\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as spmv ($2.51\times$) and spgemm ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $0.25\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
+\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as spmv ($2.51\times$) and spgemm ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.

### 2024-04-05 sections/background.tex
fix stuff
fix

@@
-\section{Background On Looplets}
+\section{Background}
+\subsection{Looplets}
 Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
@@
 \footnotesize
-\begin{tabular} {|p{0.65\linewidth}|c|} 
+\begin{tabular} {|p{0.7\linewidth}|c|} 
     \hline
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-lookup.png}}
+    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-lookup.png}}
     \\
@@
         &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-run.png}}
+    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-run.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-phase.png}}
+    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-phase.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-switch.png}}
+    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-switch.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-thunk.png}}
+    \raisebox{-0.8\totalheight}{\includegraphics[scale=0.20]{Looplets-thunk.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-sequence.png}}
+    \raisebox{-0.85\totalheight}{\includegraphics[scale=0.20]{Looplets-sequence.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-spike.png}}
+    \raisebox{-1.15\totalheight}{\includegraphics[scale=0.20]{Looplets-spike.png}}
     \\ 
@@
     &
-    \raisebox{-\totalheight}{\includegraphics[scale=0.20]{Looplets-stepper.png}}
+    \raisebox{-1\totalheight}{\includegraphics[scale=0.20]{Looplets-stepper.png}}
     \\
@@
 \end{figure}
+
+\begin{wrapfigure}{r}{0.45\linewidth}
+    \centering
+    \vspace{-28pt}
+    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
+    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
+    \vspace{-16pt}
+    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
+    \label{fig:levelsvsfibers}
+    \vspace{-24pt}
+\end{wrapfigure}
+\subsection{Fiber Trees}
+
+Fiber-tree style tensor abstractions have been the subject of extensive study
+\cite{sze2017efficient, chou2022compilation, chou2018format}.  The underlying
+idea is to represent a multi-dimensional tensor as a nested vector
+datastructure, where each level of the nesting corresponds to a dimension of the
+tensor. Thus, a matrix would be represented as a vector of vectors. This kind of
+abstraction lends itself to representing sparse tensors if we vary the type of
+vector used at each level in a tree. Thus, a sparse matrix might be represented
+as a dense vector of sparse vectors. The vector of subtensors in this
+abstraction is referred to as a \textbf{fiber}.
+
+Instead of storing the data for each subfiber separately, most sparse tensor
+formats such as CSR, DCSR, and COO usually store the data for all fibers in a
+level contiguously. In this way, we can think of a level as a bulk allocator for
+fibers. Continuing the analogy, we can think of each fiber as being
+disambiguated by a \textbf{position}, or an index into the bulk pool of
+subfibers. The mapping $f$ from indices to subfibers is thus a mapping from an
+index and a position in a level to a subposition in a sublevel.
+Figure~\ref{fig:levelsvsfibers} shows a simple example of a level as a pool of fibers.
+
+When we need to refer to a particular fiber at position $p$ in the level $l$, we
+may write $fiber(l, p)$. Note that the formation of fibers from levels is lazy,
+and the data underlying each fiber is managed entirely by the level, so the
+level may choose to overlap the storage between different fibers. Thus, the only
+unique data associated with $fiber(l, p)$ is the position $p$.

### 2024-04-05 sections/tensors.tex
Update on Overleaf.
fix

--- 2024-04-05 sections/tensors.tex
@@
 The $assemble$ function manages the allocation of new data in the tensor.
-
-\begin{wrapfigure}{R}{0.5\linewidth}
-    \centering
-    \vspace{-8pt}
-    \includegraphics[width=\linewidth]{LevelsVsFibers-matrix.png}
-    %\includegraphics[width=0.5\linewidth]{LevelsVsFibers-tensor.png}
-    \vspace{-8pt}
-    \caption{Levels in the fiber tree representation of a sparse matrix in CSC format, with a dense outer level and a sparse inner level. The element level holds the leaves of the tree.}
-    \label{fig:levelsvsfibers}
-    \vspace{-8pt}
-\end{wrapfigure}
 
@@
 
-Fiber-tree style tensor abstractions have been the subject of extensive study
-\cite{sze2017efficient, chou2022compilation, chou2018format}.  The underlying
-idea is to represent a multi-dimensional tensor as a nested vector
-datastructure, where each level of the nesting corresponds to a dimension of the
-tensor. Thus, a matrix would be represented as a vector of vectors. This kind of
-abstraction lends itself to representing sparse tensors if we vary the type of
-vector used at each level in a tree. Thus, a sparse matrix might be represented
-as a dense vector of sparse vectors. The vector of subtensors in this
-abstraction is referred to as a \textbf{fiber}. Prior fiber-tree representations
+Prior fiber-tree representations
 focus on sparsity (where only the nonzero elements are represented) and treat
@@
 %postype(lvl)
-
-Instead of storing the data for each subfiber separately, most sparse tensor
-formats such as CSR, DCSR, and COO usually store the data for all fibers in a
-level contiguously. In this way, we can think of a level as a bulk allocator for
-fibers. Continuing the analogy, we can think of each fiber as being
-disambiguated by a \textbf{position}, or an index into the bulk pool of
-subfibers. The mapping $f$ from indices to subfibers is thus a mapping from an
-index and a position in a level to a subposition in a sublevel.
-Figure~\ref{fig:levelsvsfibers} shows a simple example of a level as a pool of fibers.
-
-When we need to refer to a particular fiber at position $p$ in the level $l$, we
-may write $fiber(l, p)$. Note that the formation of fibers from levels is lazy,
-and the data underlying each fiber is managed entirely by the level, so the
-level may choose to overlap the storage between different fibers. Thus, the only
-unique data associated with $fiber(l, p)$ is the position $p$.
 
@@
 
-\vspace{3pt}
+\vspace{1pt}
 
@@
 \vspace{-12pt}
-\end{figure}
 
-\begin{figure}[ht]
-\footnotesize
-\begin{tabular} {|c|p{0.65\linewidth}|} 
-    \hline
-    $declare(lvl, init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
-read-only mode.
-\hline
+% \begin{figure}[ht]
+% \footnotesize
+% \begin{tabular} {|m{7em}|p{0.85\linewidth}|} 
+%     \hline
+%     $declare(lvl,$\newline{}$ init, dims...)$ & Declares the level to hold subtensors of size $dims$ and an initial value of $init$. Requires the level to be in
+% read-only mode.\\
+% \hline
+% $freeze(lvl)$ & Finalizes the updates in the level, and readies the
+% level for reading. Requires the level to be in update-only mode.\\
+% \hline
+% $thaw(lvl)$ & Prepares the level to accept updates. Requires the level
+% to be in read-only mode.\\
+% \hline
+% $unfurl($\newline{}$fiber(lvl, pos),$\newline{}$ [ext], mode,$ \newline{} $[op], [rhs])$ & When $ext$ is given, unfurls the fiber at position
+% $pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
+% returns a looplet nest over the values in the read-only fiber.  When $mode =
+% \finchupdate$, returns a looplet nest over mutable subfibers in the update-only
+% fiber. When applied to a scalar or leaf node, $ext$ is omitted and we may also use $op$ and $rhs$ to update the scalar value. Often, skipping over mutable locations allows the level to know which
+% locations must be stored. Often, a dirty bit may be used in $tns$ to communicate
+% whether the mutable subfiber has been written to, which allows the parent fiber
+% to know whether the subfiber must be stored explicitly.\\
+% \hline
+% $assemble(lvl,$\newline{}$ pos_{start},$\newline{}$ pos_{stop})$& Allocates subfibers in the
+% level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
+% only ever called on unassembled positions, but some levels (such as dense levels
+% or bytemaps) may support reassembly.\\
+% \hline
+%     \end{tabular}
 
-    \end{tabular}
+%\caption{The five functions that define a level.}
 
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$freeze(lvl)$: Finalizes the updates in the level, and readies the
-level for reading. Requires the level to be in update-only mode.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$thaw(lvl)$: Prepares the level to accept updates. Requires the level
-to be in read-only mode.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$unfurl(fiber(lvl, pos), [ext], mode, [op], [rhs])$: When $ext$ is given, unfurls the fiber at position
-$pos$ in the level $lvl$ over the extent $ext$. When $mode = \finchread$,
-returns a looplet nest over the values in the read-only fiber.  When $mode =
-\finchupdate$, returns a looplet nest over mutable subfibers in the update-only
-fiber. When applied to a scalar or leaf node, $ext$ is omitted and we may also use $op$ and $rhs$ to update the scalar value. Often, skipping over mutable locations allows the level to know which
-locations must be stored. Often, a dirty bit may be used in $tns$ to communicate
-whether the mutable subfiber has been written to, which allows the parent fiber
-to know whether the subfiber must be stored explicitly.}}
-
-\vspace{3pt}
-
-\fbox{\parbox{\textwidth}{$assemble(lvl, pos_{start}, pos_{stop})$: Allocates subfibers in the
-level from positions $pos_{start}$ to $pos_{stop}$. Usually, this function is
-only ever called on unassembled positions, but some levels (such as dense levels
-or bytemaps) may support reassembly.}}
-\vspace{-8pt}
-\caption{The five functions that define a level.}
-\vspace{-12pt}
 \end{figure}
@@
 
-    \begin{wrapfigure}{r}{.35\textwidth}
+    \begin{figure}
         \centering
+        \includegraphics[width=\linewidth]{Structures.png}\hfill%
         \vspace{-8pt}
+        \caption{Several examples of matrix structures represented using the
+        level structures identified in Table~\ref{tab:TypesOfStructure}.
+        Comparing this figure to \cite[Figure 3]{ahrens_looplets_2023}, we see
+        that a level-by-level structural decomposition is diagrammed together
+        with the Looplets.}
+        \label{fig:structuraldiversity}
+        \vspace{-18pt}
+    \end{figure}
+
+\subsection{Randomly Accessible Temporary Levels}
+To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
+sequential creation of formats. However, when users want to be able to write out
+of order (which is a common requirement arising from loop order or from the
+problem itself, it occurs in our spgemm algorithms and our histogram example in
+the evaluation section), we must use more complicated datastructures like hash
+tables and trees to support the random access. Because these datastructures are
+more complex and have a higher implementation burden and performance overhead,
+we only support random access construction of sparse or dense structures.  We
+can use these two more general structures as intermediates to convert to our
+more specialized structures later.
+
+\begin{wrapfigure}{r}{.5\textwidth}
+        \centering
+        \vspace{-20pt}
         \scriptsize
@@
         \label{tab:TypesOfStructure}
-        \vspace{-12pt}
+        \vspace{-36pt}
     \end{wrapfigure}
-
-    \begin{figure}
-        \centering
-        \includegraphics[width=\linewidth]{Structures.png}\hfill%
-        \vspace{-8pt}
-        \caption{Several examples of matrix structures represented using the
-        level structures identified in Table~\ref{tab:TypesOfStructure}.
-        Comparing this figure to \cite[Figure 3]{ahrens_looplets_2023}, we see
-        that a level-by-level structural decomposition is diagrammed together
-        with the Looplets.}
-        \label{fig:structuraldiversity}
-        \vspace{-18pt}
-    \end{figure}
-
-\subsection{Randomly Accessible Temporary Levels}
-To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
-sequential creation of formats. However, when users want to be able to write out
-of order (which is a common requirement arising from loop order or from the
-problem itself, it occurs in our spgemm algorithms and our histogram example in
-the evaluation section), we must use more complicated datastructures like hash
-tables and trees to support the random access. Because these datastructures are
-more complex and have a higher implementation burden and performance overhead,
-we only support random access construction of sparse or dense structures.  We
-can use these two more general structures as intermediates to convert to our
-more specialized structures later.
 
@@
 
-\begin{table}[ht]
+\begin{table}[h]
     \centering
@@
     \end{tabular}
+    \vspace{8pt}
     \caption{The main level formats supported by Finch. Note that all non-leaf

### 2024-04-05 sections/eval/spgemm.tex
Update on Overleaf.

@@
 %
-Without such hard wiring, TACO would have to use a dense intermediate to support random writes.
-%
-But without if conditions, TACO would then be forced to propagate all values in the dense vector to the sparse output matrix, turning the output dense and leading to the same asymptotic results as in the case of outer products.
+Without such hard wiring, TACO would have to use a dense intermediate to support random writes, which TACO would then propagate to the output, turning it dense and leading to the same asymptotic results as in the case of outer products.
 %
@@
     \vspace{-8pt}
-    \caption{A comparison of several matrix multiplication algorithms between
+    \caption{A comparison of several matrix multiply algorithms between
     Finch and Taco. On left, we use smaller matrices, ordered from small to big

### 2024-04-05 sections/tensors.tex
Update on Overleaf.

@@
 %
-Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional abstractions common to tensor compilers.
+Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional fiber-tree abstractions common to tensor compilers.
 % 
@@
 
-In the rest of this section, we describe this functionality in detail after we review the basics of level abstractions~\cite{sze2017efficient,chou2022compilation, chou2018format}.
-%
-Then we discuss $declare$, $freeze$, $thaw$, and $unfurl$ functions as part of a life cycle abstraction as well as the $assemble$ function.
+Additionally, prior fiber-tree representations
+focus on sparsity (where only the nonzero elements are represented) and treat
+sparse vectors as sets of represented points. Since our fiber-tree
+representation must handle other kinds of structure, such as diagonal,
+repeated, or constant values, we generalize the concept to view each fiber as a mapping from
+indices into a space of subfibers.
+
+In the rest of this section, we discuss $declare$, $freeze$, $thaw$, and $unfurl$ functions as part of a life cycle abstraction as well as the $assemble$ function.
 %
@@
 Finally, we describe several levels that we implement in this abstraction, which allow us to implement a taxonomy of known structures that appear in tensor computations.
+%
+These levels include implementations of various combinations of structures, but they also include specializations to zero dimensional iterators (scalars) and level structures that support different access patterns.
 
@@
 
-\subsection{Level Abstraction}
-
-Prior fiber-tree representations
-focus on sparsity (where only the nonzero elements are represented) and treat
-sparse vectors as sets of represented points. Since our fiber-tree
-represesentation must handle other kinds of structure, such as diagonal,
-repeated, or constant values, we instead view each fiber as a mapping from
-indices into a space of subfibers.
+%\subsection{Level Abstraction}
 
@@
 
-\subsection{The 8 Key Level Structures}
+\subsection{The 4 Key Structures}
 
-    The main benefits of specializing to structure come from the following properties of the data:
+    In the Finch programming model, the programmer relies on the Finch compiler to specialize to the sequential properties of the data.
+In our experience, the main benefits of specializing to structure come from the following properties of the data:
     \begin{itemize}
@@
     \end{itemize}
+    In the following section, we consider a set of concrete implementations of levels that expose all combinations of these structures, paying some attention to a few important special cases: random access, scalars, and leaf levels.
+We summarize the structures in Table~\ref{table:formats} and Table~\ref{tab:TypesOfStructure}.
 
-   We consider all combinations of these structural properties in Table
-    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
-    the the 8 resulting situations. While it is impossible to write code which
-    precisely addresses every possible structure, the combination of our level
-    formats can express a wide variety of hierarchical structures to a
-    sufficient granularity that we can generate code which utilizes the four
-    properties.  For example, though banded arrays are a superset of ragged
-    arrays, representing a ragged array using a banded format does not add much
-    overhead.  The structures we consider are exhaustive in the sense that they
-    address all combinations of sparsity, blocks, runs, and singletons in each
-    level. We can represent a wide variety of hierarchical tensor structures by
-    combining these level structures in a tree, as shown in
-    Figure~\ref{fig:structuraldiversity}.
-
-    \begin{figure}
+    \begin{figure}[htbp]
         \centering
@@
 
-\subsection{Randomly Accessible Temporary Levels}
+\subsection{Implementations of Structures}
+\subsubsection{Sequentially Constructed Levels}
+   We consider all combinations of these structural properties in Table
+    \ref{tab:TypesOfStructure}, resulting in 8 key level formats that correspond
+    the the 8 resulting situations. 
+    While it is impossible to write code which
+    precisely addresses every possible structure, the combination of our level
+    formats can express a wide variety of hierarchical structures to a
+    sufficient granularity that we can generate code which utilizes the four
+    properties.  For example, though banded arrays are a superset of ragged
+    arrays, representing a ragged array using a banded format does not add much
+    overhead.  The structures we consider are exhaustive in the sense that they
+    address all combinations of sparsity, blocks, runs, and singletons in each
+    level. We can represent a wide variety of hierarchical tensor structures by
+    combining these level structures in a tree, as shown in
+    Figure~\ref{fig:structuraldiversity}.
+
+\subsubsection{Non-sequentially Constructed Levels}
 To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
@@
 of order (which is a common requirement arising from loop order or from the
-problem itself, it occurs in our spgemm algorithms and our histogram example in
+problem itself, it occurs in our SpGEMM algorithms and our histogram example in
 the evaluation section), we must use more complicated datastructures like hash
-tables and trees to support the random access. Because these datastructures are
-more complex and have a higher implementation burden and performance overhead,
+tables and trees to support the random writes so that in-order levels can be constructed later.
+Because these datastructures are more complex and have a higher implementation burden and performance overhead,
 we only support random access construction of sparse or dense structures.  We
@@
 more specialized structures later.
+
+\subsubsection{Scalars}
+Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
+%
+Scalars don't have as much structure as clearly only two structures apply.
+%
+Through these structures, scalars can also affect other tensors in crucial ways.
+%
+Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
+%
+Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
+%
+We also provide early break scalars, which modify the stepper Looplets to
+re-specialize the loop whenever a reduction into an early break scalar hits an
+annihilator.
+%
+We assume that we don't need to re-specialize other Looplets because the stepper
+is the only one which repeats a non-constant number of times.
 
@@
 
-\subsection{Leaf Levels}
+\subsubsection{Leaf Levels}
 
@@
 
-\subsection{Scalars}
-Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
-%
-Scalars can also affect how the leaves are lowered.
-%
-Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
-%
-Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
-%
-We also provide early break scalars, which modify the stepper looplets to
-re-specialize the loop whenever a reduction into an early break scalar hits an
-annihilator.
-%
-We assume that we don't need to re-specialize other looplets because the stepper
-is the only one which repeats a non-constant number of times.
-
 \begin{table}[h]
@@
     \hline
-    \textbf{Sequential Access Levels} \\
+    \textbf{Sequentially Constructed Levels} \\
     \hline
@@
     \hline
-    \textbf{Random Access Levels} \\
+    \textbf{Nonsequentially Constructed Levels} \\
     \hline

### 2024-04-05 sections/background.tex
woah
Update on Overleaf.
small fixes

@@
-\section{Background}
-\subsection{Looplets}
-Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
-%
-Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
-%
-In particular, looplets are good at lifting code to the highest loop level that it's needed and subdividing iteration hierarchically in coordinate space.
-%
-Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
-can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.
-
-The looplets are described in Figure~\ref{fig:looplets}.
-We simplify the presentation to focus on the semantics, rather than precise implementation.
-Several looplets introduce or modify variables in the scope of the target language.
-It is assumed that if a looplet introduces a variable to be used in a child looplet, the child looplet will not modify that variable.
-For more background on Looplets, we recommend the original work \cite{ahrens_looplets_2023}.
-
 \begin{figure}[ht]
@@
 
+\section{Background}
+
+\subsection{Looplets}
+Finch represents iteration patterns using Looplets, a language that decomposes datastructure iterators hierarchically. 
+%
+Looplets represent the control-flow structures needed to iterate over any given datastructure, or multiple datastructures simultaneously. 
+%
+In particular, looplets are good at lifting code to the highest possible loop level and subdividing iteration hierarchically in coordinate space.
+%
+Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
+can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.
+
+The looplets are described in Figure~\ref{fig:looplets}.
+We simplify the presentation to focus on the semantics, rather than precise implementation.
+Several looplets introduce or modify variables in the scope of the target language.
+It is assumed that if a looplet introduces a variable, the child looplet will not modify that variable.
+For more background on Looplets, we recommend the original work \cite{ahrens_looplets_2023}.
+
 \begin{wrapfigure}{r}{0.45\linewidth}

### 2024-04-05 sections/tensors.tex
small fixes

@@
 
-Additionally, prior fiber-tree representations
-focus on sparsity (where only the nonzero elements are represented) and treat
-sparse vectors as sets of represented points. Since our fiber-tree
-representation must handle other kinds of structure, such as diagonal,
-repeated, or constant values, we generalize the concept to view each fiber as a mapping from
-indices into a space of subfibers.
+Additionally, prior fiber-tree representations focus on sparsity (where only the nonzero elements are represented) and treat sparse vectors as sets of represented points.
+Since our fiber-tree representation must handle other kinds of structure, such as diagonal, repeated, or constant values, we generalize our fiber abstraction to allow arbitrary mappings from indices into a space of subfibers.
 

### 2024-04-05 sections/tensors.tex
fixes

--- 2024-04-05 sections/tensors.tex
@@
 %
-First, looplets represent an instance of an iterator over a tensor, but we may access the same tensor twice with different indices.
+First, while looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
 %
@@
 %
-Next, since Finch programs are not just single einsums, they may read and write to the same data at different times.
+Next, since Finch programs go beyond just single einsums, they may read and write to the same data at different times.
 %
@@
 
-Additionally, prior fiber-tree representations
-focus on sparsity (where only the nonzero elements are represented) and treat
-sparse vectors as sets of represented points. Since our fiber-tree
-representation must handle other kinds of structure, such as diagonal,
-repeated, or constant values, we generalize the concept to view each fiber as a mapping from
-indices into a space of subfibers.
+Additionally, prior fiber-tree representations focus on sparsity (where only the nonzero elements are represented) and treat sparse vectors as sets of represented points.
+Since our fiber-tree representation must handle other kinds of structure, such as diagonal, repeated, or constant values, we must generalize our fiber abstraction to allow arbitrary mappings from indices into a space of subfibers.
 
-In the rest of this section, we discuss $declare$, $freeze$, $thaw$, and $unfurl$ functions as part of a life cycle abstraction as well as the $assemble$ function.
+In the rest of this section, we discuss how these 5 core functions ($declare$, $freeze$, $thaw$, $unfurl$, and $assemble$) function as part of a life cycle abstraction that defines a level in Finch.
 %
@@
 %
-Finally, we describe several levels that we implement in this abstraction, which allow us to implement a taxonomy of known structures that appear in tensor computations.
+We then identify a taxonomy of four key structural properties exhibited in data.
 %
-These levels include implementations of various combinations of structures, but they also include specializations to zero dimensional iterators (scalars) and level structures that support different access patterns.
+We implement several levels in this abstraction that capture all combinations of these structures, 
+including specializations to zero dimensional tensors (scalars) and level structures that support different access patterns.
 
@@
 
-The first three of these functions, declare, freeze, and thaw, \saman{Italics for keywords?}have to do with
-managing when tensors can be assumed mutable or immutable.
+The first three of these functions, $declare$, $freeze$, and $thaw$, have to do with managing when tensors can be assumed mutable or immutable.
 %
-\saman{This is because Finch goes beyond just a single einsum...the scope of Finch w.r.t. TACO etc, is not clear.}
-%
-As we use Looplets to
-represent iteration over a tensor, we must restrict the mutability of tensors
-while we iterate over them. 
+As we use Looplets to represent iteration over a tensor, we must restrict the mutability of tensors while we iterate over them. 
 %

### 2024-04-05 sections/tensors.tex
good throught tensors

@@
 %
-The declare function is used to allocate a tensor,
+The $declare$ function is used to allocate a tensor,
 initialize it to some specified size and value, and leave it in update-only
@@
 
-The unfurl function is used to manage iteration over a subfiber. 
+The $unfurl$ function is used to manage iteration over a subfiber. 
 %
 At the point when it comes time to iterate over a tensor, be in on the left or right hand
-side of an assignment, we call the unfurl function to precompute whatever state
+side of an assignment, we call $unfurl$ to precompute whatever state
 and datastructures are necessary to return a looplet nest that would iterate
@@
 %
-The unfurl function is called directly before
+We call $unfurl$ directly before
 iterating over the corresponding loop, so it has access to any state variables introduced
@@
 reused. While all of the previous functions are used to manage the lifecycle and
-iteration over a general tensor, the assemble function is quite specific to the
+iteration over a general tensor, $assemble$ is quite specific to the
 level abstraction, and the notion of positions within sublevels. Note: it was an
@@
 a sparse level might use a list of indices to store which nonzeros are present,
-and when it comes time to resize that list, it could also call assemble to resize the
-sublevel, reducing the number of branches in the code. The assemble function
+and when it comes time to resize that list, it could also call $assemble$ to resize the
+sublevel, reducing the number of branches in the code. The $assemble$ function
 lends itself particularly to a "vector doubling" allocation approach, which we
 have found to be effective and flexible when managing the allocation
-of sparse left hand sides. This benefit is made clear in our evaluation section,
-where we see that prior systems like TACO do not support all possible loop
+of sparse left hand sides. This benefit is made clear in our case studies,
+where prior systems like TACO do not support all possible loop
 orderings and format combinations for sparse matrix multiply because they do
@@
 
+\begin{wrapfigure}{r}{.4\textwidth}
+    \vspace{-12pt}
+    \centering
+    \scriptsize
+    \begin{tabular}{|c|c|c|c|l|}
+        \hline
+        \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singular } & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\ \hline
+        &  &  &  & Dense \\ \hline
+        &  &  & $\checkmark$ & n/a \\ \hline
+        &  & $\checkmark$ &  & DenseRLE \\ \hline
+        &  & $\checkmark$ & $\checkmark$ & n/a \\ \hline
+        & $\checkmark$ &  &  & n/a \\ \hline
+        & $\checkmark$ &  & $\checkmark$ & n/a \\ \hline
+        & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
+        & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
+       $\checkmark$ &  &  &  & Sparse \\ \hline
+       $\checkmark$ &  &  & $\checkmark$ & SparsePinpoint \\ \hline
+       $\checkmark$ &  & $\checkmark$ &  & SparseRLE \\ \hline
+       $\checkmark$ &  & $\checkmark$ & $\checkmark$ & SparseInterval \\ \hline
+       $\checkmark$ & $\checkmark$ &  &  & SparseVBL \\ \hline
+       $\checkmark$ & $\checkmark$ &  & $\checkmark$ & SparseBand \\ \hline
+       $\checkmark$ & $\checkmark$ & $\checkmark$ &  & n/a \\ \hline
+       $\checkmark$ & $\checkmark$ & $\checkmark$ & $\checkmark$ & n/a \\ \hline
+    \end{tabular}
+    \caption{All combinations of our 4 structural properties and the
+    corresponding formats we have chosen to represent them. Not all combinations
+    are relevant. Note that blocks and runs need not be considered together
+    because we must store a run length for each run, so there isn't a
+    significant storage benefit to combining them. Blocks and singletons only
+    make sense in the contex of sparsity.}
+    \vspace{-24pt}
+    %We omit such combinations from the exhaustive table.
+    \label{tab:TypesOfStructure}
+\end{wrapfigure}
+
     In the Finch programming model, the programmer relies on the Finch compiler to specialize to the sequential properties of the data.
 In our experience, the main benefits of specializing to structure come from the following properties of the data:
+
     \begin{itemize}
@@
 
-        \item \textbf{Singletons} When we have only one non-fill region in sparse data,
+        \item \textbf{Singular} When we have only one non-fill region in sparse data,
         we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
@@
     While it is impossible to write code which
-    precisely addresses every possible structure, the combination of our level
-    formats can express a wide variety of hierarchical structures to a
+    precisely addresses every possible structure, our level
+    formats can be combined to express a wide variety of hierarchical structures to a
     sufficient granularity that we can generate code which utilizes the four
@@
 To reduce the implementation burden and improve efficiency in the common case, the levels we described in the previous section only support bulk,
-sequential creation of formats. However, when users want to be able to write out
+sequential construction of formats. However, when users want to be able to write out
 of order (which is a common requirement arising from loop order or from the
@@
 \subsubsection{Scalars}
-Because elements are geared towards representing many leaves in a level, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
+Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
 %
@@
 
-\begin{wrapfigure}{r}{.5\textwidth}
-        \centering
-        \vspace{-20pt}
-        \scriptsize
-        \begin{tabular}{|c|c|c|c|l|}
-            \hline
-            \rothead{Sparse} & \rothead{Blocked} & \rothead{Runs} & \rothead{Singletons} & \begin{tabular}{@{}l@{}}\textbf{Corresponding} \\ \textbf{Format}\end{tabular} \\
-            \hline
-             &  &  &  & Dense \\
-            \hline
-             &  & \checkmark &  & DenseRLE \\
-            \hline
-            \checkmark &  &  &  & Sparse \\
-            \hline
-            \checkmark &  &  & \checkmark & SparsePinpoint \\
-            \hline
-            \checkmark &  & \checkmark &  & SparseRLE \\
-            \hline
-            \checkmark &  & \checkmark & \checkmark & SparseInterval \\
-            \hline
-            \checkmark & \checkmark &  &  & SparseVBL \\
-            \hline
-            \checkmark & \checkmark &  & \checkmark & SparseBand \\
-            \hline
-        \end{tabular}
-        \vspace{-8pt}
-        \caption{All combinations of relevant structural properties and their
-        corresponding formats.  Note that blocks and runs need not be considered
-        together because we must store a run length for each run, and so there
-        isn't a significant storage benefit to combining them. Blocks and
-        singletons only make sense in the contex of sparsity, so we don't
-        consider them together either. We omit such combinations
-        from the exhaustive table.}
-        \label{tab:TypesOfStructure}
-        \vspace{-36pt}
-    \end{wrapfigure}
-
-\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction.
+%\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction.
 Currently, the cost of freezing and thawing is usually proportional to the number of stored fibers, so an operation that sets a single value in a tensor would be prohibitively expensive (i.e.
@@
 
-\help{this section fails to address freeze or thaw and how it relates to the concrete levels}
+%\help{this section fails to address freeze or thaw and how it relates to the concrete levels}
 

### 2024-04-05 sections/tensors.tex
done with 3

@@
 
-\subsubsection{Scalars}
-Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
-%
-Scalars don't have as much structure as clearly only two structures apply.
-%
-Through these structures, scalars can also affect other tensors in crucial ways.
-%
-Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
-%
-Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
-%
-We also provide early break scalars, which modify the stepper Looplets to
-re-specialize the loop whenever a reduction into an early break scalar hits an
-annihilator.
-%
-We assume that we don't need to re-specialize other Looplets because the stepper
-is the only one which repeats a non-constant number of times.
-
 %\HIDE{\saman{Can we not put future work into the submission!} As future work, we also recommend the implementation of a tree-based format, as this would enable incremental construction.
@@
 %\help{this section fails to address freeze or thaw and how it relates to the concrete levels}
-
-\subsubsection{Leaf Levels}
-
-The leaf level stores the actual entries of the tensor. In most cases, it is
-sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
-values are the same, an additional optimization can be made by storing the
-identical value only once. In this work, we introduce the concept of a \textbf{pattern}
-level to handle this binary case.
-Pattern has a fill value of $false$, and returning $true$ for all ``stored'' values.
-The pattern level allows us to easily represent unweighted graphs or other boolean matrices.
 
@@
 \end{table}
+
+\subsubsection{Scalars}
+Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
+%
+Scalars don't have as much structure as clearly only two structures apply.
+%
+Through these structures, scalars can also affect other tensors in crucial ways.
+%
+Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
+%
+Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
+%
+We also provide early break scalars, which modify the stepper Looplets to
+re-specialize the loop whenever a reduction into an early break scalar hits an
+annihilator.
+%
+We don't need to re-specialize other Looplets because the stepper
+is the only one which repeats a non-constant number of times.
+%
+We represent early break as a structural property rather than a program node
+because it allows us to represent the tail of a loop where one scalar has hit an
+annihlator but another scalar hasn't.
+%
+To our knowlege, sparse and early break scalars are novel contributions of this
+work; other systems don't include them, limiting the impact of sparsity.
+
+\subsubsection{Leaf Levels}
+
+The leaf level stores the actual entries of the tensor. In most cases, it is
+sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
+values are the same, an additional optimization can be made by storing the
+identical value only once. In this work, we introduce the concept of a \textbf{pattern}
+level to handle this binary case.
+Pattern has a fill value of $false$, and returning $true$ for all ``stored'' values.
+The pattern level allows us to easily represent unweighted graphs or other boolean matrices.

### 2024-04-05 sections/lang.tex
Update on Overleaf.
good till 5

@@
+\vspace{12pt}
 \section{The Finch Language}
 
-\begin{wrapfigure}{r}{0.51\linewidth}
-    %\vspace{-26pt}
+\begin{wrapfigure}{R}{0.51\linewidth}
+    \vspace{-28pt}
     \begin{minted}{julia}
@@
 Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
-addition of multiple left hand sides through code blocks, access with expressions, and explicit initialiation.}\label{fig:syntax}
-
+addition of multiple left hand sides through code blocks, access with arbitrary expressions, and explicit declaration, as well as freeze and thaw.}\label{fig:syntax}
+\vspace{-12pt}
 \end{wrapfigure}
@@
 
-Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
+Our syntax is highly permissive: by allowing blocks of code with multiple statements, we implicitly support many features gained through complicated scheduling commands in other frameworks, such as multiple outputs, masking to avoid work, temporary tensors, and arbitrary loop fusion and nesting.
 %
-Our breadth-first seach also uses an
-\mintinline{julia}{if} statement to avoid operating on vertices outside the
-frontier. 
+These features are seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply, which simply writes to a temporary tensor in an inner loop and then reuses it, or in our breadth-first-search, which uses an \mintinline{julia}{if} statement to avoid operating on vertices outside the frontier.
 %
-We have found that \mintinline{julia}{if} is particularly useful in
-our model for masking to avoid work.
+The only restriction we impose on our syntax is that it must respect tensor life cycles.
 %
-However, to compile these constructs efficiently, we must restrict this syntax in several ways, as we have alluded to in the previous section, via life cycles.
-%
-In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.
+In the semantics section, we detail the specifics of how we compile our syntax to efficient code over structured data.
 
@@
 %
-First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
+First, we note that the life cycle rules (in combination with the definition rules) prevent programs that access tensors on both the left and the right hand side of an incrementing assignment within a loop.
 %
-Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
+Tensors may only change their mode in the scope in which they were declared.
 %
-We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
+The $\finchloop$, $\finchdeclare$, and $\finchsieve$ nodes introduce new scopes, and tensors without declarations
+are assumed to have global scope.
 %
-These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
+Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants, zeros, and repetitive values within a loop, eliminating problems that in a less restrictive language would require alias analysis.
+%(similar to the role of a pure functions in Halide~\cite{ragan-kelley_halide_2013}).
 %
-Second, loops with tensor accesses enter into a looplet system via the $Unfurl$ rule.
+%We eliminate problems that in a less restrictive language would require alias analysis to, for example, propagate a zero or repetitive value in a tensor.
+%
+Second, loops enter into a looplet system via the $Unfurl$ rule.
 %
@@
 
-Though the reader may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
+The next section will detail how the compiler adds additional features on top of this core language.  
+We rely on the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the tensor access.
 %
-In the next section, we will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
+We will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
 %
-Also in the next section, we will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
+We will also show how we can similarly exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.
 
@@
 
-\begin{figure}
+\begin{figure}[b]
     \centering

### 2024-04-05 sections/compiler.tex
fixes
fix
Update on Overleaf.
good on 5

@@
 
-Our semantics in Figure
-\ref{fig:semantics_core,fig:semantics_lifecycle,fig:semantics_looplets} is only
+Our semantics in Figure~\ref{fig:semantics_core} and Figure~\ref{fig:semantics_looplets} is only
 well-defined on some programs. We define a particular class of programs on which
@@
 
-\subsection{Wrapperization}
-    \begin{wrapfigure}{R}{0.17\textwidth}
-        \vspace{-1\intextsep}
+    \begin{wrapfigure}{r}{0.17\textwidth}
+        \vspace{-14pt}
         \begin{minted}{julia}
@@
     \end{wrapfigure}
+\subsection{Wrapperization}
 
@@
     and shifts all of the ranges declared in Figure \ref{fig:semantics_looplets} by one.
-    The implementation burden for a wrapper tensor is to implement a suitable
-    program modification during the wrapperization procedure and then unwrap the wrapper, or to 
+    The implementation burden for a wrapper is to implement a suitable
+    program rewrite during the wrapperization procedure to unwrap the wrapper, or to 
     implement the looplet functions of \ref{fig:semantics_looplets} with some minor modifications
@@
         \hline
-        A[p(i)] $\rightarrow$ protocolize(A, p)[i] & Accesses dimension n with protocol protos[n], allowing for advanced iteration protocols. \\
+        %A[p(i)] $\rightarrow$ protocolize(A, p)[i] & Accesses dimension n with protocol protos[n], allowing for advanced iteration protocols. \\
         \hline
@@
     Other languages, such as Halide, perform bounds inference where
-    known bounds are symbolically propagated to fill in unknown bounds, often from the desired output/input sizes to intermediates via some conservative approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
+    known bounds are symbolically propagated to fill in unknown bounds, often from output/input sizes to intermediates via some approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
     %
@@
 
-    In Finch, we use a straightforward algorithm to automatically determine suitable dimensions for loops and declaration statements (output tensors).
+    In Finch, we use a straightforward dimensionalization algorithm on loops and declaration statements (output tensors).
     %
@@
         \caption{Dimensionalization.}\label{fig:dimensionalization}
+        \vspace{-12pt}
     \end{wrapfigure}
@@
     \end{enumerate}
+
+    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B.
+Also, the first dimension of A must match the i loop dimension, 1:3.
+Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5).
+If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.
 
@@
       \end{minipage}\hfill
-        \caption{Examples of concordization, transforming accesses to column major with bare indices.}\label{fig:concordization}
+      \caption{Examples of concordization, transforming accesses to normal column major.}\label{fig:concordization}
+      \vspace{-20pt}
   \end{wrapfigure} 
-
-    For example, in Figure ~\ref{fig:dimensionalization}, the second dimension of A must match the first dimension of B.
-Also, the first dimension of A must match the i loop dimension, 1:3.
-Finch will also resize declared tensors to match indices used in writes, so C is resized to (1:3, 1:5).
-If no dimensions are specified elsewhere, then Finch will use the dimension of the declared tensor.
-
     Dimensionalization occurs after wrapper arrays are de-sugared. You can
@@
 
-    \begin{wrapfigure}{R}{0.34\linewidth}
+    \begin{wrapfigure}{R}{0.34\textwidth}
+        \vspace{-12pt}
         \begin{minipage}{0.15\textwidth}
@@
             for i = _
-                y[i] .= x[i] + 1
+                y[i] = x[i] + 1
             end
@@
         \end{minipage}
+
+        \vspace{-8pt}
         \caption{Life cycle automation.} \label{fig:lifecycles}
+        \vspace{-18pt}
     \end{wrapfigure}

### 2024-04-05 sections/eval/api.tex
Update on Overleaf.
fixes

@@
 %
-Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degredation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
+Further, implementing each operation on its own is insufficient. Operator fusion is required to avoid asymptotic performance degradation on sparse kernels, as we show in Figure~\ref{fig:fusion}.
 %
@@
 
-\begin{wrapfigure}{R}{0.6\linewidth}
+\begin{wrapfigure}{R}{0.57\linewidth}
   \scriptsize
+  \vspace{-12pt}
 \begin{minted}{julia}
-  EXPR := table(TENSOR, FIELDS...) |         #TENSOR is an immediate tensor
-    ALIAS | FIELD | mapjoin(OP, EXPRS...) |  #OP is a immediate function
-    aggregate(OP, INIT, EXPR, FIELDS...) |   #INIT is an immediate init value
+  EXPR := table(TENSOR, FIELDS...) |           #TENSOR is a tensor value
+    ALIAS | FIELD | mapjoin(OP, EXPRS...) |    #OP is a function
+    aggregate(OP, INIT, EXPR, FIELDS...) |     #INIT is an initial value
     reorder(EXPR, FIELDS...) |                 #FIELD is a name for an index
     relabel(EXPR, FIELDS...) |                 #ALIAS is a name for a tensor
-    reformat(TENSOR, EXPR)                     PLAN := plan(QUERIES..., PRODUCES)
+    reformat(TENSOR, EXPR)                       PLAN := plan(QUERIES...)
                                                 QUERY := query(ALIAS, EXPR)
 \end{minted}
-\caption{\scriptsize Finch Logic Syntax. }\label{fig:finch_logic_syntax}
+\vspace{-12pt}
+\caption{\footnotesize Finch Logic Syntax. }\label{fig:finch_logic_syntax}
+\vspace{-12pt}
+\end{wrapfigure}
+\begin{figure}[b]
 \begin{minted}{python}
@@
 \end{minted}
-\caption{\scriptsize Implementations of a few common functions in Finch Logic.
+\vspace{-12pt}
+\caption{\footnotesize Implementations of a few common functions in Finch Logic.
 Here, \texttt{minby(x, y)} compares \texttt{x[1]} and \texttt{y[1]} and returns the smaller \texttt{x} or \texttt{y}.
@@
 Finch can understand declared properties and rewrite rules for all of these functions and types, extending these algorithms to operate on structured arguments.} \label{fig:api_extensible}
-\end{wrapfigure}
+\vspace{-18pt}
+\end{figure}
 \subsubsection{Finch Logic and the Physical Interpreter}
 Finch Logic is used as a common representation to implement and fuse the operations in the Array API.
+%
+The syntax is designed to be minimal and expressive, taking inspiration from relational algebra with the goal of being able to express most of the operations in the Array API.
 
@@
 The syntax of Finch Logic is described in Figure~\ref{fig:finch_logic_syntax}.
-%
-Note that the syntax is designed to be minimal and expressive, taking inspiration from relational algebra with the goal of being able to express most of the operations in the Array API.
 %

### 2024-04-05 sections/eval/spmv.tex
fixes

@@
     \label{spmv_programs}
-    \vspace{-36pt}
+    \vspace{-12pt}
   \end{wrapfigure}
@@
     \centering
-    \vspace{-16pt}
+    \vspace{-14pt}
     \begin{tabular}{|r|l|l|}
@@
 
+  \begin{figure}[!h]
+    \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
+    \caption{\footnotesize Performance of SpMV by Finch format.
+The performance displayed for Finch on each dataset is the fastest among the formats we tested.
+Programs are from Figure~\ref{spmv_programs} and formats are from Figure~\ref{spmv_tensor_formats}.
+``finch\_baseline'' is the faster of row major or column major ``SparseList''.}
+    \label{fig:spmv_grouped}
+  \end{figure}
+
   We found that the SpMV performance was superior for the level format that best paralleled the structure of the tensor.
   %
-  Namely, matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had slowdowns of 0.71, 0.53, and 0.92 relative to TACO.
+  Matrices with a clear blocked structure like exdata\_1, TSOPF\_RS\_b678\_c1, and heart3 performed notably well with the SparseVBL format with speedups of 2.16, 1.55, and 1.30 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
   %
-  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had slowdowns of 0.84 and 0.51 relative to TACO.
+  Furthermore, the synthetic banded matrices we constructed performed the best with the SparseBand matrix, in particular with the large\_band and the medium\_band matrices having a speedup of 1.98 and 1.64 relative to TACO, while the baseline format had minor slowdowns relative to TACO.
   %
-  There were also significant advantages of using the Pattern format instead of the Element format to represent scalar values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
+  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
   %
@@
   %Here's a figure with spmv_performance_sorted_(faster_than_taco).png and spmv_performance_sorted_(slower_than_taco).png
-
-  \begin{figure}
-      \includegraphics[width=\linewidth]{Spmv_annotated.drawio.png}
-      \vspace{-18pt}
-      \caption{Performance of SpMV by Finch format.}
-      \label{fig:spmv_grouped}
-      \footnotesize The performance displayed for Finch on each dataset is the fastest among the formats we tested.
-Programs are from Figure~\ref{spmv_programs} and formats are from Figure~\ref{spmv_tensor_formats}.
-``finch\_baseline'' is the faster of row major or column major ``SparseList''.
-  \end{figure}

### 2024-04-06 sections/related.tex
add images

@@
 %
-In ML, several frameworks offer this type of support, most notably TorchSparse, which supports a few arrays and a few operations on them~\cite{tang2022torchsparse, tang2023torchsparse++}.
+Several machine learning frameworks support some sparse arrays and operations, most notably TorchSparse\cite{tang2022torchsparse, tang2023torchsparse++}.
 %
@@
 %
-The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
+The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
 %
 Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions.
+SparseTIR is a similar system supporting combined
+sparse formats (including block structures) \cite{ye_sparsetir_2022}.
 %

### 2024-04-06 sections/appendix.tex
cool

@@
 \end{minted}
+
+\section{Mask Images}
+We interpreted the following images from ``Digital Image Processing'' \cite{gonzalez_digital_2006} as masks:
+
+\textbf{FigP1012.png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP1012.png}}
+
+\textbf{Fig1008(a)(step edge).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1008(a)(step edge).png}}
+
+\textbf{FigP0905(d).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(d).png}}
+
+\textbf{FigP0528(c)(doughnut).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0528(c)(doughnut).png}}
+
+\textbf{FigP0616(b).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0616(b).png}}
+
+\textbf{Fig0114(c)(bottles).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0114(c)(bottles).png}}
+
+\textbf{FigP0616(c).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0616(c).png}}
+
+\textbf{FigP0433(b).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0433(b).png}}
+
+\textbf{Figp0917.png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Figp0917.png}}
+
+\textbf{FigP0905(b).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(b).png}}
+
+\textbf{Fig1059(c)(NegADI).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1059(c)(NegADI).png}}
+
+\textbf{Fig1111(a)(triangle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1111(a)(triangle).png}}
+
+\textbf{Fig1111(b)(square).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1111(b)(square).png}}
+
+\textbf{FigP0905(top).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(top).png}}
+
+\textbf{FigP1110.png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP1110.png}}
+
+\textbf{Fig0533(a)(circle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0533(a)(circle).png}}
+
+\textbf{FigP0917(noisy\_rectangle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0917(noisy\_rectangle).png}}
+
+\textbf{Fig0230(b)(dental\_xray\_mask).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0230(b)(dental\_xray\_mask).png}}
+
+\textbf{FigP0528(b)(two\_dots).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0528(b)(two\_dots).png}}
+
+\textbf{Fig1059(a)(AbsADI).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1059(a)(AbsADI).png}}
+
+\textbf{Fig1059(b)(PosADI).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1059(b)(PosADI).png}}
+
+\textbf{Fig0539(c)(shepp-logan\_phantom).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0539(c)(shepp-logan\_phantom).png}}
+
+\textbf{FigP0905(c).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(c).png}}
+
+\textbf{Fig1043(a)(yeast\_USC).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1043(a)(yeast\_USC).png}}
+
+\textbf{FigP0905(U).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(U).png}}
+
+\textbf{Fig0524(b)(blurred-impulse).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0524(b)(blurred-impulse).png}}
+
+\textbf{Fig0424(a)(rectangle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0424(a)(rectangle).png}}
+
+\textbf{Fig1008(c)(roof\_edge).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1008(c)(roof\_edge).png}}
+
+\textbf{Fig0539(a)(vertical\_rectangle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0539(a)(vertical\_rectangle).png}}
+
+\textbf{FigP0905(a).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0905(a).png}}
+
+\textbf{FigP0433(a).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0433(a).png}}
+
+\textbf{Fig0.15(a)(translated\_rectangle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0425(a)(translated\_rectangle).png}}
+
+\textbf{FigP0918(c).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0918(c).png}}
+
+\textbf{Fig0524(a)(impulse).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0524(a)(impulse).png}}
+
+\textbf{Fig0236(a)(letter\_T).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0236(a)(letter\_T).png}}
+
+\textbf{Fig0503 (original\_pattern).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0503 (original\_pattern).png}}
+
+\textbf{FigP0501.png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0501.png}}
+
+\textbf{Fig1218(airplanes).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig1218(airplanes).png}}
+
+\textbf{Fig0534(a)(ellipse\_and\_circle).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/Fig0534(a)(ellipse\_and\_circle).png}}
+
+\textbf{FigP0616(a).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0616(a).png}}
+
+\textbf{FigP0918(b).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0918(b).png}}
+
+\textbf{FigP0528(c).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0528(c).png}}
+
+\textbf{FigP0528(a)(single\_dot).png}
+
+\includegraphics[scale=0.15]{{./dip3e\_masks/FigP0528(a)(single\_dot).png}}

### 2024-04-06 sections/appendix.tex
fixed.

--- 2024-04-04 sections/appendix.tex
@@
 \end{minted}
+
+\section{Mask Images}
+We interpreted the following images from ``Digital Image Processing'' \cite{gonzalez_digital_2006} as masks:
+
+\textbf{FigP1012.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP1012.png}}
+
+\textbf{Fig1008(a)(step edge).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008(a)(step edge).png}}
+
+\textbf{FigP0905(d).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(d).png}}
+
+\textbf{FigP0528(c)(doughnut).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528(c)(doughnut).png}}
+
+\textbf{FigP0616(b).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616(b).png}}
+
+\textbf{Fig0114(c)(bottles).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0114(c)(bottles).png}}
+
+\textbf{FigP0616(c).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616(c).png}}
+
+\textbf{FigP0433(b).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0433(b).png}}
+
+\textbf{Figp0917.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Figp0917.png}}
+
+\textbf{FigP0905(b).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(b).png}}
+
+\textbf{Fig1059(c)(NegADI).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059(c)(NegADI).png}}
+
+\textbf{Fig1111(a)(triangle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1111(a)(triangle).png}}
+
+\textbf{Fig1111(b)(square).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1111(b)(square).png}}
+
+\textbf{FigP0905(top).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(top).png}}
+
+\textbf{FigP1110.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP1110.png}}
+
+\textbf{Fig0533(a)(circle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0533(a)(circle).png}}
+
+\textbf{FigP0917(noisy\_rectangle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0917(noisy\_rectangle).png}}
+
+\textbf{Fig0230(b)(dental\_xray\_mask).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0230(b)(dental\_xray\_mask).png}}
+
+\textbf{FigP0528(b)(two\_dots).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528(b)(two\_dots).png}}
+
+\textbf{Fig1059(a)(AbsADI).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059(a)(AbsADI).png}}
+
+\textbf{Fig1059(b)(PosADI).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059(b)(PosADI).png}}
+
+\textbf{Fig0539(c)(shepp-logan\_phantom).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0539(c)(shepp-logan\_phantom).png}}
+
+\textbf{FigP0905(c).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(c).png}}
+
+\textbf{Fig1043(a)(yeast\_USC).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1043(a)(yeast\_USC).png}}
+
+\textbf{FigP0905(U).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(U).png}}
+
+\textbf{Fig0524(b)(blurred-impulse).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0524(b)(blurred-impulse).png}}
+
+\textbf{Fig0424(a)(rectangle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0424(a)(rectangle).png}}
+
+\textbf{Fig1008(c)(roof\_edge).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008(c)(roof\_edge).png}}
+
+\textbf{Fig0539(a)(vertical\_rectangle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0539(a)(vertical\_rectangle).png}}
+
+\textbf{FigP0905(a).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905(a).png}}
+
+\textbf{FigP0433(a).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0433(a).png}}
+
+\textbf{Fig0.15(a)(translated\_rectangle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0425(a)(translated\_rectangle).png}}
+
+\textbf{FigP0918(c).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0918(c).png}}
+
+\textbf{Fig0524(a)(impulse).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0524(a)(impulse).png}}
+
+\textbf{Fig0236(a)(letter\_T).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0236(a)(letter\_T).png}}
+
+\textbf{Fig0503 (original\_pattern).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0503 (original\_pattern).png}}
+
+\textbf{FigP0501.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0501.png}}
+
+\textbf{Fig1218(airplanes).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1218(airplanes).png}}
+
+\textbf{Fig0534(a)(ellipse\_and\_circle).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0534(a)(ellipse\_and\_circle).png}}
+
+\textbf{FigP0616(a).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616(a).png}}
+
+\textbf{FigP0918(b).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0918(b).png}}
+
+\textbf{FigP0528(c).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528(c).png}}
+
+\textbf{FigP0528(a)(single\_dot).png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528(a)(single\_dot).png}}

### 2024-04-06 sections/related.tex
fix figures

@@
 %
-The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
+The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, and compilation in MLIR \cite{bik_compiler_2022}, all as separate extensions.
 %

### 2024-04-06 sections/eval/graphs.tex
fixes
separate

@@
 %
-In Appendix~\ref{sec:graph_listings}, we display the code for BFS and Bellman-Ford in Finch (57 and 50 LOC) and LAGraph (215 and 227 LOC), and invite readers to compare the clarity of the algorithms.
+In Appendix B, we display the code for BFS and Bellman-Ford in Finch (57 and 50 LOC) and LAGraph (215 and 227 LOC), and invite readers to compare the clarity of the algorithms.
 
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
-	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
-    \vspace{-12pt}
-    \caption{Performance of graph apps across various tools.
-finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
-Finch's support for push/pull traversal and early break facilitates direction-optimization.
-Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
-     \label{fig:graph_result}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[b]
     \begin{minipage}{0.33\linewidth}
@@
 \end{figure}
+
+\begin{figure}[t]
+	\includegraphics[width=0.5\linewidth]{bfs_speedup_over_graphs.jl.png}%
+	\includegraphics[width=0.5\linewidth]{bellmanford_speedup_over_graphs.jl.png}
+    \vspace{-12pt}
+    \caption{Performance of graph apps across various tools.
+finch\_push\_only exclusively utilizes push traversal, while finch\_push\_pull applies direction-optimization akin to GraphBLAS.
+Finch's support for push/pull traversal and early break facilitates direction-optimization.
+Among GraphBLAS's five variants for Bellman-Ford, we selected LAGraph\_BF\_full1a, consistently the fastest with our graphs.}
+     \label{fig:graph_result}
+\end{figure}

### 2024-04-06 sections/eval/morphology.tex
separate

@@
 %
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in the appendix) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
+We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
 %

### 2024-04-06 sections/background.tex
fix

@@
     \hline
-    $\finchlookup(seek, body)$: The Lookup Looplet represents a randomly accessible region of an iterator.
+    $\finchlookup(seek, body)$: The Lookup looplet represents a randomly accessible region of an iterator.
 The body of the lookup is understood to have one less dimension than the lookup itself, as we have already ``looked up'' that index in the tensor by the time we reach the body.
@@
     \hline
-    $\finchrun(body)$: The Run Looplet represents a constant region of an iterator.
+    $\finchrun(body)$: The Run looplet represents a constant region of an iterator.
 The body of the run is understood to have one less dimension than the lookup itself, as all of the bodies are identical.
@@
     \hline
-    $\finchphase(c:d, body)$: The Phase Looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
+    $\finchphase(c:d, body)$: The Phase looplet represents a restriction of the range on which a loop should execute, and allows us to succinctly express the ranges on which children of compound looplets are defined.
     &
@@
     \hline
-    $\finchswitch(cond, head, tail)$: The Switch Looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context.
+    $\finchswitch(cond, head, tail)$: The Switch looplet allows us to specialize the body of a looplet based on a condition, evaluated in the embedding context.
 If the condition is true, we use `head`, otherwise we use `tail`.
@@
 This also lifts the condition as high as possible into the loop nest.
-The condition is assumed to evaluate to a boolean.
+The condition is assumed to evaluate to a Boolean.
     &
@@
     \hline
-    $\finchthunk(preamble, body, epilogue)$: The Thunk Looplet allows us to cache certain computations in the state under which the body will execute.
+    $\finchthunk(preamble, body, epilogue)$: The Thunk looplet allows us to cache certain computations in the state under which the body will execute.
 This is useful for computing and caching the results of expensive computations.
@@
     $\finchsequence(head, tail)$: The Sequence looplet represents the concatenation of two looplets.
-Both arguments must be phase looplets, and are assumed to be nonoverlapping, covering, and in order.
+Both arguments must be phase looplets, and are assumed to be non-overlapping, covering, and in order.
     &
@@
     \hline
-    $\finchspike(body, tail)$: The Spike Looplet represents a run followed by a single value.
+    $\finchspike(body, tail)$: The Spike looplet represents a run followed by a single value.
 In this paper, Spike will be considered a shorthand for $\finchsequence(\finchphase(i:j-1, \finchrun(body)), \finchphase(j:j, \finchrun(tail)))$.
@@
 \vspace{-8pt}
-\caption{The looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
+\caption{The Looplet language, as understood in a correct execution of a Finch program.}\label{fig:looplets}
 \end{figure}
@@
 %
-In particular, looplets are good at lifting code to the highest possible loop level and subdividing iteration hierarchically in coordinate space.
+In particular, Looplets are good at lifting code to the highest possible loop level and subdividing iteration hierarchically in coordinate space.
 %
-Because looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
+Because Looplets are compiled with progressive lowering, structure-specific mathematical optimizations such as integrals, multiply by zero, etc.
 can be implemented using simple compiler passes like term rewriting and constant propagation during the intermediate lowering stages.
 
-The looplets are described in Figure~\ref{fig:looplets}.
+The Looplets are described in Figure~\ref{fig:looplets}.
 We simplify the presentation to focus on the semantics, rather than precise implementation.

### 2024-04-06 sections/compiler.tex
fix

@@
 constraints can be rewritten to programs that satisfy them, and thus how we can
-support such a wide variety of programs. For example, we can write nonconcordant
+support such a wide variety of programs. For example, we can write non-concordant
 programs like  \mintinline{julia}{for i = _; for j = _; s[] += A[i, j] end end}

### 2024-04-06 sections/eval.tex
fix

@@
 how flexible operators, formats, and indexing expressions in Finch have
-supported a flexible implementation of the Python Array API, supporting fused exection.
+supported a flexible implementation of the Python Array API, supporting fused execution.
 

### 2024-04-06 sections/eval/morphology.tex
fix

@@
 %
-IN our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
+In our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
 %
-We believe the 51.6x on MNIST is due to calling overhead in OpenCV). 
+We believe the 51.6x on MNIST is due to calling overhead in OpenCV. 
 %

### 2024-04-06 sections/eval/spmv.tex
fix

@@
         \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Element(0.0)))} & real, sparse \\
-        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & boolean, sparse \\
+        \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseList(Pattern()))} & Boolean, sparse \\
         \mintinline[fontsize=\scriptsize]{julia}{Dense(SparseVBL(Element(0.0)))} & real, blocked \\
@@
 to test the TACO library \cite{kjolstad_tensor_2017}.
-Additionally, we included the SNAP graph collection to test with boolean matrices.
+Additionally, we included the SNAP graph collection to test with Boolean matrices.
 We also created some synthetic matrices containing bands of varying sizes.
@@
   %
-  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were boolean, such as matrices in the SNAP collection which represent graph datasets are boolean.
+  The Pattern format performed better than the Element format for representing the leaf values in the matrices when these values were Boolean, such as matrices in the SNAP collection which represent graph datasets are Boolean.
   %

### 2024-04-06 sections/intro.tex
fix

@@
 For example, a convolution with a filter can be expressed as a matrix multiplication
-with the toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
+with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
 
@@
 %
-For example, TACO merge lattices represent boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
+For example, TACO merge lattices represent Boolean logic over sets of non-zero values on an integer grid~\cite{kjolstad_tensor_2017}.
 %
@@
 %
-\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as spmv ($2.51\times$) and spgemm ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
+\item We evaluate the efficiency, flexibility, and expressibility of our language in several case studies on a wide range of applications, demonstrating speedups over the state of the art in classic operations such as SpMV ($2.51\times$) and SpGEMM ($1.25\times$), to more complex applications such as graph analytics ($3.53\times$, reducing lines of code by $4\times$ over GraphBLAS), image processing ($19.5\times$), and the Python Array API ($28\times$).
 %We also demonstrate how Finch can fuse high-level operations to achieve a significant speedup over non-fused kernels.
@@
 
-This advances the state-of-the-art over the looplets work \cite{ahrens_looplets_2023}.
+This advances the state-of-the-art over the Looplets work \cite{ahrens_looplets_2023}.
 While Looplets presented a way to merge iterators over multiple single dimensional structures, we go further by combining single-dimensional looplets into multi-dimensional tensors and operating on them in a programming language with fully-featured control flow.

### 2024-04-06 sections/lang.tex
fix

@@
 %
-These features are seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply, which simply writes to a temporary tensor in an inner loop and then reuses it, or in our breadth-first-search, which uses an \mintinline{julia}{if} statement to avoid operating on vertices outside the frontier.
+These features are seen most prominently in our implementation of Gustavson's algorithm for sparse-sparse matrix multiply, which simply writes to a temporary tensor in an inner loop and then reuses it, or in our breadth-first-search, which uses an \mintinline{julia}{if} statement to avoid operating on vertices outside the frontier.
 %
@@
 
-% Finch uses looplets to lower \mintinline{julia}{for}-loops, though a sparse
+% Finch uses Looplets to lower \mintinline{julia}{for}-loops, though a sparse
 % program with complicated looplets should produce output tensors with the same
@@
 %
-In Figure~\ref{fig:semantics_looplets}, we detail the Looplet evaluation semantics, which details how acceses within a loop can be replaced with Loopets, which are progressively rewritten to host code.
+In Figure~\ref{fig:semantics_looplets}, we detail the Looplet evaluation semantics, which details how accesses within a loop can be replaced with Looplets, which are progressively rewritten to host code.
 %
@@
     %
-    Several Looplets introduce variables into the embedding
+    Several looplets introduce variables into the embedding
     language, which may be read when evaluating the {\finchvalue} node.
@@
     %
-    The nonlocal evaluations of Looplets are what allow looplets to
+    The nonlocal evaluations of Looplets are what allow Looplets to
     hoist conditions and subranges out of loops.

### 2024-04-06 sections/related.tex
fix

--- 2024-04-06 sections/related.tex
@@
 %
-Perhaps the most well-known example is Numpy~\cite{harris_array_2020},
+Perhaps the most well-known example is NumPy~\cite{harris_array_2020},
 and a classic example is the BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}. 
@@
 %
-Examples include Scipy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
+Examples include SciPy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
 %
@@
 %
-Similarly, the Mapreduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
+Similarly, the MapReduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
 algorithms in the GBASE library\cite{kang2011gbase}.
@@
 %
-Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
+Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendant such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
 %
@@
 %
-The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
+The TACO compiler originally supported just plain Einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, and compilation in MLIR \cite{bik_compiler_2022}, all as separate extensions.
 %
@@
 %
-The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
+The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to Einsums~\cite{zhao2022polyhedral}.
 %
@@
 %
-However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
+However, many SQL or dataframe systems rely on b-trees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
 %

### 2024-04-06 sections/tensors.tex
fix

--- 2024-04-05 sections/tensors.tex
@@
 %
-Our level abstraction provides a bridge between the single dimensional iterators created from looplets and the mutli-dimensional fiber-tree abstractions common to tensor compilers.
+Our level abstraction provides a bridge between the single dimensional iterators created from Looplets and the multi-dimensional fiber-tree abstractions common to tensor compilers.
 % 
@@
 %
-First, while looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
+First, while Looplets represent an instance of an iterator over a tensor, we may access the same tensor twice with different indices.
 %
@@
 %
-Next, since Finch programs go beyond just single einsums, they may read and write to the same data at different times.
+Next, since Finch programs go beyond just single Einsums, they may read and write to the same data at different times.
 %
@@
 % %
-% These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to looplets and expanding the contexts in which they can be used.
+% These interfaces add to the level abstraction, expanding the types of data that they can express via mapping to Looplets and expanding the contexts in which they can be used.
 % %
@@
 
-Our simplified view of a level is enabled by our use of looplets to represent
+Our simplified view of a level is enabled by our use of Looplets to represent
 the structure within each fiber.
@@
     significant storage benefit to combining them. Blocks and singletons only
-    make sense in the contex of sparsity.}
+    make sense in the context of sparsity.}
     \vspace{-24pt}
@@
     \textbf{DenseRLE}:
-    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $qth$ run in the $p^th$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
+    Used to represent runs of repeated values, storing two vectors, $right$ and $ptr$, with $q^{th}$ run in the $p^{th}$ subfiber starting and ending at $right[ptr[p] + q]$ and $right[ptr[p] + q + 1] - 1$, respectively.
     %
@@
     %
-    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. \\
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$. \\
     \textbf{SparsePinpoint}:
@@
     %
-    It stores three vectors $left$, $right$, and $ptr$, such that the $qth$ run in the $p^th$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
+    It stores three vectors $left$, $right$, and $ptr$, such that the $q^{th}$ run in the $p^{th}$ subfiber begins and ends at $left[ptr[p] + q]$ and $right[ptr[p] + q]$, respectively.
     %
@@
     %
-    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$. 
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$. 
     %
@@
     %
-    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $qth$ nonzero in the subfiber at position $p$.
+    It stores two vectors, $idx$ and $ptr$, such that $idx[ptr[p] + q]$ is the index of the $q^{th}$ nonzero in the subfiber at position $p$.
     %
@@
 \end{table}
+
+\subsubsection{Scalars}
+Because leaf levels are geared towards representing multiple leaves, we also introduce a much simpler Scalar format to represent 0-dimensional tensors.
+%
+Scalars don't have as much structure as clearly only two structures apply.
+%
+Through these structures, scalars can also affect other tensors in crucial ways.
+%
+Constant propagation through arrays is known to be a complex compiler pass \cite{mcmichen_representing_nodate}.
+%
+Instead, we provide sparse scalars, which allow reductions but also specialize read accesses for the possible zero value.
+%
+We also provide early break scalars, which modify the stepper looplets to
+re-specialize the loop whenever a reduction into an early break scalar hits an
+annihilator.
+%
+We don't need to re-specialize other looplets because the stepper
+is the only one which repeats a non-constant number of times.
+%
+We represent early break as a structural property rather than a program node
+because it allows us to represent the tail of a loop where one scalar has hit an
+annihilator but another scalar hasn't.
+%
+To our knowledge, sparse and early break scalars are novel contributions of this
+work; other systems don't include them, limiting the impact of sparsity.
+
+\subsubsection{Leaf Levels}
+
+The leaf level stores the actual entries of the tensor. In most cases, it is
+sufficient to store each entry at a separate position in a vector. This is accomplished by the \textbf{ElementLevel}. However, when all of the
+values are the same, an additional optimization can be made by storing the
+identical value only once. In this work, we introduce the concept of a \textbf{PatternLevel}
+ to handle this binary case.
+The PatternLevel has a fill value of $false$, and returning $true$ for all ``stored'' values.
+The PatternLevel allows us to easily represent unweighted graphs or other Boolean matrices.

### 2024-04-06 sections/eval/api.tex
spelling

@@
-\subsection{Implementing Numpy's Array API in Finch}
+\subsection{Implementing NumPy's Array API in Finch}
 In the past decade, the adoption of the Python Array API \cite{harris_array_2020} has allowed for a proliferation array programming systems, but existing implementations of this and similar APIs for structured data suffer from either incompleteness or inefficiency.
@@
 %
-Compilers with simpler interfaces (such as TACOs simple einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
+Compilers with simpler interfaces (such as TACO's simple Einsum) need a massive amount of glue code outside of the normal compilation path to support the full API, leading to a large implementation burden.
 %

### 2024-04-06 sections/eval/morphology.tex
spelling

@@
 %
-We focus on binary image morpology and the logical transformation of binary images and masks.
+We focus on binary image morphology and the logical transformation of binary images and masks.
 %
@@
 %
-We also hand-selected a subset of mask images (these images were less homogenous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
+We also hand-selected a subset of mask images (these images were less homogeneous, so we listed them in Appendix C) from a digital image processing textbook \cite{gonzalez_digital_2006}. 
 %
@@
 %
-In our erosion task, the sparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
+In our erosion task, the SparseRLE format performs the best as it is asymptotically faster and uses less memory, leading to a 19.5X speedup over OpenCV on the sketches dataset, which becomes arbitrarily large as we magnify the images (here shown as 266X).
 %

### 2024-04-06 sections/eval/spgemm.tex
fixes
spelling

@@
 \subsection{Sparse-Sparse Matrix Multiply (SpGEMM)}
-\begin{figure}
-    \begin{minipage}{0.333\linewidth}
+
+\begin{figure}[b]
+    \begin{minipage}[t]{0.32\linewidth}
+      \vspace{0pt}
     \begin{minted}{julia}
@@
     \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
+    \begin{minipage}[t]{0.33\linewidth}
+      \vspace{0pt}
     \begin{minted}{julia}
@@
     \end{minipage}%
-    \begin{minipage}{0.333\linewidth}
+    \begin{minipage}[t]{0.35\linewidth}
+      \vspace{0pt}
     \begin{minted}{julia}
@@
     \end{minipage}
-    \vspace{-12pt}
-    \caption{Inner Products, Gustavsons, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
+    \caption{Inner Products, Gustavson's, and Outer Products matrix multiply in Finch}\label{fig:spgemm_listing}
+\end{figure}
+
+\begin{figure}[b]
+	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
+	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
+    \vspace{-8pt}
+    \caption{A comparison of several matrix multiply algorithms between
+    Finch and Taco. On left, we use smaller matrices, ordered from small to big
+    dimension. Note that inner-products necessarily requires $O(MN)$ work and
+    TACO's outer-products format is dense. Finch can use a sparse outer products
+    format and thus has an asymptotic advantage that becomes evident as the
+    output dimensions grow. On right, we use only Gustavson's algorithm and
+    compare on larger matrices.}
+    \label{fig:spgemm}
     \vspace{-12pt}
@@
 %\changwan{(self)may remove this number for consistency with other exps. finch\_outer\_bytemap not explained.}
-
-\begin{figure}
-	\includegraphics[width=0.5\linewidth]{spgemm_small_speedup_log_scale.png}%
-	\includegraphics[width=0.5\linewidth]{spgemm_joel_speedup.png}
-    \vspace{-8pt}
-    \caption{A comparison of several matrix multiply algorithms between
-    Finch and Taco. On left, we use smaller matrices, ordered from small to big
-    dimension. Note that inner-products necessarily requires $O(MN)$ work and
-    TACO's outer-products format is dense. Finch can use a sparse outer products
-    format and thus has an asymptotic advantage that becomes evident as the
-    output dimensions grow. On right, we use only gustavson's algorithm and
-    compare on larger matrices.}
-    \label{fig:spgemm}
-    \vspace{-12pt}
-\end{figure}

### 2024-04-06 sections/lang.tex
nice

@@
     but these semantics enforce that each of these functions may update state in
-    the host language and flip the mode of the tensor between {\finchread} and {\finchupdate}}
+    the host language and flip the mode of the tensor between {\finchread} and {\finchupdate}.}
     \label{fig:semantics_core}
@@
     We omit the full set of rules for brevity, but
-    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples}
+    point the curious reader to \cite[Figure 5]{ahrens_looplets_2023} for some examples.}
     \label{fig:semantics_looplets}

### 2024-04-06 sections/compiler.tex
nice
fix

@@
         A[\~i] $\rightarrow$ permissive(A)[i] & Creates a PermissiveArray, allowing for out-of-bounds access or padding. This tensor returns no dimensions as its size. \\
-        \hline
+        %\hline
         %A[p(i)] $\rightarrow$ protocolize(A, p)[i] & Accesses dimension n with protocol protos[n], allowing for advanced iteration protocols. \\
@@
         \item We assign dimensions to indices.
-        \item Using an index in an access “hints” that the index should have the corresponding dimension.
-        \item Loop dimensions are equal to the “meet” of all hints in the loop body
+        \item Using an index in an access ``hints'' that the index should have the corresponding dimension.
+        \item Loop dimensions are equal to the ``meet'' of all hints in the loop body
         and any existing dimensions in the loop bounds. The meet usually asserts
@@
         \item We assign dimensions to declarations.
-        \item Left hand side (updating) tensor access “hint” the size of that tensor
-        \item The dimensions of a declaration are the “meet” of all hints from
+        \item Left hand side (updating) accesses ``hint'' the size of their tensor
+        \item The dimensions of a declaration are the ``meet'' of all hints from
         the declaration to the first read.

### 2024-04-24 main.tex
update
fixes
Update on Overleaf.
cool
quick fix
fix duplicate citations

@@
 %%
-\documentclass[acmsmall,screen,review,anonymous]{acmart}
+\documentclass[acmsmall,screen,nonacm]{acmart}
 
@@
 %% complete the rights form.
-\setcopyright{acmlicensed}
+\setcopyright{none}
 \copyrightyear{2024}
@@
 \newcommand{\willow}[1]{}
+\newcommand{\changwan}[1]{}
 \newcommand{\leftoff}{}
@@
 %% the work being presented. Separate the keywords with commas.
-\keywords{Array, Sparse, Structure, Control Flow, Language, Compiler}
+\keywords{Sparse Array, Structured Array, Control Flow, Programming Language}
 
@@
 \begin{acks}
-    To Mateusz, Hameer, and Jaeyeon for their excellent programming contributions to the Finch codebase.
+  Intel and NSF PPoSS Grant CCF-2217064; DARPA PROWESS Award HR0011-23-C-0101; NSF SHF Grant CCF-2107244
 \end{acks}
+
+\clearpage
 
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{FinchOOPSLAWillow.bib, FinchOOPSLAOverleaf.bib}
+\bibliography{FinchOOPSLAWillow.bib}
 
@@
 %% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Frequently Requested Finch kernels}
+%\appendix
+%\include{sections/appendix}
 

### 2024-04-24 sections/compiler.tex
fixes

@@
     Other languages, such as Halide, perform bounds inference where
-    known bounds are symbolically propagated to fill in unknown bounds, often from output/input sizes to intermediates via some approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013, grosser2012polly}.
+    known bounds are symbolically propagated to fill in unknown bounds, often from output/input sizes to intermediates via some approximation such as interval analysis or polyhedral methods~\cite{ragan-kelley_halide_2013,grosser_pollyperforming_2012}.
     %

### 2024-04-24 sections/eval/graphs.tex
fixes

@@
 
-Direction-optimization~\cite{beamer2012direction} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
+Direction-optimization~\cite{beamer_direction-optimizing_2012} is crucial for achieving high BFS performance in such scenarios, switching between push and pull traversals to efficiently explore graphs.
 %
@@
 
-Figure~\ref{fig:graph_result} compares performance to Graphs.jl, a Julia library, and the LAGraph Library, which implements graph algorithms with sparse linear algebra using GraphBLAS~\cite{mattson2019lagraph}.
+Figure~\ref{fig:graph_result} compares performance to Graphs.jl, a Julia library, and the LAGraph Library, which implements graph algorithms with sparse linear algebra using GraphBLAS~\cite{mattson_lagraph_2019}.
 %

### 2024-04-24 sections/intro.tex
fixes

@@
 Our world is full of structured data.
-%
-Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankarpetsc, bell2007lessons, mcauley2013hidden, balay2020petsc}.
+
+Sparse arrays (which store only nonzero elements) describe networks, databases, and simulations~\cite{abhyankar_petsc_nodate,bell_lessons_2007,mcauley_hidden_2013,balay_petsc_2020}.
 %
 Run-length encoding describes images and masks, geometry, and
-databases (such as a list of transactions with the date field all the same)~\cite{shi2020column,golomb1966run}.
+databases (such as a list of transactions with the date field all the same)~\cite{shi_column_2020,golomb_run-length_1966}.
 %
-Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{ded, saad2003iterative, o2009scientific}.
+Symmetry, bands, padding, and blocks arise due to modeling choices in scientific computing (e.g., higher order FEMs) as well as in intermediate structures in many linear solvers (e.g., GMRES)~\cite{burns_dedalus_2020,saad_iterative_2003,oleary_scientific_2009}.
 %
-In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao2022monarch}.
+In the context of machine learning, combinations of sparse and blocked matrices are increasingly under consideration~\cite{dao_monarch_2022}.
 %
@@
 For example, a convolution with a filter can be expressed as a matrix multiplication
-with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze2017efficient}.
+with the Toeplitz matrix of all the circular shifts of the filter~\cite{sze_efficient_2017}.
 
@@
 Hand-written solutions are collected in diverse libraries like
-MKL, OpenCV, LAPACK or SciPy~\cite{ bradski2000opencv, anderson1999lapack, virtanen2020scipy, psarras2022linear}. 
+MKL, OpenCV, LAPACK or SciPy~\cite{bradski_opencv_2000,anderson_lapack_1999,virtanen_scipy_2020,psarras_linear_2022}. 
 %
@@
 operations over various semi-rings always lack support for other features, such
-as tensors, fused outputs, or runs of repeated values~\cite{bulucc2017design, mattson2019lagraph}.
+as tensors, fused outputs, or runs of repeated values~\cite{buluc_design_2017,mattson_lagraph_2019}.
 %
@@
 compilers support an enormous variety of program constructs like early break and
-multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser2012polly}.  
+multiple left hand sides, they only support dense arrays~\cite{ragan-kelley_halide_2013,grosser_pollyperforming_2012}.  
 %
 Special-purpose
-compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani2023compiling}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
+compilers like TACO~\cite{kjolstad_tensor_2019}, Taichi~\cite{hu_taichi_2019}, StructTensor~\cite{ghorbani_compiling_2023}, or CoRa~\cite{fegade_cora_2022} which support a select subset of structured data
 structures (only sparse, or only ragged arrays) must compromise by greatly
@@
 %
-The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser2012polly}.
+The polyhedral model allows various compilers to represent dense computations on affine regions~\cite{grosser_pollyperforming_2012}.
 %
@@
 %
-For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens2022autoscheduling, zhang2021gamma}.
+For example, the asymptotic performance of sparse matrix multiplication can be impacted by the distribution of nonzeros, the sparse format, and the loop order~\cite{ahrens_autoscheduling_2022,zhang_gamma_2021}.
 %For example, sparse kernels don't need to compute on zeros, but this means that the precise input nonzero patterns act as computational filters, affecting the runtime as they interact with each other and the implementation.
@@
 %
-Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou2018format,  hu_taichi_2019, strout2018sparse, bernstein2016ebb} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level.
+Although many systems (TACO, Taichi, SPF, Ebb)~\cite{chou_format_2018,  hu_taichi_2019, strout_sparse_2018, bernstein_ebb_2016} feature a flexible structure description, our level abstraction is more capable and extensible because it uses Looplets \cite{ahrens_looplets_2023} to express the structure of each level.
 %
@@
 %
-For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{won2024continuous}.
+For example, Finch has been extended to support real-valued array indices with continuous arrays \cite{anonymous_authors_continuous_2024}.
+Finch is also used as a compiler backend for the Python PyData/Sparse library \cite{abbasi_plans_2023}.
 %

### 2024-04-24 sections/related.tex
fixes

--- 2024-04-06 sections/related.tex
@@
 %
-Perhaps the most well-known example is Numpy~\cite{harris_array_2020},
-and a classic example is the BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{Anderson1999}. 
+Perhaps the most well-known example is NumPy~\cite{harris_array_2020},
+and a classic example is the BLAS, though several BLAS routines are specialized to symmetric, hermitian, and triangular matrices~\cite{anderson_lapack_1999}. 
 %
-Many research projects have advanced on BLAS, such as BatchedBlas and BLIS~\cite{dongarra2017design, van2015blis}.
+Many research projects have advanced on BLAS, such as BatchedBlas and BLIS~\cite{dongarra_design_2017,van_zee_blis:_2015}.
 
@@
 %
-Examples include Scipy~\cite{virtanen2020scipy}, PETSc~\cite{abhyankarpetsc}, Armadillo~\cite{Rumengan2021}, OSKI~\cite{vuduc2005oski}, Cyclops~\cite{solomonik2013cyclops}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{eigenweb}.
+Examples include SciPy~\cite{virtanen_scipy_2020}, PETSc~\cite{abhyankar_petsc_nodate}, Armadillo~\cite{rumengan_pyarmadillo_2021}, OSKI~\cite{vuduc_oski:_2005}, Cyclops~\cite{solomonik_cyclops_2013}, MKL~\cite{noauthor_developer_2024}, and Eigen~\cite{guennebaud_eigen_2010}.
 %
-There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith2015splatt} (MTTKRP on CSF).
+There are even libraries for very specific kernels and format combinations, such as SPLATT~\cite{smith_splatt:_2015} (MTTKRP on CSF).
 %
@@
 %
-The GraphBLAS~\cite{kepner2016mathematical} supports primitive semiring operations (operations beyond $(+, *)$, such as $(min, +)$ multiplication) which can be composed to enable graph algorithms, some of which are collected in LAGraph~\cite{mattson2019lagraph}.
+The GraphBLAS~\cite{kepner_mathematical_2016} supports primitive semiring operations (operations beyond $(+, *)$, such as $(min, +)$ multiplication) which can be composed to enable graph algorithms, some of which are collected in LAGraph~\cite{mattson_lagraph_2019}.
 %
-Similarly, the Mapreduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
-algorithms in the GBASE library\cite{kang2011gbase}.
+Similarly, the MapReduce and Hadoop platforms support operations on indexed collections~\cite{dean_mapreduce_2008}, and have been used to support graph
+algorithms in the GBASE library\cite{kang_gbase_2011}.
 %
-Several machine learning frameworks support some sparse arrays and operations, most notably TorchSparse\cite{tang2022torchsparse, tang2023torchsparse++}.
+Several machine learning frameworks support some sparse arrays and operations, most notably TorchSparse\cite{tang_torchsparse_2022,tang_torchsparse_2023}.
 %
@@
 %
-Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendent such as TVM~\cite{chen2018tvm}, Exo~\cite{ikarashi2022exocompilation}, Elevate~\cite{hagedorn2020achieving}, and ATL~\cite{liu_verified_2022}.
+Perhaps the most well known example is Halide~\cite{ragan-kelley_halide_2013} and its various descendant such as TVM~\cite{chen_tvm_2018}, Exo~\cite{ikarashi_exocompilation_2022}, Elevate~\cite{hagedorn_achieving_2020}, and ATL~\cite{liu_verified_2022}.
 %
@@
 %
-Several polyhedral languages, such as Polly~\cite{grosser2012polly}, Tiramisu~\cite{baghdadi2019tiramisu}, CHiLL~\cite{chen2008framework}, Pluto~\cite{bondhugula2008pluto}, and AlphaZ~\cite{yuki2012alphaz} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
+Several polyhedral languages, such as Polly~\cite{grosser_pollyperforming_2012}, Tiramisu~\cite{baghdadi_tiramisu_2019}, CHiLL~\cite{chen_framework_2008}, Pluto~\cite{bondhugula_pluto_2008}, and AlphaZ~\cite{yuki_alphaz_2012} offer similar capabilities in terms of control flow though they often support more irregular regions that the polyhedral framework supports.
 %
-These are based on ISL~\cite{verdoolaege2010isl}.
+These are based on ISL~\cite{verdoolaege_isl_2010}.
 %
@@
 %
-The TACO compiler originally supported just plain einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake2020sparse}, and convolution~\cite{won2023unified}, all as separate extensions.
+The TACO compiler originally supported just plain Einsum computations~\cite{kjolstad_tensor_2017}, but has been extended several times to support (single dimensional) local tensors \cite{kjolstad_tensor_2019}, imperfectly nested loops \cite{dias_sparselnr_2022}, breaks via semi-rings~\cite{henry_compilation_2021}, windowing and tiling \cite{senanayake_sparse_2020}, and convolution~\cite{won_unified_2023}, and compilation in MLIR \cite{bik_compiler_2022}, all as separate extensions.
 %
-Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou2018format}, and tree like structures~\cite{chou2022compilation}, as separate extensions.
+Similarly, TACO originally support just dense and CSF like N dimensional structures, but was extended independently to support COO like structures~\cite{chou_format_2018}, and tree like structures~\cite{chou_compilation_2022}, as separate extensions.
 SparseTIR is a similar system supporting combined
-sparse formats (including block structures) \cite{ye_sparsetir_2022}.
+sparse formats (including block structures) \cite{ye_sparsetir_2023}.
 %
-The SDQL language offers a similar level of control flow~\cite{shaikhha2022functional}, but only on sparse hash tables.
+The SDQL language offers a similar level of control flow~\cite{shaikhha_functional_2022}, but only on sparse hash tables.
 %
-Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich2023optimizing} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani2023compiling}.
+Similarly, SDQL has been extended with a system that allows one to specify formats as queries on a set of base storage types~\cite{schleich_optimizing_2023} and separately by another system that describes static symmetries and other structures as predicates~\cite{ghorbani_compiling_2023}.
 %
@@
 %
-The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout2018sparse} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to einsums~\cite{zhao2022polyhedral}.
+The sparse polyhedral framework builds on CHiLL for the purpose of generating inspector/executor optimizations~\cite{strout_sparse_2018} though the branch of this work that specifies sparse formats separately from the computation (otherwise they are inlined into the computation manually) seems to apply mainly to Einsums~\cite{zhao_polyhedral_2022}.
 %
-Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar1997relational, date1989guide}.
+Second to last, SQL's classical physical/logical distinction is the classic program/format distinction, and SQL supports a huge variety of control flow constructs~\cite{kotlyar_relational_1997,date_guide_1989}.
 %
-However, many SQL or dataframe systems rely on BTrees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz2012vectorwise}, LaraDB~\cite{hutchison2017laradb}, GMAP~\cite{tsatalos1996gmap}, or SciDB~\cite{stonebraker2013scidb} building physical layouts with other constructs based in array programming.
+However, many SQL or dataframe systems rely on b-trees, columnar, or hash tables, with only a few systems, such as Vectorwise~\cite{boncz_vectorwise_2012}, LaraDB~\cite{hutchison_laradb_2017}, GMAP~\cite{tsatalos_gmap_1996}, or SciDB~\cite{stonebraker_scidb_2013} building physical layouts with other constructs based in array programming.
 %
-However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann2021array,luo2018scalable}.
+However, array based databases are a new focus given the rise of mixed ML/DB pipelines~\cite{baumann_array_2021,luo_scalable_2018}.
 %
-Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti2018spiral,franchetti2009operator}.
+Lastly, SPIRAL focuses on recursively defined datastructures and recursively define linear algebra, and can therefore express a structure and computation that none of the systems mentioned above can: a Cooley–Tukey FFT ~\cite{franchetti_spiral_2018,franchetti_operator_2009}.
 %
@@
 Sparse compilers have been extended to many architectures.
-An extension of TACO supports GPU~\cite{senanayake2020sparse}, Cyclops~\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal~\cite{yadav_spdistal_2022-1} support distributed memory, and the Sparse Abstract Machine~\cite{hsu_sparse_2022} supports custom hardware.
+An extension of TACO supports GPU~\cite{senanayake_sparse_2020}, Cyclops~\cite{solomonik_cyclops_2013,solomonik_sparse_2015} and SPDistal~\cite{yadav_spdistal_2022} support distributed memory, and the Sparse Abstract Machine~\cite{hsu_sparse_2023} supports custom hardware.
 We believe that supporting control flow is the first step towards architectural support beyond unstructured sparsity.

### 2024-04-24 sections/tensors.tex
fixes

@@
         \item \textbf{Singular} When we have only one non-fill region in sparse data,
-        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani2023compiling, ahrens_looplets_2023}.
+        we can avoid a loop entirely and reduce the complexity of iteration \cite{ghorbani_compiling_2023, ahrens_looplets_2023}.
     \end{itemize}

### 2024-04-25 cleanup.sh
fixes

@@ -1,0 +1,12 @@
+rm -f *.bak
+rm -f .*.bak
+rm -f */*.bak
+rm -f */*/*.bak
+rm -f *.pdf
+rm -f *.aux
+rm -f *.blg
+rm -f *.fdb_latexmk
+rm -f *.log
+rm -f *.synctex.gz
+rm -f *.fls
+rm -f *.out

### 2024-04-25 sections/appendix.tex
no parens

--- 2024-04-04 sections/appendix.tex
@@
 \end{minted}
+
+\section{Mask Images}
+We interpreted the following images from ``Digital Image Processing'' \cite{gonzalez_digital_2006} as masks:
+
+\textbf{FigP1012.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP1012.png}}
+
+\textbf{Fig1008\_a\_\_step edge\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008\_a\_\_step edge\_.png}}
+
+\textbf{FigP0905\_d\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_d\_.png}}
+
+\textbf{FigP0528\_c\_\_doughnut\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528\_c\_\_doughnut\_.png}}
+
+\textbf{FigP0616\_b\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616\_b\_.png}}
+
+\textbf{Fig0114\_c\_\_bottles\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0114\_c\_\_bottles\_.png}}
+
+\textbf{FigP0616\_c\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616\_c\_.png}}
+
+\textbf{FigP0433\_b\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0433\_b\_.png}}
+
+\textbf{Figp0917.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Figp0917.png}}
+
+\textbf{FigP0905\_b\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_b\_.png}}
+
+\textbf{Fig1059\_c\_\_NegADI\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059\_c\_\_NegADI\_.png}}
+
+\textbf{Fig1111\_a\_\_triangle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1111\_a\_\_triangle\_.png}}
+
+\textbf{Fig1111\_b\_\_square\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1111\_b\_\_square\_.png}}
+
+\textbf{FigP0905\_top\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_top\_.png}}
+
+\textbf{FigP1110.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP1110.png}}
+
+\textbf{Fig0533\_a\_\_circle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0533\_a\_\_circle\_.png}}
+
+\textbf{FigP0917\_noisy\_rectangle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0917\_noisy\_rectangle\_.png}}
+
+\textbf{Fig0230\_b\_\_dental\_xray\_mask\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0230\_b\_\_dental\_xray\_mask\_.png}}
+
+\textbf{FigP0528\_b\_\_two\_dots\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528\_b\_\_two\_dots\_.png}}
+
+\textbf{Fig1059\_a\_\_AbsADI\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059\_a\_\_AbsADI\_.png}}
+
+\textbf{Fig1059\_b\_\_PosADI\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1059\_b\_\_PosADI\_.png}}
+
+\textbf{Fig0539\_c\_\_shepp-logan\_phantom\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0539\_c\_\_shepp-logan\_phantom\_.png}}
+
+\textbf{FigP0905\_c\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_c\_.png}}
+
+\textbf{Fig1043\_a\_\_yeast\_USC\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1043\_a\_\_yeast\_USC\_.png}}
+
+\textbf{FigP0905\_U\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_U\_.png}}
+
+\textbf{Fig0524\_b\_\_blurred-impulse\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0524\_b\_\_blurred-impulse\_.png}}
+
+\textbf{Fig0424\_a\_\_rectangle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0424\_a\_\_rectangle\_.png}}
+
+\textbf{Fig1008\_c\_\_roof\_edge\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008\_c\_\_roof\_edge\_.png}}
+
+\textbf{Fig0539\_a\_\_vertical\_rectangle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0539\_a\_\_vertical\_rectangle\_.png}}
+
+\textbf{FigP0905\_a\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0905\_a\_.png}}
+
+\textbf{FigP0433\_a\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0433\_a\_.png}}
+
+\textbf{Fig0.15\_a\_\_translated\_rectangle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0425\_a\_\_translated\_rectangle\_.png}}
+
+\textbf{FigP0918\_c\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0918\_c\_.png}}
+
+\textbf{Fig0524\_a\_\_impulse\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0524\_a\_\_impulse\_.png}}
+
+\textbf{Fig0236\_a\_\_letter\_T\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0236\_a\_\_letter\_T\_.png}}
+
+\textbf{Fig0503 \_original\_pattern\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0503 \_original\_pattern\_.png}}
+
+\textbf{FigP0501.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0501.png}}
+
+\textbf{Fig1218\_airplanes\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1218\_airplanes\_.png}}
+
+\textbf{Fig0534\_a\_\_ellipse\_and\_circle\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0534\_a\_\_ellipse\_and\_circle\_.png}}
+
+\textbf{FigP0616\_a\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0616\_a\_.png}}
+
+\textbf{FigP0918\_b\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0918\_b\_.png}}
+
+\textbf{FigP0528\_c\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528\_c\_.png}}
+
+\textbf{FigP0528\_a\_\_single\_dot\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/FigP0528\_a\_\_single\_dot\_.png}}

### 2024-04-25 figures/dip3e_masks/foo.py

@@ -1,0 +1,19 @@
+import os
+
+def replace_parens_in_filenames(directory):
+    # Iterate over all files in the specified directory
+    for filename in os.listdir(directory):
+        # Check if the filename contains parentheses
+        if '(' in filename or ')' in filename:
+            # Replace parentheses with underscores
+            new_filename = filename.replace('(', '_').replace(')', '_')
+            # Get the full path of the original and new filenames
+            original_path = os.path.join(directory, filename)
+            new_path = os.path.join(directory, new_filename)
+            # Rename the file
+            os.rename(original_path, new_path)
+            print(f"Renamed '{filename}' to '{new_filename}'")
+
+# Specify the directory to process
+directory_path = '.'
+replace_parens_in_filenames(directory_path)

### 2024-04-25 figures/dip3e_masks/foo.py
cool

@@ -1,19 +1,0 @@
-import os
-
-def replace_parens_in_filenames(directory):
-    # Iterate over all files in the specified directory
-    for filename in os.listdir(directory):
-        # Check if the filename contains parentheses
-        if '(' in filename or ')' in filename:
-            # Replace parentheses with underscores
-            new_filename = filename.replace('(', '_').replace(')', '_')
-            # Get the full path of the original and new filenames
-            original_path = os.path.join(directory, filename)
-            new_path = os.path.join(directory, new_filename)
-            # Rename the file
-            os.rename(original_path, new_path)
-            print(f"Renamed '{filename}' to '{new_filename}'")
-
-# Specify the directory to process
-directory_path = '.'
-replace_parens_in_filenames(directory_path)

### 2024-04-25 sections/appendix.tex
fix

@@
 
-\textbf{Fig1008\_a\_\_step edge\_.png}
-
-\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008\_a\_\_step edge\_.png}}
+\textbf{Fig1008\_a\_\_step\_edge\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig1008\_a\_\_step\_edge\_.png}}
 
@@
 
-\textbf{Fig0503 \_original\_pattern\_.png}
-
-\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0503 \_original\_pattern\_.png}}
+\textbf{Fig0503\_\_original\_pattern\_.png}
+
+\includegraphics[scale=0.20]{{./dip3e\_masks/Fig0503\_\_original\_pattern\_.png}}
 

### 2024-04-29 sections/eval/api.tex

@@
 Across all three of these kernels, we see that the high level interface for Finch provides a major improvement over DuckDB, ranging from $1.2x-28x$.
-For triangle counting and SDDMM, this improvement stems from Finch's efficient intersection of the nonzero indices as opposed to DuckDB's use of binary join plans\footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang2023free}.}.
+For triangle counting and SDDMM, this improvement stems from Finch's efficient intersection of the nonzero indices as opposed to DuckDB's use of binary join plans\footnote{This matches with findings in the database literature showing that worst-case optimal joins (which are very similar to our kernel execution) are more efficient than binary joins for these queries \cite{wang_free_2023}.}.
 For element-wise operations, this improvement stems from Finch's better handling of expressions which combine index intersection and union and its compressed data representation.

### 2024-04-29 sections/eval/spgemm.tex

@@
 %
-There are three main approaches to SpGEMM \cite[Section 2.2]{zhang2021gamma}.
+There are three main approaches to SpGEMM \cite[Section 2.2]{zhang_gamma_2021}.
 %
@@
 %
-Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens2022autoscheduling}.
+Inner-products is known to be asymptotically less efficient than the others, as we must do a merge operation to compute each of the $O(MN)$ entries in the output \cite{ahrens_autoscheduling_2022}.
 %

### 2024-06-04 response.md
example response

# Response

We thank the reviewers for their thorough feedback, as it will surely improve the work. We address high-level comments first, and more detailed comments in an explicit appendix.

## Relationship to Looplets

We began by clarifying the relationship between Finch and Looplets. Finch is the only framework to support such a broad range of single and multidimensional structured data and control flow. Looplets alone could not achieve these results without Finch. While Looplets provide a powerful mechanism to simplify structured loops, our paper shows how to make this functionality practical. Finch uses Looplets as a symbolic loop simplification engine. The precise choice and implementation of level structures, the lifecycle interface between levels and looplets, and the canonicalization of fancy indexing and masking operations all serve to recombine and utilize looplets to achieve efficient computation over structured tensors.

## Improving the Explanation of the Compiler
> **Reviewer B**: “it appears that all the magic has been squirreled away in the unfurl function”, “the concordization step described in Section 5.4 looks very expensive”

> **Reviewer C**: “some important parts, like the structure implementations, do not have detailed explanation.” "Does concordization impact performance? I ask because adding another loop level even with if-conditions might cause some slowdown."

Finch relies on multiple detailed parts that all interact. While each part is explained in detail, examples would help relate the parts to the whole. We take Reviewer B's advice and propose a new section (a draft of which is given in our appendix) which relates the disparate parts through a complete example of lowering.

To summarize here: The `unfurl` function returns a Looplet nest describing the hierarchical structure of the outermost dimension of the tensor. Looplets were chosen for this purpose as a symbolic engine to ensure certain simplifications take place, but another symbolic system could have been used. We chose Looplets because they reliably process structured iterators. Here is an simplified definition of `unfurl` for a SparseList vector. It is up to the implementer of the structured level to ensure that `unfurl` returns a correct Looplet nest.

```
unfurl(SubFiber(A.lvl::SparseList, pos)) = Thunk(
    # First, we initialize our position iterator
    preamble = (q = pos),
    # The stepper is analogous to a for-loop
    Stepper(
        # This fast forwards the loop to a specific iteration
        seek = (i) -> (q = binarysearch(A.lvl.idx, i)),
        # This tells us the boundary of the loop
        stop = A.lvl.idx[q],
        body = Spike(
            body = 0,
            tail = A.lvl.val[q]
        ),
        # This advances the loop to the next iteration
        next = (q += 1)
    )
)
```

In our new section or an appendix, we plan to add an enhanced description of each mask and structured level format in terms of the looplet nests returned by `unfurl`, relating looplets to the concrete data fields of the tensor such as (`idx`, `ofs`, and `pos`), and addressing the concerns of Reviewer C.

As an example, Looplets allow us to make concordization inexpensive because the concordized loop is lowered with a mask. For example, consider the following program to calculate `A[5]` when `A` is sparse:
```
for i = _
    if i == 5
        s[] = A[i]
    end
end
```
This would lower to the following:
```
q = binarysearch(A.lvl.idx, 5)
i = A.lvl.idx[q]
if i == 5
    s[] = A.lvl.val[q]
end
```
The lowered code runs in logarithmic time. We achieve the desired code because of (1) how SparseList communicates its structure in looplet nests via unfurl, (2) how wrapperization converts `i == 5` to a symbolic `DiagMask[i, 5]` tensor, also expressed in looplets, and (3) how the two looplet nests reliably combine to fast forward the sparse iterator and eliminate the `if false` branches.

This is an example of the productivity of Finch, as we were able to implement efficient scalar reads and writes for all combinations of formats by simply combining the `unfurl` definitions with a structured tensor.

Our language allows fairly complex programs to be expressed on fairly complex structures yet still have these optimizations happen in a predictable and reliable way. Our normal form, life cycles, concordization, and wrapperization are key ideas that allow us to boil our Finch language down to looplets so that the right looplet optimizations fire.

> **Reviewer B**: “the use of wrappers ... appears to require materializing potentially expensive temporaries (e.g. Toeplitz matrices).”

None of the wrapper arrays are materialized, they are all lazy. Wrappers instead modify the behavior of `unfurl`, dimensionalization, and other looplet properties such as `stop`. To give an example in terms of the simpler `OffsetArray(1)`,

```
A = Tensor(Dense(Element(0.0)))
B = Tensor(Dense(Element(0.0)))
@finch begin
    for i = 2:4
        B[i] = A[i + 1]
    end
end
```

Would lower to

```
for i = 2:4
    B.lvl.val[i] += A.lvl.val[i + 1]
end
```

## Clarifying the Evaluation

Reviewer A asks that we compare to another SpMV and SpGEMM implementation (in addition to TACO and Julia), we agree that this would strengthen the work and plan to include a comparison against Eigen, SparseTIR, or Cora on the datasets where it is relevant. Other related work is either specialized to different architectures (such as Taichi to GPUs) or is not open source (such as MKL or STUR).

Reviewer A also requests geo-mean speedups for our SpMV and SpGEMM implementations, which were _ and _, respectively. We plan to include geo-mean speedups.
The SpGEMM dataset was chosen to be the matrices in  [101]. We separated the matrices into small matrices and large matrices. Because of asymptotic slowdowns, we couldn’t run TACO outer products or inner products on the large matrices. However, since the matrices are organized by size in the small matrix plot, we can already see the beginnings of a trend where the performance of TACO outer products and inner products perform worse as the matrix size increases. We will include a dedicated scaling plot to make these relationships more clear.

> **Reviewer B**: “In Section 6.1, the text says that the program in Figure 14 is restricting its attention to the canonical triangle using masks, but it looks like all three flavors are looping over the bounds of i and j in their entirety?”

> **Reviewer C**: "Please quantify and comment upon the overheads due to extra book keeping code generated by the compiler inside loop nests."

To Reviewer B, this is a typo in the text. For symmetric SpMV, we found it is faster to pre-compute the strict triangle of the argument and re-use it within the kernel, to simplify the logic. It is of course possible to add an `if i < j` statement and compute the triangle on the fly without accessing the whole matrix, but the runtime of SpMV is impacted because this changes the exit condition of that loop.

In general, we have found Finch is quite adept at lifting conditionals to the highest level of the loop nest, but that there is a tradeoff between the branching between specialized routines and the benefits of such specialization. Specialization is better in cases where the specialized routine is much faster and much more frequent (such as the zero region of a sparse tensor, which becomes a no-op). In practice, this has led us to sometimes ask Finch to de-specialize certain conditional expressions, and is another example of how Finch can widen the design space for structured operators. We will include a discussion of this tradeoff in the paper.

> **Reviewer A**: Which features of your language are exercised by which case study? Can you indicate which features your baseline is lacking to reach the performance you are demonstrating.

> **Reviewer C**:  “I would have liked to see quantifying data like how many times you had to use Sequentially constructed levels over Nonsequentially Constructed Levels”

Our SpMV evaluation demonstrates the performance of the SparseList, SparseVBL, SparseBand, and Pattern level formats, as well as conditionals and multiple outputs in the symmetric kernel

Our SpMM evaluation shows off the importance of sparse intermediates (SparseBytemap and SparseHash are non-sequential) and the flexibility of our language for scheduling, doing all three matrix multiplication variants.

Our Graph applications demonstrate conditionals, complex loop structures, early breaks, and user-defined functions.

Our Image Morphology benchmarks demonstrate the flexibility of the language, as well as affine indexing and `sparseRLE` and `pattern` formats, as well as user-defined functions.

The Array API benchmark shows that the language is reliable and flexible enough to support autoscheduling. It also involves user-defined functions and `SparseHash, non-sequential` format.

> **Reviewer C**: I would like to know why the formats provide the speedup: Is it just due to looplets iterators at each level? Is it due to constructed levels and if so, how do they provide benefit over other constructed levels and why over TACO’s representation?

# Appendix

## Lowering Example Section

Though Finch programs look as if they are written for dense loops, Finch specializes the code during lowering so that only the necessary elements of structure need to be processed. We illustrate the lowering of a program which sums the upper triangle of an `m` by `n` matrix, `A`:

```
A = Tensor(Dense(SparseList(Element(0.0))), m, n)
s = Tensor(Element(0.0))
@finch begin
    s .= 0.0
    for j = _
        for i = _
            if i <= j
                s[] += A[i, j]
            end
        end
    end
end
```

To begin, the wrapperization pass replaces `i <= j` with `UpTriMask[i, j]`. The dimensionalization pass assigns `i`, and `j` the dimensions `m` and `n`, respectively. All accesses are already concordant, and after adding lifecycle statements we have:

```
@finch begin
    @declare(s, 0.0)
    for j = 1:n
        for i = 1:m
            if UpTriMask[i, j]
                s[] += A[i, j]
            end
        end
    end
    @freeze(s)
end
```

We then begin lowering the program. The `@declare` statement on `y` results in a simple initialization of the `val` array, `s.lvl.val[1] = 0.0`. To process the `j` loop, we unfurl both tensors which access `j`. `unfurl(UpTriMask)` returns `Lookup(body = (j) -> UpTriMaskCol(j))`, and `unfurl(A::DenseLevel)` returns `Lookup(body = (j) -> SubFiber(A.lvl.lvl, j))`. Thus, we are left with:
```
s.lvl.val[1] = 0.0
@finch begin
    for j = 1:n
        for i = 1:m
               if (Lookup(...)[j])[i]
                 s[] += (Lookup(...)[j])[i]
            end
        end
    end
    @freeze(s)
end
```

Since we can lower Lookup looplets with a for-loop, our loop becomes:
```
s.lvl.val[1] = 0.0
for j = 1:n
    @finch begin
        for i = 1:m
            if UpTriMaskCol(j)[i]
                s[] += SubFiber(A.lvl.lvl, j)[i]
            end
        end
    end
end
@finch @freeze(s)
```

Next, we process the `i` loop. We Unfurl the tensors, replacing them with the follwing nests of looplets:

```
unfurl(UpTriMaskCol(j)) = Sequence(
    Phase(
        stop = j,
        Run(true)
    ),
    Phase(
        Run(False)
    )
)
```

```
unfurl(SubFiber(A.lvl.lvl::SparseListLevel, j)) = Thunk(
    # First, we initialize our position iterator
    preamble = (q = 1),
    # The stepper is analogous to a for-loop
    Stepper(
        # This fast forwards the loop to a specific iteration
        seek = (i) -> (q = binarysearch(A.lvl.lvl.idx, i)),
        # This tells us the boundary of the loop
        stop = A.lvl.lvl.idx[q],
        body = Spike(
            body = 0,
            tail = A.lvl.lvl.val[q]
        ),
        # This advances the loop to the next iteration
        next = (q += 1)
    )
)
```

Substituting the results of `unfurl` into the expression, we are left with:
```
s.lvl.val[1] = 0.0
for j = 1:n
    @finch begin
        for i = 1:m
            if Sequence(...)[i]
                s[] += Thunk(...)[i]
            end
        end
    end
end
@finch @freeze(s)
```

The highest lowering priority is for the Thunk, then we lower the Sequence looplet, leaving us with:
```
s.lvl.val[1] = 0.0
for j = 1:n
    q = 1
    @finch for i = 1:j
        if true
            s[] += Stepper(...)[i]
        end
    end
    @finch for i = j+1:m
        if false
            s[] += Stepper(...)[i]
        end
    end
end
@finch @freeze(s)
```

We can simplify the latter phase to a no-op, and lower the stepper in the first phase to a while loop. The Spike inside the stepper lowers similarly to a sequence, leaving us with

```
s.lvl.val[1] = 0.0
for j = 1:n
    q = 1
    i_loop = 1
    while i_loop < j
        i_loop_2 = A.lvl.lvl.idx[q]
        @finch for i = i_loop:min(i_loop_2, j)
                s[] += 0
        end
        i = i_loop_2
        @finch if i < j
                s[] += A.lvl.lvl.val[q]
        end
        q += 1
    end
end
@finch @freeze(s)
```
Finally, we simplify the first case and lower the final freeze (which is a no-op), obtaining:

```
s.lvl.val[1] = 0.0
for j = 1:n
    q = 1
    i_loop = 1
    while i_loop < j
        i_loop_2 = A.lvl.lvl.idx[q]
            i = i_loop_2
        if i < j
            s[] += A.lvl.lvl.val[q]
        end
        q += 1
    end
end
```

The final program accesses only the upper triangle of A, though the original code looks as though it loops over all `i` and `j`.

### 2024-06-05 response.md
Update on Overleaf.

@@
 The precise choice and implementation of level structures, the lifecycle interface between levels and looplets, and the canonicalization of fancy indexing and masking operations all serve to recombine and utilize looplets to achieve efficient computation over structured tensors.
+We will illustrate this further in the next section.
 
@@
 To summarize here: The `unfurl` function returns a Looplet nest describing the hierarchical structure of the outermost dimension of the tensor.
-Looplets were chosen for this purpose as a symbolic engine to ensure certain simplifications take place, but another symbolic system could have been used.
-We chose Looplets because they reliably process structured iterators.
+Looplets were chosen for this purpose as a symbolic engine to ensure certain simplifications take place, but another symbolic system could have been used (e.g.
+the sparse polyhedral framework attempts similar optimizations (cite that one Mary Hall paper from a year ago)).
+We chose Looplets because they reliably process structured iterators, predictably eliminating zero regions, using faster lookups when avaliable, and utilizing repeated work.
 Here is an simplified definition of `unfurl` for a SparseList vector.
@@
 
+We will also expand the section on wrapperization and masks to show the interface and clarify how they are allowed to modify Looplets.
+Additionally, this further shows the productivity in Finch: by adding more wrappers that reinterpret looplet nests in different ways, we can allow users to experiment with different low level strategies: for an index expression `i+j`, the compiler currently scans through i after looking up j, but other strategies are possible.
+And this is in addition to just mixing one strategy for an index expression with many different structures.
+
 ## Clarifying the Evaluation
@@
 We will include a dedicated scaling plot to make these relationships more clear.
+Moreover, we used a matrix test set designed to reveal this asymtoptic difference, as seen in (Cite Joel's paper).
 

### 2024-06-05 response.md
cool

--- 2024-06-04 response.md
@@
 Finch relies on multiple detailed parts that all interact.
-While each part is explained in detail, examples would help relate the parts to the whole.
-We take Reviewer B's advice and propose a new section (a draft of which is given in our appendix) which relates the disparate parts through a complete example of lowering.
+While each part is explained in detail, we agree with Reviewer B that an example would help relate the parts to the whole.
+We propose a new section which relates the disparate parts through a complete example of lowering.
+A complete draft is given in the appendix, and we summarize here:
 
-To summarize here: The `unfurl` function returns a Looplet nest describing the hierarchical structure of the outermost dimension of the tensor.
-Looplets were chosen for this purpose as a symbolic engine to ensure certain simplifications take place, but another symbolic system could have been used.
-We chose Looplets because they reliably process structured iterators.
+The `unfurl` function returns a Looplet nest describing the hierarchical structure of the outermost dimension of the tensor.
+Looplets were chosen for this purpose as a symbolic engine to ensure certain simplifications take place, but another symbolic system could have been used (e.g.
+polyhedral [102] or e-graph search [77]).
+We chose Looplets because they reliably process structured iterators, predictably eliminating zero regions, using faster lookups when avaliable, and utilizing repeated work.
 Here is an simplified definition of `unfurl` for a SparseList vector.
-It is up to the implementer of the structured level to ensure that `unfurl` returns a correct Looplet nest.
+It is the implementer's responsibility to ensure that `unfurl` returns a correct Looplet nest.
 
@@
 The lowered code runs in logarithmic time.
-We achieve the desired code because of (1) how SparseList communicates its structure in looplet nests via unfurl, (2) how wrapperization converts `i == 5` to a symbolic `DiagMask[i, 5]` tensor, also expressed in looplets, and (3) how the two looplet nests reliably combine to fast forward the sparse iterator and eliminate the `if false` branches.
+We achieve the desired code because of (1) how SparseList communicates its structure in looplet nests via unfurl, (2) how wrapperization converts `i == 5` to a symbolic `DiagMask[i, 5]`, which then converts to looplets, and (3) how the two looplet nests combine to fast forward the sparse iterator and eliminate the `if false` branches.
 
@@
 
-Our language allows fairly complex programs to be expressed on fairly complex structures yet still have these optimizations happen in a predictable and reliable way.
+Our language allows fairly complex programs to be expressed on fairly complex structures becayse these optimizations happen in a predictable and reliable way.
 Our normal form, life cycles, concordization, and wrapperization are key ideas that allow us to boil our Finch language down to looplets so that the right looplet optimizations fire.
@@
 
-None of the wrapper arrays are materialized, they are all lazy.
+We appreciate this concern, and assure Reviewer B that none of the wrapper arrays are materialized, they are all lazy.
 Wrappers instead modify the behavior of `unfurl`, dimensionalization, and other looplet properties such as `stop`.
+We plan to expand the section on wrapperization and masks to show the interface and clarify how they are allowed to modify Looplets.
 To give an example in terms of the simpler `OffsetArray(1)`,
@@
 
-Reviewer A also requests geo-mean speedups for our SpMV and SpGEMM implementations, which were _ and _, respectively. We plan to include geo-mean speedups.
-The SpGEMM dataset was chosen to be the matrices in  [101].
+As suggested by Reviewer A, we also plan to report our geo-mean speedups, which were:
+
+The SpGEMM dataset was chosen to be the matrices in [101], a matrix test set designed to reveal asymtoptic differences in SpGEMM algorithms.
 We separated the matrices into small matrices and large matrices.
 Because of asymptotic slowdowns, we couldn’t run TACO outer products or inner products on the large matrices.
-However, since the matrices are organized by size in the small matrix plot, we can already see the beginnings of a trend where the performance of TACO outer products and inner products perform worse as the matrix size increases.
+However, since the matrices are organized by size in the small matrix plot, we can already see the beginnings of a trend where the TACO outer products and inner products perform worse as the matrix size increases.
 We will include a dedicated scaling plot to make these relationships more clear.
@@
 For symmetric SpMV, we found it is faster to pre-compute the strict triangle of the argument and re-use it within the kernel, to simplify the logic.
-It is of course possible to add an `if i < j` statement and compute the triangle on the fly without accessing the whole matrix, but the runtime of SpMV is impacted because this changes the exit condition of that loop.
+It is of course possible to add an `if i < j` statement and compute the triangle on the fly without accessing the whole matrix, but the runtime of SpMV is impacted because this changes the exit condition of the innermost loop.
 
@@
 
-Our SpMV evaluation demonstrates the performance of the SparseList, SparseVBL, SparseBand, and Pattern level formats, as well as conditionals and multiple outputs in the symmetric kernel
+In the following list, nonsequental formats are marked with *.
 
-Our SpMM evaluation shows off the importance of sparse intermediates (SparseBytemap and SparseHash are non-sequential) and the flexibility of our language for scheduling, doing all three matrix multiplication variants.
+SpMV exercises SparseList, SparseVBL, SparseBand, and Pattern formats, as well as multiple outputs in the symmetric kernel.
 
-Our Graph applications demonstrate conditionals, complex loop structures, early breaks, and user-defined functions.
+SpMM exercies SparseBytemap* and SparseHash* formats, as well as sparse intermediates, and demonstrates the flexibility of our language for scheduling all three matrix multiplication variants.
 
-Our Image Morphology benchmarks demonstrate the flexibility of the language, as well as affine indexing and `sparseRLE` and `pattern` formats, as well as user-defined functions.
+Graph Applications exercise SparseBytemap* and Pattern formats, and conditionals, complex loop structures, early breaks, and user-defined functions.
 
-The Array API benchmark shows that the language is reliable and flexible enough to support autoscheduling. It also involves user-defined functions and `SparseHash, non-sequential` format.
+Image Morphology exercises SparseRLE and Pattern formats, as well as affine indexing, padding, and user-defined functions, and demonstrates the flexibility of our language.
+
+The Array API scheduler exercises SparseHash* format and user-defined functions, and shows that the language is reliable and flexible enough to support autoscheduling.
 

### 2024-06-05 response.md
Update response.md
fixes

@@
 
-We thank the reviewers for their thorough feedback, as it will surely improve the work. We address high-level comments first, and more detailed comments in an explicit appendix.
+We thank the reviewers for their thorough feedback, as it will surely improve the work.
+Our response addresses high-level comments first, followed by an appendix which contains a proposed "example of compilation" section and a detailed response to each reviewer.
 
@@
 Here is an simplified definition of `unfurl` for a SparseList vector.
-It is the implementer's responsibility to ensure that `unfurl` returns a correct Looplet nest.
 
@@
         seek = (i) -> (q = binarysearch(A.lvl.idx, i)),
-        # This tells us the boundary of the loop
+        # This tells us the last index in each iteration
         stop = A.lvl.idx[q],
@@
 
-In our new section or an appendix, we plan to add an enhanced description of each mask and structured level format in terms of the looplet nests returned by `unfurl`, relating looplets to the concrete data fields of the tensor such as (`idx`, `ofs`, and `pos`), and addressing the concerns of Reviewer C.
+In our new section, we plan to add an enhanced description of each mask and structured level format in terms of the looplet nests returned by `unfurl`, relating looplets to the concrete data fields of the tensor such as (`idx`, `ofs`, and `pos`), and addressing the concerns of Reviewer C.
 
-As an example, Looplets allow us to make concordization inexpensive because the concordized loop is lowered with a mask.
-For example, consider the following program to calculate `A[5]` when `A` is sparse:
+Looplets allow us to make concordization inexpensive because the concordized loop is lowered with a mask. For example, consider the following program to calculate `A[5]` when `A` is sparse:
 ```
@@
 The lowered code runs in logarithmic time.
-We achieve the desired code because of (1) how SparseList communicates its structure in looplet nests via unfurl, (2) how wrapperization converts `i == 5` to a symbolic `DiagMask[i, 5]`, which then converts to looplets, and (3) how the two looplet nests combine to fast forward the sparse iterator and eliminate the `if false` branches.
+We achieve the desired code because of (1) how SparseList communicates its structure in looplet nests via unfurl, (2) how wrapperization converts `i == 5` to a symbolic `DiagMask[i, 5]` which also uses looplets, and (3) how the two looplet nests combine to fast forward the sparse iterator and eliminate the `if false` branches.
 
-This is an example of the productivity of Finch, as we were able to implement efficient scalar reads and writes for all combinations of formats by simply combining the `unfurl` definitions with a structured tensor.
+This demonstrates the productivity of Finch, as we were able to implement efficient scalar reads and writes for all combinations of formats by simply combining the `unfurl` definitions with a structured tensor.
 
-Our language allows fairly complex programs to be expressed on fairly complex structures becayse these optimizations happen in a predictable and reliable way.
+Our language allows fairly complex programs to be expressed on fairly complex structures because these optimizations happen in a predictable and reliable way.
 Our normal form, life cycles, concordization, and wrapperization are key ideas that allow us to boil our Finch language down to looplets so that the right looplet optimizations fire.
@@
 
-## Clarifying the Evaluation
+## Improving the Evaluation
 
@@
 
-As suggested by Reviewer A, we also plan to report our geo-mean speedups, which were:
+As suggested by Reviewer A, we also plan to report our geo-mean speedups, which were as follows:
 
-The SpGEMM dataset was chosen to be the matrices in [101], a matrix test set designed to reveal asymtoptic differences in SpGEMM algorithms.
-We separated the matrices into small matrices and large matrices.
-Because of asymptotic slowdowns, we couldn’t run TACO outer products or inner products on the large matrices.
-However, since the matrices are organized by size in the small matrix plot, we can already see the beginnings of a trend where the TACO outer products and inner products perform worse as the matrix size increases.
-We will include a dedicated scaling plot to make these relationships more clear.
+- For SpMV, Finch achieves 1.08x and 1.09x geomean speedups over julia_stdlib and TACO, respectively.
+
+- For SpGEMM, for the small dataset (Figure 18 left), Finch's inner, outer, and Gustavson achieves 1.03x, 2.92x, 1.38x geomean speedups over TACO's inner, outer, and Gustavson, respectively.
+For the large dataset (Figure 18 right), Finch's Gustavson achieves 1.24x geomean speedup over TACO's Gustavson.
+
+- For BFS, Finch's pull-push achieves 3.01x and 0.88x geomean speedups over Graphs.jl and GraphBLAS, respectively.
+For SSSP, Finch achieves 1.71x and 3.43x geomean speedups over Graphs.jl and GraphBLAS, respectively.
+
+- For Triangle Count, SDDMM, and (A+B)*C kernels, Finch achieves 14.43x, 6.50x, and 19.58x geomean speedups over DuckDB, respectively.
+
+## Tradeoffs between Specialization and Conditional Overhead
 
@@
 
-In general, we have found Finch is quite adept at lifting conditionals to the highest level of the loop nest, but that there is a tradeoff between the branching between specialized routines and the benefits of such specialization.
-Specialization is better in cases where the specialized routine is much faster and much more frequent (such as the zero region of a sparse tensor, which becomes a no-op).
-In practice, this has led us to sometimes ask Finch to de-specialize certain conditional expressions, and is another example of how Finch can widen the design space for structured operators.
+Finch makes specialization more convenient than ever before.
+We have begun to find that there is a tradeoff between branching and more complicated loop structures and the benefits of such specialization.
+Specialization is better in cases where the specialized routine is much faster and the common case (such as the zero region of a sparse tensor, which becomes a no-op).
+In practice, this has led us to sometimes ask Finch to de-specialize certain conditional expressions, which is another example of how Finch can widen the design space for structured operators.
 We will include a discussion of this tradeoff in the paper.
 
-> **Reviewer A**: Which features of your language are exercised by which case study? Can you indicate which features your baseline is lacking to reach the performance you are demonstrating.
+## Freeze and Thaw
 
-> **Reviewer C**:  “I would have liked to see quantifying data like how many times you had to use Sequentially constructed levels over Nonsequentially Constructed Levels”
+We do not intend the user to insert freeze/thaw manually, but we include them in the language because it allows us to handle lifecycles with a separate, simpler, compiler pass, rather than all at once.
+Tensors can only change between read and write mode in the scope in which they were defined, so we can insert freeze/thaw automatically by checking whether the tensor is being read or written to in each child scope.
+We error if a tensor appears on both the left hand and right hand sides within the same child scope.
+We will clarify this in the paper.
 
-In the following list, nonsequental formats are marked with *.
+## Breaks
 
-SpMV exercises SparseList, SparseVBL, SparseBand, and Pattern formats, as well as multiple outputs in the symmetric kernel.
+As Reviewer B suggests, we plan to include a larger discussion of how breaks integrate into the language.
+To integrate better with other structured arguments, Finch handles breaks as a structural simplification, rather than with a `break` statement.
+The following example uses a short circuit scalar to find the first negative value in A:
 
-SpMM exercies SparseBytemap* and SparseHash* formats, as well as sparse intermediates, and demonstrates the flexibility of our language for scheduling all three matrix multiplication variants.
-
-Graph Applications exercise SparseBytemap* and Pattern formats, and conditionals, complex loop structures, early breaks, and user-defined functions.
-
-Image Morphology exercises SparseRLE and Pattern formats, as well as affine indexing, padding, and user-defined functions, and demonstrates the flexibility of our language.
-
-The Array API scheduler exercises SparseHash* format and user-defined functions, and shows that the language is reliable and flexible enough to support autoscheduling.
-
-> **Reviewer C**: I would like to know why the formats provide the speedup: Is it just due to looplets iterators at each level?
-Is it due to constructed levels and if so, how do they provide benefit over other constructed levels and why over TACO’s representation?
+```
+p = ShortCircuitScalar{0}()
+@finch begin
+    p .= 0
+    for j=_
+        if A[j] < 0
+            p[] <<choose(0)>>= j
+        end
+    end
+end
+```
 
@@
 
-## Lowering Example Section
+## Proposed Section with an Example of Compilation
 
@@
 We then begin lowering the program.
-The `@declare` statement on `y` results in a simple initialization of the `val` array, `s.lvl.val[1] = 0.0`.
+The `@declare` statement on `s` results in a simple initialization of the `val` array, `s.lvl.val[1] = 0.0`.
 To process the `j` loop, we unfurl both tensors which access `j`.
@@
 ```
-unfurl(UpTriMaskCol(j)) = Sequence(
+unfurl(UpTriMaskCol(j)) = 
+    Sequence(
         Phase(
@@
 unfurl(SubFiber(A.lvl.lvl::SparseListLevel, j)) = Thunk(
-    # First, we initialize our position iterator
-    preamble = (q = 1),
-    # The stepper is analogous to a for-loop
+    preamble = (q = A.lvl.lvl.ptr[j]),
     Stepper(
-        # This fast forwards the loop to a specific iteration
         seek = (i) -> (q = binarysearch(A.lvl.lvl.idx, i)),
-        # This tells us the boundary of the loop
         stop = A.lvl.lvl.idx[q],
@@
         ),
-        # This advances the loop to the next iteration
         next = (q += 1)
@@
 for j = 1:n
-    q = 1
+    q = A.lvl.lvl.ptr[j]
     @finch for i = 1:j
@@
 for j = 1:n
-    q = 1
+    q = A.lvl.lvl.ptr[j]
     i_loop = 1
@@
 The final program accesses only the upper triangle of A, though the original code looks as though it loops over all `i` and `j`.
+
+## Detailed Comments
+
+### Reviewer A
+
+> Which features of your language are exercised by which case study? Can you indicate which features your baseline is lacking to reach the performance you are demonstrating.
+
+The baseline implemenation usually uses a combination of SparseList and Dense formats, which do not specialize at all to the particular structures we tested against.
+
+- SpMV exercises SparseVBL, SparseBand, and Pattern formats, as well as multiple outputs in the symmetric kernel.
+
+- SpMM exercies SparseBytemap and SparseHash formats, as well as sparse intermediates, and demonstrates the flexibility of our language for scheduling all three matrix multiplication variants.
+
+- Graph Applications exercise SparseBytemap and Pattern formats, and conditionals, complex loop structures, early breaks, and user-defined functions.
+
+- Image Morphology exercises SparseRLE and Pattern formats, as well as affine indexing, padding, and user-defined functions, and demonstrates the flexibility of our language.
+
+- The Array API scheduler exercises SparseHash format and user-defined functions, and shows that the language is reliable and flexible enough to support autoscheduling.
+
+> It is unclear how the [SpGEMM] dataset has been selected. A reference to a standard matrix dataset would be appreciated
+
+The SpGEMM dataset was chosen to be the matrices in [101], a matrix test set designed to reveal asymtoptic differences in SpGEMM algorithms.
+We separated the matrices into small matrices and large matrices.
+Because of asymptotic slowdowns, we couldn’t run TACO outer products or inner products on the large matrices.
+However, since the matrices are organized by size in the small matrix plot, we can already see the beginnings of a trend where the TACO outer products and inner products perform worse as the matrix size increases.
+We will include a dedicated scaling plot to make these relationships more clear.
+
+> Do you report here the 'geo mean' or 'arith mean' speedup?
+
+We report the arithmetic mean in the paper.
+
+> I wonder if the indirection through Julia had a performance cost. I would have preferred to compare with GraphBLAS natively.
+
+We compared with the native C implementation of LAGraph, which calls GraphBLAS using the native C bindings.
+
+> I appreciate that you mention this timeout and would be curious who timed out? Finch, Graphblas, ...? Are there previously reported numbers on GAP-road and why does this time a timeout occur?
+
+[98] does not measure Bellman-ford, which is a more expensive algorithm that scales roughly as O(V*E), a quantity which is about 20x larger in GAP-road.
+GAP-road has 23.9M vertices.
+At 8 B per integer distance, this exceeds the 49.5 MB L3 cache on our processor, and we stopped measuring LAGraph/GraphBLAS after an hour.
+We will add a note to the paper.
+
+> [98] evaluates on 11 datasets, but this work only on 6. I am unsure why the others are omitted.
+
+Thank you for pointing out this discrepancy.
+We were sourcing our matrices from the University of Florida sparse matrix collection.
+Based on the labels in Figure 7, we could not locate `h09` and `i04`.
+However, upon re-reading, we see that Table 3 indicates these graphs were actually `LAW/hollywood-2009` and `LAW/indochina-04`, and we will add these to our evaluation.
+The other matrices weren't in the collection because they were randomly generated.
+We also plan to generate these and add them to the evaluation.
+
+### Reviewer B
+
+> First, Figure 7 defines DECLARE, FREEZE, and THAW, but they aren't referred to anywhere else?
+
+Thank you for this correction, DECLARE, FREEZE, and THAW should have been added as options to STMT in Figure 7.
+
+> Does "consistently better" on line 751 mean "better in every single case"?
+
+Yes, it does.
+
+> Section 6.5 looks to be introducing yet another language and an interpreter ... why is it stuffed into one page in the evaluation of Finch?
+
+This section demonstrates the potential use of Finch as a backend for higher-level languages. We will include more details of the high-level language in the appendix.
+
+### Reviewer C
+
+> Please check figure 13 since it appears freeze has been placed wrongly before l-value of x and y are encountered.
+
+Thank you for this correction, both freeze statements in Figure 13 need to be moved downward.
+
+> Due to these spreading of case studies, each study also feel quite high-level and should be discussed in details.
+
+As mentioned above, we plan to include additional details in an appendix.
+
+> The speedups and slowdowns are highlighted and typically simple explanations are given for these performance results like 787-788 for SparseList or Sparse-List Pattern formats.
+
+In this particular instance, the speedup on the symmetric kernel is because we reuse reads to the upper triangle of the matrix. We will include this explanation in the paper.
+
+> I would have liked to see quantifying data like how many times you had to use Sequentially constructed levels over Nonsequentially Constructed Levels
+
+SparseBytemap and SparseHash are the two nonsequential levels we used.
+They were used in outer products and Gustavson SpGeMM, the frontier in our graph applications, and as a conservative output format in the Array API scheduler.
+
+> I would like to know why the formats provide the speedup
+
+In general, the sparse list format stores the locations of nonzeros and values individually.
+Other structured formats improve on this through specialization.
+For example, the SparseBand format stores the location of a contiguous group of nonzeros only once.
+Iterating through a sparse list produces a while loop that looks up each nonzero index from memory, while iterating through a sparse band produces a for-loop over the range of indices in the band.
+We will measure operation counts to help further explain these speedups.

### 2024-06-05 response.md
spelling

@@
 polyhedral [102] or e-graph search [77]).
-We chose Looplets because they reliably process structured iterators, predictably eliminating zero regions, using faster lookups when avaliable, and utilizing repeated work.
+We chose Looplets because they reliably process structured iterators, predictably eliminating zero regions, using faster lookups when available, and utilizing repeated work.
 Here is an simplified definition of `unfurl` for a SparseList vector.
@@
 
-Next, we process the `i` loop. We Unfurl the tensors, replacing them with the follwing nests of looplets:
+Next, we process the `i` loop. We Unfurl the tensors, replacing them with the following nests of looplets:
 
@@
 
-The baseline implemenation usually uses a combination of SparseList and Dense formats, which do not specialize at all to the particular structures we tested against.
+The baseline implementation usually uses a combination of SparseList and Dense formats, which do not specialize at all to the particular structures we tested against.
 
@@
 
-- SpMM exercies SparseBytemap and SparseHash formats, as well as sparse intermediates, and demonstrates the flexibility of our language for scheduling all three matrix multiplication variants.
+- SpMM exercises SparseBytemap and SparseHash formats, as well as sparse intermediates, and demonstrates the flexibility of our language for scheduling all three matrix multiplication variants.
 
@@
 
-The SpGEMM dataset was chosen to be the matrices in [101], a matrix test set designed to reveal asymtoptic differences in SpGEMM algorithms.
+The SpGEMM dataset was chosen to be the matrices in [101], a matrix test set designed to reveal asymptotic differences in SpGEMM algorithms.
 We separated the matrices into small matrices and large matrices.

### 2024-06-05 response.md
update

@@
 
-[98] does not measure Bellman-ford, which is a more expensive algorithm that scales roughly as O(V*E), a quantity which is about 20x larger in GAP-road.
-GAP-road has 23.9M vertices.
-At 8 B per integer distance, this exceeds the 49.5 MB L3 cache on our processor, and we stopped measuring LAGraph/GraphBLAS after an hour.
+[98] does not measure Bellman-Ford, which is a more expensive algorithm.
+In the implementations we tested, the number of iterations of Bellman-Ford is bounded by the diameter of the graph.
+GAP-road has a very large diameter.
+For example, soc-orkut has 212.7M edges and a diameter of 9.
+GAP-road has 577.1M edges and a diameter of 6809.
+If each iteration ran in `O(nnz)`, this would result in a factor of 2059  increase in expected runtime.
+We stopped measuring LAGraph/GraphBLAS after an hour.
 We will add a note to the paper.
